### 2022-02-09 README.Rmd
intial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# GJBpottery

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/LiYingWang/GJBpottery/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your broswer or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2022-02-10 analysis/code/isotope-analysis.R
added data and R code

@@ -1,0 +1,59 @@
+library(tidyverse)
+isotope_data <-
+  readxl::read_excel(here::here("Data", "12_21_potshard_13C.xlsx"),
+                     sheet = "Summary", col_names =  FALSE)
+
+# filter the blank corrected data
+delta_blank_cor <-
+  isotope_data %>%
+  select(1, 15, 17) %>%
+  slice(2:8) %>%
+  dplyr::mutate(...1 = ifelse(is.na(...1), paste0("C13"), ...1))
+
+names(delta_blank_cor) <- delta_blank_cor[1,]
+
+delta_blank_cor <-
+  delta_blank_cor %>%
+  slice(2:7) %>%
+  dplyr::mutate(across(starts_with("13C"), as.numeric))
+
+# filter the Meth corrected data
+delta_meth_cor <-
+  isotope_data %>%
+  select(1, 7, 9) %>%
+  slice(2:8) %>%
+  dplyr::mutate(...1 = ifelse(is.na(...1), paste0("C13"), ...1))
+
+names(delta_meth_cor) <- delta_meth_cor[1,]
+
+delta_meth_cor <-
+  delta_meth_cor %>%
+  slice(2:7) %>%
+  dplyr::mutate(across(starts_with("13C"), as.numeric))
+
+# import a SVG for later ggplot overlaying with ellipses
+tem <- rsvg::rsvg(here::here("carbon-isotope-ellipses-template.svg"))
+png::writePNG(tem, "tem.png", dpi =300)
+browseURL("tem.png") # take a look
+
+# plot C16 and C18
+ggplot(delta_blank_cor, # delta_meth_cor or delta_blank_cor
+       aes(`13C C16:0`,`13C C18:0`)) +
+  geom_point(size = 1, alpha = 0.9) +
+  ggrepel::geom_text_repel(aes(label = C13)) +
+  theme_minimal(base_size = 14) +
+  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
+       y = bquote(delta*{}^13*"C 18:0 \u2030")) +
+  xlim(-40,-10) +
+  ylim(-40,-10) +
+  coord_fixed(ratio=1) +
+  #scale_colour_viridis_d(direction = -1) +
+  annotation_raster(tem, ymin = -42.8, ymax= -8 ,
+                    xmin = -44.5 , xmax = -5.65)
+
+ggsave(here::here("delta_C16_C18.png"),
+       width = 8,
+       height = 8,
+       dpi = 300,
+       units = "in")
+

### 2022-02-10 analysis/code/isotope-analysis.R
added a svg file and changed file path

@@ -1,6 +1,6 @@
 library(tidyverse)
 isotope_data <-
-  readxl::read_excel(here::here("Data", "12_21_potshard_13C.xlsx"),
+  readxl::read_excel(here::here("analysis","data", "raw_data","12_21_potshard_13C.xlsx"),
                      sheet = "Summary", col_names =  FALSE)
 
 # filter the blank corrected data
@@ -32,7 +32,7 @@ delta_meth_cor <-
   dplyr::mutate(across(starts_with("13C"), as.numeric))
 
 # import a SVG for later ggplot overlaying with ellipses
-tem <- rsvg::rsvg(here::here("carbon-isotope-ellipses-template.svg"))
+tem <- rsvg::rsvg(here::here("analysis", "figures","carbon-isotope-ellipses-template.svg"))
 png::writePNG(tem, "tem.png", dpi =300)
 browseURL("tem.png") # take a look
 

### 2022-10-29 analysis/code/isotope-analysis.R
update the isotope plot

@@ -15,7 +15,8 @@ names(delta_blank_cor) <- delta_blank_cor[1,]
 delta_blank_cor <-
   delta_blank_cor %>%
   slice(2:7) %>%
-  dplyr::mutate(across(starts_with("13C"), as.numeric))
+  dplyr::mutate(across(starts_with("13C"), as.numeric)) %>%
+  dplyr::mutate(delta = `13C C18:0`-`13C C16:0`)
 
 # filter the Meth corrected data
 delta_meth_cor <-
@@ -29,7 +30,8 @@ names(delta_meth_cor) <- delta_meth_cor[1,]
 delta_meth_cor <-
   delta_meth_cor %>%
   slice(2:7) %>%
-  dplyr::mutate(across(starts_with("13C"), as.numeric))
+  dplyr::mutate(across(starts_with("13C"), as.numeric)) %>%
+  dplyr::mutate(delta = `13C C18:0`-`13C C16:0`)
 
 # import a SVG for later ggplot overlaying with ellipses
 tem <- rsvg::rsvg(here::here("analysis", "figures","carbon-isotope-ellipses-template.svg"))
@@ -74,8 +76,9 @@ ggsave(here::here("delta_C16_C18_remove_pot6.png"),
        dpi = 300,
        units = "in")
 
-# plot carbon isotopes of 16 and C18 by following Salque et al. (2013)
-ggplot(delta_blank_cor, # delta_meth_cor or delta_blank_cor
+# plot the carbon isotopes by following the method in Salque et al. (2013)
+isotope_C16_C18 <-
+  ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
        aes(`13C C16:0`,`13C C18:0`)) +
   geom_point(size = 1, alpha = 0.9) +
   ggrepel::geom_text_repel(aes(label = C13)) +
@@ -85,12 +88,23 @@ ggplot(delta_blank_cor, # delta_meth_cor or delta_blank_cor
   xlim(-35,-20) +
   ylim(-35,-20) +
   coord_fixed(ratio = 1) +
-  geom_abline(intercept = -0.3, slope = 1, linetype = "dashed") +
+  geom_abline(intercept = -0.3, linetype = "dashed") +
   geom_abline(intercept = -3.1, linetype = "dashed") +
   annotate("text", x = -22, y = -22, angle = 45, vjust = 1.5,
[…]

### 2022-10-30 analysis/code/site-regional-map-Ch.R
update the site map

@@ -21,9 +21,9 @@ world_points <- cbind(world, st_coordinates(st_point_on_surface(world$geometry))
 
 # add site location
 site_location <-
-  data.frame(location = c("Guijuabao"),
-             lon = c(101.3617),
-             lat = c(27.2657))
+  data.frame(location = c("Guijuabao", "Chengdu"),
+             lon = c(101.3617, 104.0587),
+             lat = c(27.2657, 30.5899))
 
 China_SE_Asia <-
   ggplot(data = world) +

### 2022-11-01 analysis/paper/paper.Rmd
removed the nanopore code

---
title: "Investigating Neolithic pottery use in the Yanyuan Basin in Southwest China using organic residue analysis"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300)
```

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange or introduction of goods, species, and ideas that could result in far-reaching impacts on local societies economically and culturally. In Southwest China, the Hengduan Mountain ranges as natural passes play an important role in the migration of farming communities and the spread of crops into this region and mainland Southeast Asia. Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022]. However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood. To explore whether the new species became main food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, the oldest Neolithic site in the Yanyuan region so far excavated.

<!-- sites and research questions-->
As an early Neolithic site, Guijiabao (ca. 4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Plateau, and the Yunnan-Guizhou Plateau. The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@citation], different from the abundant rice grains found at other Neolithic sites in that area. Although the plant remains suggest the possible domestication of millet in the early Neolithic, we still lack direct evidence of culinary practices during the Neolithic transition. To explore the extent of acceptance of new crops and species in the Yanyuan Basin, we ask the questions: What was the culinary practices during the Neolithic transition? Could we identify the evidence of new crops or species incorporated into the local dietary systems in the Early Neolithic period? We answer those questions by examining potsherds collected from Guijiabao using organic residue analysis, which could offer a better understanding of food sources during the Neolithic transition in the Yanyuan Basin.

This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the use of pottery vessels at Guijiabao. Lipid analysis have proven useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. Lipid residues are the most direct evidence to investigate culinary practices and dietary patterns since specific lipid compounds can be associated with different types of foodstuffs cooked or stored in pottery vessels [@Copley2005]. It is especially informative for human subsistence in the situation where archaeological sites lack organic remains due to acidic sediments [@Kwak2020]. The growing application of organic residue analysis demonstrate it offers a better understanding of culinary practices during the transitional period to agricultural societies [@Craig2011], dietary patterns in

# Geographical and Cultural Background of Guijiabao in Yanyuan Basin
<!--Info about Yanyuan area-->
The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to Yunnan-Guizhou Plateau. Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022]. The archaeological sites in the Yanyuan Basin, therefore, are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan.

<!--Info about Guijiabao-->
Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the Basin is the only site with continuous occupation from the early Neolithic period to the historical period. The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022]. The Neolithic period can be further separated into two subphases, 5000-4500 cal. BP and 4500-3700 cal. BP, according to the cultural remains and pottery characteristics. Recent archaeological evidence indicates the spread of millet-rice mixed economy to southwest Sichuan can be dated back to around 5,000 BP.

# Materials and methods

## Samples
We selected five ceramic fragments (Figure XX) from Guijiabao for organic residue analysis, in which four samples were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season. To further investigate the pottery use in an earlier period, we focus on pottery vessels from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.). During this period, the common vessel forms include jars, urns, bowls, and basins. For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022]. Our samples are mostly fragments from jars of those two major types that were supposed to be used for food serving or storage according to the archaeological context of living area [].

## Methods
In this study, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial or marine animals and plants, have specific ranges of δ^13^C values [@Evershed1994].

We cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C). We conducted saponification by adding 10 mL 1M KOH in methanol and 1 mL water into the extracts and heated them in an oven at 70 degrees C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour. After drying under a stream of nitrogen, the derivitized sample was dissolved in toluene (20 uL) prior to GC-MS.

The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) using 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream. About 5% aliquot dissolved in toluene was analyzed by GC-FID and the remaining sample was analyzed by GC-C-IRMS.

# Results
Lipid residues from our pottery samples show that the saturated medium-chain fatty acids, palmitic (C16:0) and stearic(C18:0), are the two dominant compounds. Those two compounds were normally degraded from animal fat and plant wax that indicates the organic compounds in our samples were highly degraded

. However, for the sample five, we detected rich compounds with high intensity compared to the other four samples. This might indicate a different source.

To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples. Here you can see the stable carbon isotope measurements of C16:0 and C18:0 fatty acids. The results suggest a dominance of C3 plant for most pots, and a few shows a mixture of animal fats and C3 plants. The sample 5 where more compounds were preserved shows a source of C4 plant or marine animals instead. We think this is very interesting finding since there was just a few rice grains found at this site compared to the evidence of millets.

# Discussion
The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source. Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously. Our next steps is to increase sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.

# Conclusion
Our results show a mixture of C3 plants and terrestrial animals for most pots with one pot showing more C4 plant sources. This case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region. Our case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2022-11-01 analysis/code/DNA-nanopore.R

@@ -1,28 +1,0 @@
-# read data
-nano <- read.csv(here::here("analysis/data/raw_data/Combined_Renamed_Seq.csv"),
-                      header = F)
-
-# tidy data
-ixd <- which(str_detect(nano$V1, "seq"))
-nano_split <- split(nano, cumsum(1:nrow(nano) %in% ixd))
-names(nano_split) <- map(nano_split, ~.x$V1[1])
-nano_all <- map_df(nano_split, ~.x[-1,])
-
-# transfer the data layout
-nano_long <-
-  nano_all %>%
-  pivot_longer(cols = starts_with(">"),
-               names_to = "read",
-               values_to = "sequence") %>%
-  mutate(read = gsub('>', '', read)) %>%
-  mutate(dup_w = lapply(str_extract_all(sequence, "(?i)([a-z])\\1+"), nchar)) %>%
-  mutate(dup_w_combo = lapply(str_extract_all(sequence, "([ACGT][ACGT])\\1+"), nchar)) %>%
-  mutate(length_w = nchar(sequence)) %>%
-  mutate(dup_w_max = lapply(dup_w, max)) %>%
-  mutate(dup_w_com_max = lapply(dup_w_combo, max)) %>%
-  filter(dup_w_max < 10 & dup_w_com_max < 10) %>%
-  filter(length_w >= 200)
-
-# load a BLAST database (replace db with the location + name of the BLAST DB)
-bl <- blast(db = ".16SMicrobialDB/16S_ribosomal_RNA")
-bl

### 2022-11-23 analysis/code/GCFID-quant.R
update the result for specific GS sample

@@ -77,11 +77,14 @@ highlight_SYG2_2 <-
     retention == 27.3 ~ "C17:0",
     retention == 28.4 ~ "C18:1",
     retention == 29.1 ~ "C18:0",
+    retention == 31.0 ~ "C19:0",
     retention == 32.4 ~ "C20:0",
     retention == 35.4 ~ "C22:0",
     retention == 38.2 ~ "C24:0",
+    retention == 40.9 ~ "C26:0",
     retention == 33.9 ~ "IS",
-    retention == 41.3 ~ "IS")) %>%
+    retention == 41.3 ~ "IS",
+    retention == 47.8 ~ "C:32",)) %>%
   filter(name == "SYG-TN13-E22-2#2")
 
 # plot the first four

### 2022-11-26 analysis/paper/paper.Rmd
edit code chunk

@@
 ---
-title: "Investigating Neolithic pottery use in the Yanyuan Basin in Southwest China using organic residue analysis"
+title: "Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis"
 author:
@@
 
-This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the use of pottery vessels at Guijiabao.
+This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the vessel usage at Guijiabao.
 Lipid analysis have proven useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017].
@@
 ## Samples
-We selected five ceramic fragments (Figure XX) from Guijiabao for organic residue analysis, in which four samples were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season.
-To further investigate the pottery use in an earlier period, we focus on pottery vessels from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.).
+We selected five ceramic fragments (Table XX) from Guijiabao for organic residue analysis, in which four samples were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season.
+To further investigate the content and use of pottery vessels in an earlier period, we focus on potsherds excavated from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.).
 During this period, the common vessel forms include jars, urns, bowls, and basins.
 For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022].
-Our samples are mostly fragments from jars of those two major types that were supposed to be used for food serving or storage according to the archaeological context of living area [].
+Our samples are jar fragments from those two common types supposed to be used for food serving or storage according to their archaeological context that suggests living area [].
 
@@
 # Results
-Lipid residues from our pottery samples show that the saturated medium-chain fatty acids, palmitic (C16:0) and stearic(C18:0), are the two dominant compounds.
-Those two compounds were normally degraded from animal fat and plant wax that indicates the organic compounds in our samples were highly degraded
 
-. However, for the sample five, we detected rich compounds with high intensity compared to the other four samples. This might indicate a different source.
+```{r tidy data}
+library(tidyverse)
+
+# load the GC data
+meth <- read_csv(here::here("analysis/data/raw_data/LY_21_08_20_pot_chrom.CSV"))
+
+# data wrangling
+ixd <- which(str_detect(meth$Path, "C"))
+meths <- split(meth, cumsum(1:nrow(meth) %in% ixd))
+meth_data <- meths[c(FALSE, TRUE)]
+meth_name <- meths[c(TRUE, FALSE)]
+names(meth_data) <- map(meth_data, ~.x$Path[1])
+names(meth_name) <-map(meth_name, ~.x$Sample[1])
+meths <- map_df(meth_data, ~.x[-1, ] %>%
+                   mutate_all(as.numeric), .id = 'sample')
+
+# add sample name
+meth_all <-
+  meths %>%
+  mutate(sample = paste0(parse_number(sample),
+                         str_extract(sample, pattern = "_00[0-9]"))) %>%
+  mutate(name = case_when(
+    sample %in% c("210729_001") ~ "CDG-062",
+    sample %in% c("210729_002") ~ "SYG-3",
+    sample %in% c("210729_003") ~ "SYG-TN13-E22-2#1",
+    sample %in% c("210729_004") ~ "SYG-TN13-E22-1",
+    sample %in% c("210729_005") ~ "SYG-TN13-E22-2#2",
+    sample %in% c("210729_006") ~ "SYG-TN13-E23-3")) %>%
+  mutate(name = factor(name, levels = c("SYG-TN13-E22-1",
+                                        "SYG-TN13-E22-2#1",
+                                        "SYG-TN13-E22-2#2",
+                                        "SYG-TN13-E23-3",
+                                        "SYG-3",
+                                        "CDG-062")))
+```
+
+Lipid residues from our samples show that the saturated fatty acids with even number of carbons (C14:0, C16:0, and C18:0), particularly palmitic (C16:0) and stearic(C18:0), are the dominant compounds.
+Because palmitic and stearic fatty acids are normally degraded from animal fat or plant wax, the domination pattern with little concentration of other fatty acids suggests the organic compounds in our samples were highly degraded during post depositional processes.
+However, there is one sample contains rich compounds that include long-chain fatty acids, C20, C22, and C24, which were absent in other samples.
+This might indicate
 
@@
 The sample 5 where more compounds were preserved shows a source of C4 plant or marine animals instead.
-We think this is very interesting finding since there was just a few rice grains found at this site compared to the evidence of millets.
+We think this is very interesting finding since there was just a few rice grains found at this site compared to the evidence of millet.
 

### 2022-11-30 analysis/paper/paper.Rmd
added the code chunk for chromatogram

@@
   dpi = 300)
+
+image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
 ```
@@
 
-```{r tidy data}
+```{r tidy-data}
 library(tidyverse)
@@
                                         "CDG-062")))
+```
+
+```{r highlight-compounds}
+# highlight target compounds
+highlight_GJB_four <-
+  meth_all %>%
+  mutate(retention = round(Path, 1)) %>%
+  group_by(sample, retention) %>%
+  arrange(desc(File)) %>%
+  distinct(retention, .keep_all = TRUE) %>%
+  mutate(FA = case_when(
+    retention == 21.6 ~ "C14:0",
+    retention == 23.6 ~ "C15:0",
+    retention == 25.5 ~ "C16:0",
+    retention == 27.3 ~ "C17:0",
+    retention == 28.4 ~ "C18:1",
+    retention == 29.1 ~ "C18:0",
+    retention == 32.4 ~ "C20:0",
+    retention == 35.4 ~ "C22:0")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
+  filter(File > 100000)
+
+highlight_IS <-
+  meth_all %>%
+  mutate(retention = round(Path, 1)) %>%
+  group_by(sample, retention) %>%
+  arrange(desc(File)) %>%
+  distinct(retention, .keep_all = TRUE) %>%
+  mutate(FA = case_when(
+    retention == 33.9 ~ "IS",
+    retention == 41.3 ~ "IS")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3"))
+
+# highlight the compounds in the potsherd with high concentration
+highlight_SYG2_2 <-
+  meth_all %>%
+  mutate(retention = round(Path, 1)) %>%
+  group_by(sample, retention) %>%
+  arrange(desc(File)) %>%
+  distinct(retention, .keep_all = TRUE) %>%
+  mutate(FA = case_when(
+    retention == 21.6 ~ "C14:0",
+    retention == 25.5 ~ "C16:0",
+    retention == 27.3 ~ "C17:0",
+    retention == 28.4 ~ "C18:1",
+    retention == 29.1 ~ "C18:0",
+    retention == 31.0 ~ "C19:0",
+    retention == 32.4 ~ "C20:0",
+    retention == 35.4 ~ "C22:0",
+    retention == 36.8 ~ "C23:0",
+    retention == 38.2 ~ "C24:0",
+    retention == 40.9 ~ "C26:0",
+    retention == 33.9 ~ "IS",
+    retention == 41.3 ~ "IS",
+    retention == 45.6 ~ "C:30",
+    retention == 47.8 ~ "C:32")) %>%
+  filter(name == "SYG-TN13-E22-2#2")
+```
+
+```{r chromatogram-four}
+# plot the first four
+chroma_four <-
+  meth_all %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
+  mutate(File = ifelse(File > 550000, 550000, File)) %>%
+  ggplot(aes(Path, File)) +
+  geom_line(size = 0.3) +
+  geom_text(data = highlight_GJB_four, #ggrepel::geom_text_repel
+            aes(label = FA, angle = 90),
+            size = 3,
+            nudge_y = 87000,
+            show.legend = FALSE) +
+  geom_text(data = highlight_IS,
+            aes(label = FA),
+            size = 3,
+            nudge_y = 40000,
+            show.legend = FALSE) +
+  geom_segment(aes(x= 41.1, xend= 41.5, y= 550000, yend= 550000), size =0.3) +
+  geom_segment(aes(x= 41.1, xend= 41.5, y= 560000, yend= 560000), size =0.3) +
+  annotate("text", x = 41.3, y = 600000, label = "IS", size =3) +
+  facet_wrap(~name, ncol = 1) +
+  scale_y_continuous(labels = scales::comma_format(),
+                     limits = c(0, 600000),
+                     breaks = seq(0, 600000, 200000)) +
+  scale_x_continuous(limits = c(20, 45),
+                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
+  labs(x = "retention time", y = "Relative Intensity") +
+  theme_minimal()
+
+ggsave(plot = chroma_four,
+       here::here("analysis","figures", "chromatograms_four.png"),
+       width = 8,
+       height = 8,
+       dpi = 300,
+       units = "in")
+```
+
+```{r chromatograms-plots-four, echo=FALSE, fig.cap= "chromatograms"}
+knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
 ```

### 2022-12-01 analysis/code/GCFID-quant.R
revise code for chromatogram

@@ -1,9 +1,9 @@
 library(tidyverse)
 
-# GC-FID results after methyl
+# load the GC data
 meth <- read_csv(here::here("analysis/data/raw_data/LY_21_08_20_pot_chrom.CSV"))
 
-# tidy meth
+# tidy data
 ixd <- which(str_detect(meth$Path, "C"))
 meths <- split(meth, cumsum(1:nrow(meth) %in% ixd))
 meth_data <- meths[c(FALSE, TRUE)]
@@ -13,7 +13,7 @@ names(meth_name) <-map(meth_name, ~.x$Sample[1])
 meths <- map_df(meth_data, ~.x[-1, ] %>%
                    mutate_all(as.numeric), .id = 'sample')
 
-# join two datasets and get sample number
+# assign sample name
 meth_all <-
   meths %>%
   mutate(sample = paste0(parse_number(sample),
@@ -80,11 +80,13 @@ highlight_SYG2_2 <-
     retention == 31.0 ~ "C19:0",
     retention == 32.4 ~ "C20:0",
     retention == 35.4 ~ "C22:0",
+    retention == 36.8 ~ "C23:0",
     retention == 38.2 ~ "C24:0",
     retention == 40.9 ~ "C26:0",
     retention == 33.9 ~ "IS",
     retention == 41.3 ~ "IS",
-    retention == 47.8 ~ "C:32",)) %>%
+    retention == 45.6 ~ "C:30",
+    retention == 47.8 ~ "C:32")) %>%
   filter(name == "SYG-TN13-E22-2#2")
 
 # plot the first four
@@ -129,14 +131,14 @@ meth_all %>%
   geom_text(data = highlight_SYG2_2, #ggrepel::geom_text_repel
             aes(label = FA),
             size = 3,
-            nudge_x = 0,
             nudge_y = 250000,
             show.legend = FALSE) +
   scale_y_continuous(labels = scales::comma_format(),
                      limits = c(0, 7500000),
                      breaks = seq(0, 7500000, 1000000)) +
-  scale_x_continuous(limits = c(15, 50),
+  scale_x_continuous(limits = c(20, 50),
                      expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
+  labs(x = "retention time", y = "Relative Intensity") +
   theme_minimal()
 
 ggsave(here::here("analysis","figures", "chromatograms_specific.png"),

### 2023-01-06 analysis/code/isotope-analysis.R
combined isotopic plots

@@ -106,39 +106,45 @@ ggsave(here::here("delta_C16_C18_remove_pot6.png"),
 isotope_C16_C18 <-
   ggplot(delta_meth_cor, # delta_meth_cor/delta_blank_cor
        aes(`13C C16:0`,`13C C18:0`)) +
-  geom_point(size = 1, alpha = 0.9, aes(shape = sample)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
   #ggrepel::geom_text_repel(aes(label = C13)) +
-  theme_minimal(base_size = 14) +
+  theme_minimal() +
   labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
        y = bquote(delta*{}^13*"C 18:0 \u2030")) +
-  xlim(-35,-20) +
-  ylim(-35,-20) +
-  coord_fixed(ratio = 1) +
-  geom_abline(intercept = -0.3, linetype = "dashed") +
+  xlim(-33,-20) +
+  ylim(-33,-20) +
+  #coord_fixed(ratio = 1) +
+  geom_abline(intercept = -0.7, linetype = "dashed") + #Suryanarayan et al.2021
   geom_abline(intercept = -3.1, linetype = "dashed") +
-  annotate("text", x = -22, y = -22, angle = 45, vjust = 1.5,
-           label = bquote(Delta*{}^13*"C = -0.3 \u2030")) +
-  annotate("text", x = -22, y = -25, angle = 45, vjust = 1.5,
-           label = bquote(Delta*{}^13*"C = -3.1 \u2030")) +
-  #theme(legend.position="none")
+  annotate("text", x = -21.5, y = -22, angle = 45, vjust = 1.5,
+           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
+  annotate("text", x = -21.5, y = -24.5, angle = 45, vjust = 1.5,
+           label = bquote(Delta*{}^13*"C = -3.1 \u2030"))
 
 big_delta_C16 <-
   ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
        aes(`13C C16:0`, delta)) +
-  geom_point(size = 1, alpha = 0.9, aes(shape = sample)) +
-  #ggrepel::geom_text_repel(aes(label = C13)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
   theme_minimal() +
   labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
[…]

### 2023-01-08 analysis/code/GCFID-quant.R
changed the range for x axis

@@ -129,7 +129,7 @@ ggsave(here::here("analysis","figures", "chromatograms_four.png"),
 # plot the one with more compounds
 meth_all %>%
   filter(name == "SYG-TN13-E22-2#2") %>%
-  filter(File > 55000) %>%
+  #filter(File > 55000) %>%
   ggplot(aes(Path, File)) +
   geom_line(size = 0.3) +
   geom_text(data = highlight_SYG2_2, #ggrepel::geom_text_repel
@@ -140,8 +140,8 @@ meth_all %>%
   scale_y_continuous(labels = scales::comma_format(),
                      limits = c(0, 7500000),
                      breaks = seq(0, 7500000, 1000000)) +
-  scale_x_continuous(limits = c(20, 50),
-                     breaks = seq(20, 50, 5),
+  scale_x_continuous(limits = c(15, 50),
+                     breaks = seq(15, 50, 5),
                      expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
   labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
   theme_minimal() +

### 2023-01-08 analysis/paper/paper.Rmd
worked on the result section

--- 2022-11-26 analysis/paper/paper.Rmd
@@
 This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the vessel usage at Guijiabao.
-Lipid analysis have proven useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017].
+Lipid analysis is useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017].
 Lipid residues are the most direct evidence to investigate culinary practices and dietary patterns since specific lipid compounds can be associated with different types of foodstuffs cooked or stored in pottery vessels [@Copley2005].
@@
 
+To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples.
+
 # Results
 
-```{r tidy data}
+```{r tidy-data, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/LY_21_08_20_pot_chrom.CSV"))}
 library(tidyverse)
@@
 
-Lipid residues from our samples show that the saturated fatty acids with even number of carbons (C14:0, C16:0, and C18:0), particularly palmitic (C16:0) and stearic(C18:0), are the dominant compounds.
-Because palmitic and stearic fatty acids are normally degraded from animal fat or plant wax, the domination pattern with little concentration of other fatty acids suggests the organic compounds in our samples were highly degraded during post depositional processes.
-However, there is one sample contains rich compounds that include long-chain fatty acids, C20, C22, and C24, which were absent in other samples.
-This might indicate
+```{r highlight-compounds, cache=TRUE}
+# highlight target compounds
+highlight_GJB_four <-
+  meth_all %>%
+  mutate(retention = round(Path, 1)) %>%
+  group_by(sample, retention) %>%
+  arrange(desc(File)) %>%
+  distinct(retention, .keep_all = TRUE) %>%
+  mutate(FA = case_when(
+    retention == 21.6 ~ "C14:0",
+    retention == 23.6 ~ "C15:0",
+    retention == 25.5 ~ "C16:0",
+    retention == 27.3 ~ "C17:0",
+    retention == 28.4 ~ "C18:1",
+    retention == 29.1 ~ "C18:0",
+    retention == 32.4 ~ "C20:0",
+    retention == 35.4 ~ "C22:0")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
+  filter(File > 100000)
 
-To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples.
-Here you can see the stable carbon isotope measurements of C16:0 and C18:0 fatty acids.
-The results suggest a dominance of C3 plant for most pots, and a few shows a mixture of animal fats and C3 plants.
-The sample 5 where more compounds were preserved shows a source of C4 plant or marine animals instead.
-We think this is very interesting finding since there was just a few rice grains found at this site compared to the evidence of millet.
+highlight_IS <-
+  meth_all %>%
+  mutate(retention = round(Path, 1)) %>%
+  group_by(sample, retention) %>%
+  arrange(desc(File)) %>%
+  distinct(retention, .keep_all = TRUE) %>%
+  mutate(FA = case_when(
+    retention == 33.9 ~ "IS",
+    retention == 41.3 ~ "IS")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3"))
+
+# highlight the compounds in the potsherd with high concentration
+highlight_SYG2_2 <-
+  meth_all %>%
+  mutate(retention = round(Path, 1)) %>%
+  group_by(sample, retention) %>%
+  arrange(desc(File)) %>%
+  distinct(retention, .keep_all = TRUE) %>%
+  mutate(FA = case_when(
+    retention == 21.6 ~ "C14:0",
+    retention == 25.5 ~ "C16:0",
+    retention == 27.3 ~ "C17:0",
+    retention == 28.4 ~ "C18:1",
+    retention == 29.1 ~ "C18:0",
+    retention == 31.0 ~ "C19:0",
+    retention == 32.4 ~ "C20:0",
+    retention == 35.4 ~ "C22:0",
+    retention == 36.8 ~ "C23:0",
+    retention == 38.2 ~ "C24:0",
+    retention == 40.9 ~ "C26:0",
+    retention == 33.9 ~ "IS",
+    retention == 41.3 ~ "IS",
+    retention == 45.6 ~ "C:30",
+    retention == 47.8 ~ "C:32")) %>%
+  filter(name == "SYG-TN13-E22-2#2")
+```
+
+```{r chromatogram-four, cache=TRUE}
+# plot the first four
+chroma_four <-
+  meth_all %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
+  mutate(File = ifelse(File > 550000, 550000, File)) %>%
+  ggplot(aes(Path, File)) +
+  geom_line(size = 0.3) +
+  geom_text(data = highlight_GJB_four, #ggrepel::geom_text_repel
+            aes(label = FA, angle = 90),
+            size = 3,
+            nudge_y = 87000,
+            show.legend = FALSE) +
+  geom_text(data = highlight_IS,
+            aes(label = FA),
+            size = 3,
+            nudge_y = 40000,
+            show.legend = FALSE) +
+  geom_segment(aes(x= 41.1, xend= 41.5, y= 550000, yend= 550000), size =0.3) +
+  geom_segment(aes(x= 41.1, xend= 41.5, y= 560000, yend= 560000), size =0.3) +
+  annotate("text", x = 41.3, y = 600000, label = "IS", size =3) +
+  facet_wrap(~name, ncol = 1) +
+  scale_y_continuous(labels = scales::comma_format(),
+                     limits = c(0, 600000),
+                     breaks = seq(0, 600000, 200000)) +
+  scale_x_continuous(limits = c(20, 50),
+                     breaks = seq(20, 50, 5),
+                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
+  labs(x = "retention time", y = "relative intensity") +
+  theme_minimal() +
+  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        panel.background = element_blank(), axis.line = element_line(colour = "black"))
+
+ggsave(plot = chroma_four,
+       here::here("analysis","figures", "chromatograms_four.png"),
+       width = 7,
+       height = 8,
+       dpi = 300,
+       units = "in")
+```
+
+The GC-MS analysis for organic residues extracted from Guijiabao potsherds revealed the presence of lipid compounds, but with low amounts in most of them.
+The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid ) and C18:0 (stearic acid), and monounsaturated fatty acid  C18:1 (oleic acid), are the main compounds for most potsherds (Figure \@ref(fig:chromatograms-plots-four)).
+Those fatty acids commonly exist in most animal fat and plant oil.
+They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007].
+High concentrations of C16:0 and C18:0 fatty acids in our samples suggest that the original organic residues were highly degraded during post-depositional processes.
+In addition, most potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
+A larger content of C18:1 and C16:0 with smaller content of C18:0 could indicate an origin of plant oils [@Mayyas2018; @Copley2005].
+Although oleic acid (C18:1) has the highest concentration in most of our potsherd samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.
+
+```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao.
+Cn:x refers to fatty acid with n carbon atoms and x double bonds.
+IS indicates internal standard used for quantification, including FA C21:0 and n-alkane 34"}
+knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
+```
+
+```{r chromatogram-specific, cache=TRUE}
+# plot the specific one
+chroma_specific <-
+  meth_all %>%
+  filter(name == "SYG-TN13-E22-2#2") %>%
+  ggplot(aes(Path, File)) +
+  geom_line(size = 0.3) +
+  geom_text(data = highlight_SYG2_2, #ggrepel::geom_text_repel
+            aes(label = FA),
+            size = 3,
+            nudge_y = 250000,
+            show.legend = FALSE) +
+  scale_y_continuous(labels = scales::comma_format(),
+                     limits = c(0, 7500000),
+                     breaks = seq(0, 7500000, 1000000)) +
+  scale_x_continuous(limits = c(15, 50),
+                     breaks = seq(15, 50, 5),
+                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
+  labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
+  theme_minimal() +
+  theme(plot.title = element_text(hjust = 0.5),
+        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+        panel.background = element_blank(), axis.line = element_line(colour = "black"))
+
+ggsave(plot = chroma_specific,
+       here::here("analysis","figures", "chromatograms_specific.png"),
+       width = 8,
+       height = 5,
+       dpi = 300,
+       units = "in")
+```
+
+Long-chain fatty acids from C20:0 to C32:0 are present in only one pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)).
+The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005].
+However, it could be also associated with an origin of degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
+Overall, the GC-MS results demonstrate evidence of degraded animal fats or plant oil.
+
+```{r chromatograms-specific, echo=FALSE, fig.cap= "GC-MS chromatograms of the potshard with the presence of long-chain fatty acids.
+Cn:x refers to fatty acid with n carbon atoms and x double bonds.
+IS indicates internal standard used for quantification, including FA C21:0 and n-alkane 34"}
+knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
+```
+
+```{r isotopic-analysis}
+library(tidyverse)
+isotope_data <-
+  readxl::read_excel(here::here("analysis","data", "raw_data","12_21_potshard_13C.xlsx"),
+                     sheet = "Summary", col_names =  FALSE)
+
+# filter the Meth corrected data
+delta_meth_cor <-
+  isotope_data %>%
+  select(1, 7, 9) %>%
+  slice(2:8) %>%
+  dplyr::mutate(...1 = ifelse(is.na(...1), paste0("C13"), ...1))
+
+names(delta_meth_cor) <- delta_meth_cor[1,]
+
+delta_meth_cor <-
+  delta_meth_cor %>%
+  slice(2:7) %>%
+  dplyr::mutate(across(starts_with("13C"), as.numeric)) %>%
+  dplyr::mutate(delta = `13C C18:0`-`13C C16:0`) %>%
+  mutate(sample = case_when(
+    C13 %in% c("Pot 1") ~ "CDG-062",
+    C13 %in% c("Pot 2") ~ "SYG-3",
+    C13 %in% c("Pot 3") ~ "SYG-TN13-E22-2#1",
+    C13 %in% c("Pot 4") ~ "SYG-TN13-E22-1",
+    C13 %in% c("Pot 5") ~ "SYG-TN13-E22-2#2",
+    C13 %in% c("Pot 6") ~ "SYG-TN13-E23-3")) %>%
+  mutate(sample = factor(sample, levels = c("SYG-TN13-E22-1",
+                                            "SYG-TN13-E22-2#1",
+                                            "SYG-TN13-E22-2#2",
+                                            "SYG-TN13-E23-3",
+                                            "SYG-3",
+                                            "CDG-062")))
+
+delta_min <- round(min(delta_meth_cor$delta), 2)
+delta_max <- round(max(delta_meth_cor$delta), 2)
+
+# plot the carbon isotopes 
+isotope_C16_C18 <-
+  ggplot(delta_meth_cor, #delta_meth_cor/delta_blank_cor
+       aes(`13C C16:0`,`13C C18:0`)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
+  #ggrepel::geom_text_repel(aes(label = C13)) +
+  theme_minimal() +
+  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
+       y = bquote(delta*{}^13*"C 18:0 \u2030")) +
+  xlim(-33,-20) +
+  ylim(-33,-20) +
+  #coord_fixed(ratio = 1) +
+  geom_abline(intercept = -0.7, linetype = "dashed") + #Suryanarayan et al.2021
+  geom_abline(intercept = -3.1, linetype = "dashed") +
+  annotate("text", x = -21.5, y = -22, angle = 45, vjust = 1.5,
+           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
+  annotate("text", x = -21.5, y = -24.5, angle = 45, vjust = 1.5,
+           label = bquote(Delta*{}^13*"C = -3.1 \u2030"))
+
+big_delta_C16 <-
+  ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
+       aes(`13C C16:0`, delta)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
+  theme_minimal() +
+  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
+       y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
+  scale_x_continuous(limits = c(-33, -20),
+                     breaks = seq(-33, -20, 2)) +
+  ylim(-2.5, 2) +
+  geom_hline(yintercept = -0.7, linetype = "dashed") +
+  annotate("text", x = -22, y = -0.5,
+           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
+  annotate("text", x = -31, y = 1.8,
+           label = "C3-diet") +
+  annotate("segment", x = -29, y = 1.8, xend = -24, yend = 1.8,
+           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
+  annotate("text", x = -22, y = 1.8,
+           label = "C4-diet")
+
+library(cowplot)
+combined_isotopic_plots <-
+  plot_grid(isotope_C16_C18,
+          big_delta_C16,
+          labels = c('A', 'B'),
+          ncol =1,
+          align = 'v',
+          label_size = 12)
+
+ggsave(here::here("analysis","figures","isotopic_plots.png"),
+       width = 6,
+       height = 8,
+       dpi = 300,
+       units = "in")
+```
+
+To identify the origins of degraded animal fats, we conducted isotope analysis to obtain the δ^13^C values of C16:0 and C18:0 using GC-C-IRMS.
+The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on different ways where dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
+To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the origins of degraded fats [@Evershed2008; @Salque2013].
+The isotopic results show the Δ13C values in our samples range from `r delta_min` ‰ to `r delta_max` ‰ that were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].
+
+Most of the potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A).
+The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
+There are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003].
+Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) that falls within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020].
+In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants.
+There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.
+
+```{r isotopic-plots, echo=FALSE, fig.cap= ""}
+knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
+```
 
@@
 # Conclusion
-Our results show a mixture of C3 plants and terrestrial animals for most pots with one pot showing more C4 plant sources.
+Our investigation on the ceramics excavated from Southwest China suggests that a mixture of C3 plants and ruminant animals could be the food sources in this region during the Neolithic transition.
 This case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.

### 2023-01-10 analysis/paper/paper.Rmd
edited the result section

@@
 
-The GC-MS analysis for organic residues extracted from Guijiabao potsherds revealed the presence of lipid compounds, but with low amounts in most of them.
-The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid ) and C18:0 (stearic acid), and monounsaturated fatty acid  C18:1 (oleic acid), are the main compounds for most potsherds (Figure \@ref(fig:chromatograms-plots-four)).
-Those fatty acids commonly exist in most animal fat and plant oil.
+The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them.
+The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid) and C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds for most potsherds (Figure \@ref(fig:chromatograms-plots-four)).
+Those fatty acids commonly exist in animal fat and plant oil.
 They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007].
 High concentrations of C16:0 and C18:0 fatty acids in our samples suggest that the original organic residues were highly degraded during post-depositional processes.
-In addition, most potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
-A larger content of C18:1 and C16:0 with smaller content of C18:0 could indicate an origin of plant oils [@Mayyas2018; @Copley2005].
+In addition, some potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
+A larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005].
 Although oleic acid (C18:1) has the highest concentration in most of our potsherd samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.
@@
 The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005].
-However, it could be also associated with an origin of degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
-Overall, the GC-MS results demonstrate evidence of degraded animal fats or plant oil.
+However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
+Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.
 
@@
                                             "CDG-062")))
+
+min_carbon_isotope <- round(min(delta_meth_cor$`13C C18:0`), 2)
+max_carbon_isotope <- round(max(delta_meth_cor$`13C C16:0`), 2)
 
@@
 
-To identify the origins of degraded animal fats, we conducted isotope analysis to obtain the δ^13^C values of C16:0 and C18:0 using GC-C-IRMS.
+To identify the origins of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions δ^13^C values of C16:0 and C18:0 using GC-C-IRMS.
 The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on different ways where dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
-To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the origins of degraded fats [@Evershed2008; @Salque2013].
-The isotopic results show the Δ13C values in our samples range from `r delta_min` ‰ to `r delta_max` ‰ that were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].
+To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
+The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰.
+The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].
 
-Most of the potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A).
+The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A).
 The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
-There are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003].
-Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) that falls within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020].
-In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants.
+It should be noted that there are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003].
+Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020].
+In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants.
 There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.
 
-```{r isotopic-plots, echo=FALSE, fig.cap= ""}
+```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C16:0 and C18:0. B: The Δ13C (δ13C18:0-16:0) values plotted against δ13C16:0 values"}
 knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
@@
 Our next steps is to increase sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.
+
+1000 BC horse
+Gaushan-90 pig, plant-most are rice: 95% and a few millet
 

### 2023-01-11 analysis/code/isotope-analysis.R
added annotations to the plot

@@ -104,7 +104,7 @@ ggsave(here::here("delta_C16_C18_remove_pot6.png"),
 
 # plot the carbon isotopes by following the method in Salque et al. (2013)
 isotope_C16_C18 <-
-  ggplot(delta_meth_cor, # delta_meth_cor/delta_blank_cor
+  ggplot(delta_meth_cor, #delta_meth_cor/delta_blank_cor
        aes(`13C C16:0`,`13C C18:0`)) +
   geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
   #ggrepel::geom_text_repel(aes(label = C13)) +
@@ -128,11 +128,18 @@ big_delta_C16 <-
   theme_minimal() +
   labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
        y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
-  xlim(-33,-20) +
+  scale_x_continuous(limits = c(-33, -20),
+                     breaks = seq(-33, -20, 2)) +
   ylim(-2.5, 2) +
   geom_hline(yintercept = -0.7, linetype = "dashed") +
   annotate("text", x = -22, y = -0.5,
-           label = bquote(Delta*{}^13*"C = -0.7 \u2030"))
+           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
+  annotate("text", x = -31, y = 1.8,
+           label = "C3-diet") +
+  annotate("segment", x = -29, y = 1.8, xend = -24, yend = 1.8,
+           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
+  annotate("text", x = -22, y = 1.8,
+           label = "C4-diet")
 
 library(cowplot)
 combined_isotopic_plots <-

### 2023-01-11 analysis/code/GCFID-quant.R
edited code for chromatograms

@@ -24,13 +24,15 @@ meth_all <-
     sample %in% c("210729_003") ~ "SYG-TN13-E22-2#1",
     sample %in% c("210729_004") ~ "SYG-TN13-E22-1",
     sample %in% c("210729_005") ~ "SYG-TN13-E22-2#2",
-    sample %in% c("210729_006") ~ "SYG-TN13-E23-3")) %>%
+    sample %in% c("210729_006") ~ "SYG-TN13-E23-3",
+    sample %in% c("210729_008") ~ "blank"))%>%
   mutate(name = factor(name, levels = c("SYG-TN13-E22-1",
                                         "SYG-TN13-E22-2#1",
                                         "SYG-TN13-E22-2#2",
                                         "SYG-TN13-E23-3",
                                         "SYG-3",
-                                        "CDG-062")))
+                                        "CDG-062",
+                                        "blank")))
 
 # highlight the compounds
 highlight_GJB_four <-
@@ -50,7 +52,7 @@ highlight_GJB_four <-
     retention == 29.1 ~ "C18:0",
     retention == 32.4 ~ "C20:0",
     retention == 35.4 ~ "C22:0")) %>%
-  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "blank")) %>%
   filter(File > 100000)
 
 highlight_IS <-
@@ -91,7 +93,7 @@ highlight_SYG2_2 <-
 
 # plot the first four
 meth_all %>%
-  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3","blank")) %>%
   mutate(File = ifelse(File > 550000, 550000, File)) %>%
   ggplot(aes(Path, File)) +
   geom_line(size = 0.3) +

### 2023-01-11 analysis/paper/paper.Rmd
added a table for FA composition to the Rmd file

@@
 
-```{r chromatogram-four, cache=TRUE}
+```{r chromatogram-four, cache= TRUE}
 # plot the first four
@@
 
-The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them.
+The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them (Table \@ref(tab:FA-composition-table)).
 The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid) and C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds for most potsherds (Figure \@ref(fig:chromatograms-plots-four)).
@@
 Although oleic acid (C18:1) has the highest concentration in most of our potsherd samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.
+
+```{r FA-composition-table, cache= TRUE}
+quant_FA <-
+  readxl::read_excel(here::here("analysis", "data", "raw_data", "210824_FAMES_pot_LW_publication.xlsx"), 
+                     skip = 2, col_names = TRUE)
+
+names(quant_FA)[1] <- "number"
+names(quant_FA)[2] <- "sample"
+
+quant_FA_tidy <-
+  quant_FA %>%
+  pivot_longer(!number&!sample,
+               names_to = "FA", 
+               values_to = "weight") %>% 
+  select(-number) %>% 
+  filter(str_starts(FA, "ng") & !sample =="") %>% 
+  mutate(FA = str_remove(FA, "ng "),
+         weight = as.character(round(weight/1000, 2))) %>% 
+  mutate(weight = ifelse(is.na(weight)|weight == "0", "-", weight)) %>% 
+  pivot_wider(names_from = sample, 
+              values_from = weight) %>% 
+  select("FA", "SYG-TN13-E22-1", "SYG-TN13-E22-2#1", "SYG-TN13-E22-2#2",
+         "SYG-TN13-E23-3", "SYG-3", "CDG-062") %>% 
+  filter(!FA == "C34:0"&!FA == "C21:0")
+
+knitr::kable(quant_FA_tidy,
+             caption = "Quantitative data of fatty acid (FA) composition (ug) of potsherds from Guijiabao and Gaushan that are named starting with 'SYG' and 'CDG' respectively")
+```
 

### 2023-01-12 analysis/paper/paper.Rmd
edited the code chunk for chromatogram

@@
 # highlight target compounds
-highlight_GJB_four <-
+highlight_common <-
   meth_all %>%
@@
     retention == 29.1 ~ "C18:0",
-    retention == 32.4 ~ "C20:0",
+    retention == 32.3 ~ "C20:0",
     retention == 35.4 ~ "C22:0")) %>%
-  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062")) %>%
   filter(File > 100000)
@@
     retention == 41.3 ~ "IS")) %>%
-  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3"))
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062"))
 
@@
     retention == 41.3 ~ "IS",
-    retention == 45.6 ~ "C:30",
-    retention == 47.8 ~ "C:32")) %>%
+    retention == 45.6 ~ "C30:0",
+    retention == 47.8 ~ "C32:0")) %>%
   filter(name == "SYG-TN13-E22-2#2")
@@
   meth_all %>%
-  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
-  mutate(File = ifelse(File > 550000, 550000, File)) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062")) %>%
+  mutate(File = ifelse(File > 1000000, 1000000, File)) %>%
   ggplot(aes(Path, File)) +
   geom_line(size = 0.3) +
-  geom_text(data = highlight_GJB_four, #ggrepel::geom_text_repel
-            aes(label = FA, angle = 90),
+  ggrepel::geom_text_repel(data = highlight_common, #ggrepel::geom_text_repel
+            aes(label = FA, angle = 0),
             size = 3,
-            nudge_y = 87000,
+            nudge_y = 120000,
             show.legend = FALSE) +
@@
             size = 3,
-            nudge_y = 40000,
+            nudge_y = 100000,
             show.legend = FALSE) +
-  geom_segment(aes(x= 41.1, xend= 41.5, y= 550000, yend= 550000), size =0.3) +
-  geom_segment(aes(x= 41.1, xend= 41.5, y= 560000, yend= 560000), size =0.3) +
+  geom_segment(aes(x= 41.1, xend= 41.5, y= 1000000, yend= 1000000), size =0.3) +
+  geom_segment(aes(x= 41.1, xend= 41.5, y= 1030000, yend= 1030000), size =0.3) +
   annotate("text", x = 41.3, y = 600000, label = "IS", size =3) +
@@
   scale_y_continuous(labels = scales::comma_format(),
-                     limits = c(0, 600000),
-                     breaks = seq(0, 600000, 200000)) +
-  scale_x_continuous(limits = c(20, 50),
-                     breaks = seq(20, 50, 5),
+                     limits = c(0, 1100000),
+                     breaks = seq(0, 1100000, 200000)) +
+  scale_x_continuous(limits = c(20, 45),
+                     breaks = seq(20, 45, 5),
                      expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
@@
        here::here("analysis","figures", "chromatograms_four.png"),
-       width = 7,
+       width = 6,
        height = 8,
@@
 The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them (Table \@ref(tab:FA-composition-table)).
-The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid) and C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds for most potsherds (Figure \@ref(fig:chromatograms-plots-four)).
+The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds in most potsherds (Figure \@ref(fig:chromatograms-plots-four)).
 Those fatty acids commonly exist in animal fat and plant oil.
@@
 A larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005].
-Although oleic acid (C18:1) has the highest concentration in most of our potsherd samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.
+Although oleic acid (C18:1) is the most abundant compound for most of our samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.
 
@@
          "SYG-TN13-E23-3", "SYG-3", "CDG-062") %>% 
-  filter(!FA == "C34:0"&!FA == "C21:0")
+  filter(!FA == "C34:0" & !FA == "C21:0")
 
@@
 
-```{r chromatogram-specific, cache=TRUE}
+```{r chromatogram-specific, cache= TRUE}
 # plot the specific one
@@
   meth_all %>%
-  filter(name == "SYG-TN13-E22-2#2") %>%
+  filter(name %in% c("SYG-TN13-E22-2#2")) %>% 
   ggplot(aes(Path, File)) +
@@
 
-```{r isotopic-analysis}
+```{r isotopic-analysis, cache= TRUE}
 library(tidyverse)

### 2023-01-13 analysis/paper/paper.Rmd
worked on the method section

@@
 ## Samples
-We selected five ceramic fragments (Table XX) from Guijiabao for organic residue analysis, in which four samples were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season.
-To further investigate the content and use of pottery vessels in an earlier period, we focus on potsherds excavated from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.).
-During this period, the common vessel forms include jars, urns, bowls, and basins.
+We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX).
+The four ceramic fragments were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season.
+To understand the content and use of pottery vessels in the Yanyan region in an earlier period, we analyze potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison.
+During the early Neolithic period, the common vessel forms include jars, urns, bowls, and basins.
 For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022].
-Our samples are jar fragments from those two common types supposed to be used for food serving or storage according to their archaeological context that suggests living area [].
+Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage [].
 
@@
 After drying under a stream of nitrogen, the derivitized sample was dissolved in toluene (20 uL) prior to GC-MS.
+Internal standards were added before and after the total lipid extraction.
 
@@
 The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream.
-About 5% aliquot dissolved in toluene was analyzed by GC-FID and the remaining sample was analyzed by GC-C-IRMS.
+About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification and the remaining was analyzed by GC-C-IRMS.
 
-To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples.
+The GC-MS analysis was performed using an Agilent 7683 B series chromatograph attached to an Agilent 5975 inert mass selective detector.
+Samples were injected in splitless mode at 320 °C and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness).
+Helium carrier is in constant flow rate of 1.24 ml/min.
+The temperature of the oven was started from 90 °C for 1 minute, and increased at a rate of 5 °C/min to reach 320 °C , where it was held for 23 min.
 

### 2023-01-13 analysis/paper/paper.Rmd
added details of running parameters to the method section

@@
 
-The GC-MS analysis was performed using an Agilent 7683 B series chromatograph attached to an Agilent 5975 inert mass selective detector.
+The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector.
 Samples were injected in splitless mode at 320 °C and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness).
-Helium carrier is in constant flow rate of 1.24 ml/min.
-The temperature of the oven was started from 90 °C for 1 minute, and increased at a rate of 5 °C/min to reach 320 °C , where it was held for 23 min.
+Helium carrier is in constant flow rate of 1.3 ml/min.
+The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 min.
+The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph with an FID detector.
+The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface.
+The sample injection volume was 1 ul in toluene and was injected in splitless mode.
+Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set from 90 °C for 1.5 minutes, and then raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold.
+Helium was used as a gas carrier in constant flow mode at 1.24 ml/minute.
 

### 2023-01-13 analysis/code/site-regional-map-Ch.R
added the location of Gaoshan to the map

@@ -21,9 +21,9 @@ world_points <- cbind(world, st_coordinates(st_point_on_surface(world$geometry))
 
 # add site location
 site_location <-
-  data.frame(location = c("Guijuabao", "Chengdu"),
-             lon = c(101.3617, 104.0587),
-             lat = c(27.2657, 30.5899))
+  data.frame(location = c("Guijuabao", "Gaoshan", "Chengdu"),
+             lon = c(101.3617, 103.3446, 104.0587),
+             lat = c(27.2657, 30.2709, 30.5899))
 
 China_SE_Asia <-
   ggplot(data = world) +
@@ -50,7 +50,7 @@ library(ggmap)
 # we don't want to download every time, so let's save the map locally
 # from https://stackoverflow.com/a/52710855/1036500
 GJB_map <- ggmap(get_stamenmap(rbind(as.numeric(c(100.5, 26.5,
-                                                  102.5, 28.5))), zoom = 10))
+                                                  104.5, 30.5))), zoom = 10))
 # saveRDS(tw_map, here("analysis", "data", "raw_data", "tw_map.rds"))
 # tw_map <- readRDS(here("analysis", "data", "raw_data", "tw_map.rds"))
 pg <- ggplot_build(GJB_map)
@@ -73,8 +73,8 @@ China_map_with_site <-
                   check.overlap = TRUE) +
   #annotate(geom = "text", x = 103.95, y = 13.0, label = "Tonle Sap\nLake",
            #fontface = "italic", color = "blue", size = 2) +
-  coord_sf(xlim = c(100.5, 102.5),
-           ylim = c(26.5, 28.5),
+  coord_sf(xlim = c(100.5, 104.5),
+           ylim = c(26.5, 30.5),
            expand = FALSE) +
   #scale_x_continuous(breaks = c(121.0, 121.5, 122.0),
   #limits = c(120.9, 122.7)) +
@@ -121,7 +121,7 @@ ggplot() +
   theme_void() +
   theme(plot.background = element_rect(color = "white", fill = "white"))
 
-ggsave(here::here("analysis","figures", "prasat-location-map.png"),
+ggsave(here::here("analysis","figures", "Chengdu-sites-map.png"),
        width = 6,
        height = 3,
        dpi = 300,

### 2023-01-13 analysis/code/site-regional-map-Ch.R
worked on the site map

@@ -17,18 +17,18 @@ world_points <- sf::st_point_on_surface(world)
 world_points <- cbind(world, st_coordinates(st_point_on_surface(world$geometry))) %>%
   filter(brk_name %in% c("Thailand", "China", "Myanmar",
                          "Laos", "Vietnam", "Cambodia")) %>%
-  mutate(Y = Y+4, X = X-2)
+  mutate(Y = Y+2, X = X-1)
 
 # add site location
 site_location <-
-  data.frame(location = c("Guijuabao", "Gaoshan", "Chengdu"),
-             lon = c(101.3617, 103.3446, 104.0587),
-             lat = c(27.2657, 30.2709, 30.5899))
+  data.frame(location = c("Guijuabao", "Gaoshan"), # Chengdu(104.0587, 30.5899)
+             lon = c(101.3617, 103.3446),
+             lat = c(27.2657, 30.2709))
 
 China_SE_Asia <-
   ggplot(data = world) +
   geom_sf( fill= "antiquewhite") +
-  geom_rect(xmin = 100.5, xmax = 102.5, ymin = 26.5, ymax = 28.5,
+  geom_rect(xmin = 100, xmax = 106, ymin = 26, ymax = 32,
             fill = NA, colour = "red", size = 0.5) +
   geom_shadowtext(data= world_points,
                   aes(x = X, y = Y,
@@ -39,9 +39,9 @@ China_SE_Asia <-
                   position = position_nudge(y = - 1.7, x = 0.5)) +
   #annotate(geom = "text", x = 102.3, y = 9.4, label = "Gulf of\nTailand",
            #fontface = "italic", color = "grey22", size = 2) +
-  coord_sf(xlim = c(85, 115), ylim = c(16, 42), expand = FALSE) + #add datum = NA to remove
-  scale_x_continuous(breaks = seq(80, 120, by = 10)) +
-  scale_y_continuous(breaks = seq(10, 50, by = 10)) +
+  coord_sf(xlim = c(80, 123), ylim = c(9, 45), expand = FALSE) + #add datum = NA to remove
+  scale_x_continuous(breaks = seq(80, 125, by = 10)) +
+  scale_y_continuous(breaks = seq(5, 45, by = 10)) +
   theme(axis.title.x = element_blank(),
         axis.title.y = element_blank())
 
@@ -49,8 +49,8 @@ China_SE_Asia <-
 library(ggmap)
 # we don't want to download every time, so let's save the map locally
[…]

### 2023-01-14 analysis/paper/paper.Rmd
added details of GC-FID to the method section

@@
 As an early Neolithic site, Guijiabao (ca.
-4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Plateau, and the Yunnan-Guizhou Plateau.
+4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau.
 The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@citation], different from the abundant rice grains found at other Neolithic sites in that area.
@@
 Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022].
-The archaeological sites in the Yanyuan Basin, therefore, are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan.
+Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan.
 
@@
 The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively.
-The second fraction was methylated to produce fatty acid methyl esters (FAMEs) using 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight.
+The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight.
 The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream.
@@
 The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 min.
-The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph with an FID detector.
+The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode.
+The carrier gas was helium at 1.6 ml/min.
+The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode.
+The GC oven was held at 90°C for 1.3 minutes, and then raised to 320°C at a rate of 5°C/min.
+The final temperature was held for 22 minutes.
 The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface.
@@
   meth_all %>%
-  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", 
+                     "SYG-TN13-E23-3", "CDG-062")) %>%
   mutate(File = ifelse(File > 1000000, 1000000, File)) %>%
@@
 
-```{r FA-composition-table, cache= TRUE}
+```{r FA-composition-table, cache= TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/210824_FAMES_pot_LW_publication.xlsx"))}
 quant_FA <-
@@
 
+net sinkers
 1000 BC horse

### 2023-01-14 analysis/code/site-regional-map-Ch.R
added texts to the map

@@ -15,8 +15,7 @@ world <- ne_countries(scale = "medium", returnclass = "sf")
 # class(world)
 world_points <- sf::st_point_on_surface(world)
 world_points <- cbind(world, st_coordinates(st_point_on_surface(world$geometry))) %>%
-  filter(brk_name %in% c("Thailand", "China", "Myanmar",
-                         "Laos", "Vietnam", "Cambodia")) %>%
+  filter(brk_name %in% c("Thailand", "China", "Myanmar", "Laos", "Vietnam", "Cambodia")) %>%
   mutate(Y = Y+2, X = X-1)
 
 # add site location
@@ -72,7 +71,9 @@ China_map_with_site <-
                   position = position_nudge(y = - 0.2),
                   check.overlap = TRUE) +
   annotate(geom = "text", x = 104, y = 30.7, label = "Chengdu\nPlain",
-           fontface = "italic", color = "grey15", size = 2) +
+           fontface = "italic", color = "grey10", size = 2) +
+  annotate(geom = "text", x = 105.2, y = 30, label = "Sichuan\nBasin",
+           fontface = "italic", color = "grey10", size = 2) +
   coord_sf(xlim = c(100, 106),
            ylim = c(26, 32),
            expand = FALSE) +

### 2023-01-14 analysis/code/site-regional-map-Ch.R
Edited the site map

@@ -15,8 +15,10 @@ world <- ne_countries(scale = "medium", returnclass = "sf")
 # class(world)
 world_points <- sf::st_point_on_surface(world)
 world_points <- cbind(world, st_coordinates(st_point_on_surface(world$geometry))) %>%
-  filter(brk_name %in% c("Thailand", "China", "Myanmar", "Laos", "Vietnam", "Cambodia")) %>%
-  mutate(Y = Y+2, X = X-1)
+  filter(brk_name %in% c("Thailand", "China", "Myanmar", "Laos",
+                         "Vietnam", "Cambodia", "Bangladesh", "Bhutan")) %>%
+  mutate(Y = ifelse(brk_name %in% c("Bhutan", "Bangladesh"), Y+2, Y)) %>%
+  mutate(Y = ifelse(brk_name %in% c("Thailand", "Myanmar"), Y+3.5, Y))
 
 # add site location
 site_location <-
@@ -27,7 +29,7 @@ site_location <-
 China_SE_Asia <-
   ggplot(data = world) +
   geom_sf( fill= "antiquewhite") +
-  geom_rect(xmin = 100, xmax = 106, ymin = 26, ymax = 32,
+  geom_rect(xmin = 99.5, xmax = 105.5, ymin = 25.5, ymax = 31.5,
             fill = NA, colour = "red", size = 0.5) +
   geom_shadowtext(data= world_points,
                   aes(x = X, y = Y,
@@ -48,8 +50,8 @@ China_SE_Asia <-
 library(ggmap)
 # we don't want to download every time, so let's save the map locally
 # from https://stackoverflow.com/a/52710855/1036500
-GJB_map <- ggmap(get_stamenmap(rbind(as.numeric(c(100, 26,
-                                                  106, 32))), zoom = 10))
+#GJB_map <- ggmap(get_stamenmap(rbind(as.numeric(c(99, 25,
+                                                  #106, 32))), zoom = 10))
 # saveRDS(tw_map, here("analysis", "data", "raw_data", "tw_map.rds"))
 # tw_map <- readRDS(here("analysis", "data", "raw_data", "tw_map.rds"))
 pg <- ggplot_build(GJB_map)
@@ -67,15 +69,21 @@ China_map_with_site <-
                       label = location),
                   color='black',
                   bg.colour='white',
-                  size = 2.6,
+                  size = 2.8,
                   position = position_nudge(y = - 0.2),
                   check.overlap = TRUE) +
[…]

### 2023-01-14 analysis/paper/paper.Rmd
added a code chunk for maps to the rmd

@@
 # Geographical and Cultural Background of Guijiabao in Yanyuan Basin
+
+```{r SW-china-map, fig.cap = "The archaeological sites, Guijiabao and Gaoshan, and other locations in Southwest China that are mentioned in this study. Map data is from naturalearthdata.com"}
+library(here)
+knitr::include_graphics(here("analysis", "figures", "Sichuan-sites-map.png"))
+```
+
 <!--Info about Yanyuan area-->
-The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to Yunnan-Guizhou Plateau.
+The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)).
 Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022].

### 2023-01-18 analysis/paper/paper.Rmd
worked on the background section

@@
 <!--Info about Yanyuan area-->
-The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)).
+The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)).
 Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022].
@@
 <!--Info about Guijiabao-->
-Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the Basin is the only site with continuous occupation from the early Neolithic period to the historical period.
+Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is an unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
+Investigation in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and remains, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long term-occupation [@Hao2022].
 The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022].
-The Neolithic period can be further separated into two subphases, 5000-4500 cal.
+For the Neolithic period, it can be further separated into two subphases, 5000-4500 cal.
 BP and 4500-3700 cal.
-BP, according to the cultural remains and pottery characteristics.
-Recent archaeological evidence indicates the spread of millet-rice mixed economy to southwest Sichuan can be dated back to around 5,000 BP.
+BP, according to the artifact assemblages and ceramic forms and styles.
+The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet can be traced back to 5000 B.P and a mixed cultivation of millet and rice in a later period (3000–1700 B.C) [@Guijiabao2014; @Hao2022]
 
@@
 ## Methods
-In this study, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID).
+For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID).
 To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).

### 2023-01-18 analysis/paper/paper.Rmd
worked on the discussion section

@@
 <!--Info about Guijiabao-->
-Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is an unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
-Investigation in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and remains, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long term-occupation [@Hao2022].
+Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is a unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
+Investigation in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022].
 The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022].
@@
 BP, according to the artifact assemblages and ceramic forms and styles.
-The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet can be traced back to 5000 B.P and a mixed cultivation of millet and rice in a later period (3000–1700 B.C) [@Guijiabao2014; @Hao2022]
+The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet can be traced back to 5000 B.P and mixed cultivation of millet and rice in a later period (3000–1700 B.C) [@Guijiabao2014; @Hao2022].
+
+<!--Info about Gaoshan-->
+Gaoshan site is located in the southwestern part of Chengdu Plain.
 
@@
 For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022].
-Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage [].
+Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage due to the archaeological context of settlement [@Hao2022; @Guijiabao2017a].
 
@@
 # Discussion
-The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source.
+
+Our preliminary results show that the food categories indicated by potsherds at Guijiabao and Gaushan could be mainly dominated by terrestrial animals, such as ruminant animals.
+Although the yield of lipids in most of our potsherd samples is generally scarce, one potsherd demonstrates a relatively rich composition of long-chain fatty acids which confirms their usage as a cooking or storage pot.
+A larger amount of C16 and C18 fatty acids compared to other types of fatty acids indicate that the organic residues in our potsherds were highly degraded over the past 4500 years.
+It is also striking to find that one potsherd of Guijiabao contains residues that might originate from C3 plants.
+
+show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source.
 Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously.

### 2023-01-18 analysis/code/site-regional-map-Ch.R
code for saving spatial data

@@ -50,11 +50,11 @@ China_SE_Asia <-
 library(ggmap)
 # we don't want to download every time, so let's save the map locally
 # from https://stackoverflow.com/a/52710855/1036500
-#GJB_map <- ggmap(get_stamenmap(rbind(as.numeric(c(99, 25,
-                                                  #106, 32))), zoom = 10))
-# saveRDS(tw_map, here("analysis", "data", "raw_data", "tw_map.rds"))
-# tw_map <- readRDS(here("analysis", "data", "raw_data", "tw_map.rds"))
-pg <- ggplot_build(GJB_map)
+GJB_map <- ggmap(get_stamenmap(rbind(as.numeric(c(99, 25,
+                                                  106, 32))), zoom = 10))
+#saveRDS(GJB_map, here("analysis", "data", "raw_data", "SW_china_map.rds"))
+#SW_china_map <- readRDS(here("analysis", "data", "raw_data", "SW_china_map.rds"))
+pg <- ggplot_build(SW_china_map)
 
 China_map_with_site <-
  GJB_map +

### 2023-01-20 analysis/paper/paper.Rmd
editing the background section

@@
 Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is a unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
-Investigation in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022].
-The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022].
+Survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022].
+The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP) [@Hao2022].
 For the Neolithic period, it can be further separated into two subphases, 5000-4500 cal.
 BP and 4500-3700 cal.
-BP, according to the artifact assemblages and ceramic forms and styles.
-The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet can be traced back to 5000 B.P and mixed cultivation of millet and rice in a later period (3000–1700 B.C) [@Guijiabao2014; @Hao2022].
+BP, according to the content of artifact assemblages and ceramic forms and styles.
+The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet that can be traced back to 5000 BC and mixed cultivation of millet and rice in a later period (3000–1700 BC) [@Guijiabao2014; @Hao2022].
 
 <!--Info about Gaoshan-->
-Gaoshan site is located in the southwestern part of Chengdu Plain.
+Gaoshan site is located in the southwestern part of the Chengdu Plain.
+Excavations at Gaoshan unearthed house foundations, burials, ashpits with abundant materials that could represent the earliest phases of the Baodun culture [citations].
 

### 2023-01-21 analysis/paper/paper.Rmd
corrected a reference

@@
 Survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022].
-The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP) [@Hao2022].
-For the Neolithic period, it can be further separated into two subphases, 5000-4500 cal.
-BP and 4500-3700 cal.
+The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (ca.
+5000-3700 BP), the Bronze Age (ca.
+3200-2700 BP), and the late North Song to early Ming Dynasties (ca.
+1200-500 BP) [@Hao2022].
+For the Neolithic period, it can be further separated into two subphases, 5000-4500 ca.
+BP and 4500-3700 ca.
 BP, according to the content of artifact assemblages and ceramic forms and styles.
-The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet that can be traced back to 5000 BC and mixed cultivation of millet and rice in a later period (3000–1700 BC) [@Guijiabao2014; @Hao2022].
+
+Among those various ceramic styles, the decorations such as incised geometric patterns and impressed dot patterns are the common styles that have also been found in western Yunnan area, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a].
+The archaeobotanical remains at Guijiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet that can be traced back to 5000 BC with no finding of rice grains based on the flotation results in 2015 [@Guijiabao2016].
 
 <!--Info about Gaoshan-->
-Gaoshan site is located in the southwestern part of the Chengdu Plain.
-Excavations at Gaoshan unearthed house foundations, burials, ashpits with abundant materials that could represent the earliest phases of the Baodun culture [citations].
+The site of Gaoshan is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
+Excavations at Gaoshan revealed abundant pottery, stone artifacts, and features such as house foundations, burials, and middens.
+According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earlies Neolithic culture (ca.
+4500-4000 BP)in Chengdu Plain [citations].
 

### 2023-01-23 analysis/paper/paper.Rmd
worked on the background section

--- 2023-01-18 analysis/paper/paper.Rmd
@@
 ---
-title: "Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis"
+title: "Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis"
 author:
@@
 4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau.
-The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@citation], different from the abundant rice grains found at other Neolithic sites in that area.
+The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@Guijiabao2016], different from the abundant rice grains found at other Neolithic sites in that area.
 Although the plant remains suggest the possible domestication of millet in the early Neolithic, we still lack direct evidence of culinary practices during the Neolithic transition.
@@
 
-<!--Info about Yanyuan area-->
-The Yanyuan area in Sichuan province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)).
+<!--Yanyuan area-->
+The Yanyuan area in Sichuan Province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)).
 Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022].
-Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including northwest China, the Sichuan area, and northwest Yunnan.
+Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including the Sichuan basin, northwest China, and northwest Yunnan.
 
-<!--Info about Guijiabao-->
+<!--Guijiabao site-->
 Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is a unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
-Investigation in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022].
-The chronology of Guijiabao can be divided into three temporal phases based on 19 radiocarbon dates from plant remains and a human bone, including the Neolithic period (5000-3700 BP.), the Bronze Age (3200-2700 BP.), and the late North Song to early Ming Dynasties (1200-500 BP.) [@Hao2022].
-For the Neolithic period, it can be further separated into two subphases, 5000-4500 cal.
-BP and 4500-3700 cal.
-BP, according to the artifact assemblages and ceramic forms and styles.
-The archaeobotanical remains found at Guidiabao reveal the evidence of crop cultivation dominated by foxtail and broomcorn millet can be traced back to 5000 B.P and mixed cultivation of millet and rice in a later period (3000–1700 B.C) [@Guijiabao2014; @Hao2022].
+A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022].
+The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (ca.
+5000-3700 BP), the Bronze Age (ca.
+3200-2700 BP), and the late North Song to early Ming Dynasties (ca.
+1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022].
+For the Neolithic period, it can be further separated into two subphases, 5000-4500 ca.
+BP and 4500-3700 ca.
+BP, according to the content of artifact assemblages and ceramic forms and styles.
 
-<!--Info about Gaoshan-->
-Gaoshan site is located in the southwestern part of Chengdu Plain.
+<!--the major finding at GJB-->
+Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a].
+The archaeobotanical remains reveal evidence of crop cultivation dominated by foxtail and broomcorn millet based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016].
+The adoption of millet cultivation can be traced back to ca.
+5000 BP [@Hao2022].
+In addition, rice grains have also been found in ash pits throughout different phases where the earliest rice dates to 2881-2633 ca.
+BP [@Hao2022].
+It is suggested that mixed farming of millet and rice could spread to southwest Sichuan around 5,000 years ago and to western Yunnan before 4600 years ago based on the evidence of rice remains from Guijiabao and other sites in the neighboring areas, Baiyangcun and Henglanshan [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010].
+
+<!--info about Gaoshan-->
+Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
+Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
+According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca.
+4700-3700 BP) in the Chengdu Plain [@Guedes2013].
+Gaoshan could be even older than the Baodun culture according to the burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017].
+
+A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017].
+This finding also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China [@Chi2010].
+Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were domesticated, of which pigs are the main domestic animal [@He2020].
+The primary subsistence strategy at Gaoshan could be rice and millet cultivation with domesticated pigs.
 
@@
 Gaushan-90 pig, plant-most are rice: 95% and a few millet  
+ecological zones
 
@@
 Our case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.
+
+More systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.
 

### 2023-01-23 analysis/paper/paper.Rmd
finished the background section

@@
 Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a].
-The archaeobotanical remains reveal evidence of crop cultivation dominated by foxtail and broomcorn millet based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016].
+The archaeobotanical remains reveal evidence of crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016].
 The adoption of millet cultivation can be traced back to ca.
@@
 BP [@Hao2022].
-It is suggested that mixed farming of millet and rice could spread to southwest Sichuan around 5,000 years ago and to western Yunnan before 4600 years ago based on the evidence of rice remains from Guijiabao and other sites in the neighboring areas, Baiyangcun and Henglanshan [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010].
 
@@
 4700-3700 BP) in the Chengdu Plain [@Guedes2013].
-Gaoshan could be even older than the Baodun culture according to the burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017].
+It is proposed that Gaoshan could be even older than the Baodun culture according to a burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017].
 
-A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017].
-This finding also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China [@Chi2010].
-Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were domesticated, of which pigs are the main domestic animal [@He2020].
-The primary subsistence strategy at Gaoshan could be rice and millet cultivation with domesticated pigs.
+A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017].
+This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010].
+Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pigs are the main domestic animal [@He2020].
+The primary subsistence strategy at Gaoshan could be rice/millet cultivation with domesticated pigs.
+
+It is generally accepted that mixed farming of millet and rice spread to southwest Sichuan around 5,000 years ago and to western Yunnan before 4600 years ago [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010].
+The current finding at Guijiabao and Gaoshan support this argument.
+However, the ratio of rice to millet grains found at Guijiabao suggests that millet is the major crop in an earlier period, which is different from a clear predominance of rice (*Oryza sativa*) at Gaoshan or other sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013].
+Although the analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, evidence directly reflects food preparation or processing using ceramics still lacks.
+This study provides direct evidence from food residue in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.
 
@@
 We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX).
-The four ceramic fragments were recovered from the trench 13 excavated in the 2017 field season and one sample was from the 2016 season.
-To understand the content and use of pottery vessels in the Yanyan region in an earlier period, we analyze potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison.
-During the early Neolithic period, the common vessel forms include jars, urns, bowls, and basins.
+Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and another sample was from the 2016 season.
+To understand the content and use of pottery vessels in Yanyuan in an earlier period, we analyzed potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison.
+The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins.
 For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022].
@@
 # Discussion
-
 Our preliminary results show that the food categories indicated by potsherds at Guijiabao and Gaushan could be mainly dominated by terrestrial animals, such as ruminant animals.

### 2023-01-24 analysis/paper/paper.Rmd
edit the background and working on the discussion section

@@
 Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a].
-The archaeobotanical remains reveal evidence of crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016].
+The archaeobotanical remains reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016].
 The adoption of millet cultivation can be traced back to ca.
 5000 BP [@Hao2022].
-In addition, rice grains have also been found in ash pits throughout different phases where the earliest rice dates to 2881-2633 ca.
+In addition, rice grains (*Oryza sativa*) have also been found in ash pits throughout different phases where the earliest rice dates to 4700 ca.
 BP [@Hao2022].
+A detailed phytolith analysis for soil samples dating to Neolithic period (5000 ca.
+BP to 3700 ca.
+BP) further suggests the presence of rice phytoliths throughout the whole period in addition to millet phytoliths, indicative of a mixed millet/rice cultivation prior to 5000 ca.
+BP than previously thought [@Huan2022].
 
@@
 It is proposed that Gaoshan could be even older than the Baodun culture according to a burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017].
-
 A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017].
@@
 
-It is generally accepted that mixed farming of millet and rice spread to southwest Sichuan around 5,000 years ago and to western Yunnan before 4600 years ago [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010].
+Despite of a long-standing debate between waves of rice/millet dispersal from northwest China and from the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP.
+and to western Yunnan before 4600 BP.
+[@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010].
 The current finding at Guijiabao and Gaoshan support this argument.
-However, the ratio of rice to millet grains found at Guijiabao suggests that millet is the major crop in an earlier period, which is different from a clear predominance of rice (*Oryza sativa*) at Gaoshan or other sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013].
+However, plant remains from Guijiabao shows that millet is the major crop, which is at odds with the predominance of rice based on phytolich analysis and other macro-botanical analyses from Gaoshan or sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013].
 Although the analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, evidence directly reflects food preparation or processing using ceramics still lacks.
-This study provides direct evidence from food residue in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.
+This study aims to provide direct evidence from food residues in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.
 
@@
 # Discussion
-Our preliminary results show that the food categories indicated by potsherds at Guijiabao and Gaushan could be mainly dominated by terrestrial animals, such as ruminant animals.
-Although the yield of lipids in most of our potsherd samples is generally scarce, one potsherd demonstrates a relatively rich composition of long-chain fatty acids which confirms their usage as a cooking or storage pot.
-A larger amount of C16 and C18 fatty acids compared to other types of fatty acids indicate that the organic residues in our potsherds were highly degraded over the past 4500 years.
-It is also striking to find that one potsherd of Guijiabao contains residues that might originate from C3 plants.
+<!--A brief summary of interesting results-->
+Our results show that food categories indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by terrestrial animals, particularly ruminant fats or mixtures of terrestrial animals and plants.
+Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might be C3 plants.
+Although the lipid yields in most of our samples are generally low, interpretative amounts of yields for all potsherds support that those ceramics were used as cooking or storage pots.
+One potshard dating to around 4500 years ago demonstrates a rich composition of long-chain fatty acids, suggesting a better condition for preserving lipid molecules.
+We also recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes from both Guijiabao and Gaoshan.
+However, the general pattern of the isotopic values of lipid molecules could provide informative clues for a better understanding of the local subsistence of Guijiabao.
 
-show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source.
+Ruminant fats in the Neolithic potsherds around ca.
+4500 BP.
+from Guijiabao and Gaoshan could be derived from deer or horse according to the finding of faunal remains, accompanied with stone artifacts for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
+However, whether horses were part of local food choices in an earlier period is questionable.
+First, the Pleistocene and Neolithic sites with wild horse bones mostly concentrated in northern China [@Yuan2006].
+Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004].
+Although wild Asian water buffalo could be another food source due to their presence since the Pleistocene [@Zhang2016], no cattle bones were found at Guijiabao or Gaoshan.
+Therefore, it is more likely that deer could be the main resource of ruminant fats.
+The analysis of faunal remains from Guijiabao is still in progress which could allow us to evaluate our findings in the future.
+
+check isotopic results for horse, deer and cattle
+
+Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded over the past 4500 years.
+
+The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source.
 Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously.
@@
 ecological zones
+
+A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture.
 

### 2023-01-26 analysis/code/isotope-analysis.R
added text to the plot

@@ -111,8 +111,10 @@ isotope_C16_C18 <-
   theme_minimal() +
   labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
        y = bquote(delta*{}^13*"C 18:0 \u2030")) +
-  xlim(-33,-20) +
-  ylim(-33,-20) +
+  scale_x_continuous(limits = c(-33, -20),
+                     breaks = seq(-33, -20, 2)) +
+  scale_y_continuous(limits = c(-33, -20),
+                     breaks = seq(-33, -20, 2)) +
   #coord_fixed(ratio = 1) +
   geom_abline(intercept = -0.7, linetype = "dashed") + #Suryanarayan et al.2021
   geom_abline(intercept = -3.1, linetype = "dashed") +
@@ -139,7 +141,11 @@ big_delta_C16 <-
   annotate("segment", x = -29, y = 1.8, xend = -24, yend = 1.8,
            arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
   annotate("text", x = -22, y = 1.8,
-           label = "C4-diet")
+           label = "C4-diet") +
+  annotate("text", x = -31.5, y = -1.2,
+           label = "Ruminant") +
+  annotate("text", x = -31, y = -0.3,
+           label = "Non-ruminant")
 
 library(cowplot)
 combined_isotopic_plots <-

### 2023-01-27 analysis/code/site-regional-map-Ch.R
edit the map

@@ -20,6 +20,10 @@ world_points <- cbind(world, st_coordinates(st_point_on_surface(world$geometry))
   mutate(Y = ifelse(brk_name %in% c("Bhutan", "Bangladesh"), Y+2, Y)) %>%
   mutate(Y = ifelse(brk_name %in% c("Thailand", "Myanmar"), Y+3.5, Y))
 
+# shapefile
+ch_adm <- st_read(here::here("analysis/data/raw_data/Ch_adm/CHN_adm1.shp"))
+plot(st_geometry(ch_adm))
+
 # add site location
 site_location <-
   data.frame(location = c("Guijuabao", "Gaoshan"), # Chengdu(104.0587, 30.5899)
@@ -28,7 +32,7 @@ site_location <-
 
 China_SE_Asia <-
   ggplot(data = world) +
-  geom_sf( fill= "antiquewhite") +
+  geom_sf(fill= "antiquewhite") +
   geom_rect(xmin = 99.5, xmax = 105.5, ymin = 25.5, ymax = 31.5,
             fill = NA, colour = "red", size = 0.5) +
   geom_shadowtext(data= world_points,
@@ -40,24 +44,27 @@ China_SE_Asia <-
                   position = position_nudge(y = - 1.7, x = 0.5)) +
   #annotate(geom = "text", x = 102.3, y = 9.4, label = "Gulf of\nTailand",
            #fontface = "italic", color = "grey22", size = 2) +
-  coord_sf(xlim = c(80, 123), ylim = c(9, 45), expand = FALSE) + #add datum = NA to remove
+  coord_sf(xlim = c(85, 120), ylim = c(18, 45), expand = FALSE) + #add datum = NA to remove
   scale_x_continuous(breaks = seq(80, 125, by = 10)) +
   scale_y_continuous(breaks = seq(5, 45, by = 10)) +
   theme(axis.title.x = element_blank(),
         axis.title.y = element_blank())
 
+China_SE_Asia +
+  geom_sf(data = ch_adm)
+
 # Topographic map
 library(ggmap)
 # we don't want to download every time, so let's save the map locally
 # from https://stackoverflow.com/a/52710855/1036500
-GJB_map <- ggmap(get_stamenmap(rbind(as.numeric(c(99, 25,
-                                                  106, 32))), zoom = 10))
-#saveRDS(GJB_map, here("analysis", "data", "raw_data", "SW_china_map.rds"))
[…]

### 2023-01-27 analysis/paper/paper.Rmd
updated the isotope plot

@@
 To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
-The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial or marine animals and plants, have specific ranges of δ^13^C values [@Evershed1994].
+The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial animal, marine animals and plants, have specific ranges of δ^13^C values [@Craig2004; @Evershed1994].
 
@@
        y = bquote(delta*{}^13*"C 18:0 \u2030")) +
-  xlim(-33,-20) +
-  ylim(-33,-20) +
+  scale_x_continuous(limits = c(-33, -20),
+                     breaks = seq(-33, -20, 2)) +
+  scale_y_continuous(limits = c(-33, -20),
+                     breaks = seq(-33, -20, 2)) +
   #coord_fixed(ratio = 1) +
@@
                      breaks = seq(-33, -20, 2)) +
-  ylim(-2.5, 2) +
+  ylim(-2.5, 1.5) +
   geom_hline(yintercept = -0.7, linetype = "dashed") +
-  annotate("text", x = -22, y = -0.5,
+  annotate("text", x = -31.5, y = -0.9,
            label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
-  annotate("text", x = -31, y = 1.8,
+  annotate("text", x = -32, y = 1.3,
            label = "C3-diet") +
-  annotate("segment", x = -29, y = 1.8, xend = -24, yend = 1.8,
+  annotate("segment", x = -29, y = 1.3, xend = -24, yend = 1.3,
            arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
-  annotate("text", x = -22, y = 1.8,
-           label = "C4-diet")
+  annotate("text", x = -21, y = 1.3,
+           label = "C4-diet") +
+  annotate("text", x = -22, y = -1.1,
+           label = "Ruminant") +
+  annotate("text", x = -21.5, y = -0.3,
+           label = "Non-ruminant")
 
@@
 The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
-It should be noted that there are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003].
+It should be noted that there are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to a later period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003].
 Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020].
@@
 4500 BP.
-from Guijiabao and Gaoshan could be derived from deer or horse according to the finding of faunal remains, accompanied with stone artifacts for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
+from Guijiabao and Gaoshan could be derived from deer or horse according to the finding of faunal remains from the sites, accompanied with stone artifacts for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
 However, whether horses were part of local food choices in an earlier period is questionable.
@@
 Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004].
-Although wild Asian water buffalo could be another food source due to their presence since the Pleistocene [@Zhang2016], no cattle bones were found at Guijiabao or Gaoshan.
-Therefore, it is more likely that deer could be the main resource of ruminant fats.
-The analysis of faunal remains from Guijiabao is still in progress which could allow us to evaluate our findings in the future.
+Although wild Asian water buffalo could be another food source due to their presence since the Pleistocene [@Zhang2016], no cattle bones were found at Guijiabao or Gaoshan in an early period.
+Therefore, it is more likely that deer could be the resource of ruminant fats.
+The analysis of faunal remains from Guijiabao is still in progress which could allow us to evaluate our findings.
 
-check isotopic results for horse, deer and cattle
+check isotopic results for horse, deer
 

### 2023-01-27 analysis/code/isotope-analysis.R

@@ -132,19 +132,19 @@ big_delta_C16 <-
        y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
   scale_x_continuous(limits = c(-33, -20),
                      breaks = seq(-33, -20, 2)) +
-  ylim(-2.5, 2) +
+  ylim(-2.5, 1.5) +
   geom_hline(yintercept = -0.7, linetype = "dashed") +
-  annotate("text", x = -22, y = -0.5,
+  annotate("text", x = -31.5, y = -0.9,
            label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
-  annotate("text", x = -31, y = 1.8,
+  annotate("text", x = -32, y = 1.3,
            label = "C3-diet") +
-  annotate("segment", x = -29, y = 1.8, xend = -24, yend = 1.8,
+  annotate("segment", x = -29, y = 1.3, xend = -24, yend = 1.3,
            arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
-  annotate("text", x = -22, y = 1.8,
+  annotate("text", x = -21, y = 1.3,
            label = "C4-diet") +
-  annotate("text", x = -31.5, y = -1.2,
+  annotate("text", x = -22, y = -1.1,
            label = "Ruminant") +
-  annotate("text", x = -31, y = -0.3,
+  annotate("text", x = -21.5, y = -0.3,
            label = "Non-ruminant")
 
 library(cowplot)

### 2023-01-28 analysis/paper/paper.Rmd
working on the discussion section

@@
 
-Long-chain fatty acids from C20:0 to C32:0 are present in only one pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)).
+Long-chain fatty acids from C20:0 to C32:0 are present in a pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)).
 The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005].
@@
 The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A).
-The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
-It should be noted that there are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to a later period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003].
-Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020].
+The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
+A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003].
+Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish  [@Shoda2018; @Kwak2020].
 In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants.
@@
 4500 BP.
-from Guijiabao and Gaoshan could be derived from deer or horse according to the finding of faunal remains from the sites, accompanied with stone artifacts for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
-However, whether horses were part of local food choices in an earlier period is questionable.
-First, the Pleistocene and Neolithic sites with wild horse bones mostly concentrated in northern China [@Yuan2006].
+from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, with stone tools associated with hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
+However, whether horses were part of local food choices in the Neolithic period is questionable.
+First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006].
 Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004].
-Although wild Asian water buffalo could be another food source due to their presence since the Pleistocene [@Zhang2016], no cattle bones were found at Guijiabao or Gaoshan in an early period.
-Therefore, it is more likely that deer could be the resource of ruminant fats.
-The analysis of faunal remains from Guijiabao is still in progress which could allow us to evaluate our findings.
+The evidence from Guijuabao also confirms a later presence of horse remains.
+In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value located in ranges of deer or sheep fats based on a study from England [@Evershed2002].
+Although the referenced isotopic values were from modern animals and should be ideally adjusted in terms of local environments, they could still provide useful information and have been used as references in many studies [].
+It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were parts of food sources during this period [@Zhang2016].
+A detailed analysis of faunal remains from Guijiabao is still in progress that will allow evaluating our findings.
 
-check isotopic results for horse, deer
-
-Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded over the past 4500 years.
+It is interesting to find a non-ruminant origin that could be C3 plants.
 
@@
 
+Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded over the past 4500 years.
 A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture.

### 2023-01-29 analysis/paper/paper.Rmd
working on the discussion section

@@
 They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007].
-High concentrations of C16:0 and C18:0 fatty acids in our samples suggest that the original organic residues were highly degraded during post-depositional processes.
+High concentrations of Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
 In addition, some potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
@@
 The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
-A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003].
+A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998; @Suryanarayan2021].
 Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
@@
 Our results show that food categories indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by terrestrial animals, particularly ruminant fats or mixtures of terrestrial animals and plants.
-Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might be C3 plants.
-Although the lipid yields in most of our samples are generally low, interpretative amounts of yields for all potsherds support that those ceramics were used as cooking or storage pots.
-One potshard dating to around 4500 years ago demonstrates a rich composition of long-chain fatty acids, suggesting a better condition for preserving lipid molecules.
-We also recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes from both Guijiabao and Gaoshan.
-However, the general pattern of the isotopic values of lipid molecules could provide informative clues for a better understanding of the local subsistence of Guijiabao.
+Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from terrestrial plants.
+Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots.
+The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020].
+But we also found a base sample dating to ca.
+4500 BP.
+contains a rich composition of long-chain fatty acids.
+We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes.
+However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence of Guijiabao.
 
+<!--discussion about possible plant origins-->
+One aim of our study is to explore whether new crops, such as C3 plants (eg.
+rice), C4 plants (eg.
+millet), or new species (eg.
+pigs/horses) became an important part of local diet in Yanyuan during the Neolithic transition.
+We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies [@Shoda2018; @Kwak2020].
+Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.
+Also, recent macro-botanical and phytolith studies at Guijiabao suggest the presence of a mixing farming of rice and millet, indicating that the origin of C3 plants from our sample could come from rice [@Hao2022 @Huan2022].
+It is also possible other grasses classified as C3 plants, such as Pennisetum, were part of the local diet [@Guijiabao2016].
+We should note that freshwater fish could also be the origin of non-ruminant fats since C3 plants and freshwater fish have similar isotopic values [@Lucquin2016].
+Fish bones and net sinkers were also found at Guijiabao which suggests the possibility of fishing.
+There are no C4 plants, such as millets, directly detected in our sample, which contradicts the predominance of millet grains found at Guijiabao.
+But it is hard to make an inference about the role of rice/millet because the results might be caused by small sample effects or sampling strategy.
+
+<!--discussion about possible ruminant fat origin-->
 Ruminant fats in the Neolithic potsherds around ca.
@@
 from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, with stone tools associated with hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
-However, whether horses were part of local food choices in the Neolithic period is questionable.
+But whether horses were part of local food choices in the Neolithic period is questionable.
 First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006].
@@
 In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value located in ranges of deer or sheep fats based on a study from England [@Evershed2002].
-Although the referenced isotopic values were from modern animals and should be ideally adjusted in terms of local environments, they could still provide useful information and have been used as references in many studies [].
-It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were parts of food sources during this period [@Zhang2016].
-A detailed analysis of faunal remains from Guijiabao is still in progress that will allow evaluating our findings.
+It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016].
+A detailed analysis of faunal remains from Guijiabao is still in progress, allowing further evaluation of our findings.
 
-It is interesting to find a non-ruminant origin that could be C3 plants.
+In general, our results suggest that Guijiabao residents relied mostly on ruminants fats with terrestrial plants as supplements.
+The overall depleted δ^13^C value shows that the diet of ruminants was mainly C3 plants that could come from wild weeds [@Guijiabao2016].
 
-The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant source or marine source.
-Our striking finding might suggest that consumption of rice could be part of the daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously.
-Our next steps is to increase sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.
+--Gaushan-90 pig, plant-most are rice: 95% and a few millet  
+--ecological zones
 
-net sinkers
-1000 BC horse
-Gaushan-90 pig, plant-most are rice: 95% and a few millet  
-ecological zones
-
-Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded over the past 4500 years.
 A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture.
+Our next step is to increase the sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.
 
 # Conclusion
-Our investigation on the ceramics excavated from Southwest China suggests that a mixture of C3 plants and ruminant animals could be the food sources in this region during the Neolithic transition.
+
+Our investigation of the ceramics excavated from Southwest China suggests that a mixture of C3 plants and ruminant animals could be the food sources in this region during the Neolithic transition.
 This case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.
 Our case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.
+
+The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant sources or marine sources.
+Our striking finding might suggest that the consumption of rice could be part of daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously.
 

### 2023-01-31 analysis/paper/paper.Rmd
working on the discussion section

@@
 This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010].
-Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pigs are the main domestic animal [@He2020].
-The primary subsistence strategy at Gaoshan could be rice/millet cultivation with domesticated pigs.
+Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pig is the dominant animal [@He2020].
+The primary subsistence strategy at Gaoshan could be rice cultivation with domesticated pigs, supplemented by millet cultivation.
 
@@
 Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
+Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.
+
 In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants.
@@
 <!--A brief summary of interesting results-->
-Our results show that food categories indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by terrestrial animals, particularly ruminant fats or mixtures of terrestrial animals and plants.
-Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from terrestrial plants.
+Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period.
+Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from C3 plants.
 Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots.
@@
 4500 BP.
-contains a rich composition of long-chain fatty acids.
+contains a rich composition of long-chain fatty acids, reflecting an excellent preservation.
 We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes.
-However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence of Guijiabao.
+However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.
 
@@
 rice), C4 plants (eg.
-millet), or new species (eg.
-pigs/horses) became an important part of local diet in Yanyuan during the Neolithic transition.
+millets), or new species (eg.
+pigs/horses) became an essential part of local diet in Yanyuan during the Neolithic transition.
 We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies [@Shoda2018; @Kwak2020].
-Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.
-Also, recent macro-botanical and phytolith studies at Guijiabao suggest the presence of a mixing farming of rice and millet, indicating that the origin of C3 plants from our sample could come from rice [@Hao2022 @Huan2022].
-It is also possible other grasses classified as C3 plants, such as Pennisetum, were part of the local diet [@Guijiabao2016].
-We should note that freshwater fish could also be the origin of non-ruminant fats since C3 plants and freshwater fish have similar isotopic values [@Lucquin2016].
-Fish bones and net sinkers were also found at Guijiabao which suggests the possibility of fishing.
-There are no C4 plants, such as millets, directly detected in our sample, which contradicts the predominance of millet grains found at Guijiabao.
-But it is hard to make an inference about the role of rice/millet because the results might be caused by small sample effects or sampling strategy.
+It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016].
+Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022].
+Due to our small samples, it is hard to determine the role of rice/millet during the Neolithic period.
+Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Kwak2017].
 
 <!--discussion about possible ruminant fat origin-->
-Ruminant fats in the Neolithic potsherds around ca.
+Ruminant fats identified in the Neolithic potsherds around ca.
 4500 BP.
-from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, with stone tools associated with hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
-But whether horses were part of local food choices in the Neolithic period is questionable.
+from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
+But whether horse was part of local food choices in the Neolithic period is questionable.
 First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006].
@@
 The evidence from Guijuabao also confirms a later presence of horse remains.
-In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value located in ranges of deer or sheep fats based on a study from England [@Evershed2002].
+In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats based on global references [@Evershed2002].
 It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016].
-A detailed analysis of faunal remains from Guijiabao is still in progress, allowing further evaluation of our findings.
+A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.
 
-In general, our results suggest that Guijiabao residents relied mostly on ruminants fats with terrestrial plants as supplements.
-The overall depleted δ^13^C value shows that the diet of ruminants was mainly C3 plants that could come from wild weeds [@Guijiabao2016].
-
---Gaushan-90 pig, plant-most are rice: 95% and a few millet  
---ecological zones
-
-A large-scale analysis of ceramic across the region may offer insight into regional variations in food choices and the spread of agriculture.
-Our next step is to increase the sample size to include pots from different time periods and different types of pottery to identify any variation through time and also try different methods, such as phytolith analysis to further support our statement about rice cultivation.
+We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca.
+1000 BC.).
+However, we are unable to determine whether there is a shift in diet over time due to our limited sample size.
+But the overall depleted δ^13^C values for most potsherds suggest C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition.
+This corresponds with the timing when new crops spread into southwest China.
+Despite of no C4 plant origin in our samples, one potshard implies a source of C4 plant-fed ruminant.
+In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife.
+This might be related to local ecological zones and environments where wild animals were easier to obtain.
+Since zooarchaeological analysis and other studies on stone tools are still in progress, we could only rely on preliminary excavation reports with limited details about associated artifacts.
+A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.
 

### 2023-02-01 analysis/paper/paper.Rmd
worked on the conclusion section

@@
 
-Our investigation of the ceramics excavated from Southwest China suggests that a mixture of C3 plants and ruminant animals could be the food sources in this region during the Neolithic transition.
-This case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.
-Our case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.
-
-The results show a mixture of C3 plants, such as rice, and terrestrial animals for most pots with one pot showing more C4 plant sources or marine sources.
-Our striking finding might suggest that the consumption of rice could be part of daily life at Guijiabao and the adoption of rice agriculture could be as important as millet cultivation which was thought the main crop previously.
-
-More systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.
+Our study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in a relatively low amount for most potsherds.
+The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China.
+It is most likely the terrestrial plant signal is C3 plant, implying rice processing in the pot.
+However, the evidence of direct processing of terrestrial plants is not as much as we expected according to the large quantity of plant remains, such as millet grains, found at Guijuabao.
+Instead, food processing in the early Neolithic is mostly ruminant fats that could be from local wild animals, such as deer.
+This might indicate that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming.
+This study offers a perspective from organic residue analysis to explore food choices and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period.
+We recognize the limitation of our interpretation of the local diet due to the small sample size.
+A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.
 

### 2023-02-03 analysis/paper/paper.Rmd
worked on the intro section

@@
 <!-- intro to the topic and research aim-->
-Movements of people between geographically distinct areas usually accompany the exchange or introduction of goods, species, and ideas that could result in far-reaching impacts on local societies economically and culturally.
-In Southwest China, the Hengduan Mountain ranges as natural passes play an important role in the migration of farming communities and the spread of crops into this region and mainland Southeast Asia.
+Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could result in far-reaching economic or cultural impacts on local societies.
+In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022].
+Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP.
+and to western Yunnan before 4600 BP.
+[@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022].
+Therefore, the Neolithic Hengduan Mountain areas are key to understanding the processes of agriculture expansion.
 Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022].
 However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood.
-To explore whether the new species became main food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, the oldest Neolithic site in the Yanyuan region so far excavated.
 
-<!-- sites and research questions-->
-As an early Neolithic site, Guijiabao (ca.
-4800-4300 BP) is crucial to understanding cultural interactions and the spread of crops into the Yanyuan Basin due to its location at the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau.
+To explore whether the new crops or species became important food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a].
 The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@Guijiabao2016], different from the abundant rice grains found at other Neolithic sites in that area.
-Although the plant remains suggest the possible domestication of millet in the early Neolithic, we still lack direct evidence of culinary practices during the Neolithic transition.
-To explore the extent of acceptance of new crops and species in the Yanyuan Basin, we ask the questions: What was the culinary practices during the Neolithic transition?
-Could we identify the evidence of new crops or species incorporated into the local dietary systems in the Early Neolithic period?
-We answer those questions by examining potsherds collected from Guijiabao using organic residue analysis, which could offer a better understanding of food sources during the Neolithic transition in the Yanyuan Basin.
+Although the analysis of plant remains suggests millet/rice farming in the early Neolithic, we still lack direct evidence of culinary practices during this period.
+To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use?
+Could we identify the evidence of new crops or species incorporated into the local food choices in the Early Neolithic period?
+We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels during the Neolithic transition in the Yanyuan Basin.
 
-This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the vessel usage at Guijiabao.
-Lipid analysis is useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017].
-Lipid residues are the most direct evidence to investigate culinary practices and dietary patterns since specific lipid compounds can be associated with different types of foodstuffs cooked or stored in pottery vessels [@Copley2005].
-It is especially informative for human subsistence in the situation where archaeological sites lack organic remains due to acidic sediments [@Kwak2020].
-The growing application of organic residue analysis demonstrate it offers a better understanding of culinary practices during the transitional period to agricultural societies [@Craig2011], dietary patterns in
+The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies or culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022].
+Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021].
+This is because lipid molecules can be accumulated and well preserved in the porous fabric of pottery over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017].
+In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020].
 
@@
 
-<!--Yanyuan area-->
-The Yanyuan area in Sichuan Province is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)).
-Although the ranges isolate areas with their own unique geographic conditions and cultures, river valleys provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022].
-Therefore, the archaeological sites in the Yanyuan Basin are important to understanding the cultural interactions between southwest China and other regions, including the Sichuan basin, northwest China, and northwest Yunnan.
-
-<!--Guijiabao site-->
-Among those archaeological sites in Yanyuan, Guijiabao located in the southwestern part of the basin is a unique site with relatively complete and continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
-A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guidiabao is a settlement site with long-term occupation [@Hao2022].
+<!--Yanyuan area and Guijiabao site-->
+We focus on Guijiabao site located in the Yanyuan Basin in Sichuan Province.
+Yanyuan Basin is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)).
+River valleys of Hengduan Mountain provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022].
+Among the archaeological sites in Yanyuan, Guijiabao (ca.
+4800-4300 BP) located in the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau is a unique site with continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
+A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site with long-term occupation [@Hao2022].
 The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (ca.
@@
 <!--the major finding at GJB-->
-Among the various ceramic styles, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a].
+At Guijiabao, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a].
 The archaeobotanical remains reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016].
@@
 
-Despite of a long-standing debate between waves of rice/millet dispersal from northwest China and from the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP.
-and to western Yunnan before 4600 BP.
-[@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010].
  The current finding at Guijiabao and Gaoshan support this argument.
@@
 High concentrations of Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
-In addition, some potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
-A larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005].
-Although oleic acid (C18:1) is the most abundant compound for most of our samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.
+In addition, a number of potsherds presents higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
+Although a larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005], it is hard to directly relate the pattern we observed to a plant origin based on GC-MS results alone.
 
@@
 names(quant_FA)[2] <- "sample"
+
+quant_FA_number <-
+  quant_FA %>%
+  pivot_longer(!number&!sample,
+               names_to = "FA", 
+               values_to = "weight") %>% 
+  select(-number) %>% 
+  filter(str_starts(FA, "ng") & !sample =="") %>% 
+  mutate(FA = str_remove(FA, "ng "),
+         weight = round(weight/1000, 2)) %>% 
+  pivot_wider(names_from = sample, 
+              values_from = weight)
+
+FA_max_sum <-
+  sum(quant_FA_number$"SYG-TN13-E22-2#2")
 
@@
 
-Long-chain fatty acids from C20:0 to C32:0 are present in a pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)).
+One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of samples, suggesting a good preservation of lipids.
+Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)).
 The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005].
@@
 The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
-A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998; @Suryanarayan2021].
+A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
 Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
+This also corresponds to the lipid distribution of the same potshard which reflects a possible plant origin based on the GC-MS results.
 Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.
@@
 Due to our small samples, it is hard to determine the role of rice/millet during the Neolithic period.
-Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Kwak2017].
+Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
 
@@
 The evidence from Guijuabao also confirms a later presence of horse remains.
-In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats based on global references [@Evershed2002].
+In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on C3 diet based on global references [@Evershed2002].
 It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016].
@@
 
+-adoption of farming in a relatively enclosed circumstances?
+
 # Conclusion
 
-Our study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in a relatively low amount for most potsherds.
+Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds.
 The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China.
 It is most likely the terrestrial plant signal is C3 plant, implying rice processing in the pot.
-However, the evidence of direct processing of terrestrial plants is not as much as we expected according to the large quantity of plant remains, such as millet grains, found at Guijuabao.
-Instead, food processing in the early Neolithic is mostly ruminant fats that could be from local wild animals, such as deer.
-This might indicate that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming.
-This study offers a perspective from organic residue analysis to explore food choices and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period.
+However, the evidence of direct processing of terrestrial plants is not as much as we expected according to a relatively large quantity of plant remains found at Guijiabao, such as millet grains.
+Instead, food processing in the early Neolithic is mostly ruminant fats that could originate from local wild animals, such as deer.
+This indicates that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming.
 We recognize the limitation of our interpretation of the local diet due to the small sample size.
+However, our study can still offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period.
 A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.

### 2023-02-04 analysis/paper/paper.Rmd
edited the background section

@@
 
-The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies or culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022].
+The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies, and culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022].
 Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021].
@@
 In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020].
+We attempt to explore the culinary practice under the circumstances during the time of frequent cultural interactions and expansion of agriculture by our case study.
 
@@
 <!--Yanyuan area and Guijiabao site-->
-We focus on Guijiabao site located in the Yanyuan Basin in Sichuan Province.
-Yanyuan Basin is a pull-apart basin located in the middle eastern part of the Hengduan Mountain, which is a group of parallel mountain ranges in a north-south direction from the Tibetan Plateau to the Yunnan-Guizhou Plateau (Figure \@ref(fig:SW-china-map)).
-River valleys of Hengduan Mountain provide pathways for the movements of people and the exchange of goods between cultural groups since the prehistoric period [@Lin2022; @Hao2022].
-Among the archaeological sites in Yanyuan, Guijiabao (ca.
-4800-4300 BP) located in the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau is a unique site with continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
+We focus on Guijiabao (ca.
+4800-4300 BP) located in the Yanyuan Basin in Sichuan Province (Figure \@ref(fig:SW-china-map)).
+Yanyuan Basin is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges.
+Guijiabao is especially unique due to its location as the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau, with a continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
 A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site with long-term occupation [@Hao2022].
@@
 <!--the major finding at GJB-->
-At Guijiabao, ceramics with surface decorations such as incised geometric patterns and impressed dot patterns are the two common styles that have also been found in western Yunnan, indicative of frequent interactions between Guijiabao and sites in Yunnan [@Guijiabao2017a].
-The archaeobotanical remains reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on the analysis of flotation samples from ash pits and stratigraphic layers in 2015 [@Guijiabao2016].
-The adoption of millet cultivation can be traced back to ca.
-5000 BP [@Hao2022].
-In addition, rice grains (*Oryza sativa*) have also been found in ash pits throughout different phases where the earliest rice dates to 4700 ca.
-BP [@Hao2022].
-A detailed phytolith analysis for soil samples dating to Neolithic period (5000 ca.
+The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins.
+Most ceramics have surface decorations, such as incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a].
+In addition, cord-marked pottery vessels are also commonly found in the early period.
+About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016].
+The earliest domesticated millet grain can be traced back to ca.
+5000 BP according to its radiocarbon date [@Hao2022].
+A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 ca.
+BP.
+[@Hao2022].
+Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000 ca.
 BP to 3700 ca.
-BP) further suggests the presence of rice phytoliths throughout the whole period in addition to millet phytoliths, indicative of a mixed millet/rice cultivation prior to 5000 ca.
-BP than previously thought [@Huan2022].
+BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 ca.
+BP.
+[@Huan2022].
 
 <!--info about Gaoshan-->
-Gaoshan site is located in the southwestern margin of the Chengdu Plain, an alluvial plain between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
-Excavations at Gaoshan revealed abundant pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
+To have a comparison, we also look at a sample from Gaoshan, which is located on the southwestern margin of the Chengdu Plain (Figure \@ref(fig:SW-china-map)).
+The Chengdu Plain is an alluvial plain between the Sichuan Basin and the Tibetan Plateau.
+Excavations at Gaoshan revealed abundant materials, including pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
 According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca.
-4700-3700 BP) in the Chengdu Plain [@Guedes2013].
-It is proposed that Gaoshan could be even older than the Baodun culture according to a burial context, which was found in the stratigraphic layer earlier than the Baodun period [@Gaoshan2017].
+4700-3700 BP) in the Chengdu Plain, or even earlier than that [@Guedes2013; @Gaoshan2017].
 A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017].
 This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010].
-Moreover, recent research on animal bones in 2015 and 2016 shows that pigs and dogs were already domesticated, of which pig is the dominant animal [@He2020].
-The primary subsistence strategy at Gaoshan could be rice cultivation with domesticated pigs, supplemented by millet cultivation.
+Moreover, recent research on animal bones in 2015 and 2016 shows the predominance of domesticated pigs, with a small number of domesticated dogs [@He2020].
+The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation.
 
- The current finding at Guijiabao and Gaoshan support this argument.
-However, plant remains from Guijiabao shows that millet is the major crop, which is at odds with the predominance of rice based on phytolich analysis and other macro-botanical analyses from Gaoshan or sites in Sichuan and Yunnan, such as Baodun, Henglanshan and Baiyangcun [@Hao2022; @Guedes2013].
-Although the analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, evidence directly reflects food preparation or processing using ceramics still lacks.
-This study aims to provide direct evidence from food residues in ceramics to investigate food choices associated with new practices of crop cultivation and local subsistence.
+Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP., what their proportions are in local culinary practices during the early time are unclear.
+For example, plant remains from Guijiabao show that millet is the major crop, which is at odds with the predominance of rice based on phytolith analysis.
+For Gaoshan and neighboring sites in Sichuan and Yunnan, such as Baodun, Henglanshan, and Baiyangcun, it appears that rice accounts for the vast majority of plaint remain [@Hao2022; @Guedes2013].
+The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks.
+This study aims to provide direct evidence from food residues in ceramics to investigate culinary practices associated with the cultivation of new crops.
 
@@
 We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX).
-Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and another sample was from the 2016 season.
-To understand the content and use of pottery vessels in Yanyuan in an earlier period, we analyzed potsherds from the stratigraphic layer which dates to the early Neolithic period (5000-4500 BP.) with potsherds from a later period (1000 BP.) for comparison.
-The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins.
-For the pottery types, cord-marked grey pottery vessels with fine inclusions and cord-marked sand-tempered vessels in reddish-brown color with coarse inclusions are the two dominant types [@Hao2022].
-Our samples are jar fragments from those two common types that are supposed to be used for food serving or storage due to the archaeological context of settlement [@Hao2022; @Guijiabao2017a].
+Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season.
+Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP.) and the historical period (1000 BP.).
+The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022].
+Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them has incised cord-marked decoration.
+Although the original function of those jars is uncertain, they should be supposed to be used for cooking or storage according to the char on the surfaces and their archaeological contexts from a settlement [@Hao2022; @Guijiabao2017a].
 
@@
 The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
-A few number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
+A number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
 Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
@@
 4500 BP.
-contains a rich composition of long-chain fatty acids, reflecting an excellent preservation.
+contains a rich composition of long-chain fatty acids, reflecting a good preservation.
 We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes.
+Also, the lack of local modern references for animal fats and plants has impacts on our explanation.
 However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.

### 2023-02-04 analysis/paper/paper.Rmd
editing the sample section

@@
 Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP., what their proportions are in local culinary practices during the early time are unclear.
-For example, plant remains from Guijiabao show that millet is the major crop, which is at odds with the predominance of rice based on phytolith analysis.
-For Gaoshan and neighboring sites in Sichuan and Yunnan, such as Baodun, Henglanshan, and Baiyangcun, it appears that rice accounts for the vast majority of plaint remain [@Hao2022; @Guedes2013].
+For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in neighboring sites, such as Gaoshan and Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013].
 The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks.
@@
 The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022].
-Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them has incised cord-marked decoration.
+Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them have incised cord-marked decoration.
 Although the original function of those jars is uncertain, they should be supposed to be used for cooking or storage according to the char on the surfaces and their archaeological contexts from a settlement [@Hao2022; @Guijiabao2017a].
+This research also attempts to confirm that those ceramics were cooking pots used in the daily life of the local residents.
 
@@
 To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
-The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial animal, marine animals and plants, have specific ranges of δ^13^C values [@Craig2004; @Evershed1994].
+The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial animal, marine animals and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].
 
-We cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis.
+For the sample processing, we cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis.
 Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C).
@@
 
-One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of samples, suggesting a good preservation of lipids.
+One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of ceramic or a different use.
 Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)).
@@
 pigs/horses) became an essential part of local diet in Yanyuan during the Neolithic transition.
-We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies [@Shoda2018; @Kwak2020].
+We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
 It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016].
+However, the seeds of wild weeds only account for a small part of plant remains.
 Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022].
-Due to our small samples, it is hard to determine the role of rice/millet during the Neolithic period.
 Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].

### 2023-02-05 analysis/paper/paper.Rmd
minor edits on the paper

@@
 In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022].
-Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP.
-and to western Yunnan before 4600 BP.
-[@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022].
+Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022].
 Therefore, the Neolithic Hengduan Mountain areas are key to understanding the processes of agriculture expansion.
@@
 <!--Yanyuan area and Guijiabao site-->
-We focus on Guijiabao (ca.
-4800-4300 BP) located in the Yanyuan Basin in Sichuan Province (Figure \@ref(fig:SW-china-map)).
-Yanyuan Basin is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges.
+We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)).
 Guijiabao is especially unique due to its location as the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau, with a continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
-A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site with long-term occupation [@Hao2022].
-The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (ca.
-5000-3700 BP), the Bronze Age (ca.
-3200-2700 BP), and the late North Song to early Ming Dynasties (ca.
-1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022].
-For the Neolithic period, it can be further separated into two subphases, 5000-4500 ca.
-BP and 4500-3700 ca.
-BP, according to the content of artifact assemblages and ceramic forms and styles.
+A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022].
+The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming Dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022].
+For the Neolithic period, it can be further separated into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles.
 
@@
 The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins.
-Most ceramics have surface decorations, such as incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a].
-In addition, cord-marked pottery vessels are also commonly found in the early period.
+For early Neolithic, ceramics have surface decorations, such as incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a].
+In addition, cord-marked vessels are also found.
 About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016].
-The earliest domesticated millet grain can be traced back to ca.
-5000 BP according to its radiocarbon date [@Hao2022].
-A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 ca.
-BP.
-[@Hao2022].
-Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000 ca.
-BP to 3700 ca.
-BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 ca.
-BP.
-[@Huan2022].
+The earliest domesticated millet grain can be traced back to 5000 BP according to its radiocarbon date [@Hao2022].
+A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 BP [@Hao2022].
+Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].
 
@@
 
-Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP., what their proportions are in local culinary practices during the early time are unclear.
+Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP, what their proportions are in local culinary practices during the early time are unclear.
 For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in neighboring sites, such as Gaoshan and Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013].
@@
 Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season.
-Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP.) and the historical period (1000 BP.).
+Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP) and the historical period (1000 BP).
 The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022].
@@
 The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020].
-But we also found a base sample dating to ca.
-4500 BP.
-contains a rich composition of long-chain fatty acids, reflecting a good preservation.
+But we also found a base sample dates to early Neolithic period contains a rich composition of long-chain fatty acids, reflecting a good preservation.
 We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes.
@@
 <!--discussion about possible ruminant fat origin-->
-Ruminant fats identified in the Neolithic potsherds around ca.
-4500 BP.
-from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
+Ruminant fats identified in the Neolithic potsherds (ca.
+4500 BP) from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
 But whether horse was part of local food choices in the Neolithic period is questionable.
@@
 We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca.
-1000 BC.).
+1000 BP).
 However, we are unable to determine whether there is a shift in diet over time due to our limited sample size.

### 2023-02-05 analysis/code/site-regional-map-Ch.R
change the save location for rds file

@@ -59,8 +59,8 @@ library(ggmap)
 # from https://stackoverflow.com/a/52710855/1036500
 #GJB_map <- ggmap(get_stamenmap(rbind(as.numeric(c(99, 25,
                                                   #106, 32))), zoom = 10))
-# saveRDS(GJB_map, here("analysis", "data", "raw_data", "SW_china_map.rds"))
-SW_china_map <- readRDS(here::here("analysis", "data", "raw_data", "SW_china_map.rds"))
+# saveRDS(GJB_map, here("analysis", "data", "derived_data", "SW_china_map.rds"))
+SW_china_map <- readRDS(here::here("analysis", "data", "derived_data", "SW_china_map.rds"))
 pg <- ggplot_build(SW_china_map)
 
 China_map_with_site <-

### 2023-02-05 analysis/paper/paper.Rmd
added a table for potsherds

@@
 author:
-  - Myfirstname Mylastname:
+  - Li-Ying Wang Mylastname:
       email: fl@oneplace.edu
@@
       correspondence: true
-  - Yourfirstname Yourlastname:
+  - Yourlastname:
       email: fl@another.edu
@@
 In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020].
-We attempt to explore the culinary practice under the circumstances during the time of frequent cultural interactions and expansion of agriculture by our case study.
+We attempt to explore the culinary practice under the circumstances during the time of frequent cultural interactions and expansion of agriculture through our case study.
 
@@
 The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins.
-For early Neolithic, ceramics have surface decorations, such as incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a].
-In addition, cord-marked vessels are also found.
+The common ceramic styles in the early Neolithic period include grayish high-tempered fine ware and reddish brown coarse ware with surface decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a].
+Among the ceramics, incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a].
 About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016].
@@
 ## Samples
-We selected five ceramic fragments from Guijiabao and one ceramic fragment from Gaoshan for organic residue analysis (Table XX).
+We selected six ceramic fragments from Guijiabao and Gaoshan for organic residue analysis (Table \@ref(tab:sherds-table)).
 Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season.
-Those potsherds can be divided into two temporal phases, the early Neolithic period (5000-4500 BP) and the historical period (1000 BP).
-The common ceramic is cord-marked sand-tempered vessels unearthed at Guijiabao, and their forms include jars, urns, bowls, and basins [@Guijiabao2017a; @Hao2022].
-Our samples are mostly reddish-brown jar fragments with medium-size sand temper and some of them have incised cord-marked decoration.
-Although the original function of those jars is uncertain, they should be supposed to be used for cooking or storage according to the char on the surfaces and their archaeological contexts from a settlement [@Hao2022; @Guijiabao2017a].
-This research also attempts to confirm that those ceramics were cooking pots used in the daily life of the local residents.
+Those potsherds can be divided into two temporal phases, the early Neolithic period (ca.
+4500 BP) and the historical period (ca.
+1000 BP).
+Our samples are mostly grayish base fragments with fine sand temper and a number of them have impressed decoration.
+Although their original forms could not be recovered, those ceramics are supposed to be used in daily life according to the evidence of chars on some fragments and their archaeological contexts in cultural layers [@Hao2022; @Guijiabao2017a].
+This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.
+
+```{r sherds-table, cache= TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/SW-china-potsherd.xlsx"))}
+Ch_sherds <-
+  readxl::read_excel(here::here("analysis", "data", "raw_data", "SW-china-potsherd.xlsx"), col_names = TRUE)
+
+knitr::kable(Ch_sherds,
+             caption = "Quantitative data of fatty acid (FA) composition (ug) of potsherds from Guijiabao and Gaushan that are named starting with 'SYG' and 'CDG' respectively")
+```
 
@@
 High concentrations of Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
-In addition, a number of potsherds presents higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
+In addition, a number of potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
 Although a larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005], it is hard to directly relate the pattern we observed to a plant origin based on GC-MS results alone.
@@
 
-```{r chromatograms-specific, echo=FALSE, fig.cap= "GC-MS chromatograms of the potshard with the presence of long-chain fatty acids.
+```{r chromatograms-specific, echo=FALSE, fig.cap= "GC-MS chromatograms of the potsherd with the presence of long-chain fatty acids.
 Cn:x refers to fatty acid with n carbon atoms and x double bonds.
@@
 Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
-This also corresponds to the lipid distribution of the same potshard which reflects a possible plant origin based on the GC-MS results.
+This also corresponds to the lipid distribution of the same potsherd which reflects a possible plant origin based on the GC-MS results.
 Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.
@@
 Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period.
-Moreover, it is striking to identify a non-ruminant source from a Guijiabao potshard that might originate from C3 plants.
+Moreover, it is striking to identify a non-ruminant source from a Guijiabao potsherd that might originate from C3 plants.
 Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots.
 The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020].
-But we also found a base sample dates to early Neolithic period contains a rich composition of long-chain fatty acids, reflecting a good preservation.
-We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes.
-Also, the lack of local modern references for animal fats and plants has impacts on our explanation.
+But we also found a base sample dating to the early Neolithic period contains a rich composition of long-chain fatty acids, reflecting good preservation.
+We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes, and a lack of local modern references for animal fats and plants.
 However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.
@@
 millets), or new species (eg.
-pigs/horses) became an essential part of local diet in Yanyuan during the Neolithic transition.
-We successfully identified non-ruminant fats from a Guijiabao potshard that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
+pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic transition.
+We successfully identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
 It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016].
@@
 Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022].
-Although freshwater fish could be another origin of non-ruminant fats that can suggested by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
+Freshwater fish could be another origin of non-ruminant fats based on the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], but we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
 
@@
 The evidence from Guijuabao also confirms a later presence of horse remains.
-In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on C3 diet based on global references [@Evershed2002].
+In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002].
 It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016].
@@
 This corresponds with the timing when new crops spread into southwest China.
-Despite of no C4 plant origin in our samples, one potshard implies a source of C4 plant-fed ruminant.
+Despite of no C4 plant origin in our samples, one potsherd implies a source of C4 plant-fed ruminant.
 In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife.
@@
 The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China.
-It is most likely the terrestrial plant signal is C3 plant, implying rice processing in the pot.
+It is most likely the terrestrial plant signal from C3 plant, implying rice processing in the pot.
 However, the evidence of direct processing of terrestrial plants is not as much as we expected according to a relatively large quantity of plant remains found at Guijiabao, such as millet grains.

### 2023-02-06 analysis/paper/paper.Rmd
added an abstract and minor editing

@@
 author:
-  - Li-Ying Wang Mylastname:
-      email: fl@oneplace.edu
+  - Li-Ying Wang:
+      email: liyingw@asihp.net
       institute: [UofO]
       correspondence: true
-  - Yourlastname:
-      email: fl@another.edu
+  - Kuei-Chen Lin:
+      email: kueichenlin@asihp.net
       institute: [UofA]
-      correspondence: false
+      correspondence: true
 institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
+  - UofO: Institute of History and Philology, Academia Sinica 
+  - UofA: Institute of History and Philology, Academia Sinica 
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 abstract: |
-  Text of abstract
+    The spread of new crops or species across areas during the Neolithic transition is one of the important archaeological issues to understand the process of adoption of new practices.
+It is widely accepted that mixed farming of rice/millet was introduced into Southwest China from Central and Northern China around 5000 BP.
+However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear.
+The excavation of Guijiabao (ca.
+5000-800 BP) in the Yanyuan basin in Southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture.
+To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarkers identification and compounds-specific stable isotope analysis.
+Our results show the signals of terrestrial animals for most potsherds with one showing a possible source from C3 plants during the Neolithic transition.
+Our case study indicates hunting might be the primary subsistence, supplemented by rice/millet farming, and also demonstrates a perspective from organic residue analysis to explore food consumption during the Neolithic period.
+A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diet and subsistence.
+
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  Lipid residue analysis; Ceramics; Southwest China; Stable isotope analysis; Food practices
 highlights: |
@@
 <!-- intro to the topic and research aim-->
-Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could result in far-reaching economic or cultural impacts on local societies.
+Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000].
 In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022].
@@
 Therefore, the Neolithic Hengduan Mountain areas are key to understanding the processes of agriculture expansion.
-Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022].
+Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022; @Weisskopf2015].
 However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood.
@@
 
-The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies, and culinary practices across a region [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022].
+The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies or culinary practices across a region or over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
 Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021].
@@
 About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016].
+This coresponds with a C4 based diet based on isotopic analysis on four human bones [@Lin2022].
 The earliest domesticated millet grain can be traced back to 5000 BP according to its radiocarbon date [@Hao2022].
@@
 <!--info about Gaoshan-->
-To have a comparison, we also look at a sample from Gaoshan, which is located on the southwestern margin of the Chengdu Plain (Figure \@ref(fig:SW-china-map)).
-The Chengdu Plain is an alluvial plain between the Sichuan Basin and the Tibetan Plateau.
-Excavations at Gaoshan revealed abundant materials, including pottery, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
-According to the pottery forms and decorations, Gaoshan can be categorized into the earliest phase of the Baodun period, the earliest Neolithic culture (ca.
-4700-3700 BP) in the Chengdu Plain, or even earlier than that [@Guedes2013; @Gaoshan2017].
-A large quantity of carbonized rice grains was found at Gaoshan with a relatively small number of foxtail/broomcorn millets, suggesting rice is the main crop supplemented by millets [@JiangGaoshan2017].
+To have a comparison, we also look at a sample from Gaoshan in southwestern Chengdu Plain, which is an alluvial plain located between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
+Excavations at Gaoshan revealed abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
+According to the pottery forms and decorations, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca.
+4700-3700 BP) in the Chengdu Plain [@Guedes2013; @Gaoshan2017].
+A large quantity of carbonized rice grains and isotopic analysis results suggests that rice is the main crop supplemented by millets [@JiangGaoshan2017; @Lee2020].
 This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010].
@@
 
-Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP, what their proportions are in local culinary practices during the early time are unclear.
-For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in neighboring sites, such as Gaoshan and Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013].
+Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP, what their proportions are in local culinary practices during the early time remains unclear.
+For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in the neighboring sites, such as Gaoshan, Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013].
 The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks.
@@
 knitr::kable(Ch_sherds,
-             caption = "Quantitative data of fatty acid (FA) composition (ug) of potsherds from Guijiabao and Gaushan that are named starting with 'SYG' and 'CDG' respectively")
+             caption = "A list of the pottery sherds analyzed in this study. The sherd labeld with 'SYG' means potsherds from Guijiabao, while the label 'CDG' refers to Gaoshan")
 ```
@@
 To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
-The stable carbon isotope ratio for individual lipid compounds is powerful indicator of the origins of organic residues since different food categories, such as terrestrial animal, marine animals and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].
+The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of organic residues since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].
 
@@
 A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour.
-After drying under a stream of nitrogen, the derivitized sample was dissolved in toluene (20 uL) prior to GC-MS.
+After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS.
 Internal standards were added before and after the total lipid extraction.
@@
 The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively.
-The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degree C overnight.
-The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen stream.
+The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degrees C overnight.
+The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen streams.
 About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification and the remaining was analyzed by GC-C-IRMS.
@@
 The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode.
-The GC oven was held at 90°C for 1.3 minutes, and then raised to 320°C at a rate of 5°C/min.
+The GC oven was held at 90°C for 1.3 minute, and then raised to 320°C at a rate of 5°C/min.
 The final temperature was held for 22 minutes.
@@
 knitr::kable(quant_FA_tidy,
-             caption = "Quantitative data of fatty acid (FA) composition (ug) of potsherds from Guijiabao and Gaushan that are named starting with 'SYG' and 'CDG' respectively")
+             caption = "Quantitative data of fatty acid (FA) composition (μg) for each pottery sample based on GC-FID analysis")
 ```
@@
 
-One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of ceramic or a different use.
+One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of the ceramic or a different use.
+This sample also has charred interior surface, indicating its function to be associated with food processing.
 Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)).
@@
 But we also found a base sample dating to the early Neolithic period contains a rich composition of long-chain fatty acids, reflecting good preservation.
-We recognize the limitation for interpreting food choices over time or differences across regions because of our small sample sizes, and a lack of local modern references for animal fats and plants.
+We recognize limitations and challenges for interpreting lipid source over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plants.
 However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.
@@
 Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004].
-The evidence from Guijuabao also confirms a later presence of horse remains.
+The evidence from Guijiabao also confirms a later presence of horse remains.
 In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002].
@@
 However, we are unable to determine whether there is a shift in diet over time due to our limited sample size.
-But the overall depleted δ^13^C values for most potsherds suggest C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition.
-This corresponds with the timing when new crops spread into southwest China.
-Despite of no C4 plant origin in our samples, one potsherd implies a source of C4 plant-fed ruminant.
+But the overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition.
+This corresponds with the timing when new crops were introduced into southwest China.
+Despite no C4 plant origin was found, one potsherd shows the signal of C4 plant-fed ruminant.
 In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife.
-This might be related to local ecological zones and environments where wild animals were easier to obtain.
-Since zooarchaeological analysis and other studies on stone tools are still in progress, we could only rely on preliminary excavation reports with limited details about associated artifacts.
+This might be related to local ecological zones and environments where wild animals were easier to obtained compared widely adoption of agriculture.
+Since zooarchaeological analysis and studies of other artifacts such as stone tools are still in progress, we could only rely on preliminary excavation reports with preliminary results of other lines of evidence.
 A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.
 
--adoption of farming in a relatively enclosed circumstances?
+- discuss the conflict with the results of isotopic analysis for human bones
 
 # Conclusion
-
 Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds.
@@
 # Acknowledgements
+We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples.
+We thank Julian Sachs in the Department of Oceanography, University of Washington for his support to let us use their laboratory for sample preparation and conducting lipid extraction and analysis.
+We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines.
+We also want to express our gratitude to the anonymous reviewers for their comments that help us to improve this paper.
 

### 2023-02-09 analysis/paper/paper.Rmd
edited the whole paper

@@
 ---
-title: "Investigating Neolithic pottery usage from Guijiabao in the Yanyuan Basin in Southwest China using organic residue analysis"
+title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
 author:
@@
     The spread of new crops or species across areas during the Neolithic transition is one of the important archaeological issues to understand the process of adoption of new practices.
-It is widely accepted that mixed farming of rice/millet was introduced into Southwest China from Central and Northern China around 5000 BP.
+It is widely accepted that mixed farming of rice/millet was introduced into southwest China from central and northern China around 5000 BP.
 However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear.
-The excavation of Guijiabao (ca.
-5000-800 BP) in the Yanyuan basin in Southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture.
-To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarkers identification and compounds-specific stable isotope analysis.
-Our results show the signals of terrestrial animals for most potsherds with one showing a possible source from C3 plants during the Neolithic transition.
-Our case study indicates hunting might be the primary subsistence, supplemented by rice/millet farming, and also demonstrates a perspective from organic residue analysis to explore food consumption during the Neolithic period.
+The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture.
+To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis.
+The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic transition.
+Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming.
+This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period.
 A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diet and subsistence.
@@
 Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022].
-Therefore, the Neolithic Hengduan Mountain areas are key to understanding the processes of agriculture expansion.
-Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the introduction of cultivated millet and rice simultaneously during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022; @Weisskopf2015].
-However, how and to what extent the two crops were incorporated into the local cultures is still not fully understood.
+As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes.
+Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the simultaneous introduction of millet and rice during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022; @Weisskopf2015].
+However, to what extent the two crops were incorporated into the local diet is still not fully understood.
 
 To explore whether the new crops or species became important food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a].
-The recent evidence at Guijiabao reveals millet as the main crop with little rice according to the results of flotation [@Guijiabao2016], different from the abundant rice grains found at other Neolithic sites in that area.
-Although the analysis of plant remains suggests millet/rice farming in the early Neolithic, we still lack direct evidence of culinary practices during this period.
+The recent evidence at Guijiabao reveals that millet and rice were introduced in the early Neolithic period around 5000 BP.
+and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022].
+Although the analyses suggest the introduction of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
 To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use?
-Could we identify the evidence of new crops or species incorporated into the local food choices in the Early Neolithic period?
-We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels during the Neolithic transition in the Yanyuan Basin.
+Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic transition?
+We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.
 
-The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies or culinary practices across a region or over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
-Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers to trace their original foodstuffs [@Copley2005; @Whelton2021].
-This is because lipid molecules can be accumulated and well preserved in the porous fabric of pottery over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017].
-In this study, we detect and identify lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020].
-We attempt to explore the culinary practice under the circumstances during the time of frequent cultural interactions and expansion of agriculture through our case study.
+The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
+Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing their original foodstuffs [@Copley2005; @Whelton2021].
+This is due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017].
+In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020].
+We attempt to investigate culinary practice during a period of frequent cultural interactions with agricultural expansion through our case study.
 
@@
 
-```{r SW-china-map, fig.cap = "The archaeological sites, Guijiabao and Gaoshan, and other locations in Southwest China that are mentioned in this study. Map data is from naturalearthdata.com"}
+```{r SW-china-map, fig.cap = "The archaeological sites, Guijiabao and Gaoshan, and other locations in southwest China that are mentioned in this study. Map data is from naturalearthdata.com"}
 library(here)
@@
 We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)).
-Guijiabao is especially unique due to its location as the intersection of the middle Hengduan Mountain ranges, the Sichuan Basin, and the Yunnan-Guizhou Plateau, with a continuous occupation from the early Neolithic period to the historical period [@Guijiabao2017a; @Guijiabao2017b].
-A survey in 2015 and subsequent excavations from 2016 to 2019 reveal abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022].
+Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the early Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
+A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022].
 The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming Dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022].
-For the Neolithic period, it can be further separated into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles.
+For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles.
 
@@
 The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins.
-The common ceramic styles in the early Neolithic period include grayish high-tempered fine ware and reddish brown coarse ware with surface decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a].
-Among the ceramics, incised geometric patterns and impressed dot patterns that resemble ceramic decorations found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan [@Guijiabao2017a].
-About the plant remains, the flotation analysis in 2015 reveal crop cultivation dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on samples collected from ash pits and stratigraphic layers [@Guijiabao2016].
-This coresponds with a C4 based diet based on isotopic analysis on four human bones [@Lin2022].
-The earliest domesticated millet grain can be traced back to 5000 BP according to its radiocarbon date [@Hao2022].
-A smaller number of rice grains (*Oryza sativa*) have also been found in ash pits where the earliest one dates to 4700 BP [@Hao2022].
-Later, a detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, indicative of a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].
+The representative ceramic styles in the early Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a].
+Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and southern part of the Hengduan Mountains [@Guijiabao2017a].
+Plant remains reveal the cultivation of crops dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022].
+This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022].
+In addition to millet, a smaller number of rice grains (*Oryza sativa*) have also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022].
+A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, suggesting a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].
 
 <!--info about Gaoshan-->
-To have a comparison, we also look at a sample from Gaoshan in southwestern Chengdu Plain, which is an alluvial plain located between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
-Excavations at Gaoshan revealed abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
-According to the pottery forms and decorations, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca.
+To compare, we examine a sample from Gaoshan in the southwestern Chengdu Plain, an alluvial plain located between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
+Excavations at Gaoshan reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
+According to similar pottery forms and styles, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca.
 4700-3700 BP) in the Chengdu Plain [@Guedes2013; @Gaoshan2017].
-A large quantity of carbonized rice grains and isotopic analysis results suggests that rice is the main crop supplemented by millets [@JiangGaoshan2017; @Lee2020].
-This also indicates the spread of rice cultivation from the Middle and Lower Yangzi valley to Sichuan in southwest China occurred at least 4500 years ago [@Chi2010].
-Moreover, recent research on animal bones in 2015 and 2016 shows the predominance of domesticated pigs, with a small number of domesticated dogs [@He2020].
-The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation.
+The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020].
+It is suggested that the spread of rice farming into the Sichuan area occurred at least 4500 years ago from the Middle and Lower Yangzi valleys [@Chi2010].
 
-Although the current finding at Guijiabao and Gaoshan support the spread of mixed farming of millet/rice into southwest China around 5000 BP, what their proportions are in local culinary practices during the early time remains unclear.
-For example, the analysis of plant remains and phytolith analysis for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in the neighboring sites, such as Gaoshan, Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013].
-The analysis of plant remains offers information on the adoption of crop cultivation and subsistence strategies, but evidence directly reflects food preparation or processing using ceramics still lacks.
-This study aims to provide direct evidence from food residues in ceramics to investigate culinary practices associated with the cultivation of new crops.
+Although the current findings at Guijiabao and Gaoshan support the spread of mixed agriculture into southwest China around 5000 BP, the proportions of millet or rice included in local culinary practices during the early Neolithic period remain unclear.
+The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in the neighboring sites, such as Gaoshan, Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013].
+The approach of residue analysis for ceramics reflecting food preparation and processing could be helpful to understand the adoption of new practices.
 
@@
 ## Samples
-We selected six ceramic fragments from Guijiabao and Gaoshan for organic residue analysis (Table \@ref(tab:sherds-table)).
-Four ceramic fragments from Guijiabao were recovered from trench 13 excavated in the 2017 field season and one fragment was from the 2016 season.
-Those potsherds can be divided into two temporal phases, the early Neolithic period (ca.
-4500 BP) and the historical period (ca.
-1000 BP).
-Our samples are mostly grayish base fragments with fine sand temper and a number of them have impressed decoration.
-Although their original forms could not be recovered, those ceramics are supposed to be used in daily life according to the evidence of chars on some fragments and their archaeological contexts in cultural layers [@Hao2022; @Guijiabao2017a].
+We selected ceramic fragments that were identified as typical types from Guijiabao and Gaoshan.
+The fragments were from known contexts with radiocarbon dates: early Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)).
+Ceramic fragments from Guijiabao were recovered from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season.
+The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations.
+Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that those ceramics were used in daily life [@Hao2022, @Guijiabao2017a].
 This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.
@@
 knitr::kable(Ch_sherds,
-             caption = "A list of the pottery sherds analyzed in this study. The sherd labeld with 'SYG' means potsherds from Guijiabao, while the label 'CDG' refers to Gaoshan")
+             caption = "A list of the pottery sherds that were analyzed in this study. The sherd labeled 'SYG' means potsherds from Guijiabao, while the label 'CDG' refers to Gaoshan")
 ```
@@
 ## Methods
-For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS), and quantify lipids by gas chromatography with flame-ionization detection (GC-FID).
+For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS) and quantified lipids by gas chromatography with flame-ionization detection (GC-FID).
 To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
-The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of organic residues since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].
+The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].
 
 For the sample processing, we cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis.
-Approximately 3.5 grams powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 degrees C).
-We conducted saponification by adding 10 mL 1M KOH in methanol and 1 mL water into the extracts and heated them in an oven at 70 degrees C overnight.
+Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C).
+We conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight.
 Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL).
-A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 degrees C for one hour.
+A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour.
 After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS.
@@
 
-The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol respectively.
-The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 degrees C overnight.
-The FAME derivatives were extracted with hexane ($3\times10$ mL) and blown down with nitrogen streams.
-About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification and the remaining was analyzed by GC-C-IRMS.
+The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol, respectively.
+The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight.
+The FAME derivatives were extracted with hexane ($3\times10$ mL) and then blown down with nitrogen streams.
+About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed by GC-C-IRMS.
 
 The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector.
-Samples were injected in splitless mode at 320 °C and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness).
-Helium carrier is in constant flow rate of 1.3 ml/min.
-The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 min.
+Samples were injected in splitless mode at 320 °C, and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness).
+Helium carrier is at a constant flow rate of 1.3 ml/min.
+The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 minutes.
 The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode.
@@
 The sample injection volume was 1 ul in toluene and was injected in splitless mode.
-Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set from 90 °C for 1.5 minutes, and then raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold.
+Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set at 90 °C for 1.5 minutes and then raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold.
 Helium was used as a gas carrier in constant flow mode at 1.24 ml/minute.
+
+The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at XXX to enable research reproducibility and transparency.
 
@@
 
-The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but with low amounts in most of them (Table \@ref(tab:FA-composition-table)).
-The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid), are the main compounds in most potsherds (Figure \@ref(fig:chromatograms-plots-four)).
+The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 g/g) [@Evershed2008experimental], but in relatively small amounts in the majority of them (Table \@ref(tab:FA-composition-table)].
+The lipid profiles show that the main compounds in most potsherds are saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)).
 Those fatty acids commonly exist in animal fat and plant oil.
-They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007].
-High concentrations of Medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
+They are also generally degraded from larger molecules, such as triacylglycerols, after long-term burial processes due to oxidation and microbial degradation [@Dudd1998; @Eerkens2007].
+High concentrations of medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
 In addition, a number of potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
-Although a larger content of C18:1 and C16:0 with smaller content of C18:0 might indicate an origin of plant oils [@Mayyas2018; @Copley2005], it is hard to directly relate the pattern we observed to a plant origin based on GC-MS results alone.
+Although a higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005], it is hard to directly relate the pattern we observed to a plant origin based on GC-MS results alone.
 
@@
 
-```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao.
+```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao abd Gaoshan.
 Cn:x refers to fatty acid with n carbon atoms and x double bonds.
@@
 
-One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of the ceramic or a different use.
-This sample also has charred interior surface, indicating its function to be associated with food processing.
-Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)).
+One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of ceramic or a different use.
+This sample also has a charred interior surface, indicative of a function associated with food processing.
+Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than medium-chain components (Figure \@ref(fig:chromatograms-specific)).
 The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005].
-However, it could be also associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
+However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
 Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.
 
-```{r chromatograms-specific, echo=FALSE, fig.cap= "GC-MS chromatograms of the potsherd with the presence of long-chain fatty acids.
+```{r chromatograms-specific, echo=FALSE, fig.cap= "The MS chromatogram of the Guijiabao potsherd shows the presence of long-chain fatty acids.
 Cn:x refers to fatty acid with n carbon atoms and x double bonds.
-IS indicates internal standard used for quantification, including FA C21:0 and n-alkane 34"}
+IS indicates internal standards used for quantification, including FA C21:0 and n-alkane 34"}
 knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
@@
 To identify the origins of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions δ^13^C values of C16:0 and C18:0 using GC-C-IRMS.
-The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on different ways where dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
+The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
 To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
@@
 
-The results show that most potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A).
+The results show that the majority of Guijiabao potsherd samples contain fatty acids from ruminant adipose fats, with one potsherd (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A).
 The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
-A number of potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
-Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) falling within the range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
-This also corresponds to the lipid distribution of the same potsherd which reflects a possible plant origin based on the GC-MS results.
+The potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
+Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low 13C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
+This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results.
 Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.
@@
 In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants.
-There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.
+There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively enriched δ^13^C value, suggesting a C4 plant diet for the ruminant animal.
 
@@
 <!--A brief summary of interesting results-->
-Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were mainly dominated by ruminant fats during the Neolithic period.
+Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were dominated by ruminant fats during the Neolithic period.
 Moreover, it is striking to identify a non-ruminant source from a Guijiabao potsherd that might originate from C3 plants.
-Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support that those ceramics were used as cooking or storage pots.
-The low yield might be related to the part of pottery samples analyzed in this study are mostly base fragments [@Evershed2008experimental; @Kwak2020].
-But we also found a base sample dating to the early Neolithic period contains a rich composition of long-chain fatty acids, reflecting good preservation.
-We recognize limitations and challenges for interpreting lipid source over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plants.
-However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices or the local subsistence at Guijiabao.
+Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots.
+The low yield could be attributed to the base parts that we selected for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993].
+But we also found a base sample dating to the early Neolithic period contained a rich composition of long-chain fatty acids, reflecting good preservation.
+We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils.
+Also, a detailed analysis of faunal remains from Guijiabao is still in progress, which constrains our discussion.
+However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices in Guijiabao.
 
 <!--discussion about possible plant origins-->
-One aim of our study is to explore whether new crops, such as C3 plants (eg.
-rice), C4 plants (eg.
-millets), or new species (eg.
-pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic transition.
+One aim of our study is to explore whether new crops, such as C3 plants (e.g., rice), C4 plants (e.g., millets), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic transition.
 We successfully identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
-It is also possible other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016].
-However, the seeds of wild weeds only account for a small part of plant remains.
-Surprisingly, we only observed signals of C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022].
+It is also possible that other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016].
+However, the seeds of wild weeds only account for a small part of the plant remain.
+Surprisingly, we only observed signals from C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022].
 Freshwater fish could be another origin of non-ruminant fats based on the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], but we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
@@
 Ruminant fats identified in the Neolithic potsherds (ca.
-4500 BP) from Guijiabao and Gaoshan could be derived from deer or horse according to unearthed faunal remains from the sites, accompanied with stone tools for hunting, such as stone balls and arrow-heads [@Hao2022; @Guijiabao2020a].
-But whether horse was part of local food choices in the Neolithic period is questionable.
-First, the Pleistocene and Neolithic sites with wild horse bones primarily concentrated in northern China [@Yuan2006].
-Second, domesticated horses first found in southwest China were much later from the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004].
+4500 BP) from Guijiabao and Gaoshan could be derived from deer or horses according to unearthed faunal remains from the sites, which were accompanied with stone tools for hunting, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a].
+However, whether horse was a part of local food choices during the Neolithic period is uncertain.
+First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China [@Yuan2006].
+Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the early Neolithic.
 The evidence from Guijiabao also confirms a later presence of horse remains.
 In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002].
-It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were part of food sources during this period [@Zhang2016].
-A detailed analysis of faunal remains from Guijiabao is still in progress that can provide further information.
+It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were food sources during this period [@Zhang2016].
 
-We observed a mixture of ruminant fats and terrestrial plants based on a number of Guijiabao potsherds dating to the historical period (ca.
+We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds dating to the historical period (ca.
 1000 BP).
-However, we are unable to determine whether there is a shift in diet over time due to our limited sample size.
-But the overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important diet for local animals and could also be included in the local diet in Yanyuan during the Neolithic transition.
-This corresponds with the timing when new crops were introduced into southwest China.
-Despite no C4 plant origin was found, one potsherd shows the signal of C4 plant-fed ruminant.
-In general, the food source detected from our samples comprised mainly of ruminant fats that are more likely from wildlife.
-This might be related to local ecological zones and environments where wild animals were easier to obtained compared widely adoption of agriculture.
-Since zooarchaeological analysis and studies of other artifacts such as stone tools are still in progress, we could only rely on preliminary excavation reports with preliminary results of other lines of evidence.
-A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in food choices during the Neolithic transition.
+It appears that C3 plants could be processed more in the historical period compared to the early Neolithic.
+But it needs more studies to explore whether there is a shift in diet over time.
+The overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important food source for local animals and could also be included in the local diet at Guijiabao and Gaoshan during the Neolithic transition.
+Despite no C4 plant origin being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants.
+The evidence of stone knives and tools also indicates the practice of agricultural harvesting [@Zhou2019; @Hao2022; @Lin2022].
+We also note that our finding of a more C3-based diet based on ceramic residues is not consistent with the result of a C4-based diet according to isotopic analysis of human bones [@Lin2022].
+The contradiction could be caused by different sampling strategies yielding contradictory results or by residue analysis reflecting only a portion of the content of food practices.
+Nevertheless, all pieces of evidence support the existing theory of mixed rice/millet farming in southwest China.
+However, farming practices may not have been adopted as much as we initially thought.
 
-- discuss the conflict with the results of isotopic analysis for human bones
+In general, during the Neolithic transition, the food source detected from our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals.
+Unearthed stone tools, including arrowheads and balls, also suggest hunting activities.
+Considering that Guijiabao is located in highland areas and Gaoshan is close to Hengduan mountain ranges, a subsistence relying on hunting to a certain degree could be expected.
+Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the early Neolithic, and it might only account for a small proportion.
+The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison.
+An organic residue approach may only provide a narrow perspective on understanding local diets, but it could improve our current understanding of food sources through direct evidence cooked in ceramics.
+A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in the local diet during the Neolithic transition.
 
@@
 Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds.
-The food cooked in those ceramics include ruminant fats and terrestrial plants during the Neolithic transition when the mixing farming of rice/millet spread into southwest China.
+The food cooked in those ceramics included ruminant fats and terrestrial plants during the Neolithic transition, when the mixed farming of rice/millet spread into southwest China.
 It is most likely the terrestrial plant signal from C3 plant, implying rice processing in the pot.
-However, the evidence of direct processing of terrestrial plants is not as much as we expected according to a relatively large quantity of plant remains found at Guijiabao, such as millet grains.
-Instead, food processing in the early Neolithic is mostly ruminant fats that could originate from local wild animals, such as deer.
-This indicates that the proportion of rice/millet in the local diet is not high during the Neolithic transition, suggesting that hunting might be the primary subsistence, supplemented by rice/millet farming.
-We recognize the limitation of our interpretation of the local diet due to the small sample size.
-However, our study can still offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period.
+Surprisingly, the direct processing of either C3 or C4 plants is not as much as we expected, according to the relatively large quantity of plant remains found at Guijiabao, such as millet grains.
+Instead, ruminant fats from local wild animals, likely deer, accounted for a major part of early Neolithic food processing.
+This indicates that the proportion of rice/millet in the local diet was not high during the Neolithic transition, suggesting that hunting might have been an important part of local subsistence, supplemented by rice/millet farming.
+We recognize the limitations of our interpretation of the local diet due to the small sample size.
+Nevertheless, our study could offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period.
 A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.
@@
 We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples.
-We thank Julian Sachs in the Department of Oceanography, University of Washington for his support to let us use their laboratory for sample preparation and conducting lipid extraction and analysis.
+We thank Julian Sachs in the Department of Oceanography, University of Washington for his support in letting us use their laboratory for sample preparation and conducting lipid extraction and analysis.
 We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines.
-We also want to express our gratitude to the anonymous reviewers for their comments that help us to improve this paper.
+We also want to express our gratitude to the anonymous reviewers for their comments that help us improve this paper.
 

### 2023-02-10 analysis/paper/paper.Rmd
minor edits

@@
       email: kueichenlin@asihp.net
-      institute: [UofA]
+      institute: [UofO]
       correspondence: true
@@
 
-The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at XXX to enable research reproducibility and transparency.
+The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at XXX to enable research reproducibility and transparency [@Marwick2017].
 
@@
 
-The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 g/g) [@Evershed2008experimental], but in relatively small amounts in the majority of them (Table \@ref(tab:FA-composition-table)].
+The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but in relatively small amounts in the majority of them (Table \@ref(tab:FA-composition-table)].
 The lipid profiles show that the main compounds in most potsherds are saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)).

### 2023-02-12 analysis/paper/paper.Rmd
added color to the plot

@@
 In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022].
-Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2015; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022].
+Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022].
 As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes.
@@
                                             "SYG-3",
-                                            "CDG-062")))
+                                            "CDG-062"))) %>% 
+  mutate(period = ifelse(sample %in% c("SYG-TN13-E22-1", "SYG-3" ),
+                         "Historical", "Neolithic"))
 
@@
        aes(`13C C16:0`,`13C C18:0`)) +
-  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = sample, color = period)) +
   #ggrepel::geom_text_repel(aes(label = C13)) +
@@
        aes(`13C C16:0`, delta)) +
-  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = sample, color = period)) +
   theme_minimal() +

### 2023-02-12 analysis/code/isotope-analysis.R

@@ -29,7 +29,9 @@ delta_data <-
                                         "SYG-TN13-E22-2#2",
                                         "SYG-TN13-E23-3",
                                         "SYG-3",
-                                        "CDG-062")))
+                                        "CDG-062"))) %>%
+  mutate(period = ifelse(sample %in% c("SYG-TN13-E22-1", "SYG-3" ),
+                         "Historical", "Neolithic"))
 
 # filter the Meth corrected data
 delta_meth_cor <-
@@ -57,7 +59,9 @@ delta_meth_cor <-
                                             "SYG-TN13-E22-2#2",
                                             "SYG-TN13-E23-3",
                                             "SYG-3",
-                                            "CDG-062")))
+                                            "CDG-062"))) %>%
+  mutate(period = ifelse(sample %in% c("SYG-TN13-E22-1", "SYG-3" ),
+                         "Historical", "Neolithic"))
 
 # import a SVG for later ggplot overlaying with ellipses
 tem <- rsvg::rsvg(here::here("analysis", "figures","carbon-isotope-ellipses-template.svg"))
@@ -106,7 +110,7 @@ ggsave(here::here("delta_C16_C18_remove_pot6.png"),
 isotope_C16_C18 <-
   ggplot(delta_meth_cor, #delta_meth_cor/delta_blank_cor
        aes(`13C C16:0`,`13C C18:0`)) +
-  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = sample, color = period)) +
   #ggrepel::geom_text_repel(aes(label = C13)) +
   theme_minimal() +
   labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
@@ -126,7 +130,7 @@ isotope_C16_C18 <-
 big_delta_C16 <-
   ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
        aes(`13C C16:0`, delta)) +
-  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = sample, color = period)) +
   theme_minimal() +
   labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
        y = bquote(Delta*{}^13*"C 18:0 \u2030")) +

### 2023-02-14 analysis/paper/paper.Rmd
edit the YAML

@@
   - UofO: Institute of History and Philology, Academia Sinica 
-  - UofA: Institute of History and Philology, Academia Sinica 
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2023-02-21 analysis/paper/paper.Rmd
corrected the IS

@@
 Cn:x refers to fatty acid with n carbon atoms and x double bonds.
-IS indicates internal standard used for quantification, including FA C21:0 and n-alkane 34"}
+IS indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
 knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
@@
 Cn:x refers to fatty acid with n carbon atoms and x double bonds.
-IS indicates internal standards used for quantification, including FA C21:0 and n-alkane 34"}
+IS indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
 knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))

### 2023-02-25 analysis/paper/paper.Rmd
modified the FA table

@@
   quant_FA %>%
-  pivot_longer(!number&!sample,
+  pivot_longer(!number&!sample&!`g extracted`&!`%reco`,
                names_to = "FA", 
                values_to = "weight") %>% 
-  select(-number) %>% 
   filter(str_starts(FA, "ng") & !sample =="") %>% 
   mutate(FA = str_remove(FA, "ng "),
-         weight = round(weight/1000, 2)) %>% 
+         weight = round(weight/`%reco`/`g extracted`/1000, 2)) %>% 
+  select(-number, -`%reco`, -`g extracted`) %>% 
   pivot_wider(names_from = sample, 
@@
   quant_FA %>%
-  pivot_longer(!number&!sample,
+  pivot_longer(!number&!sample&!`g extracted`&!`%reco`,
                names_to = "FA", 
                values_to = "weight") %>% 
-  select(-number) %>% 
   filter(str_starts(FA, "ng") & !sample =="") %>% 
   mutate(FA = str_remove(FA, "ng "),
-         weight = as.character(round(weight/1000, 2))) %>% 
+         weight = round(weight/`%reco`/`g extracted`/1000, 2)) %>% 
   mutate(weight = ifelse(is.na(weight)|weight == "0", "-", weight)) %>% 
+  select(-number, -`%reco`, -`g extracted`) %>% 
   pivot_wider(names_from = sample, 
@@
 knitr::kable(quant_FA_tidy,
-             caption = "Quantitative data of fatty acid (FA) composition (μg) for each pottery sample based on GC-FID analysis")
+             caption = "Quantitative data of fatty acid (FA) composition (μg/g) for each pottery sample based on GC-FID analysis")
 ```

### 2023-02-27 analysis/paper/paper.Rmd
updated author info

@@
       email: liyingw@asihp.net
-      institute: [UofO]
+      institute: [AcS]
       correspondence: true
@@
       email: kueichenlin@asihp.net
-      institute: [UofO]
-      correspondence: true
+      institute: [AcS]
+      correspondence: false
+  - Zhiqing Zhou:
+      email: zhouzhiQ2000@163.com
+      institute: [InsC]
+      correspondence: false
 institute:
-  - UofO: Institute of History and Philology, Academia Sinica 
+  - AcS: Institute of History and Philology, Academia Sinica 
+  - InsC: Chengdu Institute of Cultural Relics and Archaeology
 date: "`r format(Sys.time(), '%d %B, %Y')`"

### 2023-02-27 analysis/paper/paper.Rmd
edit the intro based on KC's comments

--- 2023-02-25 analysis/paper/paper.Rmd
@@
 abstract: |
-    The spread of new crops or species across areas during the Neolithic transition is one of the important archaeological issues to understand the process of adoption of new practices.
-It is widely accepted that mixed farming of rice/millet was introduced into southwest China from central and northern China around 5000 BP.
+    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices.
+It is widely accepted that mixed farming of rice/millet was introduced into southwest China from central and northern China around 4700 BP.
 However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear.
@@
 This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period.
-A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diet and subsistence.
+A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.
 
@@
 Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000].
-In southwestern China, the Hengduan Mountain ranges and valleys connecting the Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022].
-Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 5,000 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Chi2010; @Dal2022].
+In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
+Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022].
 As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes.
-Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the simultaneous introduction of millet and rice during the early Neolithic period, similar to the mixed farming found in Neolithic central China [@Chen2015; @Hao2022; @Weisskopf2015].
+Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the presence of millet and rice at around the same time as mixed farming appeared in Sichuan [@Chen2015; @Hao2022; @Weisskopf2015].
 However, to what extent the two crops were incorporated into the local diet is still not fully understood.
@@
 To explore whether the new crops or species became important food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a].
-The recent evidence at Guijiabao reveals that millet and rice were introduced in the early Neolithic period around 5000 BP.
-and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022].
-Although the analyses suggest the introduction of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
+The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022].
+Although the analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
 To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use?

### 2023-03-01 analysis/paper/paper.Rmd
editing based on KC's comments

@@
 We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)).
-Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the early Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
+Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
 A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022].
 The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming Dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022].
-For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles.
+For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].
 
@@
 The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins.
-The representative ceramic styles in the early Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a].
+The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a].
 Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and southern part of the Hengduan Mountains [@Guijiabao2017a].
@@
 <!--info about Gaoshan-->
-To compare, we examine a sample from Gaoshan in the southwestern Chengdu Plain, an alluvial plain located between the Sichuan Basin and the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
+To compare culinary practices for different ecological zones, we examine a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
 Excavations at Gaoshan reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
 According to similar pottery forms and styles, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca.
-4700-3700 BP) in the Chengdu Plain [@Guedes2013; @Gaoshan2017].
+4700-3700 BP) in the Chengdu Plain [@Gaoshan2017].
 The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020].
-It is suggested that the spread of rice farming into the Sichuan area occurred at least 4500 years ago from the Middle and Lower Yangzi valleys [@Chi2010].
+It is suggested that the spread of rice farming into the Sichuan area occurred at least 4700 years ago from the Middle and Lower Yangzi valleys [@Zhang2010].
 
-Although the current findings at Guijiabao and Gaoshan support the spread of mixed agriculture into southwest China around 5000 BP, the proportions of millet or rice included in local culinary practices during the early Neolithic period remain unclear.
-The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice in the neighboring sites, such as Gaoshan, Baodun, Henglanshan, and Baiyangcun in Sichuan and Yunnan [@Hao2022; @Guedes2013].
+Although the current findings at Guijiabao and Gaoshan support mixed agriculture, the proportions of millet or rice included in local culinary practices during the Neolithic period remain unclear.
+The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice at most sites in southwest China, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017], and Henglanshan and Baiyangcun in Yunnan [@Hao2022].
 The approach of residue analysis for ceramics reflecting food preparation and processing could be helpful to understand the adoption of new practices.
@@
 We selected ceramic fragments that were identified as typical types from Guijiabao and Gaoshan.
-The fragments were from known contexts with radiocarbon dates: early Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)).
+The fragments were from known contexts with radiocarbon dates: Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)).
 Ceramic fragments from Guijiabao were recovered from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season.
@@
 
-The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at XXX to enable research reproducibility and transparency [@Marwick2017].
+The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at https://github.com/LiYingWang/GJBpottery to enable research reproducibility and transparency [@Marwick2017].
 
@@
 In addition, a number of potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
-Although a higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005], it is hard to directly relate the pattern we observed to a plant origin based on GC-MS results alone.
+A higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005].
+However, we may still need to examine their isotopic values to confirm their sources.
 
@@
 
-```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao abd Gaoshan.
+```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao and Gaoshan.
 Cn:x refers to fatty acid with n carbon atoms and x double bonds.
@@
 
-To identify the origins of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions δ^13^C values of C16:0 and C18:0 using GC-C-IRMS.
+To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions δ^13^C values of C16:0 and C18:0 using GC-C-IRMS.
 The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
@@
 The results show that the majority of Guijiabao potsherd samples contain fatty acids from ruminant adipose fats, with one potsherd (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A).
-The potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well.
+The potsherd from Gaoshan (CDG-062) demonstrates ruminant adipose fats as well.
 The potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
@@
 
-In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants.
-There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively enriched δ^13^C value, suggesting a C4 plant diet for the ruminant animal.
+In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting those ruminant animals seemed to have consumed mainly C3 plants.
+There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively enriched δ^13^C value, suggesting a C4-based diet for the ruminant animal.
 
@@
 <!--A brief summary of interesting results-->
+The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets), C3 plants (e.g., rice), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period.
 Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were dominated by ruminant fats during the Neolithic period.
@@
 We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils.
-Also, a detailed analysis of faunal remains from Guijiabao is still in progress, which constrains our discussion.
+Also, a detailed analysis for faunal remains from Guijiabao is still in progress, which constrains the comparison between different lines of evidence.
 However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices in Guijiabao.
@@
 <!--discussion about possible plant origins-->
-One aim of our study is to explore whether new crops, such as C3 plants (e.g., rice), C4 plants (e.g., millets), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic transition.
-We successfully identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
-It is also possible that other native grasses classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet [@Guijiabao2016].
-However, the seeds of wild weeds only account for a small part of the plant remain.
-Surprisingly, we only observed signals from C3 plants, despite the fact that C4 plants (foxtail and broomcorn millets) are the major crop based on the recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022].
-Freshwater fish could be another origin of non-ruminant fats based on the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b], but we did not detect any presence of biomarkers for aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
+While it has been indicated that millets (C4 plants), including foxtail and broomcorn millets, were dominated cereal according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples.
+Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
+It is also possible that native wild weeds classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet, but those seeds only account for a small part of the plant remain according to flotation results [@Guijiabao2016].
+Despite stone knives and tools were also found at the sites, it is uncertain if they were used for the practice of agricultural harvesting [@Zhou2019; @Hao2022; @Lin2022].
+Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b].
+However, we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
 
@@
 However, whether horse was a part of local food choices during the Neolithic period is uncertain.
-First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China [@Yuan2006].
-Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the early Neolithic.
-The evidence from Guijiabao also confirms a later presence of horse remains.
-In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with those values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002].
+First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China with only sporadic presence in the south [@Yuan2006].
+Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on.
+In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with the values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002].
 It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were food sources during this period [@Zhang2016].
@@
 It appears that C3 plants could be processed more in the historical period compared to the early Neolithic.
-But it needs more studies to explore whether there is a shift in diet over time.
-The overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important food source for local animals and could also be included in the local diet at Guijiabao and Gaoshan during the Neolithic transition.
-Despite no C4 plant origin being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants.
-The evidence of stone knives and tools also indicates the practice of agricultural harvesting [@Zhou2019; @Hao2022; @Lin2022].
-We also note that our finding of a more C3-based diet based on ceramic residues is not consistent with the result of a C4-based diet according to isotopic analysis of human bones [@Lin2022].
+The source of C3 plants might also include wheat since it was introduced in the historical period [@JiangGaoshan2017].
+More studies and larger sample size are needed to explore whether there is a shift in diet over time.
+The overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important food source for local animals and could also be included in the local diet at Guijiabao during the Neolithic period.
+Despite no C4 plant signal being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants.
+We also note that that our finding of C3 plant signal based on ceramic residues is not consistent with the result of a C4-based diet according to isotopic analysis of human bones [@Lin2022].
 The contradiction could be caused by different sampling strategies yielding contradictory results or by residue analysis reflecting only a portion of the content of food practices.
 Nevertheless, all pieces of evidence support the existing theory of mixed rice/millet farming in southwest China.
-However, farming practices may not have been adopted as much as we initially thought.
+Our results indicate that farming practices may not have been adopted as much as we initially thought.
 
-In general, during the Neolithic transition, the food source detected from our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals.
+In general, during the Neolithic period, the food source detected from our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals such as deer.
 Unearthed stone tools, including arrowheads and balls, also suggest hunting activities.
-Considering that Guijiabao is located in highland areas and Gaoshan is close to Hengduan mountain ranges, a subsistence relying on hunting to a certain degree could be expected.
-Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the early Neolithic, and it might only account for a small proportion.
+Considering the abundance of wild animal resources in the natural environment of the Hengduan mountain ranges suggested by the case of Yingpanshan site [@He2020], a subsistence relying on hunting to a certain degree could be expected.
+Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, and it might only account for a small proportion.
 The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison.
 An organic residue approach may only provide a narrow perspective on understanding local diets, but it could improve our current understanding of food sources through direct evidence cooked in ceramics.
-A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in the local diet during the Neolithic transition.
+A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in the local diet during the Neolithic period.
 

### 2023-03-05 analysis/paper/paper.Rmd
minor edits

@@
       correspondence: true
-  - Kuei-Chen Lin:
+  - Kuei-chen Lin:
       email: kueichenlin@asihp.net
@@
 Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000].
-In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
+In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
 Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022].
@@
 To explore whether the new crops or species became important food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a].
-The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022].
+The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022].
 Although the analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
@@
 A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022].
-The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming Dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022].
+The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022].
 For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].
@@
 The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a].
-Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and southern part of the Hengduan Mountains [@Guijiabao2017a].
+Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and the southern part of the Hengduan Mountains [@Guijiabao2017a].
 Plant remains reveal the cultivation of crops dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022].
@@
 <!--info about Gaoshan-->
-To compare culinary practices for different ecological zones, we examine a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
+To compare culinary practices for different ecological zones, we examined a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
 Excavations at Gaoshan reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
@@
 The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations.
-Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that those ceramics were used in daily life [@Hao2022, @Guijiabao2017a].
+Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that those ceramics were used in daily life [@Guijiabao2017a; @Hao2022].
 This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.
@@
 
-To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions δ^13^C values of C16:0 and C18:0 using GC-C-IRMS.
+To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS.
 The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
@@
 Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were dominated by ruminant fats during the Neolithic period.
-Moreover, it is striking to identify a non-ruminant source from a Guijiabao potsherd that might originate from C3 plants.
+Moreover, it is striking to identify non-ruminant fats in a Guijiabao potsherd that might originate from C3 plants.
 Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots.
 The low yield could be attributed to the base parts that we selected for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993].
-But we also found a base sample dating to the early Neolithic period contained a rich composition of long-chain fatty acids, reflecting good preservation.
+But we also found a base sample dating to the Neolithic period contained a rich composition of long-chain fatty acids, reflecting good preservation.
 We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils.
-Also, a detailed analysis for faunal remains from Guijiabao is still in progress, which constrains the comparison between different lines of evidence.
-However, the general pattern of the isotopic values of lipid molecules could still provide informative clues for a better understanding of culinary practices in Guijiabao.
+Also, a detailed analysis of faunal remains from Guijiabao is still in progress, which constrains the comparison between different types of evidence.
+However, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices in Guijiabao.
 
 <!--discussion about possible plant origins-->
-While it has been indicated that millets (C4 plants), including foxtail and broomcorn millets, were dominated cereal according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples.
-Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plant/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
-It is also possible that native wild weeds classified as C3 plants, such as pennisetum and chenopodiaceae, were part of the local diet, but those seeds only account for a small part of the plant remain according to flotation results [@Guijiabao2016].
-Despite stone knives and tools were also found at the sites, it is uncertain if they were used for the practice of agricultural harvesting [@Zhou2019; @Hao2022; @Lin2022].
+While it has been indicated that millets (C4 plants), including foxtail and broomcorn millets, were the dominant cereals according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples.
+Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plants/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
+Based on the flotation results from Guijiabao, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae; however, the seeds of wild weeds only account for a small part of the plant [@Guijiabao2016].
+Despite stone knives and tools were also found at the sites that might relate to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022].
 Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b].
-However, we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
+But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
 
@@
 Ruminant fats identified in the Neolithic potsherds (ca.
-4500 BP) from Guijiabao and Gaoshan could be derived from deer or horses according to unearthed faunal remains from the sites, which were accompanied with stone tools for hunting, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a].
-However, whether horse was a part of local food choices during the Neolithic period is uncertain.
-First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China with only sporadic presence in the south [@Yuan2006].
+4500 BP) from Guijiabao and Gaoshan could be derived from deer or horses, according to unearthed faunal remains from the sites and the local natural environments.
+Stone artifacts, including stone balls and arrowheads, were also found at the site, indicative of hunting activities [@Hao2022; @Guijiabao2020a].
+However, whether horse was part of local food choices in southwest China around 4500 years ago is skeptical.
+First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China, with only a sporadic presence in the south [@Yuan2006].
 Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on.
-In addition, one potsherd (SYG-TN13-E22-2#1) shows the isotopic value consistent with the values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002].
+In addition, our sample (SYG-TN13-E22-2#1) shows an isotopic value consistent with the values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002].
+Thus, deer is the most likely part of the local food choices.
 It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were food sources during this period [@Zhang2016].
@@
 1000 BP).
-It appears that C3 plants could be processed more in the historical period compared to the early Neolithic.
-The source of C3 plants might also include wheat since it was introduced in the historical period [@JiangGaoshan2017].
-More studies and larger sample size are needed to explore whether there is a shift in diet over time.
-The overall depleted δ^13^C values for most potsherds suggest that C3 plants might be an important food source for local animals and could also be included in the local diet at Guijiabao during the Neolithic period.
-Despite no C4 plant signal being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants.
-We also note that that our finding of C3 plant signal based on ceramic residues is not consistent with the result of a C4-based diet according to isotopic analysis of human bones [@Lin2022].
-The contradiction could be caused by different sampling strategies yielding contradictory results or by residue analysis reflecting only a portion of the content of food practices.
-Nevertheless, all pieces of evidence support the existing theory of mixed rice/millet farming in southwest China.
-Our results indicate that farming practices may not have been adopted as much as we initially thought.
+It appears that C3 plants could be processed more in the historical period compared to the Neolithic period.
+In addition to rice, the source of C3 plants could also include wheat since it was introduced during the historical period [@JiangGaoshan2017].
+A larger sample size are needed to explore whether there is a shift in diet over time.
+The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might be an important food source for local animals.
+Despite direct C4 plant signals not being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants during the Neolithic period.
+This indicates either C3 or C4 plants could be the food source for ruminants, but whether they were domesticated is unknown.
+The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison.
 
-In general, during the Neolithic period, the food source detected from our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals such as deer.
-Unearthed stone tools, including arrowheads and balls, also suggest hunting activities.
-Considering the abundance of wild animal resources in the natural environment of the Hengduan mountain ranges suggested by the case of Yingpanshan site [@He2020], a subsistence relying on hunting to a certain degree could be expected.
-Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, and it might only account for a small proportion.
-The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison.
-An organic residue approach may only provide a narrow perspective on understanding local diets, but it could improve our current understanding of food sources through direct evidence cooked in ceramics.
-A large-scale organic residue analysis of ceramics across different pottery forms in a wider region may offer insight into temporal and regional variations in the local diet during the Neolithic period.
+In general, during the Neolithic period, the food source detected in our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals such as deer.
+Considering the abundance of wild animal resources in the natural environment of the Hengduan mountain ranges, as suggested by the case of the Yingpanshan site [@He2020], a subsistence relying on hunting to a certain degree could be expected.
+Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, but it might only account for a small proportion.
+For the dietary pattern at Guijiabao previous study suggested a C4-based diet according to the isotopic analysis of human bones [@Lin2022], but we only identified a C3 plant signal from one pottery sample without any signal directly from C4 plants.
+This inconsistency could result from different sampling strategies or from the sample size issue, where our results only reflect a small portion of food content.
+Nevertheless, our results suggest that farming practices during the Neolithic period may not have been adopted as much as we previously thought.
+An organic residue approach may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through direct evidence that has been cooked in ceramics.
+A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period.
 
 # Conclusion
-Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids but in low concentrations for most potsherds.
-The food cooked in those ceramics included ruminant fats and terrestrial plants during the Neolithic transition, when the mixed farming of rice/millet spread into southwest China.
-It is most likely the terrestrial plant signal from C3 plant, implying rice processing in the pot.
-Surprisingly, the direct processing of either C3 or C4 plants is not as much as we expected, according to the relatively large quantity of plant remains found at Guijiabao, such as millet grains.
-Instead, ruminant fats from local wild animals, likely deer, accounted for a major part of early Neolithic food processing.
-This indicates that the proportion of rice/millet in the local diet was not high during the Neolithic transition, suggesting that hunting might have been an important part of local subsistence, supplemented by rice/millet farming.
+Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids that could be associated with food types.
+The food processed in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwest China.
+The source of terrestrial plants could be C3 plants, implying the processing of rice or other wild weeds in the pot.
+However, the evidence of direct processing of either C3 or C4 plants is not in proportion to the large quantity of plant remains found at Guijiabao, especially millet grains.
+Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the early Neolithic.
+This suggests that hunting might have been an important part of local subsistence, supplemented by rice/millet farming when they were introduced.
+Our results from Guijiabao might represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains.
 We recognize the limitations of our interpretation of the local diet due to the small sample size.
@@
 We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples.
-We thank Julian Sachs in the Department of Oceanography, University of Washington for his support in letting us use their laboratory for sample preparation and conducting lipid extraction and analysis.
+We thank Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for sample preparation and conducting lipid extraction and analysis.
 We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines.
-We also want to express our gratitude to the anonymous reviewers for their comments that help us improve this paper.
+We also want to express our gratitude to the anonymous reviewers for their comments that helped us improve this paper.
+This research is sponsored by the National Science and Technology Council (PE11006-0107).
 

### 2023-03-06 analysis/paper/paper.Rmd
minor edits

@@
 To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis.
-The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic transition.
+The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic period.
 Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming.
@@
 
-To explore whether the new crops or species became important food sources during the Neolithic transition in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a].
+To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a].
 The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022].
@@
 To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use?
-Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic transition?
+Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period?
 We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

### 2023-03-12 analysis/paper/paper.Rmd
minor edits

@@
     The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices.
-It is widely accepted that mixed farming of rice/millet was introduced into southwest China from central and northern China around 4700 BP.
+It is widely accepted that both millet and rice have been introduced into southwestern China at the latest in 4700 BP.
 However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear.
@@
 In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
-Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming spread to southwest Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022].
+Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015; @Weisskopf2015].
 As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes.
-Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the presence of millet and rice at around the same time as mixed farming appeared in Sichuan [@Chen2015; @Hao2022; @Weisskopf2015].
+Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the presence of millet and rice at around the same time as mixed farming appeared in Sichuan [@Huan2022; @Hao2022].
 However, to what extent the two crops were incorporated into the local diet is still not fully understood.
@@
 The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022].
-Although the analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
+Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
 To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use?
@@
 A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022].
-The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone [@Hao2022].
+The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone.
 For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].

### 2023-03-13 analysis/code/GCFID-quant.R
edit the texts on the chromatograph

@@ -134,9 +134,9 @@ meth_all %>%
   #filter(File > 55000) %>%
   ggplot(aes(Path, File)) +
   geom_line(size = 0.3) +
-  geom_text(data = highlight_SYG2_2, #ggrepel::geom_text_repel
+  ggrepel::geom_text_repel(data = highlight_SYG2_2, #ggrepel::geom_text_repel
             aes(label = FA),
-            size = 3,
+            size = 3.5,
             nudge_y = 250000,
             show.legend = FALSE) +
   scale_y_continuous(labels = scales::comma_format(),

### 2023-03-13 analysis/paper/paper.Rmd
minor edits

@@
     The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices.
-It is widely accepted that both millet and rice have been introduced into southwestern China at the latest in 4700 BP.
+It is widely accepted that both millet and rice have been introduced into southwest China at the latest in 4700 BP.
 However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear.
@@
 In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
-Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern Sichuan around 4,700 BP and to western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015; @Weisskopf2015].
-As a result, the Neolithic Hengduan Mountain areas are critical for understanding agricultural expansion processes.
-Recent archaeological evidence from the Yanyuan Basin, located in the middle Hengduan Mountains, reveals the presence of millet and rice at around the same time as mixed farming appeared in Sichuan [@Huan2022; @Hao2022].
+Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4,700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015; @Weisskopf2015].
+The Neolithic Hengduan Mountain regions are therefore essential for understanding the processes of agricultural expansion.
+Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as those grains found in Sichuan [@Huan2022; @Hao2022].
 However, to what extent the two crops were incorporated into the local diet is still not fully understood.
@@
 To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a].
-The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought, and became part of the local diet [@Guijiabao2016; @Lin2022; @Huan2022].
+The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Lin2022], and millet could have already become part of the local diet [@Lin2022].
 Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
@@
 
-Although the current findings at Guijiabao and Gaoshan support mixed agriculture, the proportions of millet or rice included in local culinary practices during the Neolithic period remain unclear.
-The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, which is at odds with the predominance of rice at most sites in southwest China, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017], and Henglanshan and Baiyangcun in Yunnan [@Hao2022].
-The approach of residue analysis for ceramics reflecting food preparation and processing could be helpful to understand the adoption of new practices.
+Although the recent findings at Guijiabao and Gaoshan support mixed agriculture, it is still unclear how much millet or rice was included in local culinary practices during the Neolithic period.
+The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011].
+However, plant remains from many sites in southwest China reveal a predominance of rice, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017] and Henglanshan and Baiyangcun in Yunnan [@Hao2022].
+This indicates there were variations in agricultural practices in southwest China.
+To better understand the extent of the adoption of new crops, organic residue analysis for ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking.
+This provides another way to examine the expansion of agriculture in the Yanyuan area.
 
@@
 ## Methods
-For organic residue analysis, we detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS) and quantified lipids by gas chromatography with flame-ionization detection (GC-FID).
+We detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS) and quantified lipids by gas chromatography with flame-ionization detection (GC-FID).
 To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
@@
 
-For the sample processing, we cleaned the surface of each potsherd to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis.
+For the sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis.
 Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C).
@@
 After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS.
-Internal standards were added before and after the total lipid extraction.
 
@@
 About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed by GC-C-IRMS.
+With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might from solvent and reagent.
+Internal standards were added before and after the lipid extraction for quantification.
 
@@
   geom_line(size = 0.3) +
-  geom_text(data = highlight_SYG2_2, #ggrepel::geom_text_repel
+  ggrepel::geom_text_repel(data = highlight_SYG2_2, #ggrepel::geom_text_repel
             aes(label = FA),
-            size = 3,
+            size = 3.5,
             nudge_y = 250000,
@@
 The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
-To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
+To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
 The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰.
@@
 The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets), C3 plants (e.g., rice), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period.
-Our results show that food choices indicated by lipid residues extracted from Guijiabao and Gaoshan potsherds were dominated by ruminant fats during the Neolithic period.
-Moreover, it is striking to identify non-ruminant fats in a Guijiabao potsherd that might originate from C3 plants.
+Our results from lipid residues in Guijiabao and Gaoshan potsherds show that food categories were dominated by ruminant fats during the Neolithic period.
+Moreover, it is striking to identify non-ruminant fats in a Guijiabao potsherd that might originate from C3 plants (see below).
 Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots.
-The low yield could be attributed to the base parts that we selected for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993].
-But we also found a base sample dating to the Neolithic period contained a rich composition of long-chain fatty acids, reflecting good preservation.
+The low yield could be attributed to the bottom parts of the vessels for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993].
+But we also found that one bottom sample dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples, reflecting better preservation or a much higher lipid yield.
 We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils.
-Also, a detailed analysis of faunal remains from Guijiabao is still in progress, which constrains the comparison between different types of evidence.
-However, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices in Guijiabao.
+Besides, a detailed analysis of faunal remains from Guijiabao is still in progress, which may constrain a comprehensive comparison between different types of evidence.
+Nevertheless, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices in Guijiabao.
 
@@
 Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plants/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
-Based on the flotation results from Guijiabao, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae; however, the seeds of wild weeds only account for a small part of the plant [@Guijiabao2016].
+According to the flotation results from Guijiabao, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae; however, the seeds of wild weeds only account for a small part of the floral assemblage derived from flotation [@Guijiabao2016].
 Despite stone knives and tools were also found at the sites that might relate to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022].
-Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Lucquin2016; @Guijiabao2017b].
-But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Kwak2017].
+Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b].
+But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016].
 
 <!--discussion about possible ruminant fat origin-->
-Ruminant fats identified in the Neolithic potsherds (ca.
-4500 BP) from Guijiabao and Gaoshan could be derived from deer or horses, according to unearthed faunal remains from the sites and the local natural environments.
-Stone artifacts, including stone balls and arrowheads, were also found at the site, indicative of hunting activities [@Hao2022; @Guijiabao2020a].
+The ruminant fats identified in the Neolithic potsherds (ca.
+4500 BP) from Guijiabao and Gaoshan could be derived from deer, sheep, or horses fed on a C3 diet based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017].
+This also corresponds with unearthed faunal remains from the sites and the local natural environments.
+Stone artifacts found at the site, including stone balls and arrowheads, further indicates hunting activities [@Hao2022; @Guijiabao2020a].
 However, whether horse was part of local food choices in southwest China around 4500 years ago is skeptical.
@@
 Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on.
-In addition, our sample (SYG-TN13-E22-2#1) shows an isotopic value consistent with the values of deer and sheep fats fed on a C3 diet based on global references [@Evershed2002].
-Thus, deer is the most likely part of the local food choices.
-It is also possible that other ruminant animals, such as wild Asian water buffalo or yak, were food sources during this period [@Zhang2016].
+Thus, supported by faunal evidence, deer is the most likely part of the local food choices.
+It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these species have been found in southwest China [@Zhang2016; @Chen2020].
 
-We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds dating to the historical period (ca.
+We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds that date to the historical period (ca.
 1000 BP).
-It appears that C3 plants could be processed more in the historical period compared to the Neolithic period.
+It appears that C3 plants could be processed more in the historical period than they were in the Neolithic period.
 In addition to rice, the source of C3 plants could also include wheat since it was introduced during the historical period [@JiangGaoshan2017].
-A larger sample size are needed to explore whether there is a shift in diet over time.
+A larger sample size is needed to explore whether there is a shift in diet over time.
 The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might be an important food source for local animals.
 Despite direct C4 plant signals not being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants during the Neolithic period.
-This indicates either C3 or C4 plants could be the food source for ruminants, but whether they were domesticated is unknown.
+This indicates either C3 or C4 plants could be the food source for ruminants, but whether those animals were domesticated is unknown.
 The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison.
 
-In general, during the Neolithic period, the food source detected in our pottery samples consisted mainly of ruminant fats, which are most likely from local wild animals such as deer.
-Considering the abundance of wild animal resources in the natural environment of the Hengduan mountain ranges, as suggested by the case of the Yingpanshan site [@He2020], a subsistence relying on hunting to a certain degree could be expected.
+In general, ruminant fats from local wild animals, such as deer, were the predominant food source during the Neolithic period based on our results.
+A subsistence that relied on hunting to a certain degree could be anticipated given the abundance of wild animals in the natural settings of the Hengduan mountain ranges, as indicated by the faunal remains from the Yingpanshan site [@He2020].
 Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, but it might only account for a small proportion.
-For the dietary pattern at Guijiabao previous study suggested a C4-based diet according to the isotopic analysis of human bones [@Lin2022], but we only identified a C3 plant signal from one pottery sample without any signal directly from C4 plants.
+For the dietary pattern at Guijiabao, previous study suggested a C4-based diet according to the isotopic analysis of human bones [@Lin2022], but we only identified a C3 plant signal from one pottery sample without any signal directly from C4 plants.
 This inconsistency could result from different sampling strategies or from the sample size issue, where our results only reflect a small portion of food content.
@@
 # Conclusion
-Our case study on organic residues preserved in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids that could be associated with food types.
+Our case study on organic residues in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids that was associated with food categories.
 The food processed in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwest China.
@@
 However, the evidence of direct processing of either C3 or C4 plants is not in proportion to the large quantity of plant remains found at Guijiabao, especially millet grains.
-Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the early Neolithic.
+Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the Neolithic.
 This suggests that hunting might have been an important part of local subsistence, supplemented by rice/millet farming when they were introduced.
-Our results from Guijiabao might represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains.
+Our results from Guijiabao could represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains.
 We recognize the limitations of our interpretation of the local diet due to the small sample size.
-Nevertheless, our study could offer a perspective from organic residue analysis to explore food preferences and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period.
+Nevertheless, our study could offer a perspective from organic residue analysis to explore food choices and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period.
 A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.

### 2023-03-13 analysis/paper/paper.Rmd
minor edits

@@
 In addition to millet, a smaller number of rice grains (*Oryza sativa*) have also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022].
-A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences, suggesting a mixed millet/rice cultivation prior to 5000 BP [@Huan2022].
+A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences [@Huan2022].
 

### 2023-03-15 analysis/paper/paper.Rmd
corrected some references

@@
 In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
-Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4,700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015; @Weisskopf2015].
+Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4,700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015].
 The Neolithic Hengduan Mountain regions are therefore essential for understanding the processes of agricultural expansion.
@@
 To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a].
-The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Lin2022], and millet could have already become part of the local diet [@Lin2022].
+The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already become part of the local diet [@Lin2022].
 Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
@@
 The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011].
-However, plant remains from many sites in southwest China reveal a predominance of rice, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017] and Henglanshan and Baiyangcun in Yunnan [@Hao2022].
+However, plant remains from many sites in southwest China reveal a predominance of rice, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017] and Henglanshan and Baiyangcun in Yunnan [@Jiang2016; @Henglanshan2014].
 This indicates there were variations in agricultural practices in southwest China.
@@
 4500 BP) from Guijiabao and Gaoshan could be derived from deer, sheep, or horses fed on a C3 diet based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017].
-This also corresponds with unearthed faunal remains from the sites and the local natural environments.
+This finding, especially the consumption of deer, also corresponds with unearthed faunal remains from the sites and the local natural environments.
 Stone artifacts found at the site, including stone balls and arrowheads, further indicates hunting activities [@Hao2022; @Guijiabao2020a].
@@
 Thus, supported by faunal evidence, deer is the most likely part of the local food choices.
-It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these species have been found in southwest China [@Zhang2016; @Chen2020].
+It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these species have been found in sites of southwest China [@Zhang2016; @Chen2020].
 

### 2023-03-31 analysis/paper/paper.Rmd
edit the intro

@@
 <!-- intro to the topic and research aim-->
-Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000].
-In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
-Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4,700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015].
-The Neolithic Hengduan Mountain regions are therefore essential for understanding the processes of agricultural expansion.
-Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as those grains found in Sichuan [@Huan2022; @Hao2022].
+Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on any people groups that are involved in the interactions that result from these movements [@Burmeister2000].
+The Hengduan Mountain ranges and valleys in southwestern China are geographical areas that connect the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau and by which movement to these plateaus must occur.
+They play an important role in the cultural exchanges and spread of crops into southwestern China and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
+Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015].
+The Hengduan Mountain regions in the Neolithic period are therefore essential for understanding the processes of agricultural expansion.
+Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as the grains found in Sichuan [@Huan2022; @Hao2022].
 However, to what extent the two crops were incorporated into the local diet is still not fully understood.
 
-To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a].
-The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already become part of the local diet [@Lin2022].
+To explore whether new crops or terrestrial animals became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues within ceramic vessels from Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a].
+The new evidence at Guijiabao reveals that millet and rice were both present there in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already been part of the local diet [@Lin2022].
 Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
-To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use?
-Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period?
-We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.
+To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery?
+Can we identify the new crops or terrestrial animals that were incorporated into the local food diets during the Neolithic period?
+We answer these questions by examining potsherds using organic residue analysis, which allows us to directly explore food cooked or processed in ceramic vessels.
 
-The recent growing application of organic residue analysis demonstrates its usefulness to understand ancient pottery use and function [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
-Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing their original foodstuffs [@Copley2005; @Whelton2021].
-This is due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017].
-In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reflect the original contents cooked or stored in a vessel during its use-life [@Dudd1999; @Miller2020].
-We attempt to investigate culinary practice during a period of frequent cultural interactions with agricultural expansion through our case study.
+The recent growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
+Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing food origins [@Copley2005; @Whelton2021].
+This is possible due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery for over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017].
+In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reveal the original contents cooked or stored in a vessel during its lifetime of use [@Dudd1999; @Miller2020].
+Through our case study, We attempt to investigate culinary practices during a period of frequent cultural interactions that accompanied agricultural expansion .
 
@@
 
-```{r SW-china-map, fig.cap = "The archaeological sites, Guijiabao and Gaoshan, and other locations in southwest China that are mentioned in this study. Map data is from naturalearthdata.com"}
+```{r SW-china-map, fig.cap = "The archaeological sites Guijiabao and Gaoshan and other locations in southwest China that are mentioned in this study. Map data is from naturalearthdata.com"}
 library(here)
@@
 <!--Yanyuan area and Guijiabao site-->
-We focus on Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)).
-Guijiabao is unique for understanding the early adoption of agriculture due to its rarity, which contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
+We focus on the site at Guijiabao located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)).
+The Guijiabao site allows us to understand the early adoption of agriculture because of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
 A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022].
-The chronology of Guijiabao can be divided into three temporal phases, including the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone.
+The chronology of the Guijiabao site can be divided into three temporal phases, the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone.
 For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].

### 2023-04-02 analysis/paper/paper.Rmd
minor edit and grammar check

@@
 abstract: |
-    The spread of new crops or species across areas during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new practices.
-It is widely accepted that both millet and rice have been introduced into southwest China at the latest in 4700 BP.
-However, to what extent the new crops were utilized by local cultures in an earlier time remains unclear.
-The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records, providing a unique opportunity to understand the local adoption of agriculture.
-To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds through lipid biomarker identification and compounds-specific stable isotope analysis.
-The results show the signals of terrestrial animals for most potsherds, with one showing a C3 plant source during the Neolithic period.
-Our case study indicates that hunting might have been the primary source of subsistence, supplemented by rice/millet farming.
-This study also demonstrates a perspective from organic residue analysis to explore pottery use and food practices during the Neolithic period.
-A more systematic sampling of potsherds across different types and temporal phases with larger samples is needed to provide a clear picture of ancient diets and subsistence.
+    The spread of new crops or terrestial animals across areas in southwest China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in those areas.
+It is widely accepted that both millet and rice were introduced into southwest China in 4700 BP at the latest.
+However, to what extent new crops were used by local peoples at an earlier time remains unclear.
+The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops.
+To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis.
+The results show signals of terrestrial animals in most potsherds and a C3 plant source in one potshard, during the Neolithic period.
+Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming.
+This study also demonstrates an approach using organic residue analysis to explore pottery use and culinary practices during the Neolithic period in southwest China.
+A more systematic sampling of different types of potsherds and temporal phases with larger samples is needed to provide a clearer picture of ancient diets and subsistence.
 
@@
 <!--Yanyuan area and Guijiabao site-->
-We focus on the site at Guijiabao located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)).
-The Guijiabao site allows us to understand the early adoption of agriculture because of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
-A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao is a settlement site [@Hao2022].
-The chronology of the Guijiabao site can be divided into three temporal phases, the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone.
-For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles [@Hao2022].
+We focus on the Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)).
+The Guijiabao site allows us to understand the early adoption of agriculturebecause of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
+A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao was a settlement site [@Hao2022].
+The chronology of the Guijiabao site can be divided into three phases, the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone.
+For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles that were identified [@Hao2022].
 
 <!--the major finding at GJB-->
-The common vessel forms unearthed at Guijiabao include jars, urns, bowls, and basins.
+The common vessel forms unearthed at the Guijiabao site include jars, urns, bowls, and basins.
 The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a].
-Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between Guijiabao and sites in Yunnan and the southern part of the Hengduan Mountains [@Guijiabao2017a].
-Plant remains reveal the cultivation of crops dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022].
+Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between the Guijiabao site and sites in Yunnan and the southern part of the Hengduan Mountains [@Guijiabao2017a].
+Plant remains reveal that cultivated crops were dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022].
 This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022].
-In addition to millet, a smaller number of rice grains (*Oryza sativa*) have also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022].
+In addition to millet, a smaller number of rice grains (*Oryza sativa*) has also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022].
 A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences [@Huan2022].
@@
 <!--info about Gaoshan-->
-To compare culinary practices for different ecological zones, we examined a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
-Excavations at Gaoshan reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
-According to similar pottery forms and styles, Gaoshan can be categorized into the early Baodun period, the earliest Neolithic culture (ca.
+To compare the culinary practices in different ecological zones, we examined a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
+Excavations at the Gaoshan site reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and archaeological features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
+According to similar pottery forms and styles, the Gaoshan site can be dated to the early Baodun period, the earliest Neolithic culture (ca.
 4700-3700 BP) in the Chengdu Plain [@Gaoshan2017].
-The primary subsistence strategy at Gaoshan could be rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020].
-It is suggested that the spread of rice farming into the Sichuan area occurred at least 4700 years ago from the Middle and Lower Yangzi valleys [@Zhang2010].
+The primary subsistence strategy at the Gaoshan site could have been rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020].
+It is suggested that the spread of rice farming into the Sichuan area occurred at least 4700 years ago from people moving from the Middle and Lower Yangzi valleys to the Sichuan area [@Zhang2010].
 
-Although the recent findings at Guijiabao and Gaoshan support mixed agriculture, it is still unclear how much millet or rice was included in local culinary practices during the Neolithic period.
-The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011].
-However, plant remains from many sites in southwest China reveal a predominance of rice, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017] and Henglanshan and Baiyangcun in Yunnan [@Jiang2016; @Henglanshan2014].
-This indicates there were variations in agricultural practices in southwest China.
-To better understand the extent of the adoption of new crops, organic residue analysis for ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking.
+Although the recent findings at the Guijiabao and Gaoshan sites support mixed agriculture as the primary subsistence strategy, it is still unclear how much millet or rice was included in the local diets during the Neolithic period.
+The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011].
+However, plant remains from many sites in southwest China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and at Henglanshan and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014].
+This indicates that there were variations in agricultural practices in southwest China.
+To better understand the extent of the adoption of new crops, organic residue analysis of ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking.
 This provides another way to examine the expansion of agriculture in the Yanyuan area.
@@
 ## Samples
-We selected ceramic fragments that were identified as typical types from Guijiabao and Gaoshan.
-The fragments were from known contexts with radiocarbon dates: Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)).
-Ceramic fragments from Guijiabao were recovered from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season.
+We selected ceramic fragments that were identified as typical from the Guijiabao and Gaoshan sites.
+The fragments date to the Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)), which were determined from known contexts based on radiocarbon dating.
+Ceramic fragments from the Guijiabao site were excavated from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season.
 The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations.
-Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that those ceramics were used in daily life [@Guijiabao2017a; @Hao2022].
-This research also attempts to confirm that those ceramics were used as cooking/storage wares for food.
+Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that these ceramics were used in daily life [@Guijiabao2017a; @Hao2022].
+This research also attempts to confirm that these pottery vessels were used as cooking/storage wares for food.
 
@@
 knitr::kable(Ch_sherds,
-             caption = "A list of the pottery sherds that were analyzed in this study. The sherd labeled 'SYG' means potsherds from Guijiabao, while the label 'CDG' refers to Gaoshan")
+             caption = "A list of the pottery sherds that were analyzed in this study.
+The sherd labeled 'SYG' indicates potsherds from the Guijiabao site, while the label 'CDG' refers to those from the Gaoshan site")
 ```
@@
 ## Methods
-We detected and identified lipid compounds by gas chromatography mass spectrometry (GC-MS) and quantified lipids by gas chromatography with flame-ionization detection (GC-FID).
-To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), found in most animals and plants, we determined their stable carbon isotope characteristics by gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
+We detected and identified lipid compounds using gas chromatography mass spectrometry (GC-MS) and quantified lipids using gas chromatography with flame-ionization detection (GC-FID).
+To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), which are found in most animals and plants, we identified their stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
 The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].
 
-For the sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis.
+For sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis.
 Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C).
@@
 
-The remaining 95% aliquot was fractioned by solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol, respectively.
+The remaining 95% aliquot was fractioned using solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol.
 The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight.
 The FAME derivatives were extracted with hexane ($3\times10$ mL) and then blown down with nitrogen streams.
-About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed by GC-C-IRMS.
-With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might from solvent and reagent.
+About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed using GC-C-IRMS.
+With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might have come from solvents and reagents.
 Internal standards were added before and after the lipid extraction for quantification.
@@
 Samples were injected in splitless mode at 320 °C, and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness).
-Helium carrier is at a constant flow rate of 1.3 ml/min.
-The temperature of the oven was started from 90 °C for 1 minute and increased at a rate of 5 °C/min to 320 °C, where it was held for 23 minutes.
+The carrier gas was helium set at a constant flow rate of 1.3 ml/min.
+The temperature of the oven was set at 90 °C and held for 1 minute and then was increased at a rate of 5 °C/min to 320 °C, where it was held for 23 minutes.
 The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode.
-The carrier gas was helium at 1.6 ml/min.
+The carrier gas was helium flowing at a rate of 1.6 ml/min.
 The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode.
-The GC oven was held at 90°C for 1.3 minute, and then raised to 320°C at a rate of 5°C/min.
-The final temperature was held for 22 minutes.
+The GC oven was set at 90°C and held for 1.3 minute, and then was raised to 320°C at a rate of 5°C/min.
+The final temperature was set for 22 minutes.
 The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface.
 The sample injection volume was 1 ul in toluene and was injected in splitless mode.
-Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set at 90 °C for 1.5 minutes and then raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold.
-Helium was used as a gas carrier in constant flow mode at 1.24 ml/minute.
+Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set at 90 °C for 1.5 minutes and then was raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold.
+Helium was used as a gas carrier with a constant flow rate of 1.24 ml/minute.
 
@@
 
-The GC-MS analysis for organic residues extracted from Guijiabao and Gaoshan potsherds revealed the presence of lipid compounds above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but in relatively small amounts in the majority of them (Table \@ref(tab:FA-composition-table)].
-The lipid profiles show that the main compounds in most potsherds are saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)).
-Those fatty acids commonly exist in animal fat and plant oil.
-They are also generally degraded from larger molecules, such as triacylglycerols, after long-term burial processes due to oxidation and microbial degradation [@Dudd1998; @Eerkens2007].
-High concentrations of medium-chain fatty acids (C16:0 and C18:0) with an even-over-old carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
-In addition, a number of potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
+The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the presence of lipid compounds were above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)].
+The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)).
+These fatty acids commonly exist in animal fat and plant oil.
+They are also generally degraded from larger molecules, such as triacylglycerols, due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007].
+High concentrations of medium-chain fatty acids (C16:0 and C18:0) with an even-over-odd carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
+In addition, a number of potsherds had higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
 A higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005].
@@
 
-```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao and Gaoshan.
-Cn:x refers to fatty acid with n carbon atoms and x double bonds.
-IS indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
+```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from the Guijiabao and Gaoshan sites.
+"Cn:x" refers to fatty acid with n carbon atoms and x double bonds.
+"IS" indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
 knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
@@
 
-One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the sample, suggesting a better preservation of lipids that might relate to the type of ceramic or a different use.
+One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the samples.
+A high yield of lipids might relate to the texture and type of ceramic used or better archaeological contexts for preservation.
 This sample also has a charred interior surface, indicative of a function associated with food processing.
-Long-chain fatty acids from C20:0 to C32:0 are present in that sample with an even-over-odd carbon chain length predominance, but in lower abundance than medium-chain components (Figure \@ref(fig:chromatograms-specific)).
-The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005].
+Long-chain fatty acids from C20:0 to C32:0 are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)).
+The existence of long-chain fatty acids with a high amount of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005].
 However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
@@
 
-```{r chromatograms-specific, echo=FALSE, fig.cap= "The MS chromatogram of the Guijiabao potsherd shows the presence of long-chain fatty acids.
-Cn:x refers to fatty acid with n carbon atoms and x double bonds.
-IS indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
+```{r chromatograms-specific, echo=FALSE, fig.cap= "The MS chromatogram of the potsherd from the Guijiabao site shows the presence of long-chain fatty acids.
+"Cn:x" refers to fatty acid with n carbon atoms and x double bonds.
+"IS" indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
 knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
@@
 
-To identify the sources of degraded animal fats, we conducted isotope analysis to obtain the stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS.
+To identify the sources of degraded animal fats, we conducted isotope analysis to obtain stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS.
 The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
 To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
-The carbon isotope values for C16:0 and C18:0 distribute between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰.
+The carbon isotope values for C16:0 and C18:0 have a distribution between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰.
 The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].
 
-The results show that the majority of Guijiabao potsherd samples contain fatty acids from ruminant adipose fats, with one potsherd (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A).
-The potsherd from Gaoshan (CDG-062) demonstrates ruminant adipose fats as well.
-The potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of those two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
+The results show that the majority of the potsherd samples from the Guijiabao site contain fatty acids from ruminant adipose fats, with one potsherd sample (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A).
+Analysis of the potsherd from the Gaoshan site (CDG-062) indicates ruminant adipose fats as well.
+The potsherds (SYG-TN13-E22-1 and SYG-3) from the Guijiabao site dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of these two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
 Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low 13C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
 This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results.
-Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for accessing food categories.
+Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for determining food categories.
 
-In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting those ruminant animals seemed to have consumed mainly C3 plants.
-There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively enriched δ^13^C value, suggesting a C4-based diet for the ruminant animal.
+In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting these ruminant animals seemed to have consumed mainly C3 plants.
+There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively large δ^13^C value, suggesting a C4-based diet for the ruminant animal.
 
-```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C16:0 and C18:0. B: The Δ13C (δ13C18:0-16:0) values plotted against δ13C16:0 values"}
+```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C16:0 and C18:0. B: The Δ13C (δ13C18:0-16:0) values are plotted against δ13C16:0 values"}
 knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
@@
 <!--A brief summary of interesting results-->
-The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets), C3 plants (e.g., rice), or new species (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period.
-Our results from lipid residues in Guijiabao and Gaoshan potsherds show that food categories were dominated by ruminant fats during the Neolithic period.
-Moreover, it is striking to identify non-ruminant fats in a Guijiabao potsherd that might originate from C3 plants (see below).
-Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that those ceramics were used as cooking or storage pots.
-The low yield could be attributed to the bottom parts of the vessels for analysis, which may absorb fewer residues than the rim of a vessel [@Evershed2008experimental; @Kwak2020; @Charters1993].
-But we also found that one bottom sample dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples, reflecting better preservation or a much higher lipid yield.
-We recognize limitations and challenges for interpreting lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils.
-Besides, a detailed analysis of faunal remains from Guijiabao is still in progress, which may constrain a comprehensive comparison between different types of evidence.
-Nevertheless, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices in Guijiabao.
+The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets) or C3 plants (e.g., rice), or animals (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period.
+Our results show that food residues in potsherds from the sites at Guijiabao and Gaoshan were dominated by ruminant fats during the Neolithic period.
+Moreover, it is striking to discover that non-ruminant fats in a Guijiabao potsherd might have originated from C3 plants (see below).
+Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that these ceramics were used as cooking or storage vesssels.
+The low lipid yield could be attributed to the bottom of the vessels, where there were fewer residues than other parts of the vessel, for example, the rim [@Evershed2008experimental; @Kwak2020; @Charters1993].
+But we also found that one sample from the vessel bottom dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples, reflecting better preservation or a much higher lipid yield.
+We recognize that there are limitations and challenges for determining lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils.
+In addition, a detailed analysis of faunal remains from the Guijiabao site is still in progress, which may constrain a comprehensive comparison between different types of evidence.
+Nevertheless, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices at the Guijiabao site.
 
 <!--discussion about possible plant origins-->
-While it has been indicated that millets (C4 plants), including foxtail and broomcorn millets, were the dominant cereals according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples.
-Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might be from terrestrial plants/C3 plants based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
-According to the flotation results from Guijiabao, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae; however, the seeds of wild weeds only account for a small part of the floral assemblage derived from flotation [@Guijiabao2016].
-Despite stone knives and tools were also found at the sites that might relate to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022].
-Another possible source for non-ruminant fats could be freshwater fish, which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b].
+While it has been determined that millets (C4 plants), including foxtail and broomcorn millets, were the dominant cereals, according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples.
+Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might have been from terrestrial plants/C3 plants, based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
+According to the results from flotation analysis at the Guijiabao site, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae.
+However, the seeds of wild weeds only account for a small part of the floral assemblage derived from flotation [@Guijiabao2016].
+Despite stone knives and tools that were also found at the sites that might have been related to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022].
+Another possible source for non-ruminant fats could have been freshwater fish, a possibility which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b].
 But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016].
@@
 The ruminant fats identified in the Neolithic potsherds (ca.
-4500 BP) from Guijiabao and Gaoshan could be derived from deer, sheep, or horses fed on a C3 diet based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017].
+4500 BP) from the Guijiabao and Gaoshan sites could have been derived from deer, sheep, or horses that were fed on a C3 diet, based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017].
 This finding, especially the consumption of deer, also corresponds with unearthed faunal remains from the sites and the local natural environments.
-Stone artifacts found at the site, including stone balls and arrowheads, further indicates hunting activities [@Hao2022; @Guijiabao2020a].
-However, whether horse was part of local food choices in southwest China around 4500 years ago is skeptical.
-First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China, with only a sporadic presence in the south [@Yuan2006].
-Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on.
-Thus, supported by faunal evidence, deer is the most likely part of the local food choices.
-It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these species have been found in sites of southwest China [@Zhang2016; @Chen2020].
+Stone artifacts found at the site, including stone balls and arrowheads, further indicate hunting activities [@Hao2022; @Guijiabao2020a].
+However, whether horse was part of local food choices in southwest China around 4500 years ago is doubtful.
+First, the Pleistocene and Neolithic sites having wild horse bones are primarily concentrated in northern China, with only sporadic sites in the south [@Yuan2006].
+Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period that we are focusing on.
+Thus, supported by faunal evidence, deer was most likely the predominant local food choices.
+It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these animals have been found in sites in southwest China [@Zhang2016; @Chen2020].
 
-We observed a mixture of ruminant fats and C3 plants based on a number of Guijiabao potsherds that date to the historical period (ca.
+We observed a mixture of ruminant fats and C3 plants, based on analyses of a number of Guijiabao potsherds that date to the historical period (ca.
 1000 BP).
-It appears that C3 plants could be processed more in the historical period than they were in the Neolithic period.
-In addition to rice, the source of C3 plants could also include wheat since it was introduced during the historical period [@JiangGaoshan2017].
-A larger sample size is needed to explore whether there is a shift in diet over time.
-The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might be an important food source for local animals.
-Despite direct C4 plant signals not being found, one Guijiabao potsherd demonstrates the source of C4 plant-fed ruminants during the Neolithic period.
-This indicates either C3 or C4 plants could be the food source for ruminants, but whether those animals were domesticated is unknown.
-The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, which could provide more details about local animal species for comparison.
+It appears that C3 plants could have been processed more in the historical period than they could have been in the Neolithic period.
+In addition to rice, the source of C3 plants could also have included wheat, since it was introduced during the historical period [@JiangGaoshan2017].
+A larger sample size is needed to determine whether there was a change in diet over time.
+The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might have been an important food source for local animals.
+Despite direct C4 plant signals not being found, analysis of one Guijiabao potsherd shows the source of C4 plant-fed ruminants during the Neolithic period.
+This indicates that either C3 or C4 plants could have been the food source for ruminants, but whether those animals were domesticated is unknown.
+The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, the results of which which could provide more details about local animal species for comparison.
 
-In general, ruminant fats from local wild animals, such as deer, were the predominant food source during the Neolithic period based on our results.
-A subsistence that relied on hunting to a certain degree could be anticipated given the abundance of wild animals in the natural settings of the Hengduan mountain ranges, as indicated by the faunal remains from the Yingpanshan site [@He2020].
-Crop cultivation could be adopted as a supplement to local food sources while it was introduced in the Neolithic period, but it might only account for a small proportion.
-For the dietary pattern at Guijiabao, previous study suggested a C4-based diet according to the isotopic analysis of human bones [@Lin2022], but we only identified a C3 plant signal from one pottery sample without any signal directly from C4 plants.
-This inconsistency could result from different sampling strategies or from the sample size issue, where our results only reflect a small portion of food content.
-Nevertheless, our results suggest that farming practices during the Neolithic period may not have been adopted as much as we previously thought.
-An organic residue approach may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through direct evidence that has been cooked in ceramics.
-A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period.
+In general, based on our results, ruminant fats from local wild animals, such as deer, were the predominant food source during the Neolithic period for peoples at Guijiabao.
+A subsistence that relied on hunting to a certain degree could be plausible, given the abundance of wild animals in the natural settings of the Hengduan mountain ranges, as indicated by the faunal remains from the Yingpanshan site [@He2020].
+Crop cultivation could have been adopted as a supplement to local food sources when it was introduced in the Neolithic period, but crops might have only accounted for a small proportion of the food sources.
+A previous study using isotopic analysis of human bones suggests that the diet at the Guijiabao site was a C4-based diet [@Lin2022], but we only identified a signal from C3 plants from one pottery sample without any signal from C4 plants.
+This inconsistency could have resulted from different sampling strategies or from the sample size issue, where our results reflect only a small portion of food content.
+Nevertheless, our results suggest that farming practices during the Neolithic period might not have been adopted as much as we previously thought.
+An approach analyzing organic residues may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through presence of food residues that were cooked in ceramic vessels.
+A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period in southwest China.
 
 # Conclusion
-Our case study on organic residues in pottery fragments from Guijiabao and Gaoshan in southwest China reveals the presence of lipids that was associated with food categories.
-The food processed in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwest China.
-The source of terrestrial plants could be C3 plants, implying the processing of rice or other wild weeds in the pot.
-However, the evidence of direct processing of either C3 or C4 plants is not in proportion to the large quantity of plant remains found at Guijiabao, especially millet grains.
-Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the Neolithic.
-This suggests that hunting might have been an important part of local subsistence, supplemented by rice/millet farming when they were introduced.
-Our results from Guijiabao could represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains.
+Our case study on organic residues in pottery fragments from the sites at Guijiabao and Gaoshan in southwest China reveals the presence of lipids that can be analyzed to determine food sources.
+The foods cooked in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwest China.
+The source of terrestrial plants could have been C3 plants, implying the cooking of rice or other wild weeds in the pots.
+However, the evidence of direct cooking of either C3 or C4 plants shows that the amount of C3/C4 plants cooked is not in proportion to the large quantity of plant remains found at the Guijiabao site, especially C4 plants such as millet.
+Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the Neolithic period.
+This suggests that hunting might have been an important part of local subsistence strategy, supplemented by rice/millet farming when they were introduced.
+Our results from the Guijiabao site could represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains.
 We recognize the limitations of our interpretation of the local diet due to the small sample size.
-Nevertheless, our study could offer a perspective from organic residue analysis to explore food choices and the extent of farming adoption when rice/millet agriculture first spread into the Yanyuan during the Neolithic period.
-A more systematic sampling of potsherds across different phases and pottery types may eliminate a clear picture of diet and subsistence during the Neolithic transition.
+Nevertheless, our study could offer an approach using organic residue analysis to determine food choices and explore the extent of the adoption of farming when rice/millet agriculture first spread into the Yanyuan area during the Neolithic period.
+A more systematic sampling of potsherds across different phases and pottery types may give a clear picture of diet and subsistence in southwest China during the Neolithic period of transition.
 
@@
 We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples.
-We thank Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for sample preparation and conducting lipid extraction and analysis.
+We thank Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for sample preparation and for conducting lipid extraction and analysis.
 We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines.

### 2023-04-15 analysis/code/GCFID-quant.R
adjust the text size and scale

@@ -136,14 +136,14 @@ meth_all %>%
   geom_line(size = 0.3) +
   ggrepel::geom_text_repel(data = highlight_SYG2_2, #ggrepel::geom_text_repel
             aes(label = FA),
-            size = 3.5,
+            size = 3,
             nudge_y = 250000,
             show.legend = FALSE) +
   scale_y_continuous(labels = scales::comma_format(),
                      limits = c(0, 7500000),
                      breaks = seq(0, 7500000, 1000000)) +
-  scale_x_continuous(limits = c(15, 50),
-                     breaks = seq(15, 50, 5),
+  scale_x_continuous(limits = c(20, 50),
+                     breaks = seq(20, 50, 5),
                      expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
   labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
   theme_minimal() +

### 2023-04-15 analysis/paper/paper.Rmd
minor edits before submission

@@
 abstract: |
-    The spread of new crops or terrestial animals across areas in southwest China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in those areas.
-It is widely accepted that both millet and rice were introduced into southwest China in 4700 BP at the latest.
-However, to what extent new crops were used by local peoples at an earlier time remains unclear.
+    The spread of new crops or terrestial animals across areas in southwest China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region.
+It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and milletes came even earlier.
+However, to what extent new crops were used by local peoples at that time remains unclear.
 The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops.
@@
 <!-- intro to the topic and research aim-->
-Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on any people groups that are involved in the interactions that result from these movements [@Burmeister2000].
+Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on local people groups involved in the interactions [@Burmeister2000].
 The Hengduan Mountain ranges and valleys in southwestern China are geographical areas that connect the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau and by which movement to these plateaus must occur.
@@
 The new evidence at Guijiabao reveals that millet and rice were both present there in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already been part of the local diet [@Lin2022].
-Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period.
+Although some of these analyses suggest the presence of mixed millet/rice farming [@Huan2022], we still lack direct evidence of culinary practices during this period.
 To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery?
@@
 In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reveal the original contents cooked or stored in a vessel during its lifetime of use [@Dudd1999; @Miller2020].
-Through our case study, We attempt to investigate culinary practices during a period of frequent cultural interactions that accompanied agricultural expansion .
+Through our case study, we attempt to investigate culinary practices during a period of frequent cultural interactions that accompanied agricultural expansion.
 
@@
 We focus on the Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)).
-The Guijiabao site allows us to understand the early adoption of agriculturebecause of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
+The Guijiabao site allows us to understand the early adoption of agriculture because of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
 A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao was a settlement site [@Hao2022].
@@
 The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011].
-However, plant remains from many sites in southwest China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and at Henglanshan and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014].
+However, plant remains from many sites in southwest China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014].
 This indicates that there were variations in agricultural practices in southwest China.
@@
 A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour.
-After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene (20 uL) prior to GC-MS.
+After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene prior to GC-MS.
 
@@
 
-The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at https://github.com/LiYingWang/GJBpottery to enable research reproducibility and transparency [@Marwick2017].
+The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at https://github.com/LiYingWang/GJBpottery to enable research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].
 
@@
 ```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from the Guijiabao and Gaoshan sites.
-"Cn:x" refers to fatty acid with n carbon atoms and x double bonds.
-"IS" indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
+'Cn:x' refers to fatty acid with n carbon atoms and x double bonds.
+'IS' indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
 knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
@@
                      breaks = seq(0, 7500000, 1000000)) +
-  scale_x_continuous(limits = c(15, 50),
-                     breaks = seq(15, 50, 5),
+  scale_x_continuous(limits = c(20, 50),
+                     breaks = seq(20, 50, 5),
                      expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
@@
 One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the samples.
-A high yield of lipids might relate to the texture and type of ceramic used or better archaeological contexts for preservation.
+A higher yield of lipids might relate to the texture and type of ceramic used or better archaeological contexts for preservation.
 This sample also has a charred interior surface, indicative of a function associated with food processing.
@@
 ```{r chromatograms-specific, echo=FALSE, fig.cap= "The MS chromatogram of the potsherd from the Guijiabao site shows the presence of long-chain fatty acids.
-"Cn:x" refers to fatty acid with n carbon atoms and x double bonds.
-"IS" indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
+'Cn:x' refers to fatty acid with n carbon atoms and x double bonds.
+'IS' indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
 knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
@@
 <!--A brief summary of interesting results-->
-The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets) or C3 plants (e.g., rice), or animals (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period.
+The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets) and C3 plants (e.g., rice), or animals (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period.
 Our results show that food residues in potsherds from the sites at Guijiabao and Gaoshan were dominated by ruminant fats during the Neolithic period.
@@
 Another possible source for non-ruminant fats could have been freshwater fish, a possibility which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b].
-But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid, from our potsherds [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016].
+But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016], from our potsherds.
 
@@
 It appears that C3 plants could have been processed more in the historical period than they could have been in the Neolithic period.
-In addition to rice, the source of C3 plants could also have included wheat, since it was introduced during the historical period [@JiangGaoshan2017].
+In addition to rice, the source of C3 plants could also have included wheat, since it was introduced into the region during the historical period [@JiangGaoshan2017].
 A larger sample size is needed to determine whether there was a change in diet over time.
@@
 This indicates that either C3 or C4 plants could have been the food source for ruminants, but whether those animals were domesticated is unknown.
-The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, the results of which which could provide more details about local animal species for comparison.
+The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, the results of which could provide more details about local animal species for comparison.
 
@@
 We also want to express our gratitude to the anonymous reviewers for their comments that helped us improve this paper.
-This research is sponsored by the National Science and Technology Council (PE11006-0107).
+This research is sponsored by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).
 

### 2023-04-16 analysis/paper/paper.Rmd
edit the abstract and keywords

@@
 ---
-title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis"
+title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwestern China using organic residue analysis"
 author:
@@
 abstract: |
-    The spread of new crops or terrestial animals across areas in southwest China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region.
-It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and milletes came even earlier.
+    The spread of new crops or terrestrial animals across areas in southwestern China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region.
+It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and millets came even earlier.
 However, to what extent new crops were used by local peoples at that time remains unclear.
-The excavation of Guijiabao in the Yanyuan basin in southwest China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops.
+The excavation of Guijiabao in the Yanyuan basin in southwestern China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops.
 To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis.
@@
 Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming.
-This study also demonstrates an approach using organic residue analysis to explore pottery use and culinary practices during the Neolithic period in southwest China.
-A more systematic sampling of different types of potsherds and temporal phases with larger samples is needed to provide a clearer picture of ancient diets and subsistence.
+This study also demonstrates an approach using organic residue analysis to explore the extent of the adoption of farming during the Neolithic period in southwestern China.
 
 keywords: |
-  Lipid residue analysis; Ceramics; Southwest China; Stable isotope analysis; Food practices
+  Southwestern China; Neolithic period; Lipid residue analysis; Stable isotope analysis; Ceramics; Culinary practices
 highlights: |
@@
 They play an important role in the cultural exchanges and spread of crops into southwestern China and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
-Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015].
+Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015].
 The Hengduan Mountain regions in the Neolithic period are therefore essential for understanding the processes of agricultural expansion.
@@
 
-```{r SW-china-map, fig.cap = "The archaeological sites Guijiabao and Gaoshan and other locations in southwest China that are mentioned in this study. Map data is from naturalearthdata.com"}
+```{r SW-china-map, fig.cap = "The archaeological sites Guijiabao and Gaoshan and other locations in southwestern China that are mentioned in this study. Map data is from naturalearthdata.com"}
 library(here)
@@
 The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011].
-However, plant remains from many sites in southwest China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014].
-This indicates that there were variations in agricultural practices in southwest China.
+However, plant remains from many sites in southwestern China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014].
+This indicates that there were variations in agricultural practices in southwestern China.
 To better understand the extent of the adoption of new crops, organic residue analysis of ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking.
@@
 Stone artifacts found at the site, including stone balls and arrowheads, further indicate hunting activities [@Hao2022; @Guijiabao2020a].
-However, whether horse was part of local food choices in southwest China around 4500 years ago is doubtful.
+However, whether horse was part of local food choices in southwestern China around 4500 years ago is doubtful.
 First, the Pleistocene and Neolithic sites having wild horse bones are primarily concentrated in northern China, with only sporadic sites in the south [@Yuan2006].
-Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period that we are focusing on.
+Second, domesticated horses first found in southwestern China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period that we are focusing on.
 Thus, supported by faunal evidence, deer was most likely the predominant local food choices.
-It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these animals have been found in sites in southwest China [@Zhang2016; @Chen2020].
+It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these animals have been found in sites in southwestern China [@Zhang2016; @Chen2020].
 
@@
 An approach analyzing organic residues may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through presence of food residues that were cooked in ceramic vessels.
-A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period in southwest China.
+A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period in southwestern China.
 
 # Conclusion
-Our case study on organic residues in pottery fragments from the sites at Guijiabao and Gaoshan in southwest China reveals the presence of lipids that can be analyzed to determine food sources.
-The foods cooked in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwest China.
+Our case study on organic residues in pottery fragments from the sites at Guijiabao and Gaoshan in southwestern China reveals the presence of lipids that can be analyzed to determine food sources.
+The foods cooked in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwestern China.
 The source of terrestrial plants could have been C3 plants, implying the cooking of rice or other wild weeds in the pots.
@@
 Nevertheless, our study could offer an approach using organic residue analysis to determine food choices and explore the extent of the adoption of farming when rice/millet agriculture first spread into the Yanyuan area during the Neolithic period.
-A more systematic sampling of potsherds across different phases and pottery types may give a clear picture of diet and subsistence in southwest China during the Neolithic period of transition.
+A more systematic sampling of potsherds across different phases and pottery types may give a clear picture of diet and subsistence in southwestern China during the Neolithic period of transition.
 

### 2023-04-16 analysis/code/site-regional-map-Ch.R
edit the code for the map

@@ -20,10 +20,6 @@ world_points <- cbind(world, st_coordinates(st_point_on_surface(world$geometry))
   mutate(Y = ifelse(brk_name %in% c("Bhutan", "Bangladesh"), Y+2, Y)) %>%
   mutate(Y = ifelse(brk_name %in% c("Thailand", "Myanmar"), Y+3.5, Y))
 
-# shapefile
-ch_adm <- st_read(here::here("analysis/data/raw_data/Ch_adm/CHN_adm1.shp"))
-plot(st_geometry(ch_adm))
-
 # add site location
 site_location <-
   data.frame(location = c("Guijuabao", "Gaoshan"), # Chengdu(104.0587, 30.5899)
@@ -49,9 +45,6 @@ China_SE_Asia <-
   scale_y_continuous(breaks = seq(5, 45, by = 10)) +
   theme(axis.title.x = element_blank(),
         axis.title.y = element_blank())
-
-China_SE_Asia +
-  geom_sf(data = ch_adm)
 
 # Topographic map
 library(ggmap)

### 2023-04-16 analysis/paper/paper.Rmd
update data directory

@@
 quant_FA <-
-  readxl::read_excel(here::here("analysis", "data", "raw_data", "210824_FAMES_pot_LW_publication.xlsx"), 
+  readxl::read_excel(here::here("analysis", "data", "raw_data", "210824_FAMES_pot_LW_pub.xlsx"), 
                      skip = 2, col_names = TRUE)
@@
 isotope_data <-
-  readxl::read_excel(here::here("analysis","data", "raw_data","12_21_potshard_13C.xlsx"),
+  readxl::read_excel(here::here("analysis","data", "raw_data","12_21_potshard_13C_pub.xlsx"),
                      sheet = "Summary", col_names =  FALSE)

### 2023-04-16 analysis/code/isotope-analysis.R

@@ -1,6 +1,6 @@
 library(tidyverse)
 isotope_data <-
-  readxl::read_excel(here::here("analysis","data", "raw_data","12_21_potshard_13C.xlsx"),
+  readxl::read_excel(here::here("analysis","data", "raw_data","12_21_potshard_13C_pub.xlsx"),
                      sheet = "Summary", col_names =  FALSE)
 
 # filter the blank corrected data

### 2023-04-16 analysis/paper/paper.Rmd
adjust the background color of figures

@@
        here::here("analysis","figures", "chromatograms_four.png"),
+       bg = "white",
        width = 6,
@@
        here::here("analysis","figures", "chromatograms_specific.png"),
+       bg = "white",
        width = 8,
@@
 ggsave(here::here("analysis","figures","isotopic_plots.png"),
+       bg = "white",
        width = 6,

### 2023-04-16 analysis/code/GCFID-quant.R

@@ -123,6 +123,7 @@ meth_all %>%
         panel.background = element_blank(), axis.line = element_line(colour = "black"))
 
 ggsave(here::here("analysis","figures", "chromatograms_four.png"),
+       bg = "white",
        width = 7,
        height = 8,
        dpi = 300,
@@ -152,6 +153,7 @@ meth_all %>%
         panel.background = element_blank(), axis.line = element_line(colour = "black"))
 
 ggsave(here::here("analysis","figures", "chromatograms_specific.png"),
+       bg = "white",
        width = 8,
        height = 5,
        dpi = 360,

### 2023-04-16 analysis/code/isotope-analysis.R

@@ -161,6 +161,7 @@ combined_isotopic_plots <-
           label_size = 12)
 
 ggsave(here::here("analysis","figures","isotopic_plots.png"),
+       bg = "white",
        width = 6,
        height = 8,
        dpi = 300,

### 2023-11-14 analysis/paper/paper.Rmd
edit based on comments

@@
 To explore whether new crops or terrestrial animals became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues within ceramic vessels from Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a].
-The new evidence at Guijiabao reveals that millet and rice were both present there in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already been part of the local diet [@Lin2022].
+The archaeobotanical evidence at Guijiabao reveals that foxtail millet, broomcorn millet, and rice were present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Huan2022], and millet could have already been part of the local diet [@Lin2022].
 Although some of these analyses suggest the presence of mixed millet/rice farming [@Huan2022], we still lack direct evidence of culinary practices during this period.
@@
 We detected and identified lipid compounds using gas chromatography mass spectrometry (GC-MS) and quantified lipids using gas chromatography with flame-ionization detection (GC-FID).
-To distinguish the source of the common fatty acids, palmitic (C16:0) and stearic (C18:0), which are found in most animals and plants, we identified their stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
+To distinguish the source of the common fatty acids, palmitic (C~16:0~) and stearic (C~18:0~), which are found in most animals and plants, we identified their stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
 The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].
@@
 With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might have come from solvents and reagents.
-Internal standards were added before and after the lipid extraction for quantification.
+Internal standards, including FA C~21:0~ and 5-α Cholestane, were added before and after the lipid extraction for quantification.
 
@@
 library(tidyverse)
+library(ggtext)
 
@@
 
-The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the presence of lipid compounds were above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)].
-The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid), C18:0 (stearic acid), and monounsaturated fatty acid C18:1 (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)).
+The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the presence of lipid compounds were above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)).
+The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid), C~18:0~ (stearic acid), and monounsaturated fatty acid C~18:1~ (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)).
 These fatty acids commonly exist in animal fat and plant oil.
 They are also generally degraded from larger molecules, such as triacylglycerols, due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007].
-High concentrations of medium-chain fatty acids (C16:0 and C18:0) with an even-over-odd carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
-In addition, a number of potsherds had higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3.
-A higher content of C18:1 and C16:0 with a lower content of C18:0 might indicate the origin of plant oils [@Mayyas2018; @Copley2005].
+High concentrations of medium-chain fatty acids (C~16:0~ and C~18:0~) with an even-over-odd carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
+In addition, a number of potsherds had higher contents of C~18:1~ compared to C~18:0~ and C~16:0~, such as SYG-TN13-E23-3.
+A higher content of C~18:1~ and C~16:0~ with a lower content of C~18:0~ might indicate the origin of plant oils [@Mayyas2018; @Copley2005].
 However, we may still need to examine their isotopic values to confirm their sources.
@@
 This sample also has a charred interior surface, indicative of a function associated with food processing.
-Long-chain fatty acids from C20:0 to C32:0 are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)).
-The existence of long-chain fatty acids with a high amount of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005].
+Long-chain fatty acids from C~20:0~ to C~32:0~ are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)).
+The existence of long-chain fatty acids with a high amount of C~16:0~ relative to C~18:0~ might be related to plant waxes [@Post1996; @Copley2005].
 However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
@@
 
-To identify the sources of degraded animal fats, we conducted isotope analysis to obtain stable carbon isotope compositions, δ^13^C values of C16:0 and C18:0, using GC-C-IRMS.
-The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
-To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C18:0 - δ^13^C16:0) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
-The carbon isotope values for C16:0 and C18:0 have a distribution between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ13C values range from `r delta_min` ‰ to `r delta_max` ‰.
-The Δ13C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].
+To identify the sources of degraded animal fats, we conducted isotope analysis to obtain stable carbon isotope compositions, δ^13^C values of C~16:0~ and C~18:0~, using GC-C-IRMS.
+The δ^13^C values of C~16:0~ and C~18:0~ have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
+To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C~18:0~ - δ^13^C~16:0~) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
+The carbon isotope values for C~16:0~ and C~18:0~ have a distribution between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ^13^C values range from `r delta_min` ‰ to `r delta_max` ‰.
+The Δ^13^C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].
 
@@
 The potsherds (SYG-TN13-E22-1 and SYG-3) from the Guijiabao site dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of these two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
-Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low 13C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
+Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low ^13^C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
 This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results.
@@
 
-```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C16:0 and C18:0. B: The Δ13C (δ13C18:0-16:0) values are plotted against δ13C16:0 values"}
+```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C~16:0~ and C~18:0~. B: The Δ13C (δ^13^C~18:0~-~16:0~) values are plotted against δ^13^C~16:0~ values"}
 knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))

### 2023-11-26 analysis/paper/paper.Rmd
revising based on comments

@@
 ---
-title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwestern China using organic residue analysis"
+title: "Preliminary insights into the use of pottery and culinary pratices in southwestern China"
 author:
@@
 To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis.
-The results show signals of terrestrial animals in most potsherds and a C3 plant source in one potshard, during the Neolithic period.
+The results show signals of terrestrial animals in most potsherds and a C~3~ plant source in one potshard, during the Neolithic period.
 Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming.
@@
 <!-- intro to the topic and research aim-->
-Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on local people groups involved in the interactions [@Burmeister2000].
-The Hengduan Mountain ranges and valleys in southwestern China are geographical areas that connect the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau and by which movement to these plateaus must occur.
-They play an important role in the cultural exchanges and spread of crops into southwestern China and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986].
-Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015].
-The Hengduan Mountain regions in the Neolithic period are therefore essential for understanding the processes of agricultural expansion.
-Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as the grains found in Sichuan [@Huan2022; @Hao2022].
-However, to what extent the two crops were incorporated into the local diet is still not fully understood.
+Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on local populations involved in the interactions [@Burmeister2000].
+In prehistoric southwest China, the Hengduan Mountain ranges have served as passages for people to move between the Qinghai-Tibetan Plateau in the north, the Yunnan–Guizhou Plateau in the south, and the Sichuan Basin in the east.
+The movements in the Hengduan Mountain ranges facilitated not only cultural exchange, but also the spread of new subsistence practices, such as farming and animal domestication [@Tong1986; @Hein2014; @Chen2015].
+With more archaeological sites have been found recently along the Hengduan Mountain valleys, researchers are able to dive deeper into local cultural developments and diversity, the timing and routes of exchange network, and the spread of farming practices during Neolithic period across the Hengduan Mountain ranges and neighboring areas [@Hein2014; @Hein2016; @Ma2021; @Lin2022; @Hao2022; @Ma2021].
 
-To explore whether new crops or terrestrial animals became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues within ceramic vessels from Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a].
+Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern China as early as 4700 BP, where the adoption of farming in southwestern Sichuan is around 4700 BP and in western Yunnan is around 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022].
+Recent excavations at Guijiabao in the middle Hengduan Mountain ranges bridges the gap between southwestern Sichuan and Yunnan through the finding of millet and rice around 4700 BP [@Huan2022; @Hao2022].
+However, it is still not fully understood the extent of the adoption of farming practices by the ancient people in the Hengduan Mountain ranges.
+Did crop cultivation become the essential part of local subsistence when they were introduced?
+What other resources were obtained in addition to those grains?
+
+In this study, we focus on archaeological sites
+
+aim to to explore the culinary practices through food residues within ceramic vessels from Guijiabao and Gaoshan sites
+
+Also, the discussion of local subsistence mostly relies on paleobotanical evidence and zooarchaeological assemblages,  The preservation of animal or plant remains could be heavily impacted by environmental conditions.
+
+to explore the culinary practices in the Yanyuan Basin through food residues within ceramic vessels from Guijiabao and Gaoshan sites.
+We analyzed small samples to test whether those pottery were used for cooking, and if we could detect new crops or species such as rice, millet, or pigs, as food sources when they spread into the Yanyuan Basin.
+Guijiabao is a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a].
 The archaeobotanical evidence at Guijiabao reveals that foxtail millet, broomcorn millet, and rice were present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Huan2022], and millet could have already been part of the local diet [@Lin2022].
@@
 The recent growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
-Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing food origins [@Copley2005; @Whelton2021].
-This is possible due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery for over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017].
+Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing food origins; for example, millet can be identified through the biomarker miliacin [@Copley2005; @Whelton2021; @Heron2016].
+This is possible for prehistoric ceramics due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery for over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017].
 In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reveal the original contents cooked or stored in a vessel during its lifetime of use [@Dudd1999; @Miller2020].
@@
 Plant remains reveal that cultivated crops were dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022].
-This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022].
+This corresponds to a C~4~-based diet based on isotopic analysis of four human bones [@Lin2022].
 In addition to millet, a smaller number of rice grains (*Oryza sativa*) has also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022].
@@
 ## Samples
-We selected ceramic fragments that were identified as typical from the Guijiabao and Gaoshan sites.
+As a preliminary research to accesses the feasibility of this approach, we selected ceramic fragments that were identified as typical from the Guijiabao and Gaoshan sites.
 The fragments date to the Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)), which were determined from known contexts based on radiocarbon dating.
@@
 We detected and identified lipid compounds using gas chromatography mass spectrometry (GC-MS) and quantified lipids using gas chromatography with flame-ionization detection (GC-FID).
-To distinguish the source of the common fatty acids, palmitic (C~16:0~) and stearic (C~18:0~), which are found in most animals and plants, we identified their stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
-The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].
+To distinguish the source of the common fatty acids, palmitic (C~16:0~) and stearic (C~18:0~), which are found in most animals and plants, we conducted compound-specific isotope analysis to identify stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
+The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2007; @Evershed1994].
 
 For sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis.
-Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C).
-We conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight.
+Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C) [@Mcclure2018].
+To recover free fatty acids from bound lipids for a higher yield, we conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight [@Kaluzna2016; @Craig2004].
 Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL).
@@
 
-The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector.
-Samples were injected in splitless mode at 320 °C, and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness).
+The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector (Agilent technologies, USA).
+Samples were injected in splitless mode at 320 °C, and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness; J&W Scientific, USA).
 The carrier gas was helium set at a constant flow rate of 1.3 ml/min.
 The temperature of the oven was set at 90 °C and held for 1 minute and then was increased at a rate of 5 °C/min to 320 °C, where it was held for 23 minutes.
-The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode.
+The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode (Agilent technologies, USA).
 The carrier gas was helium flowing at a rate of 1.6 ml/min.
@@
 The final temperature was set for 22 minutes.
-The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface.
+The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface (Thermo Fisher Scientific, USA).
 The sample injection volume was 1 ul in toluene and was injected in splitless mode.
@@
   mutate(FA = case_when(
-    retention == 21.6 ~ "C14:0",
-    retention == 23.6 ~ "C15:0",
-    retention == 25.5 ~ "C16:0",
-    retention == 27.3 ~ "C17:0",
-    retention == 28.4 ~ "C18:1",
-    retention == 29.1 ~ "C18:0",
-    retention == 32.3 ~ "C20:0",
-    retention == 35.4 ~ "C22:0")) %>%
+    retention == 21.6 ~ "C[14:0]",
+    retention == 23.6 ~ "C[15:0]",
+    retention == 25.5 ~ "C[16:0]",
+    retention == 27.3 ~ "C[17:0]",
+    retention == 28.4 ~ "C[18:1]",
+    retention == 29.1 ~ "C[18:0]",
+    retention == 32.3 ~ "C[20:0]",
+    retention == 35.4 ~ "C[22:0]")) %>%
   filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062")) %>%
@@
   mutate(FA = case_when(
-    retention == 21.6 ~ "C14:0",
-    retention == 25.5 ~ "C16:0",
-    retention == 27.3 ~ "C17:0",
-    retention == 28.4 ~ "C18:1",
-    retention == 29.1 ~ "C18:0",
-    retention == 31.0 ~ "C19:0",
-    retention == 32.4 ~ "C20:0",
-    retention == 35.4 ~ "C22:0",
-    retention == 36.8 ~ "C23:0",
-    retention == 38.2 ~ "C24:0",
-    retention == 40.9 ~ "C26:0",
+    retention == 21.6 ~ "C[14:0]",
+    retention == 25.5 ~ "C[16:0]",
+    retention == 27.3 ~ "C[17:0]",
+    retention == 28.4 ~ "C[18:1]",
+    retention == 29.1 ~ "C[18:0]",
+    retention == 31.0 ~ "C[19:0]",
+    retention == 32.4 ~ "C[20:0]",
+    retention == 35.4 ~ "C[22:0]",
+    retention == 36.8 ~ "C[23:0]",
+    retention == 38.2 ~ "C[24:0]",
+    retention == 40.9 ~ "C[26:0]",
     retention == 33.9 ~ "IS",
     retention == 41.3 ~ "IS",
-    retention == 45.6 ~ "C30:0",
-    retention == 47.8 ~ "C32:0")) %>%
+    retention == 45.6 ~ "C[30:0]",
+    retention == 47.8 ~ "C[32:0]")) %>%
   filter(name == "SYG-TN13-E22-2#2")
@@
             nudge_y = 120000,
-            show.legend = FALSE) +
+            show.legend = FALSE,
+            parse = TRUE) +
   geom_text(data = highlight_IS,
@@
 
-The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the presence of lipid compounds were above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)).
-The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid), C~18:0~ (stearic acid), and monounsaturated fatty acid C~18:1~ (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)).
-These fatty acids commonly exist in animal fat and plant oil.
-They are also generally degraded from larger molecules, such as triacylglycerols, due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007].
-High concentrations of medium-chain fatty acids (C~16:0~ and C~18:0~) with an even-over-odd carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes.
+The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the amount of lipid compounds ranges from 7 to 48.6 μg/g, above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)).
+The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid), C~18:0~ (stearic acid) (Figure \@ref(fig:chromatograms-plots-four)).
+The fatty acids C~16:0~, C~18:0, and C~18:1~ commonly exist in animal fat and plant oil.
+Unsaturated fatty acid C~18:1~ was also found in all samples, but we are unable to distinguish the C~18:1~ types based on current method.
 In addition, a number of potsherds had higher contents of C~18:1~ compared to C~18:0~ and C~16:0~, such as SYG-TN13-E23-3.
-A higher content of C~18:1~ and C~16:0~ with a lower content of C~18:0~ might indicate the origin of plant oils [@Mayyas2018; @Copley2005].
-However, we may still need to examine their isotopic values to confirm their sources.
 
@@
   quant_FA %>%
+  rename("ng total" = "total FAs ug/g") %>%
   pivot_longer(!number&!sample&!`g extracted`&!`%reco`,
@@
   filter(str_starts(FA, "ng") & !sample =="") %>% 
-  mutate(FA = str_remove(FA, "ng "),
-         weight = round(weight/`%reco`/`g extracted`/1000, 2)) %>% 
-  mutate(weight = ifelse(is.na(weight)|weight == "0", "-", weight)) %>% 
+  filter(!FA == "ng C34:0" & !FA == "ng C21:0") %>% 
+  mutate(FA = str_remove(FA, "ng")) %>% 
+  mutate(weight = ifelse(str_detect(FA, "C"), 
+                         round(weight/`%reco`/`g extracted`/1000, 2), weight)) %>% 
   select(-number, -`%reco`, -`g extracted`) %>% 
+  mutate(weight = round(weight, 2),
+         weight = ifelse(is.na(weight), "-", weight),
+         weight = ifelse(weight == 0, "-", weight)) %>% 
   pivot_wider(names_from = sample, 
@@
          "SYG-TN13-E23-3", "SYG-3", "CDG-062") %>% 
-  filter(!FA == "C34:0" & !FA == "C21:0")
+  mutate(FA= str_replace(FA, "C(.*)", "C~\\1~"))
 
@@
 'Cn:x' refers to fatty acid with n carbon atoms and x double bonds.
-'IS' indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
+'IS' indicates internal standard used for quantification, including FA C~21:0~ and 5-α Cholestane"}
 knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
@@
             nudge_y = 250000,
-            show.legend = FALSE) +
+            show.legend = FALSE,
+            parse = TRUE) +
   scale_y_continuous(labels = scales::comma_format(),
@@
 Long-chain fatty acids from C~20:0~ to C~32:0~ are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)).
-The existence of long-chain fatty acids with a high amount of C~16:0~ relative to C~18:0~ might be related to plant waxes [@Post1996; @Copley2005].
-However, it could also be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
+Although the existence of even-numbered long-chain fatty acids, such as C~30:0~ and C~32:0~, with a high amount of C~16:0~ relative to C~18:0~ might be related to plant waxes [@Post1996; @Copley2005], it is more likely to be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
 Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.
@@
 'Cn:x' refers to fatty acid with n carbon atoms and x double bonds.
-'IS' indicates internal standards used for quantification, including FA C21:0 and 5-α Cholestane"}
+'IS' indicates internal standards used for quantification, including FA C~21:0~ and 5-α Cholestane"}
 knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
@@
 
+E23_C16 <- round(delta_meth_cor$`13C C16:0`[6], 2)
+E23_C18 <- round(delta_meth_cor$`13C C18:0`[6], 2)
+
+SYG3_C16 <- round(delta_meth_cor$`13C C16:0`[2], 2)
+SYG3_C18 <- round(delta_meth_cor$`13C C18:0`[2], 2)
+
+E22_1_C16 <- round(delta_meth_cor$`13C C16:0`[4], 2)
+E22_1_C18 <- round(delta_meth_cor$`13C C18:0`[4], 2)
+
+E22_22_C16 <- round(delta_meth_cor$`13C C16:0`[5], 2)
+E22_22_C18 <- round(delta_meth_cor$`13C C18:0`[5], 2)
+
 # plot the carbon isotopes 
@@
   theme_minimal() +
-  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
-       y = bquote(delta*{}^13*"C 18:0 \u2030")) +
+  labs(x = expression(paste(delta*{}^13* C[16:0], "\u2030")),
+       y = expression(paste(delta*{}^13* C[18:0], "\u2030"))) +
   scale_x_continuous(limits = c(-33, -20),
@@
   theme_minimal() +
-  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
-       y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
+  labs(x = expression(paste(delta*{}^13* C[16:0], "\u2030")),
+       y = expression(paste(delta*{}^13* C[18:0], "\u2030"))) +
   scale_x_continuous(limits = c(-33, -20),
@@
   annotate("text", x = -32, y = 1.3,
-           label = "C3-diet") +
+           label = "C[3]-diet", parse=TRUE) +
   annotate("segment", x = -29, y = 1.3, xend = -24, yend = 1.3,
@@
   annotate("text", x = -21, y = 1.3,
-           label = "C4-diet") +
+           label = "C[4]-diet", parse=TRUE) +
   annotate("text", x = -22, y = -1.1,
@@
 To identify the sources of degraded animal fats, we conducted isotope analysis to obtain stable carbon isotope compositions, δ^13^C values of C~16:0~ and C~18:0~, using GC-C-IRMS.
-The δ^13^C values of C~16:0~ and C~18:0~ have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008].
-To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C (δ^13^C~18:0~ - δ^13^C~16:0~) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
-The carbon isotope values for C~16:0~ and C~18:0~ have a distribution between `r min_carbon_isotope` and `r max_carbon_isotope`, and their Δ^13^C values range from `r delta_min` ‰ to `r delta_max` ‰.
-The Δ^13^C values were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].
+The δ^13^C values of C~16:0~ and C~18:0~ have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by distinct fat types [@Dunne2012; @Copley2003; @Evershed2008].
+Considering the bias might be introduced from differential δ^13^C values across geographical regions, we also used the Δ^13^C (δ^13^C~18:0~ - δ^13^C~16:0~) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
+The carbon isotope values for C~16:0~ and C~18:0~ from our samples have a distribution between `r min_carbon_isotope` ‰ and `r max_carbon_isotope` ‰, and their Δ^13^C values range from `r delta_min` ‰ to `r delta_max` ‰.
+The δ^13^C values were plotted based on the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].
+Although the referenced isotopic values should be ideally adjusted in terms of local environments, the known reference ranges especially for East Asia could still provide useful information about food categories with cautious comparison.
 
-The results show that the majority of the potsherd samples from the Guijiabao site contain fatty acids from ruminant adipose fats, with one potsherd sample (SYG-TN13-E23-3) indicating a non-ruminant adipose fat origin (Figure \@ref(fig:isotopic-plots): A).
-Analysis of the potsherd from the Gaoshan site (CDG-062) indicates ruminant adipose fats as well.
-The potsherds (SYG-TN13-E22-1 and SYG-3) from the Guijiabao site dating to the historical period were plotted on the reference line between ruminant adipose and non-ruminant, indicative of the mixing of these two origins [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
-Furthermore, the potsherd representing non-ruminant adipose fats has a relatively low ^13^C value (Figure \@ref(fig:isotopic-plots): B), which is consistent with the value range of C3/terrestrial plants or freshwater fish [@Shoda2018; @Kwak2020].
-This also corresponds to the lipid distribution of the same potsherd detected by GC-MS, reflecting a possible plant origin based on the GC-MS results.
-Although the referenced isotopic values should be ideally adjusted in terms of local environments, they could still provide useful information for determining food categories.
+The results show that the majority of the Neolithic potsherd samples from Guijiabao and Gaoshan contain saturated fatty acids derived from ruminant adipose fats, with one Guijiabao potsherd sample, SYG-TN13-E23-3, indicating a non-ruminant adipose fat origin according to the Δ^13^C value 0.06 ‰ (Figure \@ref(fig:isotopic-plots)).
+This potsherd (SYG-TN13-E23-3) has the mean δ^13^C values `r E23_C16`‰ and `r E23_C18`‰ for the C~16:0~ and C~18:0~ fatty acids respectively, consistent with the isotopic value ranges of C~3~ plants, terrestrial or freshwater fish (ca.
+−26.5‰ to -33‰ for both compounds) reported for Neolithic potsherds in southern China [@Shoda2018], and freshwater aquatic products (-27.3‰ to 31.6‰ for the C~16:0~ and -26.4‰ to -29.9 for the C~18:0~) or wild boar (ca.
+-27‰ to -29.5‰ for both compounds) reported for the Jōmon pottery in western Japan [@Lucquin2016].
+However, we are unable to distinguish between the C~3~ plants, terrestrial and freshwater fish by this isotopic approach alone.
+A further detection of biomarkers for fish such as isoprenoid fatty acids, or for starchy plants and millet will be needed for an better understanding [@Shoda2018; @Lucquin2016; @Heron2016].
 
-In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with the δ^13^C values of C3/terrestrial plants, suggesting these ruminant animals seemed to have consumed mainly C3 plants.
-There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively large δ^13^C value, suggesting a C4-based diet for the ruminant animal.
+For the pottery fragments (SYG-TN13-E22-1 and SYG-3) dating to the historical period, we found that the mean δ^13^C values of the extracts were plotted on the reference line between ruminant adipose and non-ruminant fats.
+Specifically, SYG-TN13-E22-1 and SYG-3 have the mean δ^13^C values `r E22_1_C16` and `r SYG3_C16` for the C~16:0~ fatty acids, and the mean δ^13^C values `r E22_1_C18` and `r SYG3_C18` for the C~18:0~ fatty acids.
+It is unsure to know what the origin of these values are, but it is possible that there is a mixing of both non-ruminant sources and ruminant fat [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
+It should be noted that the δ13C values could be influenced by multiple different products, such as ruminant products and plant products in the same cooking event, or over different cooking events with different products.
 
-```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C~16:0~ and C~18:0~. B: The Δ13C (δ^13^C~18:0~-~16:0~) values are plotted against δ^13^C~16:0~ values"}
+There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively enriched δ^13^C value (`r E22_22_C16` for the C~16:0~ and `r E22_22_C18` for the C~18:0~) compared with other samples.
+It is widely known that the δ^13^C values of fatty acids of marine source or C~4~ plants are more enriched than those of ruminant fats [@Lucquin2016].
+Given that Guijiabao is situated inland with finding of millet grains, the enriched δ^13^C values might indicate the processing of C~4~ plants instead of marine resources, or the mixture of C~4~ plants and ruminant adipose products due to a slightly depleted δ^13^C value for the C~18:0~ fatty acid.
+However, in general, most samples have depleted δ^13^C values, which consistent with the δ^13^C values of C~3~ plants or ruminant fats, suggesting the processing of C~3~ plants or C~3~ plants were cooked with ruminant adipose meat.
+We also could not rule out the possibility that these depleted δ13C values came from mainly ruminant animals fed on C~3~ plants.
+
+```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C~16:0~ and C~18:0~.
+B: The Δ^13^C (δ^13^C~18:0~-~16:0~) values are plotted against δ^13^C~16:0~ values.
+The dotted lines represent cut-offs for ruminant, non-ruminant, and ruminant dairy fats according to global references"}
 knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
@@
 <!--A brief summary of interesting results-->
-The aim of this paper is to explore whether new crops, such as C4 plants (e.g., millets) and C3 plants (e.g., rice), or animals (e.g., pigs/horses) became an essential part of the local diet in Yanyuan during the Neolithic period.
-Our results show that food residues in potsherds from the sites at Guijiabao and Gaoshan were dominated by ruminant fats during the Neolithic period.
-Moreover, it is striking to discover that non-ruminant fats in a Guijiabao potsherd might have originated from C3 plants (see below).
-Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that these ceramics were used as cooking or storage vesssels.
+This paper is a preliminary study to investigate the pottery use at Guijiabao and Gaushan in Yanyuan area to explore whether those pots were used for cooking and what food categories could be processed in the pots.
+This will provide a direct evidence of the culinary practices in the Yanyuan region.
+Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that these ceramics were used as cooking or food storage vessels.
 The low lipid yield could be attributed to the bottom of the vessels, where there were fewer residues than other parts of the vessel, for example, the rim [@Evershed2008experimental; @Kwak2020; @Charters1993].
-But we also found that one sample from the vessel bottom dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples, reflecting better preservation or a much higher lipid yield.
-We recognize that there are limitations and challenges for determining lipid sources over time or differences across regions due to our small sample sizes and a lack of local modern references for animal fats and plant oils.
+However, we also found that one sample from the vessel bottom dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples.
+We recognize that there are limitations and challenges for interpretation due to our small sample sizes and a lack of local modern references for animal fats and plant oils.
 In addition, a detailed analysis of faunal remains from the Guijiabao site is still in progress, which may constrain a comprehensive comparison between different types of evidence.
-Nevertheless, the general pattern of lipid profiles could still provide informative clues for a better understanding of culinary practices at the Guijiabao site.
+Nevertheless, the lipid profiles and their isotopic ratios could still provide informative clues for a better understanding of culinary practices.
+
+Saturated fatty acids are the major lipid molecules identified in our samples that normally come from animal fats or plant oils, indicative of cooking events.
+In addition to saturated fatty acids, unsaturated fatty acids such as C~18:1~ was also identified in all samples, and three (SYG-TN13-E22-1, SYG-TN13-E22-2#2, and SYG-TN13-E23-4) out of the six potsherds show a more abundant C~18:1~ than C~18:0~ and C~16:0.
+Although the specific type of these unsaturated fatty acids can not be further identified in this study without the preparation of specific derivatives [@Berstan2004], their presences might still hint a possibility of plant oils.
+It is suggested that a higher ratio of C~18:1~ and C~16:0~ to C~18:0~ is associated with a greater likelihood of the origin of plant oils [@Mayyas2018; @Copley2005].
+In our samples, the potsherd SYG-TN13_E23-3 has a higher ratio of C~18:1~ and C~16:0~ to C~18:0~, which also corresponds to the category of non-ruminant fats based on the isotopic values.
+However, we also acknowledge that C18:1 might have originate from animal fats, or there is a potential for modern contamination since monounsaturated fatty acids normally present limited preservation in archaeological contexts [@Whelton2021].
 
 <!--discussion about possible plant origins-->
-While it has been determined that millets (C4 plants), including foxtail and broomcorn millets, were the dominant cereals, according to recent macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022], we did not detect any lipid residues associated with C4 plants in our samples.
-Surprisingly, we identified non-ruminant fats from a Guijiabao potsherd that might have been from terrestrial plants/C3 plants, based on similar isotopic values reported by other studies in East Asia [@Shoda2018; @Kwak2020].
-According to the results from flotation analysis at the Guijiabao site, it is possible that the source of C3 plants is rice or native wild weeds, such as pennisetum and chenopodiaceae.
+According to the δ^13^C values of fatty acids, the food residues in Neolithic potsherds could derive from non-ruminant fats, ruminant fats, and mixtures of ruminant fats and C~3~ plants or C~4~ plants.
+Previous studies suggested that millets (C~4~ plants), including foxtail and broomcorn millets, were the dominant cereals, according to macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022].
+The processing of millet might also be suggested by one Guijiabao potsherd dating to Neolithic period based on its isotopic ratios.
+However, a further confirmation would require to identify of any lipid residues associated with C~4~ plants, such as the biomarker for broomcorn millet, miliacin (olean-18-en-3β-ol methyl ether) [@Heron2016].
+The non-ruminant fats we identified from a Guijiabao potsherd that might have been from C~3~ plants or freshwater aquatic products, based on similar isotopic values reported by other studies in East Asia [@Shoda2018].
+According to the results from flotation analysis at the Guijiabao site, it is possible that the source of C~3~ plants is rice or native wild weeds, such as *Pennisetum* and *Chenopodiaceae*.
 However, the seeds of wild weeds only account for a small part of the floral assemblage derived from flotation [@Guijiabao2016].
-Despite stone knives and tools that were also found at the sites that might have been related to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022].
-Another possible source for non-ruminant fats could have been freshwater fish, a possibility which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b].
-But we did not identify any lipid biomarkers related to aquatic resources, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids or phytanic acid [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016], from our potsherds.
+Given that a more depleted isotopic values for 5 out of 6 samples, ~C~3~ plants might consumed more than ~C~4 plants either by local people or by animals that were food source.
+
+We can not rule out other sources for non-ruminant fats, such as freshwater aqautic products, a possibility which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b].
+This will also need identification of lipid biomarkers related to aquatic oils, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids and isoprenoid fatty acids [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016].
+Although horse could be another non-ruminant fat source, whether it was part of local food choices in southwestern China around 4500 years ago during the Neolithic period is doubtful.
+First, the Pleistocene and Neolithic sites having wild horse bones are primarily concentrated in northern China, with only sporadic sites in the south [@Yuan2006].
+Second, domesticated horses first found in southwestern China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period.
+Thus, if horse meats is one of the source for non-ruminant fats, then it could only be present in historical ceramics.
 
@@
 The ruminant fats identified in the Neolithic potsherds (ca.
-4500 BP) from the Guijiabao and Gaoshan sites could have been derived from deer, sheep, or horses that were fed on a C3 diet, based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017].
-This finding, especially the consumption of deer, also corresponds with unearthed faunal remains from the sites and the local natural environments.
-Stone artifacts found at the site, including stone balls and arrowheads, further indicate hunting activities [@Hao2022; @Guijiabao2020a].
-However, whether horse was part of local food choices in southwestern China around 4500 years ago is doubtful.
-First, the Pleistocene and Neolithic sites having wild horse bones are primarily concentrated in northern China, with only sporadic sites in the south [@Yuan2006].
-Second, domesticated horses first found in southwestern China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period that we are focusing on.
-Thus, supported by faunal evidence, deer was most likely the predominant local food choices.
+4500 BP) from the Guijiabao and Gaoshan sites could have been derived from deer or sheep that were fed on a C~3~ diet, based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017].
 It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these animals have been found in sites in southwestern China [@Zhang2016; @Chen2020].
+It is hard to identify the exact ruminant category in this case due to the limited accessibility to local animal bones or fats as reference samples.
+However, the finding of fatty acid source from ruminant fats, couple with stone artifacts found at the site such as stone balls and arrowheads, suggesting hunting could be an important subsistence activity at the site [@Hao2022; @Guijiabao2020a].
+This also corresponds with resources that could be accessed in a highland environment.
 
-We observed a mixture of ruminant fats and C3 plants, based on analyses of a number of Guijiabao potsherds that date to the historical period (ca.
+We observed mixtures of ruminant fats and C~3~ plants/freshwater aquatic source, based on analyses of two Guijiabao potsherds that date to the historical period (ca.
 1000 BP).
-It appears that C3 plants could have been processed more in the historical period than they could have been in the Neolithic period.
-In addition to rice, the source of C3 plants could also have included wheat, since it was introduced into the region during the historical period [@JiangGaoshan2017].
+In addition to rice, the source of C~3~ plants could also have included wheat, since it was introduced into the region during the historical period [@JiangGaoshan2017].
 A larger sample size is needed to determine whether there was a change in diet over time.
-The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C3 plants might have been an important food source for local animals.
-Despite direct C4 plant signals not being found, analysis of one Guijiabao potsherd shows the source of C4 plant-fed ruminants during the Neolithic period.
-This indicates that either C3 or C4 plants could have been the food source for ruminants, but whether those animals were domesticated is unknown.
+The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C~3~ plants might have been an important food source for local animals.
+Despite direct C~4~ plant signals not being found, analysis of one Guijiabao potsherd shows the source of C~4~ plant-fed ruminants during the Neolithic period.
+This indicates that either C~3~ or C~4~ plants could have been the food source for ruminants, but whether those animals were domesticated is unknown.
 The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, the results of which could provide more details about local animal species for comparison.
 
-In general, based on our results, ruminant fats from local wild animals, such as deer, were the predominant food source during the Neolithic period for peoples at Guijiabao.
-A subsistence that relied on hunting to a certain degree could be plausible, given the abundance of wild animals in the natural settings of the Hengduan mountain ranges, as indicated by the faunal remains from the Yingpanshan site [@He2020].
-Crop cultivation could have been adopted as a supplement to local food sources when it was introduced in the Neolithic period, but crops might have only accounted for a small proportion of the food sources.
-A previous study using isotopic analysis of human bones suggests that the diet at the Guijiabao site was a C4-based diet [@Lin2022], but we only identified a signal from C3 plants from one pottery sample without any signal from C4 plants.
-This inconsistency could have resulted from different sampling strategies or from the sample size issue, where our results reflect only a small portion of food content.
-Nevertheless, our results suggest that farming practices during the Neolithic period might not have been adopted as much as we previously thought.
+Mixed isotopic signals were also identified from two
+
+Our preliminary results shows that culinary practices in the Neolithic period at Guijiabao include the processing of ruminant adipose products that might be from local wild animals, such as deer.
+A subsistence that relies on hunting to a certain degree could be possible, given the abundance of wild animals in the natural settings of the Hengduan mountain ranges, as indicated by the faunal remains from the Yingpanshan site [@He2020].
+
+It is also likely that C~3~ plants (eg.
+rice) was processed based on the isotopic ratios showing non-ruminant fats from one Neolithic potsherds at Guijiabao.
+In addition to C~3 plants, another possibility for the source of non-ruminant fats could be freshwater aquatic products, which would require confirmation of the presence or absence of aquatic products.
+It is interesting to detect that one potsherd has a slightly higher values of carbon isotope ratios, indicating a contribution from C4 plants.
+The detection of C~3~ plants or C~4~ plants might suggest that farming practices were adopted during the time when those crops were first introduced to the Yanyuan area in the Neolithic period.
+The finding of C~4~ plant signal also corresponds with a previous study that suggests that a C~4~-based diet at the Guijiabao site using isotopic analysis of human bones [@Lin2022].
+However, the scale of farming practices and crop consumption is unclear.
+Despite stone knives and tools that were also found at the sites that might have been related to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022].
+
 An approach analyzing organic residues may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through presence of food residues that were cooked in ceramic vessels.
 A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period in southwestern China.
+
+An further investigation on the aquatic source should be conducted through selected ion monitoring (SIM) methods in the future to distinguish whether the δ13C
+values are the result of processing C3 plants or freshwater aquatic products in the pottery vessel.
+
+sampling bias, poor preservation
+
+<!--- removed
+such as C~4~ plants (e.g., millets) and C~3~ plants (e.g., rice), or animals (e.g., pigs) became an essential part of the local diet in Yanyuan during the Neolithic period.
+while the food residues in historical potsherds present mixtures of ruminant and C~3~ plants.
+-->
 
@@
 The foods cooked in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwestern China.
-The source of terrestrial plants could have been C3 plants, implying the cooking of rice or other wild weeds in the pots.
-However, the evidence of direct cooking of either C3 or C4 plants shows that the amount of C3/C4 plants cooked is not in proportion to the large quantity of plant remains found at the Guijiabao site, especially C4 plants such as millet.
+The source of terrestrial plants could have been C~3~ plants, implying the cooking of rice or other wild weeds in the pots.
+However, the evidence of direct cooking of either C~3~ or C~4~ plants shows that the amount of C~3~/C~4~ plants cooked is not in proportion to the large quantity of plant remains found at the Guijiabao site, especially C~4~ plants such as millet.
 Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the Neolithic period.

### 2024-01-08 analysis/paper/paper.Rmd
edits based on comments

@@
 ---
-title: "Preliminary insights into the use of pottery and culinary pratices in southwestern China"
+title: "Preliminary insights into the use of pottery and culinary pratices at Guijiabao site in southwestern China"
 author:
@@
 <!-- intro to the topic and research aim-->
-Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on local populations involved in the interactions [@Burmeister2000].
-In prehistoric southwest China, the Hengduan Mountain ranges have served as passages for people to move between the Qinghai-Tibetan Plateau in the north, the Yunnan–Guizhou Plateau in the south, and the Sichuan Basin in the east.
-The movements in the Hengduan Mountain ranges facilitated not only cultural exchange, but also the spread of new subsistence practices, such as farming and animal domestication [@Tong1986; @Hein2014; @Chen2015].
-With more archaeological sites have been found recently along the Hengduan Mountain valleys, researchers are able to dive deeper into local cultural developments and diversity, the timing and routes of exchange network, and the spread of farming practices during Neolithic period across the Hengduan Mountain ranges and neighboring areas [@Hein2014; @Hein2016; @Ma2021; @Lin2022; @Hao2022; @Ma2021].
 
-Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern China as early as 4700 BP, where the adoption of farming in southwestern Sichuan is around 4700 BP and in western Yunnan is around 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022].
-Recent excavations at Guijiabao in the middle Hengduan Mountain ranges bridges the gap between southwestern Sichuan and Yunnan through the finding of millet and rice around 4700 BP [@Huan2022; @Hao2022].
-However, it is still not fully understood the extent of the adoption of farming practices by the ancient people in the Hengduan Mountain ranges.
-Did crop cultivation become the essential part of local subsistence when they were introduced?
-What other resources were obtained in addition to those grains?
+The movement of Neolithic populations has long been discussed in archaeology to understand the processes of agricultural diffusion, regional variations, and their consequences [@bellwood2023; @Davison2006].
+In southwest China, the emergence of farming practices was thought to be associated with the migrating population from central or northern China in the Neolithic period [@Deng2019].
+Crop remains, including foxtail millet, broomcorn millet, and rice, found in prehistoric southwest China that can be traced back to as early as 4700 BP in Sichuan and around 4600 BP in western Yunnan [@Jiang2016; @Deng2019; @YunnanMuseum1981; @Zhang2010; @Dal2022].
+Despite different views on the routes of agricultural dispersal [@Guedes2013; @Guedes2018; @Liu2022], it is generally believed that mixed farming was the main economy in southwest China after the introduction of agriculture [@Deng2019].
+However, more studies still needed to understand how migrant communities interacted with new environments, especially in an area with high diversity of landscape, such as high mountain, plateau, basin, and terrace.
 
-In this study, we focus on archaeological sites
+This study focuses mainly on archaeological sites located in the Hengduan Mountain ranges, a region in southwest China connecting the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south.
+As a critical passage for population movement, the Hengduan Mountain ranges and valleys facilitate not only material exchange [@Tong1986; @Hein2014; @Hein2016; @Ma2021], but also the spread of agriculture even further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017].
+Recent excavations at the Guijiabao site in Yanyuan Basin, the middle Hengduan Mountain ranges, is a key site with the earliest Neolithic human occupation in this region dating to 5000 B.P., continuing to the historical period [@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022].
+The complete stratigraphy at Guijiabao provides a unique opportunity to investigate questions relative to the subsistence since the Neolithic period in a basin surrounded by high mountains.
+For example, what is the role of agriculture in local environments?
+What other resources were obtained in addition to crops?
+In this study, we present our preliminary analyses on archaeological potsherds from the Guijiabao site and the Gaoshan site in southwest China using organic residue analysis of lipids.
 
-aim to to explore the culinary practices through food residues within ceramic vessels from Guijiabao and Gaoshan sites
+Although the remains of animal bones and plants offer evidence for subsistence, their preservation heavily depend on environmental conditions.
+Geographical regions with seasonal heavy precipitation and acidic sediments, such as Southwest Sichuan and the middle Hunduan Mountain [@Li1989], tend to lead to poor preservation of archaeological remains.
+In such case, organic residue analysis is a powerful tool for studying ancient food practices by directly examining residues in pottery.
+Organic residues, specifically lipids, can be well preserved in the porous fabric of pottery for over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017; @Dudd1999; @Miller2020].
+The growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies [@Kwak2020], and culinary practices across a region over time [@Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
 
-Also, the discussion of local subsistence mostly relies on paleobotanical evidence and zooarchaeological assemblages,  The preservation of animal or plant remains could be heavily impacted by environmental conditions.
-
-to explore the culinary practices in the Yanyuan Basin through food residues within ceramic vessels from Guijiabao and Gaoshan sites.
-We analyzed small samples to test whether those pottery were used for cooking, and if we could detect new crops or species such as rice, millet, or pigs, as food sources when they spread into the Yanyuan Basin.
-Guijiabao is a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a].
-The archaeobotanical evidence at Guijiabao reveals that foxtail millet, broomcorn millet, and rice were present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Huan2022], and millet could have already been part of the local diet [@Lin2022].
-Although some of these analyses suggest the presence of mixed millet/rice farming [@Huan2022], we still lack direct evidence of culinary practices during this period.
-To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery?
-Can we identify the new crops or terrestrial animals that were incorporated into the local food diets during the Neolithic period?
-We answer these questions by examining potsherds using organic residue analysis, which allows us to directly explore food cooked or processed in ceramic vessels.
-
-The recent growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
-Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing food origins; for example, millet can be identified through the biomarker miliacin [@Copley2005; @Whelton2021; @Heron2016].
-This is possible for prehistoric ceramics due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery for over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017].
-In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reveal the original contents cooked or stored in a vessel during its lifetime of use [@Dudd1999; @Miller2020].
-Through our case study, we attempt to investigate culinary practices during a period of frequent cultural interactions that accompanied agricultural expansion.
+Through identification of lipid biomarkers and compound-specific carbon isotope analysis based on geochemical techniques, archaeologists are able to distinguish food origins between plant oils, ruminant fats, diary products, and aquatic/marine resources [@Copley2005; @Craig2007; @Evershed1994].
+Recent studies further successfully identified different types of plants processed in archaeological vessels, such as broomcorn millet [@Bossard2013; @Heron2016; @Ganzarolli2018], starchy plants [@Shoda2018], and cereals [@Hammann2018; @Colonese2017].
+A growing case studies for ceramics in China present the robustness of organic residue analysis for understanding regional cooking practices [@Shoda2018; @Lyu2024; @Han2022; @Lyu2024], distinct subsistence strategies between low- and high altitude zones [@Zhang2022], and the impact of nomadic pastoralism on northwest China [@Sun2023].
+In this study, we attempt to investigate pottery use and culinary practices in the middle Hengduan mountains to understand the local subsistence by focusing on the ceramics from the Guijiabao and Gaoshan sites.
 
@@
 <!--Yanyuan area and Guijiabao site-->
-We focus on the Guijiabao site located in the Yanyuan Basin, which is a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)).
-The Guijiabao site allows us to understand the early adoption of agriculture because of its uniqueness and rarity: it contains complete cultural components from the Neolithic period to the historical period in that region [@Guijiabao2017a; @Guijiabao2017b].
-A survey in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean houses, ground houses, post holes, ash pits, ditches, and burials, suggesting that Guijiabao was a settlement site [@Hao2022].
-The chronology of the Guijiabao site can be divided into three phases, the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the late North Song to early Ming dynasties (1200-500 BP) based on 19 radiocarbon dates from plant remains and a human bone.
-For the Neolithic period, it can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the content of artifact assemblages and ceramic forms and styles that were identified [@Hao2022].
+The Guijiabao site is located in the southwest of Yanyuan Basin, a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)), with an elevation of 2400 meters.
+The first survey at Guijiabao in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean and ground houses, post holes, ash pits, ditches, and burials, suggesting a settlement site [@Hao2022; @Guijiabao2017a].
+The site was occupied since the Neolithic period (5000-3700 BP), continued to the Bronze Age (3200-2700 BP) and the late North Song to early Ming dynasties (1200-500 BP), based on 19 radiocarbon dates [@Hao2022].
+The Neolithic period can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the artifact assemblages and ceramic styles
 
 <!--the major finding at GJB-->
-The common vessel forms unearthed at the Guijiabao site include jars, urns, bowls, and basins.
-The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a].
-Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between the Guijiabao site and sites in Yunnan and the southern part of the Hengduan Mountains [@Guijiabao2017a].
-Plant remains reveal that cultivated crops were dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022].
-This corresponds to a C~4~-based diet based on isotopic analysis of four human bones [@Lin2022].
-In addition to millet, a smaller number of rice grains (*Oryza sativa*) has also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022].
-A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences [@Huan2022].
+Ceramics is the major finding at Guijiabao.
+The common vessel forms include jars, urns, bowls, and basins.
+The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with cord-marked or basket-impressed patterns [@Hao2022; @Guijiabao2017a].
+The evidence of chars on their surfaces suggest that they might be used for cooking [@Guijiabao2017a; @Hao2022].
+This research also attempts to confirm that these pottery vessels were used as cooking/storage wares.
+Some ceramic decorations present northern styles, while others with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of close relationships or interaction between Guijiabao and neighboring areas, the northern and southern parts of the Hengduan Mountains [@Guijiabao2017a].
+The early excavation reveals the presence of cultivated crops at Guijiabao around 4700 BP, including foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*), earlier than previously thought in Yanyuan [@Guijiabao2016].
+In later excavations, rice grains (*Oryza sativa*) were also found throughout the Neolithic period, but in a smaller number than millet [@Hao2022; @Huan2022].
 
 <!--info about Gaoshan-->
-To compare the culinary practices in different ecological zones, we examined a sample from the Gaoshan site in the southwestern Chengdu Plain, an alluvial plain in the western Sichuan Basin at the foot of the Qinghai-Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
+To compare the culinary practices in different ecological zones, we examined a sample from the Gaoshan site in southwestern Chengdu Plain, an alluvial plain in Sichuan close to the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
 Excavations at the Gaoshan site reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and archaeological features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
@@
 The primary subsistence strategy at the Gaoshan site could have been rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020].
-It is suggested that the spread of rice farming into the Sichuan area occurred at least 4700 years ago from people moving from the Middle and Lower Yangzi valleys to the Sichuan area [@Zhang2010].
+It is suggested that the spread of rice farming into Sichuan occurred at least 4700 years ago through population migration from the Middle and Lower Yangzi valleys [@Zhang2010].
 
-Although the recent findings at the Guijiabao and Gaoshan sites support mixed agriculture as the primary subsistence strategy, it is still unclear how much millet or rice was included in the local diets during the Neolithic period.
+Although mixed agriculture of millet and rice observed in southwest China, it seems there are variations in crop cultivation across the region.
 The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011].
-However, plant remains from many sites in southwestern China reveal a predominance of rice, such as at the Gaoshan and Baodun sites in Sichuan [@JiangGaoshan2017] and Baiyangcun sites in Yunnan [@Jiang2016; @Henglanshan2014].
-This indicates that there were variations in agricultural practices in southwestern China.
-To better understand the extent of the adoption of new crops, organic residue analysis of ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking.
-This provides another way to examine the expansion of agriculture in the Yanyuan area.
+However, botanical evidence from most sites in the Chengdu plain reveal a predominance of rice, such as at the Gaoshan and Baodun sites [@JiangGaoshan2017], and the Baiyangcun site in Yunnan [@Jiang2016; @Henglanshan2014].
+Moreover, most research focuses on the agricultural practices, what other resources utilized by early farmer in new environments remains unclear.
+Here we seek to examine direct evidence from pottery residues to provide clues on culinary practices through case study in southwest China using geochemical techniques.
 
@@
 
-## Samples
-As a preliminary research to accesses the feasibility of this approach, we selected ceramic fragments that were identified as typical from the Guijiabao and Gaoshan sites.
-The fragments date to the Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)), which were determined from known contexts based on radiocarbon dating.
-Ceramic fragments from the Guijiabao site were excavated from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season.
-The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations.
-Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that these ceramics were used in daily life [@Guijiabao2017a; @Hao2022].
-This research also attempts to confirm that these pottery vessels were used as cooking/storage wares for food.
+## Sampling
+As a preliminary research on food residues preserved in ceramic vessels, we sampled pottery fragments from cultural layers at the residential area of Guijiabao (n = 5) to test the feasibility of organic geochemical approaches (Table \@ref(tab:sherds-table)).
+As a comparison, we sampled one fragment from Goashan.
+Those pottery fragments were identified as typical types with well-defined contexts, as indicated by radiocarbon dates ranging from the Neolithic period (ca.
+4500 BP) to the historical period (ca.
+1000 BP).
+The pottery samples are mostly base fragments and fine sand-tempered, with a few displaying impressed decorations.
+Through examining potsherds from early to late period at Guijiabao, we seek to explore the potential of organic geochemical approach across different time period.
 
@@
 
-## Methods
-We detected and identified lipid compounds using gas chromatography mass spectrometry (GC-MS) and quantified lipids using gas chromatography with flame-ionization detection (GC-FID).
-To distinguish the source of the common fatty acids, palmitic (C~16:0~) and stearic (C~18:0~), which are found in most animals and plants, we conducted compound-specific isotope analysis to identify stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS).
-The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2007; @Evershed1994].
-
-For sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis.
+## Lipid extraction
+The surface of each pottery sample was removed to prevent external contamination from environments and handling.
 Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C) [@Mcclure2018].
-To recover free fatty acids from bound lipids for a higher yield, we conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight [@Kaluzna2016; @Craig2004].
+To obtain a higher yield, free fatty acids were recovered from bound lipids through saponification, conducted by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heated at 70 °C overnight [@Kaluzna2016; @Craig2004].
 Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL).
-A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour.
-After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene prior to GC-MS.
+A 5% aliquot of the extract was treated with 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) containing 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour.
+After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene for gas chromatography mass spectrometry (GC-MS).
 
@@
 The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight.
-The FAME derivatives were extracted with hexane ($3\times10$ mL) and then blown down with nitrogen streams.
-About 5% aliquot dissolved in toluene was analyzed by GC-FID for quantification, and the remaining sample was analyzed using GC-C-IRMS.
-With each batch of samples, we prepared analytical blanks to check for potential contamination sources that might have come from solvents and reagents.
-Internal standards, including FA C~21:0~ and 5-α Cholestane, were added before and after the lipid extraction for quantification.
+The FAME derivatives were extracted with hexane ($3\times10$ mL).
+About 5% aliquot was analyzed by gas chromatography with flame-ionization detection (GC-FID) for quantification, and the remaining sample was analyzed using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS) [@Craig2007; @Evershed1994].
+Analytical blanks were prepared and analyzed using the same procedure as the pottery samples to monitor potential contamination sources from solvents and reagents.
+Internal standards, FA C~21:0~ and 5-α Cholestane or n-tetratriacontane, were added before and after the lipid extraction.
 
+## GC methods
 The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector (Agilent technologies, USA).
-Samples were injected in splitless mode at 320 °C, and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness; J&W Scientific, USA).
-The carrier gas was helium set at a constant flow rate of 1.3 ml/min.
-The temperature of the oven was set at 90 °C and held for 1 minute and then was increased at a rate of 5 °C/min to 320 °C, where it was held for 23 minutes.
-The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode (Agilent technologies, USA).
+Samples were injected in splitless mode at 320 °C.
+Compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness; J&W Scientific, USA).
+The carrier gas was helium flowing at a rate of 1.3 ml/min.
+The oven temperature was set at 90 °C for 1 minute, then increased at a rate of 5 °C/min to 320 °C and held for 23 minutes.
+In addition to scan mode, a selected ion monitoring (SIM) mode was used to further detect miliacin (olean-18-en-3β-ol methyl ether; *m/z* 189, 204, 231, 425, 440) [@Heron2016].
+The oven temperature was set at 50 °C for 1 min, then raised at a rate of 10°C/min to 310 °C and held for 15 min.
+A authentic standard of miliacin was also ran to monitor the retention time.
+
+The GC-FID analysis of FAMEs was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode (Agilent technologies, USA).
 The carrier gas was helium flowing at a rate of 1.6 ml/min.
 The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode.
-The GC oven was set at 90°C and held for 1.3 minute, and then was raised to 320°C at a rate of 5°C/min.
-The final temperature was set for 22 minutes.
-The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface (Thermo Fisher Scientific, USA).
-The sample injection volume was 1 ul in toluene and was injected in splitless mode.
-Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set at 90 °C for 1.5 minutes and then was raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold.
+The GC oven was set at 90°C for 1.3 minute, then raised at a rate of 5°C/min to 320°C and held for 22 minutes.
+The GC-C-IRMS analyses of FAMEs of palmitic (C~16:0~) and stearic (C~18:0~) were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface (Thermo Fisher Scientific, USA) in splitless mode.
+Separation was carried out using the same model of column for GC-MS.
 Helium was used as a gas carrier with a constant flow rate of 1.24 ml/minute.
+The oven temperature was set at 90 °C for 1.5 minutes, then raised at a rate of 5 °C/min to 320 °C and held for 18 minutes.
 
-The raw data and R code used for visualizations contained in this paper and R Markdown documents are available online at https://github.com/LiYingWang/GJBpottery to enable research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].
+The data generated by these analyses and R code used for visualizations are openly available at https://github.com/LiYingWang/GJBpottery to enable research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].
 
@@
             size = 3,
-            nudge_y = 120000,
+            nudge_y = 115000,
             show.legend = FALSE,
@@
   geom_segment(aes(x= 41.1, xend= 41.5, y= 1030000, yend= 1030000), size =0.3) +
-  annotate("text", x = 41.3, y = 600000, label = "IS", size =3) +
+  annotate("text", x = 41.3, y = 1100000, label = "IS", size =3) +
   facet_wrap(~name, ncol = 1) +
   scale_y_continuous(labels = scales::comma_format(),
-                     limits = c(0, 1100000),
+                     limits = c(0, 1160000),
                      breaks = seq(0, 1100000, 200000)) +
@@
 
-The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the amount of lipid compounds ranges from 7 to 48.6 μg/g, above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)).
-The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid), C~18:0~ (stearic acid) (Figure \@ref(fig:chromatograms-plots-four)).
-The fatty acids C~16:0~, C~18:0, and C~18:1~ commonly exist in animal fat and plant oil.
+The total lipid concentrations in our potsherds ranges from 7 to 48.6 μg/g, above interpretable concentrations (5 μg/g) [@Evershed2008experimental] (Table \@ref(tab:FA-composition-table)).
+This suggests that those ceramics should be associated with food processing.
+The main compounds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid) and C~18:0~ (stearic acid), commonly derived from animal fat and plant oil (Figure \@ref(fig:chromatograms-plots-four)).
 Unsaturated fatty acid C~18:1~ was also found in all samples, but we are unable to distinguish the C~18:1~ types based on current method.
-In addition, a number of potsherds had higher contents of C~18:1~ compared to C~18:0~ and C~16:0~, such as SYG-TN13-E23-3.
+The domination of C~16:0~ and C~18:0~ suggests that the lipids in our samples were greatly degraded.
+However, one pottery sample (SYG-TN13-E22-2#2) has relatively high lipid yield (`r FA_max_sum` μg/g), with the presence of long-chain fatty acids (C~24:0~ to C~32:0~) (Figure \@ref(fig:chromatograms-specific)).
+The most interesting compound in this sample is miliain, which is the biomarker of broomcorn millet [@Heron2016; @Bossard2013; @Ganzarolli2018].
+The finding of miliacin also corresponds with the presence of very long-chain fatty acids, such as C~30:0~ and C~32:0~, indicative of plant waxes [@Post1996; @Copley2005].
 
@@
 ```
-
-One pottery sample (SYG-TN13-E22-2#2) has much higher lipid concentrations (about `r FA_max_sum` μg/g) than the rest of the samples.
-A higher yield of lipids might relate to the texture and type of ceramic used or better archaeological contexts for preservation.
-This sample also has a charred interior surface, indicative of a function associated with food processing.
-Long-chain fatty acids from C~20:0~ to C~32:0~ are present in that sample, with an even-over-odd carbon chain length predominance, but in lower amounts than in medium-chain components (Figure \@ref(fig:chromatograms-specific)).
-Although the existence of even-numbered long-chain fatty acids, such as C~30:0~ and C~32:0~, with a high amount of C~16:0~ relative to C~18:0~ might be related to plant waxes [@Post1996; @Copley2005], it is more likely to be associated with degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998].
-Overall, the GC-MS results demonstrate strong evidence of degraded animal fats.
 
@@
 The processing of millet might also be suggested by one Guijiabao potsherd dating to Neolithic period based on its isotopic ratios.
-However, a further confirmation would require to identify of any lipid residues associated with C~4~ plants, such as the biomarker for broomcorn millet, miliacin (olean-18-en-3β-ol methyl ether) [@Heron2016].
+However, a further confirmation would require to identify of any lipid residues associated with C~4~ plants, such as the biomarker for broomcorn millet, miliacin [@Heron2016].
 The non-ruminant fats we identified from a Guijiabao potsherd that might have been from C~3~ plants or freshwater aquatic products, based on similar isotopic values reported by other studies in East Asia [@Shoda2018].
@@
 sampling bias, poor preservation
+These studies and the ability/limitations of ORA should be mentioned.
+
+The domination of millet at Guijiabao corresponds to a C~4~-based diet based on isotopic analysis of four human bones [@Lin2022].
 

### 2024-01-21 analysis/paper/paper.Rmd
substantial edits

@@
 abstract: |
-    The spread of new crops or terrestrial animals across areas in southwestern China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region.
-It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and millets came even earlier.
-However, to what extent new crops were used by local peoples at that time remains unclear.
-The excavation of Guijiabao in the Yanyuan basin in southwestern China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops.
-To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis.
-The results show signals of terrestrial animals in most potsherds and a C~3~ plant source in one potshard, during the Neolithic period.
-Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming.
-This study also demonstrates an approach using organic residue analysis to explore the extent of the adoption of farming during the Neolithic period in southwestern China.
+
+     The emergence of mixed agriculture of millet and rice in Neolithic southwestern China has been a topic of extensive discussion.
+Despite the recovery of both millet and rice grains, the varying proportions observed across the area underscore the complexity of the process of agricutural dispersal.
+Further investigation is needed to understand how new crops were integrated into southwestern China, characterized by diverse environmental conditions and landscapes.
+This study contributes to this understanding by exploring subsistence strategies in southwestern Sichuan through a preliminary analysis of lipid residues in pottery from the Guijiabao site using organic geochemical analyses.
+Situated in the middle Hengduan mountain ranges, the Guijiabao site presents well-preserved stratigraphic records spanning from the Neolithic to the historical period, providing valuable insights into local subsistence practices in a highland environment.
+Results reveal the detection of ruminant adipose fats in the majority of potsherds, with a few showing plant inputs such as broomcorn millet.
+This suggests that animals played a substantial role in the local diet, alongside the consumption of millet.
+These culinary practices also indicate that hunting may have remained a significant subsistence activity in the high-altitude region of southwestern China, even after the introduction of crops.
 
 keywords: |
-  Southwestern China; Neolithic period; Lipid residue analysis; Stable isotope analysis; Ceramics; Culinary practices
+  Southwestern China; Neolithic period; Lipid residue analysis; Compound-specific carbon isotope analysis; Pottery; Culinary practices
 highlights: |
@@
 # Introduction
+
 <!-- intro to the topic and research aim-->
+The movement of Neolithic populations has long been discussed in archaeology to understand the processes of agricultural diffusion, regional variations, and their consequences [@Bellwood2023; @Davison2006].
+In southwest China, the emergence of farming practices was thought to be associated with the migrating population from central or northern China in the Neolithic period [@Deng2019].
+Crop remains found in prehistoric southwest China, including foxtail millet, broomcorn millet, and rice, can be traced back to as early as 5000 BP in southwest Sichuan and around 4600 BP in western Yunnan [@Jiang2016; @Deng2019; @YunnanMuseum1981; @Zhang2010; @Dal2022].
+Despite different views on the routes of agricultural dispersal [@Guedes2013; @Guedes2018; @Liu2022; @Huan2022], it is generally believed that mixed farming was the main agricultural practice in southwest China [@Deng2019].
+However, more studies are still needed to understand how migrant communities interacted with new environments, especially in an area with a high diversity of landscapes, such as high mountains, plateaus, basins, and terraces.
 
-The movement of Neolithic populations has long been discussed in archaeology to understand the processes of agricultural diffusion, regional variations, and their consequences [@bellwood2023; @Davison2006].
-In southwest China, the emergence of farming practices was thought to be associated with the migrating population from central or northern China in the Neolithic period [@Deng2019].
-Crop remains, including foxtail millet, broomcorn millet, and rice, found in prehistoric southwest China that can be traced back to as early as 4700 BP in Sichuan and around 4600 BP in western Yunnan [@Jiang2016; @Deng2019; @YunnanMuseum1981; @Zhang2010; @Dal2022].
-Despite different views on the routes of agricultural dispersal [@Guedes2013; @Guedes2018; @Liu2022], it is generally believed that mixed farming was the main economy in southwest China after the introduction of agriculture [@Deng2019].
-However, more studies still needed to understand how migrant communities interacted with new environments, especially in an area with high diversity of landscape, such as high mountain, plateau, basin, and terrace.
-
+<!--Guijaibao as a research area-->
 This study focuses mainly on archaeological sites located in the Hengduan Mountain ranges, a region in southwest China connecting the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south.
 As a critical passage for population movement, the Hengduan Mountain ranges and valleys facilitate not only material exchange [@Tong1986; @Hein2014; @Hein2016; @Ma2021], but also the spread of agriculture even further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017].
-Recent excavations at the Guijiabao site in Yanyuan Basin, the middle Hengduan Mountain ranges, is a key site with the earliest Neolithic human occupation in this region dating to 5000 B.P., continuing to the historical period [@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022].
-The complete stratigraphy at Guijiabao provides a unique opportunity to investigate questions relative to the subsistence since the Neolithic period in a basin surrounded by high mountains.
+The Guijiabao site in Yanyuan Basin, the middle Hengduan Mountain ranges, are a key site with the earliest Neolithic human occupation in this region dating to 5000 B.P., continuing to the historical period [@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022].
+The complete stratigraphy at Guijiabao provides a unique opportunity to investigate questions relative to subsistence since the Neolithic period in a basin surrounded by high mountains.
 For example, what is the role of agriculture in local environments?
 What other resources were obtained in addition to crops?
-In this study, we present our preliminary analyses on archaeological potsherds from the Guijiabao site and the Gaoshan site in southwest China using organic residue analysis of lipids.
+In this study, we present our preliminary analyses of archaeological potsherds from the Guijiabao site and the Gaoshan site in southwest China using organic residue analysis of lipids.
 
-Although the remains of animal bones and plants offer evidence for subsistence, their preservation heavily depend on environmental conditions.
+<!--justifying organic residue analysis-->
+Although the remains of animal bones and plants offer evidence of subsistence, their preservation heavily depends on environmental conditions.
 Geographical regions with seasonal heavy precipitation and acidic sediments, such as Southwest Sichuan and the middle Hunduan Mountain [@Li1989], tend to lead to poor preservation of archaeological remains.
-In such case, organic residue analysis is a powerful tool for studying ancient food practices by directly examining residues in pottery.
-Organic residues, specifically lipids, can be well preserved in the porous fabric of pottery for over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017; @Dudd1999; @Miller2020].
+In such cases, organic residue analysis is a powerful tool for studying ancient food practices by directly examining residues in pottery.
+Organic residues, specifically lipids, can be well preserved in the porous fabric of pottery for thousands of years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017; @Dudd1999; @Miller2020].
 The growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies [@Kwak2020], and culinary practices across a region over time [@Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
 
-Through identification of lipid biomarkers and compound-specific carbon isotope analysis based on geochemical techniques, archaeologists are able to distinguish food origins between plant oils, ruminant fats, diary products, and aquatic/marine resources [@Copley2005; @Craig2007; @Evershed1994].
-Recent studies further successfully identified different types of plants processed in archaeological vessels, such as broomcorn millet [@Bossard2013; @Heron2016; @Ganzarolli2018], starchy plants [@Shoda2018], and cereals [@Hammann2018; @Colonese2017].
-A growing case studies for ceramics in China present the robustness of organic residue analysis for understanding regional cooking practices [@Shoda2018; @Lyu2024; @Han2022; @Lyu2024], distinct subsistence strategies between low- and high altitude zones [@Zhang2022], and the impact of nomadic pastoralism on northwest China [@Sun2023].
-In this study, we attempt to investigate pottery use and culinary practices in the middle Hengduan mountains to understand the local subsistence by focusing on the ceramics from the Guijiabao and Gaoshan sites.
+<!--regional application-->
+By identifying lipid compositions and conducting compound-specific carbon isotope analysis, archaeologists are able to distinguish between plant oils, ruminant fats, dairy products, and aquatic/marine resources [@Copley2005; @Craig2007; @Evershed1994].
+Recent studies have successfully identified various plant types processed in archaeological vessels, including broomcorn millet [@Bossard2013; @Heron2016; @Ganzarolli2018], starchy plants [@Shoda2018], and cereals [@Hammann2018; @Colonese2017].
+The increasing use of ceramics in China highlights the robustness of organic residue analysis for comprehending regional cooking practices [@Shoda2018; @Lyu2024; @Han2022], discerning distinct subsistence strategies between low- and high-altitude zones [@Zhang2022], and exploring the impact of nomadic pastoralism on northwest China [@Sun2023].
+In this study, we aim to investigate pottery use and culinary practices in the middle Hengduan mountains, focusing on ceramics from the Guijiabao and Gaoshan sites to understand local subsistence.
 
@@
 The first survey at Guijiabao in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean and ground houses, post holes, ash pits, ditches, and burials, suggesting a settlement site [@Hao2022; @Guijiabao2017a].
-The site was occupied since the Neolithic period (5000-3700 BP), continued to the Bronze Age (3200-2700 BP) and the late North Song to early Ming dynasties (1200-500 BP), based on 19 radiocarbon dates [@Hao2022].
-The Neolithic period can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the artifact assemblages and ceramic styles
+The site has been occupied since the Neolithic period (5000-3700 BP), continued to the Bronze Age (3200-2700 BP) and the historical period (1200-500 BP) corresponding to the late North Song to early Ming dynasties, based on radiocarbon dates [@Hao2022; @Huan2022].
+The Neolithic period can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the artifact assemblages and ceramic styles.
 
 <!--the major finding at GJB-->
-Ceramics is the major finding at Guijiabao.
+Ceramics are the major finding at Guijiabao.
 The common vessel forms include jars, urns, bowls, and basins.
 The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with cord-marked or basket-impressed patterns [@Hao2022; @Guijiabao2017a].
-The evidence of chars on their surfaces suggest that they might be used for cooking [@Guijiabao2017a; @Hao2022].
-This research also attempts to confirm that these pottery vessels were used as cooking/storage wares.
+Ceramics with charred surfaces suggest that they might be used for cooking.
+This research also attempts to confirm that these pottery vessels were used as cooking or storage wares.
 Some ceramic decorations present northern styles, while others with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of close relationships or interaction between Guijiabao and neighboring areas, the northern and southern parts of the Hengduan Mountains [@Guijiabao2017a].
-The early excavation reveals the presence of cultivated crops at Guijiabao around 4700 BP, including foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*), earlier than previously thought in Yanyuan [@Guijiabao2016].
-In later excavations, rice grains (*Oryza sativa*) were also found throughout the Neolithic period, but in a smaller number than millet [@Hao2022; @Huan2022].
+The flotation analysis of soil samples from the 2015 excavation demonstrates the predominance of cultivated broomcorn millet (*Panicum miliaceum*) and foxtail millet (*Setaria italica*) at Guijiabao during the Neolithic period [@Guijiabao2016].
+Rice grains (*Oryza sativa*) were also identified in later excavations from the Neolithic period based on macro botanical evidence and phytolith analysis [@Hao2022; @Huan2022], though in a smaller number than millet.
 
@@
 
-Although mixed agriculture of millet and rice observed in southwest China, it seems there are variations in crop cultivation across the region.
-The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011].
-However, botanical evidence from most sites in the Chengdu plain reveal a predominance of rice, such as at the Gaoshan and Baodun sites [@JiangGaoshan2017], and the Baiyangcun site in Yunnan [@Jiang2016; @Henglanshan2014].
-Moreover, most research focuses on the agricultural practices, what other resources utilized by early farmer in new environments remains unclear.
-Here we seek to examine direct evidence from pottery residues to provide clues on culinary practices through case study in southwest China using geochemical techniques.
+While mixed agriculture of millet and rice was practiced in southwestern China, variations in their proportions across the region have been observed.
+Macro-botanical and phytolith analyses of the Guijiabao site indicate that millet was the primary crop, aligning with findings at the Yingpanshan site in Sichuan [@Zhao2011].
+Conversely, botanical evidence from numerous sites in the Chengdu plain suggests a prevalence of rice, as seen at the Gaoshan and Baodun sites [@JiangGaoshan2017], and the Baiyangcun site in Yunnan [@Jiang2016; @Henglanshan2014].
+Moreover, while most research focuses on agricultural practices, there is a lack of clarity regarding other resources utilized by early farmers in new environments.
+This study aims to address this gap by examining direct evidence from pottery residues, using geochemical techniques to provide insights into culinary practices through a case study in southwest China.
 
@@
 ## Sampling
-As a preliminary research on food residues preserved in ceramic vessels, we sampled pottery fragments from cultural layers at the residential area of Guijiabao (n = 5) to test the feasibility of organic geochemical approaches (Table \@ref(tab:sherds-table)).
-As a comparison, we sampled one fragment from Goashan.
-Those pottery fragments were identified as typical types with well-defined contexts, as indicated by radiocarbon dates ranging from the Neolithic period (ca.
+
+As a preliminary study to assess the feasibility of employing organic geochemical approaches, we collected five pottery fragments from cultural layers in the residential area of Guijiabao (sample code starting with SYG) (Table \@ref(tab:sherds-table)).
+For comparison, one fragment was sampled from Goashan (sample code starting with CDG).
+Those pottery fragments represent typical types with well-defined contexts, with their radiocarbon dates ranging from the Neolithic period (ca.
 4500 BP) to the historical period (ca.
@@
 The pottery samples are mostly base fragments and fine sand-tempered, with a few displaying impressed decorations.
-Through examining potsherds from early to late period at Guijiabao, we seek to explore the potential of organic geochemical approach across different time period.
+By examining potsherds from the early and late periods at Guijiabao, we aim to explore pottery use across time periods and the potential of utilizing an organic geochemical approach.
 
@@
 ## Lipid extraction
-The surface of each pottery sample was removed to prevent external contamination from environments and handling.
+
+The surface of each pottery sample was removed to prevent external contamination from the environment and handling.
 Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C) [@Mcclure2018].
-To obtain a higher yield, free fatty acids were recovered from bound lipids through saponification, conducted by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heated at 70 °C overnight [@Kaluzna2016; @Craig2004].
+To obtain a higher yield, free fatty acids were recovered from bound lipids through saponification, which was conducted by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating at 70 °C overnight [@Kaluzna2016; @Craig2004].
 Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL).
@@
 The FAME derivatives were extracted with hexane ($3\times10$ mL).
-About 5% aliquot was analyzed by gas chromatography with flame-ionization detection (GC-FID) for quantification, and the remaining sample was analyzed using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS) [@Craig2007; @Evershed1994].
+About a 5% aliquot was analyzed by gas chromatography with flame-ionization detection (GC-FID) for quantification, and the remaining sample was analyzed using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS) [@Craig2007; @Evershed1994].
 Analytical blanks were prepared and analyzed using the same procedure as the pottery samples to monitor potential contamination sources from solvents and reagents.
@@
 The oven temperature was set at 50 °C for 1 min, then raised at a rate of 10°C/min to 310 °C and held for 15 min.
-A authentic standard of miliacin was also ran to monitor the retention time.
+An authentic standard of miliacin was also run to monitor the retention time.
 
-The GC-FID analysis of FAMEs was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode (Agilent technologies, USA).
+The GC-FID analysis of FAMEs was carried out using an Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode (Agilent technologies, USA).
 The carrier gas was helium flowing at a rate of 1.6 ml/min.
@@
                          str_extract(sample, pattern = "_00[0-9]"))) %>%
-  mutate(name = case_when(
-    sample %in% c("210729_001") ~ "CDG-062",
+  mutate(name = case_when(sample %in% c("210729_001") ~ "CDG-062",
                           sample %in% c("210729_002") ~ "SYG-3",
@@
 
-```{r highlight-compounds, cache=TRUE}
-# highlight target compounds
-highlight_common <-
-  meth_all %>%
-  mutate(retention = round(Path, 1)) %>%
-  group_by(sample, retention) %>%
-  arrange(desc(File)) %>%
-  distinct(retention, .keep_all = TRUE) %>%
-  mutate(FA = case_when(
-    retention == 21.6 ~ "C[14:0]",
-    retention == 23.6 ~ "C[15:0]",
-    retention == 25.5 ~ "C[16:0]",
-    retention == 27.3 ~ "C[17:0]",
-    retention == 28.4 ~ "C[18:1]",
-    retention == 29.1 ~ "C[18:0]",
-    retention == 32.3 ~ "C[20:0]",
-    retention == 35.4 ~ "C[22:0]")) %>%
-  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062")) %>%
-  filter(File > 100000)
-
-highlight_IS <-
-  meth_all %>%
-  mutate(retention = round(Path, 1)) %>%
-  group_by(sample, retention) %>%
-  arrange(desc(File)) %>%
-  distinct(retention, .keep_all = TRUE) %>%
-  mutate(FA = case_when(
-    retention == 33.9 ~ "IS",
-    retention == 41.3 ~ "IS")) %>%
-  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062"))
-
-# highlight the compounds in the potsherd with high concentration
-highlight_SYG2_2 <-
-  meth_all %>%
-  mutate(retention = round(Path, 1)) %>%
-  group_by(sample, retention) %>%
-  arrange(desc(File)) %>%
-  distinct(retention, .keep_all = TRUE) %>%
-  mutate(FA = case_when(
-    retention == 21.6 ~ "C[14:0]",
-    retention == 25.5 ~ "C[16:0]",
-    retention == 27.3 ~ "C[17:0]",
-    retention == 28.4 ~ "C[18:1]",
-    retention == 29.1 ~ "C[18:0]",
-    retention == 31.0 ~ "C[19:0]",
-    retention == 32.4 ~ "C[20:0]",
-    retention == 35.4 ~ "C[22:0]",
-    retention == 36.8 ~ "C[23:0]",
-    retention == 38.2 ~ "C[24:0]",
-    retention == 40.9 ~ "C[26:0]",
-    retention == 33.9 ~ "IS",
-    retention == 41.3 ~ "IS",
-    retention == 45.6 ~ "C[30:0]",
-    retention == 47.8 ~ "C[32:0]")) %>%
-  filter(name == "SYG-TN13-E22-2#2")
-```
-
-```{r chromatogram-four, cache= TRUE}
-# plot the first four
-chroma_four <-
-  meth_all %>%
-  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", 
-                     "SYG-TN13-E23-3", "CDG-062")) %>%
-  mutate(File = ifelse(File > 1000000, 1000000, File)) %>%
-  ggplot(aes(Path, File)) +
-  geom_line(size = 0.3) +
-  ggrepel::geom_text_repel(data = highlight_common, #ggrepel::geom_text_repel
-            aes(label = FA, angle = 0),
-            size = 3,
-            nudge_y = 115000,
-            show.legend = FALSE,
-            parse = TRUE) +
-  geom_text(data = highlight_IS,
-            aes(label = FA),
-            size = 3,
-            nudge_y = 100000,
-            show.legend = FALSE) +
-  geom_segment(aes(x= 41.1, xend= 41.5, y= 1000000, yend= 1000000), size =0.3) +
-  geom_segment(aes(x= 41.1, xend= 41.5, y= 1030000, yend= 1030000), size =0.3) +
-  annotate("text", x = 41.3, y = 1100000, label = "IS", size =3) +
-  facet_wrap(~name, ncol = 1) +
-  scale_y_continuous(labels = scales::comma_format(),
-                     limits = c(0, 1160000),
-                     breaks = seq(0, 1100000, 200000)) +
-  scale_x_continuous(limits = c(20, 45),
-                     breaks = seq(20, 45, 5),
-                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
-  labs(x = "retention time", y = "relative intensity") +
-  theme_minimal() +
-  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
-        panel.background = element_blank(), axis.line = element_line(colour = "black"))
-
-ggsave(plot = chroma_four,
-       here::here("analysis","figures", "chromatograms_four.png"),
-       bg = "white",
-       width = 6,
-       height = 8,
-       dpi = 300,
-       units = "in")
-```
-
-The total lipid concentrations in our potsherds ranges from 7 to 48.6 μg/g, above interpretable concentrations (5 μg/g) [@Evershed2008experimental] (Table \@ref(tab:FA-composition-table)).
-This suggests that those ceramics should be associated with food processing.
-The main compounds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid) and C~18:0~ (stearic acid), commonly derived from animal fat and plant oil (Figure \@ref(fig:chromatograms-plots-four)).
-Unsaturated fatty acid C~18:1~ was also found in all samples, but we are unable to distinguish the C~18:1~ types based on current method.
-The domination of C~16:0~ and C~18:0~ suggests that the lipids in our samples were greatly degraded.
-However, one pottery sample (SYG-TN13-E22-2#2) has relatively high lipid yield (`r FA_max_sum` μg/g), with the presence of long-chain fatty acids (C~24:0~ to C~32:0~) (Figure \@ref(fig:chromatograms-specific)).
-The most interesting compound in this sample is miliain, which is the biomarker of broomcorn millet [@Heron2016; @Bossard2013; @Ganzarolli2018].
-The finding of miliacin also corresponds with the presence of very long-chain fatty acids, such as C~30:0~ and C~32:0~, indicative of plant waxes [@Post1996; @Copley2005].
-
-```{r FA-composition-table, cache= TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/210824_FAMES_pot_LW_publication.xlsx"))}
+```{r FA-composition-tidy, cache= TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/210824_FAMES_pot_LW_publication.xlsx"))}
 quant_FA <-
@@
   pivot_wider(names_from = sample, 
-              values_from = weight)
+              values_from = weight) %>% 
+  filter(!FA %in%c("C34:0", "C21:0"))
 
-FA_max_sum <-
-  sum(quant_FA_number$"SYG-TN13-E22-2#2")
+FA_sum_min <- sum(quant_FA_number$"SYG-TN13-E22-2#1", na.rm = T)
+FA_sum_max <- sum(quant_FA_number$"SYG-TN13-E22-2#2")
+```
 
+The total lipid concentrations of potsherds at Guijiabao and Gaoshan range from `r FA_sum_min` to `r FA_sum_max` μg/g (Table \@ref(tab:FA-composition-table)).
+Lipid yields from all potsherds are above interpretable concentrations (5 μg/g) [@Evershed2008experimental], suggesting those ceramics should be associated with food processing.
+The main compounds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid) and C~18:0~ (stearic acid), commonly derived from animal fats and plant oils (Figure \@ref(fig:chromatograms-plots-four)).
+Unsaturated fatty acid, C~18:1~, was detected in all samples and dominant in three, but we are unable to distinguish the C~18:1~ types based on current methods.
+The most interesting finding is that milliacin, the compound specific to broomcorn millet [@Heron2016; @Bossard2013; @Ganzarolli2018], was identified in one Neolithic sample (SYG-TN13-E22-2#2) from Guijiabao (Figure \@ref(fig:chromatograms-specific-std)).
+This sample has a wider range of fatty acids from C~12:0~ to C~30:0~ (`r FA_sum_max` μg/g), where long-chain fatty acids could originate from plant sources [@Post1996; @Copley2005].
+Overall, the dominance of C~16:0~ and C~18:0~ in most samples suggests that the lipids extracted from our samples were greatly degraded.
+
+```{r FA-composition-table, cache= TRUE}
 quant_FA_tidy <-
@@
 
-```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from the Guijiabao and Gaoshan sites.
+```{r highlight-compounds, cache=TRUE}
+# round all peaks in order
+all_peaks <-
+  meth_all %>%
+  mutate(retention = round(Path, 1)) %>%
+  group_by(sample, retention) %>%
+  arrange(desc(File)) %>%
+  distinct(retention, .keep_all = TRUE)
+
+# highlight target compounds
+highlight_common <-
+  all_peaks %>% 
+  mutate(FA = case_when(retention == 21.6 ~ "C[14:0]",
+                        retention == 23.6 ~ "C[15:0]",
+                        retention == 25.5 ~ "C[16:0]",
+                        retention == 27.3 ~ "C[17:0]",
+                        retention == 28.4 ~ "C[18:1]",
+                        retention == 29.1 ~ "C[18:0]",
+                        retention == 32.3 ~ "C[20:0]",
+                        retention == 35.4 ~ "C[22:0]")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062")) %>%
+  filter(File > 100000)
+
+highlight_IS <-
+  all_peaks %>% 
+  mutate(FA = case_when(retention == 33.9 ~ "IS",
+                        retention == 41.3 ~ "IS")) %>%
+  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062"))
+```
+
+```{r chromatogram-four, cache= TRUE}
+# to insert ticks in plot
+insert_minor <- function(major_labs, n_minor) {labs <-
+  c( sapply( major_labs, function(x) c(x, rep("", 4) ) ) )
+labs[1:(length(labs)-n_minor)]}
+
+# plot the first group
+chroma_four <-
+  meth_all %>%
+  filter(name %in% c("SYG-3","SYG-TN13-E22-2#1","SYG-TN13-E22-1","SYG-TN13-E23-3","CDG-062")) %>%
+  mutate(File = ifelse(File > 1000000, 1000000, File)) %>%
+  ggplot(aes(Path, File)) +
+  geom_line(size = 0.3) +
+  ggrepel::geom_text_repel(data = highlight_common, #ggrepel::geom_text_repel
+            aes(label = FA, angle = 0),
+            size = 3,
+            nudge_y = 115000,
+            show.legend = FALSE,
+            parse = TRUE) +
+  geom_text(data = highlight_IS,
+            aes(label = FA),
+            size = 3,
+            nudge_y = 100000,
+            show.legend = FALSE) +
+  geom_segment(aes(x= 41.1, xend= 41.5, y= 1000000, yend= 1000000), size =0.3) +
+  geom_segment(aes(x= 41.1, xend= 41.5, y= 1030000, yend= 1030000), size =0.3) +
+  annotate("text", x = 41.3, y = 1100000, label = "IS", size =3) +
+  facet_wrap(~name, ncol = 1) +
+  scale_y_continuous(limits = c(0, 1160000),
+                     breaks = seq(0, 1100000, 200000),
+                     expand = c(0, 0)) +
+  scale_x_continuous(limits = c(18, 47),
+                     breaks = seq(15, 50, 1),
+                     labels = insert_minor(seq(15, 50, 5), 4),
+                     expand = c(0, 0)) +
+  labs(x = "Retention time (mins)", y = expression("Relative abundance " %->% "")) +
+  theme_classic() +
+  theme(axis.line.y = element_blank(), strip.background = element_blank(), # remove title box
+        axis.ticks.y=element_blank(), axis.text.y = element_blank())
+
+ggsave(plot = chroma_four,
+       here::here("analysis","figures", "chromatograms_four.png"),
+       bg = "white",
+       width = 6,
+       height = 8,
+       dpi = 300,
+       units = "in")
+```
+
+```{r chromatograms-plots-four, echo=FALSE, fig.cap= "Partial total ion cuurent (TIC) chromatograms of pottery samples from Guijiabao and Gaoshan.
 'Cn:x' refers to fatty acid with n carbon atoms and x double bonds.
 'IS' indicates internal standard used for quantification, including FA C~21:0~ and 5-α Cholestane"}
+
 knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
@@
 ```{r chromatogram-specific, cache= TRUE}
-# plot the specific one
-chroma_specific <-
-  meth_all %>%
-  filter(name %in% c("SYG-TN13-E22-2#2")) %>% 
+## miliacin
+miliacin <- read_csv(here::here("analysis/data/raw_data/LY_miliacin_std_chrom.CSV"))
+ixd_3 <- which(str_detect(miliacin$Path, "D"))
+meths_3 <- split(miliacin, cumsum(1:nrow(miliacin) %in% ixd))
+meth_data_3 <- meths_3[c(FALSE, TRUE)]
+meth_name_3 <- meths_3[c(TRUE, FALSE)]
+names(meth_data_3) <- map(meth_data_3, ~.x$Path[1])
+names(meth_name_3) <-map(meth_name_3, ~.x$Sample[1])
+meths_3 <- map_df(meth_data_3, ~.x[-1, ] %>%
+                    mutate_all(as.numeric), .id = 'sample')
+
+highlight_m <-
+  meths_3  %>%
+  mutate(retention = round(Path, 1)) %>%
+  group_by(sample, retention) %>%
+  arrange(desc(File)) %>%
+  distinct(retention, .keep_all = TRUE) %>%
+  mutate(std = case_when(retention == 29.4 ~ "M", retention == 30.5 ~ "IS"))
+
+std_miliacin <-
+  meths_3%>%
   ggplot(aes(Path, File)) +
   geom_line(size = 0.3) +
-  ggrepel::geom_text_repel(data = highlight_SYG2_2, #ggrepel::geom_text_repel
+  ggrepel::geom_text_repel(data = highlight_m, #ggrepel::geom_text_repel
+                           aes(label = std, angle = 0), size = 3.5, nudge_y = 100000,
+                           show.legend = FALSE, parse = TRUE) +
+  scale_y_continuous(limits = c(0, 4600000),
+                     breaks = seq(0, 4600000, 1000000),
+                     expand = c(0,0)) +
+  scale_x_continuous(limits = c(13.5, 32.5),
+                     labels = insert_minor(seq(10, 35, 5), 4),
+                     breaks = seq(10, 35, 1),
+                     expand = c(0, 0)) +
+  labs(x = "Retention time (mins)", y = expression("Relative abundance " %->% "")) +
+  theme_classic() +
+  theme(plot.title = element_text(hjust = 0.5), axis.line.y = element_blank(),
+        axis.ticks.y=element_blank(), axis.text.y = element_blank())
+
+# load data produced by program 2
+meth_SYG22 <- read_csv(here::here("analysis/data/raw_data/LY_pot_SYG_E22_22_chrom.CSV"))
+
+# tidy data
+ixd_2 <- which(str_detect(meth_SYG22$Path, "D"))
+meths_2 <- split(meth_SYG22, cumsum(1:nrow(meth_SYG22) %in% ixd))
+meth_data_2 <- meths_2[c(FALSE, TRUE)]
+meth_name_2 <- meths_2[c(TRUE, FALSE)]
+names(meth_data_2) <- map(meth_data_2, ~.x$Path[1])
+names(meth_name_2) <-map(meth_name_2, ~.x$Sample[1])
+meths_2 <- map_df(meth_data_2, ~.x[-1, ] %>%
+                  mutate_all(as.numeric), .id = 'sample')
+
+# peaks in SYG-E22-2#2
+peaks_SYGE22 <-
+  meths_2 %>%
+  mutate(retention = round(Path, 1)) %>%
+  group_by(sample, retention) %>%
+  arrange(desc(File)) %>%
+  distinct(retention, .keep_all = TRUE)
+
+highlight_SYG_FA <-
+  peaks_SYGE22 %>%
+  mutate(FA = case_when(retention == 17.7 ~ "C[15:0]",
+                        retention == 18.7 ~ "C[16:0]",
+                        retention == 19.7 ~ "C[17:0]",
+                        retention == 20.4 ~ "C[18:1]",
+                        retention == 20.6 ~ "C[18:0]",
+                        retention == 21.5 ~ "C[19:0]",
+                        retention == 22.4 ~ "C[20:0]",
+                        retention == 23.3 ~ "IS",
+                        retention == 24.0 ~ "C[22:0]",
+                        retention == 24.8 ~ "C[23:0]",
+                        retention == 26.3 ~ "C[25:0]",
+                        retention == 27.0 ~ "C[26:0]",
+                        retention == 28.3 ~ "C[28:0]",
+                        retention == 30.5~ "IS"))
+
+highlight_std_small <-
+  peaks_SYGE22 %>%
+  mutate(FA = case_when(retention == 14.2 ~ "C[12:0]",
+                        retention == 15.4 ~ "C[13:0]",
+                        retention == 16.6 ~ "C[14:0]",
+                        retention == 25.6 ~ "C[24:0]",
+                        retention == 29.4 ~ "M"))
+
+## plot the one with high yield (GC program 2)
+chroma_specific_2 <-
+  meths_2 %>%
+  ggplot(aes(Path, File)) +
+  geom_line(size = 0.3) +
+  ggrepel::geom_text_repel(data = highlight_SYG_FA , #ggrepel::geom_text_repel
+                           aes(label = FA, angle = 0), size = 3.5, nudge_y = 100000,
+                           show.legend = FALSE, parse = TRUE) +
+  ggrepel::geom_text_repel(data = highlight_std_small, #ggrepel::geom_text_repel
                            aes(label = FA),
                            size = 3.5,
-            nudge_y = 250000,
+                           segment.alpha = .25,
+                           nudge_y = 320000, #min.segment.length = 300000,
                            show.legend = FALSE,
-            parse = TRUE) +
+                           parse = TRUE) + #
   scale_y_continuous(labels = scales::comma_format(),
-                     limits = c(0, 7500000),
-                     breaks = seq(0, 7500000, 1000000)) +
-  scale_x_continuous(limits = c(20, 50),
-                     breaks = seq(20, 50, 5),
-                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
-  labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
-  theme_minimal() +
-  theme(plot.title = element_text(hjust = 0.5),
-        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
-        panel.background = element_blank(), axis.line = element_line(colour = "black"))
+                     limits = c(0, 4600000),
+                     breaks = seq(0, 4600000, 1000000),
+                     expand = c(0, 0)) +
+  scale_x_continuous(limits = c(13.5, 32.5),
+                     labels = insert_minor(seq(10, 35, 5), 4),
+                     breaks = seq(10, 35, 1),
+                     expand = c(0, 0)) +
+  labs(x = "Retention time (mins)", y = expression("Relative abundance " %->% "")) +
+  theme_classic() +
+  theme(plot.title = element_text(hjust = 0.5), axis.line.y = element_blank(),
+        axis.ticks.y=element_blank(), axis.text.y = element_blank())
 
-ggsave(plot = chroma_specific,
-       here::here("analysis","figures", "chromatograms_specific.png"),
-       bg = "white",
-       width = 8,
-       height = 5,
-       dpi = 300,
-       units = "in")
+library(cowplot)
+SYG_with_miliacin <-
+  plot_grid(chroma_specific_2, std_miliacin ,
+            labels = c('A', 'B'), ncol =1, align = 'v', label_size = 12)
+
+ggsave(plot = SYG_with_miliacin,
+       here::here("analysis","figures", "chromatograms_specific_std.png"),
+       bg = "white", width = 6, height = 8, dpi = 300, units = "in")
 ```
 
-```{r chromatograms-specific, echo=FALSE, fig.cap= "The MS chromatogram of the potsherd from the Guijiabao site shows the presence of long-chain fatty acids.
-'Cn:x' refers to fatty acid with n carbon atoms and x double bonds.
-'IS' indicates internal standards used for quantification, including FA C~21:0~ and 5-α Cholestane"}
-knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
+```{r chromatograms-specific-std, echo=FALSE, fig.cap= "A: Partial total ion current (TIC) chromatogram of a Guijiabao sample, SYG-TN13-E22-2#2, showing the presence of long-chain fatty acids and miliacin.
+Cn:x refers to fatty acid with n carbon atoms and x double bonds; IS indicates internal standards (FA C~21:0~ and n-tetratriacontane); M refers to miliacin.
+B: Partial TIC chromatogram of miliacin (olean-18-en-3β-ol ME) with n-tetratriacontane as internal standard"}
+knitr::include_graphics(here::here("analysis/figures/chromatograms_specific_std.png"))
 ```
@@
   slice(2:8) %>%
-  dplyr::mutate(...1 = ifelse(is.na(...1), paste0("C13"), ...1))
+  dplyr::mutate(...1 = ifelse(is.na(...1), paste0("C13"), ...1)) %>% 
+  janitor::row_to_names(1) %>% 
+  dplyr::mutate(across(starts_with("13C"), as.numeric))
 
-names(delta_meth_cor) <- delta_meth_cor[1,]
-
-delta_meth_cor <-
+delta_meth_data <-
   delta_meth_cor %>%
-  slice(2:7) %>%
-  dplyr::mutate(across(starts_with("13C"), as.numeric)) %>%
   dplyr::mutate(delta = `13C C18:0`-`13C C16:0`) %>%
-  mutate(sample = case_when(
-    C13 %in% c("Pot 1") ~ "CDG-062",
+  mutate(name = case_when(C13 %in% c("Pot 1") ~ "CDG-062",
                           C13 %in% c("Pot 2") ~ "SYG-3",
@@
                           C13 %in% c("Pot 6") ~ "SYG-TN13-E23-3")) %>%
-  mutate(sample = factor(sample, levels = c("SYG-TN13-E22-1",
+  mutate(name = factor(name, levels = c("SYG-TN13-E22-1",
                                         "SYG-TN13-E22-2#1",
@@
                                         "CDG-062"))) %>%
-  mutate(period = ifelse(sample %in% c("SYG-TN13-E22-1", "SYG-3" ),
-                         "Historical", "Neolithic"))
+  mutate(period = ifelse(name %in% c("SYG-TN13-E22-1", "SYG-3" ), "Historical", "Neolithic"))
 
-min_carbon_isotope <- round(min(delta_meth_cor$`13C C18:0`), 2)
-max_carbon_isotope <- round(max(delta_meth_cor$`13C C16:0`), 2)
+rg_C16_isotope <- round(c(min(delta_meth_data$`13C C16:0`), max(delta_meth_data$`13C C16:0`)), 2)
+rg_C18_isotope <- round(c(min(delta_meth_data$`13C C18:0`), max(delta_meth_data$`13C C18:0`)), 2)
 
-delta_min <- round(min(delta_meth_cor$delta), 2)
-delta_max <- round(max(delta_meth_cor$delta), 2)
+delta_range <- round(c(min(delta_meth_data$delta), max(delta_meth_data$delta)), 2)
 
-E23_C16 <- round(delta_meth_cor$`13C C16:0`[6], 2)
-E23_C18 <- round(delta_meth_cor$`13C C18:0`[6], 2)
+SYG3_isotope <- round(c(delta_meth_data$`13C C16:0`[2], delta_meth_data$`13C C18:0`[2]), 2)
+E22_1_isotope <- round(c(delta_meth_data$`13C C16:0`[4], delta_meth_data$`13C C18:0`[4]), 2)
+E22_22_isotope <- round(c(delta_meth_data$`13C C16:0`[5], delta_meth_data$`13C C18:0`[5],
+delta_meth_data$delta[5]), 2)
+E22_21_isotope <- round(c(delta_meth_data$`13C C16:0`[3], delta_meth_data$`13C C18:0`[3], delta_meth_data$delta[3]), 2)
+E23_isotope <- round(c(delta_meth_data$`13C C16:0`[6], delta_meth_data$`13C C18:0`[6], delta_meth_data$delta[6]), 2)
 
-SYG3_C16 <- round(delta_meth_cor$`13C C16:0`[2], 2)
-SYG3_C18 <- round(delta_meth_cor$`13C C18:0`[2], 2)
-
-E22_1_C16 <- round(delta_meth_cor$`13C C16:0`[4], 2)
-E22_1_C18 <- round(delta_meth_cor$`13C C18:0`[4], 2)
-
-E22_22_C16 <- round(delta_meth_cor$`13C C16:0`[5], 2)
-E22_22_C18 <- round(delta_meth_cor$`13C C18:0`[5], 2)
+library(png)
+tem3 <- readPNG(here::here("analysis","figures", "isotope_range.png"), native = FALSE, info = FALSE)
 
@@
 isotope_C16_C18 <-
-  ggplot(delta_meth_cor, #delta_meth_cor/delta_blank_cor
-       aes(`13C C16:0`,`13C C18:0`)) +
-  geom_point(size = 2, alpha = 0.9, aes(shape = sample, color = period)) +
+  delta_meth_data %>% #delta_meth_cor/delta_blank_cor
+  ggplot(aes(`13C C16:0`,`13C C18:0`)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = name, color = period)) +
   #ggrepel::geom_text_repel(aes(label = C13)) +
-  theme_minimal() +
   labs(x = expression(paste(delta*{}^13* C[16:0], "\u2030")),
        y = expression(paste(delta*{}^13* C[18:0], "\u2030"))) +
-  scale_x_continuous(limits = c(-33, -20),
-                     breaks = seq(-33, -20, 2)) +
-  scale_y_continuous(limits = c(-33, -20),
-                     breaks = seq(-33, -20, 2)) +
-  #coord_fixed(ratio = 1) +
+  scale_x_continuous(limits = c(-33, -20), breaks = seq(-33, -20, 2)) +
+  scale_y_continuous(limits = c(-33, -20), breaks = seq(-33, -20, 2)) +#coord_fixed(ratio = 1)
   geom_abline(intercept = -0.7, linetype = "dashed") + #Suryanarayan et al.2021
@@
   annotate("text", x = -21.5, y = -24.5, angle = 45, vjust = 1.5,
-           label = bquote(Delta*{}^13*"C = -3.1 \u2030"))
+           label = bquote(Delta*{}^13*"C = -3.1 \u2030")) +
+  annotation_raster(tem3, ymin = -38, ymax= -17.9, xmin = -36, xmax = -18.3) +
+  theme_minimal()
 
 big_delta_C16 <-
-  ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
-       aes(`13C C16:0`, delta)) +
-  geom_point(size = 2, alpha = 0.9, aes(shape = sample, color = period)) +
-  theme_minimal() +
+  delta_meth_data %>%
+  ggplot(aes(`13C C16:0`, delta)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = name, color = period)) +
   labs(x = expression(paste(delta*{}^13* C[16:0], "\u2030")),
        y = expression(paste(delta*{}^13* C[18:0], "\u2030"))) +
-  scale_x_continuous(limits = c(-33, -20),
-                     breaks = seq(-33, -20, 2)) +
-  ylim(-2.5, 1.5) +
+  scale_x_continuous(limits = c(-33, -20), breaks = seq(-33, -20, 2)) +
+  ylim(-3, 1.5) +
   geom_hline(yintercept = -0.7, linetype = "dashed") +
@@
            label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
-  annotate("text", x = -32, y = 1.3,
-           label = "C[3]-diet", parse=TRUE) +
+  annotate("text", x = c(-32, -21), y = c(1.3, 1.3),
+           label = c("C[3]~diet", "C[4]~diet"), parse= TRUE) +
   annotate("segment", x = -29, y = 1.3, xend = -24, yend = 1.3,
            arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
-  annotate("text", x = -21, y = 1.3,
-           label = "C[4]-diet", parse=TRUE) +
-  annotate("text", x = -22, y = -1.1,
-           label = "Ruminant") +
-  annotate("text", x = -21.5, y = -0.3,
-           label = "Non-ruminant")
+  annotate("text", x = c(-22, -21.5), y = c(-1.1, -0.3),
+           label = c("Ruminant", "Non-ruminant")) +
+  theme_minimal()
 
@@
 combined_isotopic_plots <-
-  plot_grid(isotope_C16_C18,
-          big_delta_C16,
-          labels = c('A', 'B'),
-          ncol =1,
-          align = 'v',
-          label_size = 12)
+  plot_grid(isotope_C16_C18, big_delta_C16,
+          labels = c('A', 'B'), ncol =1, align = 'v', label_size = 12)
 
 ggsave(here::here("analysis","figures","isotopic_plots.png"),
-       bg = "white",
-       width = 6,
-       height = 8,
-       dpi = 300,
-       units = "in")
+       bg = "white", width = 6, height = 8, dpi = 300, units = "in")
 ```
 
-To identify the sources of degraded animal fats, we conducted isotope analysis to obtain stable carbon isotope compositions, δ^13^C values of C~16:0~ and C~18:0~, using GC-C-IRMS.
-The δ^13^C values of C~16:0~ and C~18:0~ have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on the different ways dietary carbon sources are utilized by distinct fat types [@Dunne2012; @Copley2003; @Evershed2008].
-Considering the bias might be introduced from differential δ^13^C values across geographical regions, we also used the Δ^13^C (δ^13^C~18:0~ - δ^13^C~16:0~) values to determine the sources of degraded fats [@Evershed2008; @Salque2013].
-The carbon isotope values for C~16:0~ and C~18:0~ from our samples have a distribution between `r min_carbon_isotope` ‰ and `r max_carbon_isotope` ‰, and their Δ^13^C values range from `r delta_min` ‰ to `r delta_max` ‰.
-The δ^13^C values were plotted based on the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].
-Although the referenced isotopic values should be ideally adjusted in terms of local environments, the known reference ranges especially for East Asia could still provide useful information about food categories with cautious comparison.
+To identify the sources of degraded lipids, we conducted compound-specific carbon isotope analysis of C~16:0~ and C~18:0~ using GC-C-IRMS.
+This technique has proven useful to distinguish fat origin between ruminant, non-ruminant, ruminant dairy products, C~3~ plants, C~4~ plants, and marine resources based on their stable carbon isotope values (hereafter δ13C ) [@Dunne2012; @Evershed2008].
+For example, the δ13C values from marine sources or C~4~ plants are more enriched than those from ruminant adipose fat and C~3~ plants due to different metabolic or photosynthetic pathways [@Copley2003; @Meier2002; @Cramp2014].
+The values of δ^13^C~16:0~ and δ^13^C~18:0~ were plotted based on global modern references for the assessment of their origin.
+To minimize potential bias introduced from differential δ^13^C values across geographical regions, the values of δ^13^C~16:0~ and Δ^13^C (δ^13^C~18:0~ - δ^13^C~16:0~) were also plotted for identification of animal fats [@Copley2003; @Salque2013].
+Although it is hard to identify the exact food category in this case due to the limited accessibility to local animal bones or fats as reference samples, references from neighboring areas with cautious comparison still allow meaningful interpretation.
 
-The results show that the majority of the Neolithic potsherd samples from Guijiabao and Gaoshan contain saturated fatty acids derived from ruminant adipose fats, with one Guijiabao potsherd sample, SYG-TN13-E23-3, indicating a non-ruminant adipose fat origin according to the Δ^13^C value 0.06 ‰ (Figure \@ref(fig:isotopic-plots)).
-This potsherd (SYG-TN13-E23-3) has the mean δ^13^C values `r E23_C16`‰ and `r E23_C18`‰ for the C~16:0~ and C~18:0~ fatty acids respectively, consistent with the isotopic value ranges of C~3~ plants, terrestrial or freshwater fish (ca.
-−26.5‰ to -33‰ for both compounds) reported for Neolithic potsherds in southern China [@Shoda2018], and freshwater aquatic products (-27.3‰ to 31.6‰ for the C~16:0~ and -26.4‰ to -29.9 for the C~18:0~) or wild boar (ca.
--27‰ to -29.5‰ for both compounds) reported for the Jōmon pottery in western Japan [@Lucquin2016].
-However, we are unable to distinguish between the C~3~ plants, terrestrial and freshwater fish by this isotopic approach alone.
-A further detection of biomarkers for fish such as isoprenoid fatty acids, or for starchy plants and millet will be needed for an better understanding [@Shoda2018; @Lucquin2016; @Heron2016].
+The isotope data shows that δ^13^C~16:0~ values range from `r rg_C16_isotope[1]` ‰ to `r rg_C16_isotope[2]` ‰, δ^13^C~18:0~ values range from `r rg_C18_isotope[1]` ‰ to `r rg_C18_isotope[2]`, and Δ^13^C values are between `r delta_range[2]` ‰ and `r delta_range[1]` ‰.
+A half of the samples (n= 3) at Guijiabao and Gaushan fall within the global reference range for ruminant adipose fat, where one sample from Guijiabao (SYG-TN13-E22-2#2) has a more enriched δ^13^C~16:0~ (`r E22_22_isotope[1]`) than others.
+The enriched δ^13^C suggests that the diet of ruminants contains both C~4~ and C~3~ plants (Figure \@ref(fig:isotopic-plots): B), while the depleted δ^13^C is related to a more C~3~ plant diet.
+While marine resources could potentially contribute to enriched δ^13^C values, the inland location of Guijiabao, situated approximately 1000 km away from the current South China Sea coast, restricts this possibility.
 
-For the pottery fragments (SYG-TN13-E22-1 and SYG-3) dating to the historical period, we found that the mean δ^13^C values of the extracts were plotted on the reference line between ruminant adipose and non-ruminant fats.
-Specifically, SYG-TN13-E22-1 and SYG-3 have the mean δ^13^C values `r E22_1_C16` and `r SYG3_C16` for the C~16:0~ fatty acids, and the mean δ^13^C values `r E22_1_C18` and `r SYG3_C18` for the C~18:0~ fatty acids.
-It is unsure to know what the origin of these values are, but it is possible that there is a mixing of both non-ruminant sources and ruminant fat [@Copley2003; @Dudd1998milk; @Suryanarayan2021].
-It should be noted that the δ13C values could be influenced by multiple different products, such as ruminant products and plant products in the same cooking event, or over different cooking events with different products.
-
-There is only one Neolithic potsherd (SYG-TN13-E22-2#2) with a relatively enriched δ^13^C value (`r E22_22_C16` for the C~16:0~ and `r E22_22_C18` for the C~18:0~) compared with other samples.
-It is widely known that the δ^13^C values of fatty acids of marine source or C~4~ plants are more enriched than those of ruminant fats [@Lucquin2016].
-Given that Guijiabao is situated inland with finding of millet grains, the enriched δ^13^C values might indicate the processing of C~4~ plants instead of marine resources, or the mixture of C~4~ plants and ruminant adipose products due to a slightly depleted δ^13^C value for the C~18:0~ fatty acid.
-However, in general, most samples have depleted δ^13^C values, which consistent with the δ^13^C values of C~3~ plants or ruminant fats, suggesting the processing of C~3~ plants or C~3~ plants were cooked with ruminant adipose meat.
-We also could not rule out the possibility that these depleted δ13C values came from mainly ruminant animals fed on C~3~ plants.
+One sample (SYG-TN13-E23-3) with Δ^13^C of `r E23_isotope[3]` falls within the range for non-ruminant fats (Figure \@ref(fig:isotopic-plots): B).
+Its relatively depleted δ^13^C~ values (δ^13^C~16:0~ of `r E23_isotope[1]` ‰ and δ^13^C~18:0~ of `r `E23_isotope[2]` ‰) reflect input of C~3~ plants, terrestrial animals, or freshwater resources [@Shoda2018; @Lucquin2016].
+The two samples (SYG-TN13-E22-1 and SYG-3) dating to the historical period clustered within a narrow range of δ^13^C~16:0~ (`r SYG3_isotope[1]` ‰ to `r E22_1_isotope[1]`) and δ^13^C~18:0~ (`r SYG3_isotope[2]` ‰ to `r E22_1_isotope[2]`) near the reference line, showing a mixing source of ruminant and non-ruminant fat [@Copley2003].
+A mixture of lipids from different sources is possible because δ13C values could be influenced by multiple distinct products, such as ruminant fats and plants, in the same cooking event or separate cooking events over time [@Evershed2002].
 
 ```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C~16:0~ and C~18:0~.
+The ellipses represent modern animal references fed on a C~3~diet (Copley et al.
+2003).
 B: The Δ^13^C (δ^13^C~18:0~-~16:0~) values are plotted against δ^13^C~16:0~ values.
-The dotted lines represent cut-offs for ruminant, non-ruminant, and ruminant dairy fats according to global references"}
+The dash lines represent cut-offs for ruminant, non-ruminant, and ruminant dairy fats based on Suryanarayan et al.
+2021"}
+
 knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
@@
 # Discussion
-<!--A brief summary of interesting results-->
-This paper is a preliminary study to investigate the pottery use at Guijiabao and Gaushan in Yanyuan area to explore whether those pots were used for cooking and what food categories could be processed in the pots.
-This will provide a direct evidence of the culinary practices in the Yanyuan region.
-Although the lipid yields in most of our samples are generally low, interpretative amounts of yields from all of our samples support the assumption that these ceramics were used as cooking or food storage vessels.
-The low lipid yield could be attributed to the bottom of the vessels, where there were fewer residues than other parts of the vessel, for example, the rim [@Evershed2008experimental; @Kwak2020; @Charters1993].
-However, we also found that one sample from the vessel bottom dated to the Neolithic period contains a richer composition of long-chain fatty acids compared with those in other samples.
-We recognize that there are limitations and challenges for interpretation due to our small sample sizes and a lack of local modern references for animal fats and plant oils.
-In addition, a detailed analysis of faunal remains from the Guijiabao site is still in progress, which may constrain a comprehensive comparison between different types of evidence.
-Nevertheless, the lipid profiles and their isotopic ratios could still provide informative clues for a better understanding of culinary practices.
-
-Saturated fatty acids are the major lipid molecules identified in our samples that normally come from animal fats or plant oils, indicative of cooking events.
-In addition to saturated fatty acids, unsaturated fatty acids such as C~18:1~ was also identified in all samples, and three (SYG-TN13-E22-1, SYG-TN13-E22-2#2, and SYG-TN13-E23-4) out of the six potsherds show a more abundant C~18:1~ than C~18:0~ and C~16:0.
-Although the specific type of these unsaturated fatty acids can not be further identified in this study without the preparation of specific derivatives [@Berstan2004], their presences might still hint a possibility of plant oils.
-It is suggested that a higher ratio of C~18:1~ and C~16:0~ to C~18:0~ is associated with a greater likelihood of the origin of plant oils [@Mayyas2018; @Copley2005].
-In our samples, the potsherd SYG-TN13_E23-3 has a higher ratio of C~18:1~ and C~16:0~ to C~18:0~, which also corresponds to the category of non-ruminant fats based on the isotopic values.
-However, we also acknowledge that C18:1 might have originate from animal fats, or there is a potential for modern contamination since monounsaturated fatty acids normally present limited preservation in archaeological contexts [@Whelton2021].
-
-<!--discussion about possible plant origins-->
-According to the δ^13^C values of fatty acids, the food residues in Neolithic potsherds could derive from non-ruminant fats, ruminant fats, and mixtures of ruminant fats and C~3~ plants or C~4~ plants.
-Previous studies suggested that millets (C~4~ plants), including foxtail and broomcorn millets, were the dominant cereals, according to macro-botanical and phytolith studies at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022].
-The processing of millet might also be suggested by one Guijiabao potsherd dating to Neolithic period based on its isotopic ratios.
-However, a further confirmation would require to identify of any lipid residues associated with C~4~ plants, such as the biomarker for broomcorn millet, miliacin [@Heron2016].
-The non-ruminant fats we identified from a Guijiabao potsherd that might have been from C~3~ plants or freshwater aquatic products, based on similar isotopic values reported by other studies in East Asia [@Shoda2018].
-According to the results from flotation analysis at the Guijiabao site, it is possible that the source of C~3~ plants is rice or native wild weeds, such as *Pennisetum* and *Chenopodiaceae*.
-However, the seeds of wild weeds only account for a small part of the floral assemblage derived from flotation [@Guijiabao2016].
-Given that a more depleted isotopic values for 5 out of 6 samples, ~C~3~ plants might consumed more than ~C~4 plants either by local people or by animals that were food source.
-
-We can not rule out other sources for non-ruminant fats, such as freshwater aqautic products, a possibility which can be supported by the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b].
-This will also need identification of lipid biomarkers related to aquatic oils, such as long-chain $\omega$-(o-alkylphenyl) alkanoic acids and isoprenoid fatty acids [@Evershed2008marine; @Hansel2004; @Cramp2014; @Lucquin2016].
-Although horse could be another non-ruminant fat source, whether it was part of local food choices in southwestern China around 4500 years ago during the Neolithic period is doubtful.
-First, the Pleistocene and Neolithic sites having wild horse bones are primarily concentrated in northern China, with only sporadic sites in the south [@Yuan2006].
-Second, domesticated horses first found in southwestern China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period.
-Thus, if horse meats is one of the source for non-ruminant fats, then it could only be present in historical ceramics.
+<!--A brief summary of the lipid yield and function of pottery-->
+Lipids extracted from potsherds at Guijiabao and Gaoshan all exceed interpretative amounts, with their profiles revealing the processing of both animal and plant products This confirms the role of the pottery as vessels for cooking or food storage at both sites.
+The generally low lipid concentrations (<20 μg/g) in five out of the six samples may be attributed to the bottom parts of vessels, which might absorb fewer residues compared to other parts, such as the rim, during cooking events [@Evershed2008experimental; @Kwak2020; @Charters1993].
+Additional factors contributing to the low lipid yields could be post-depositional processes [@Correa2014], or diverse methods of food preparation, including grilling or air-drying [@Zhang2022].
+We recognize challenges for interpreting our results due to the small sample size and the absence of a direct local reference database from faunal bones or modern animals.
+Furthermore, the detailed contexts of faunal remains at Guijiabao are still in progress, thus they are not fully accessible.
+These limitations constrain a thorough comparison of various types of evidence at the site.
+Despite these restrictions, the overall pattern of lipid profiles with isotopic values offers valuable insights into culinary practices.
 
 <!--discussion about possible ruminant fat origin-->
-The ruminant fats identified in the Neolithic potsherds (ca.
-4500 BP) from the Guijiabao and Gaoshan sites could have been derived from deer or sheep that were fed on a C~3~ diet, based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017].
-It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these animals have been found in sites in southwestern China [@Zhang2016; @Chen2020].
-It is hard to identify the exact ruminant category in this case due to the limited accessibility to local animal bones or fats as reference samples.
-However, the finding of fatty acid source from ruminant fats, couple with stone artifacts found at the site such as stone balls and arrowheads, suggesting hunting could be an important subsistence activity at the site [@Hao2022; @Guijiabao2020a].
-This also corresponds with resources that could be accessed in a highland environment.
+Neolithic potsherds from Guijiabao and Gaushan (ca.
+4500 BP) reveal food contents, including ruminant adipose fat (n= 3) and non-ruminant fat (n= 1).
+The ruminant adipose (Δ^13^C from `r E22_21_isotope[3]` to `r E22_22_isotope [3]` ) could have originated from goat/sheep, cattle, and deer based on a modern database [@Evershed2002], alongside archaeological contexts in northwestern China [@Sun2023].
+At Guijiabao, the ruminants' diet includes C~3~ plants (SYG-TN13-E22-2#1), and a mixture of C~3~ and C~4~ plants (SYG-TN13-E22-2#2).
+A C~3~ diet is supported by the predominance of C~3~ weeds at Guijiabao, such as *Pennisetum* and *Chenopodiaceae* [@Guijiabao2016].
+Some species in *Chenopodiaceae* can undergo C~4~ photosynthesis in arid environments [@Rudov2020; @Chasan2022], which may explain the mixture of plant types in SYG-TN13-E22-2#2.
+The direct input from C~4~ plants could be another reason for the mixed signal, as we have identified millet in the same sample.
+The domestication status of these ruminants detected in the samples remains unclear.
+However, considering the finding of hunting tools at Guijiabao, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a], with the abundance of wild ruminants in the natural settings of the Hengduan mountain ranges, such as deer, mutton, and yaks [@Zhang2022; @Zhang2016; @Chen2020; @He2020], it is plausible that subsistence at Guijiabao involving hunting to a certain degree.
 
-We observed mixtures of ruminant fats and C~3~ plants/freshwater aquatic source, based on analyses of two Guijiabao potsherds that date to the historical period (ca.
-1000 BP).
+<!--Plant as possible non-ruminant fat origin-->
+The non-ruminant fat (SYG-TN13-E23-3) could originate from C~3~ plants and freshwater resources, based on similar isotopic ranges reported for Neolithic pottery in eastern China [@Shoda2018; @Lyu2024], and a global database [@Steele2010].
+Although fish bones and net sinkers were found at Guijiabao [@Guijiabao2017b], lipid molecular evidence for fish, such as ω-(o-alkylphenyl)alkanoic acids and isoprenoid fatty acids [@Cramp2014], was absence in our samples.
+This suggests that fish might not have been commonly consumed, or alternative cooking methods for fish were employed.
+Possible source C~3~ plants could be rice or wild weeds based on botanical evidence [@Guijiabao2016; @Hao2022; @Huan2022].
+Due to a small portion of seeds of wild weeds in the floral assemblage [@Guijiabao2016], coupled with rice phytolichs identified from the residential area [@Huan2022], rice appears to be the more likely major source of C~3~ plants.
+In addition, the non-ruminant sample has a higher ratio of C~18:1~ and C~16:0~ to C~18:0~, associated with a greater likelihood of plant oils [@Mayyas2018; @Copley2005].
+However, we acknowledge that C18:1 might also originate from animal fats, or there is a potential for modern contamination, as monounsaturated fatty acids typically have limited preservation in archaeological contexts [@Whelton2021].
+A further detection of biomarkers for starchy plants with a specific derivatization method for lipid compounds will be necessary to assess the presence of rice oil [@Shoda2018].
+
+<!--Wild non-ruminant as possible non-ruminant fat origin-->
+The possibility of terrestrial animal input for the non-ruminant fat, such as wild boar, cannot be ruled out.
+The isotopic values of the non-ruminant sample align with those of reference fats from modern wild boars in western Japan [@Lucquin2016] and in the southeast Tibetan Plateau [@Zhang2022].
+Although horses could be another non-ruminant fat source, it is less likely to have been part of the local diet in southwestern China during the Neolithic period.
+This is because wild horse bones are primarily concentrated at the Pleistocene and Neolithic sites in northern China, with only sporadic occurrences in the south [@Yuan2006].
+Moreover, substantial evidence of domesticated horses in southwestern China dates back to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], a period approximately 3000 years later than the Neolithic period.
+Ongoing research on faunal remains at Guijiabao will provide more details for further discussion.
+
+<!--ruminants and crops-->
+Our preliminary data reveals the predominance of ruminants fed on a C~3~ and C~3~/C~4~ diet in pottery vessels during the Neolithic period, suggesting a potentially more significant role of ruminants in local culinary practices than previously thought.
+Most studies have primarily focused on mixed farming of millet and rice as the main economy in southwest China after the introduction of agricultural in the Neolithic period [@Deng2019].
+At Guijaibao, domesticated crops, especially broomcorn millet, constitute the majority of plants based on macro- and micro- botanical analyses [@Guijiabao2016; @Huan2022], and nitrogen and carbon isotope analysis of human bones suggests a more C~4~ diet [@Lin2022].
+Additionally, stone knives found at the site might be potentially be used for farming practices [@Zhou2019; @Hao2022; @Lin2022].
+In our study, a Neolithic potsherd containing the lipid compound of broomcorn millet [@Heron2016], provides direct evidence of millet processing at Guijiabao.
+Although a depleted δ^13^C~16:0~ in the non-ruminant sample suggest the possibility of C~3~plants related to rice, further analysis is needed for confirmation.
+It is important to note that lipid signatures of plants tend to be weak due to microbial degradation or being masked by animal fats [@Colonese2017; @Evershed2008experimental], possibly explaining the low lipid yield of plants in one sample.
+Nevertheless, the frequent processing of ruminant meats in pottery, with unearthed hunting tools, suggests that animal hunting could be an essential subsistence strategy alongside crop consumption.
+This may also reflect a subsistence strategy adapted to highlands with cold environments where meat consumption is common [@Zhang2022].
+
+The two historical potsherds (ca.
+1000 BP) from Guijiabao demonstrates mixtures of ruminant adipose and non-ruminant fats, implying a potential increase in the exploitation of C3 plants, pigs, or freshwater resources compared to the Neolithic period.
 In addition to rice, the source of C~3~ plants could also have included wheat, since it was introduced into the region during the historical period [@JiangGaoshan2017].
-A larger sample size is needed to determine whether there was a change in diet over time.
-The overall depleted δ^13^C values for potsherd samples associated with ruminant fats suggest that C~3~ plants might have been an important food source for local animals.
-Despite direct C~4~ plant signals not being found, analysis of one Guijiabao potsherd shows the source of C~4~ plant-fed ruminants during the Neolithic period.
-This indicates that either C~3~ or C~4~ plants could have been the food source for ruminants, but whether those animals were domesticated is unknown.
-The zooarchaeological analysis of faunal bones from Guijiabao is still in progress, the results of which could provide more details about local animal species for comparison.
-
-Mixed isotopic signals were also identified from two
-
-Our preliminary results shows that culinary practices in the Neolithic period at Guijiabao include the processing of ruminant adipose products that might be from local wild animals, such as deer.
-A subsistence that relies on hunting to a certain degree could be possible, given the abundance of wild animals in the natural settings of the Hengduan mountain ranges, as indicated by the faunal remains from the Yingpanshan site [@He2020].
-
-It is also likely that C~3~ plants (eg.
-rice) was processed based on the isotopic ratios showing non-ruminant fats from one Neolithic potsherds at Guijiabao.
-In addition to C~3 plants, another possibility for the source of non-ruminant fats could be freshwater aquatic products, which would require confirmation of the presence or absence of aquatic products.
-It is interesting to detect that one potsherd has a slightly higher values of carbon isotope ratios, indicating a contribution from C4 plants.
-The detection of C~3~ plants or C~4~ plants might suggest that farming practices were adopted during the time when those crops were first introduced to the Yanyuan area in the Neolithic period.
-The finding of C~4~ plant signal also corresponds with a previous study that suggests that a C~4~-based diet at the Guijiabao site using isotopic analysis of human bones [@Lin2022].
-However, the scale of farming practices and crop consumption is unclear.
-Despite stone knives and tools that were also found at the sites that might have been related to farming practices, more studies are needed to confirm their usage for harvesting crops [@Zhou2019; @Hao2022; @Lin2022].
-
-An approach analyzing organic residues may only provide a narrow perspective on local diets, but it could improve our current understanding of food practices through presence of food residues that were cooked in ceramic vessels.
-A large-scale organic residue analysis of ceramics across different pottery forms in a wider region would offer insight into temporal and regional variations in the local diet during the Neolithic period in southwestern China.
-
-An further investigation on the aquatic source should be conducted through selected ion monitoring (SIM) methods in the future to distinguish whether the δ13C
-values are the result of processing C3 plants or freshwater aquatic products in the pottery vessel.
-
-sampling bias, poor preservation
-These studies and the ability/limitations of ORA should be mentioned.
-
-The domination of millet at Guijiabao corresponds to a C~4~-based diet based on isotopic analysis of four human bones [@Lin2022].
-
-<!--- removed
-such as C~4~ plants (e.g., millets) and C~3~ plants (e.g., rice), or animals (e.g., pigs) became an essential part of the local diet in Yanyuan during the Neolithic period.
-while the food residues in historical potsherds present mixtures of ruminant and C~3~ plants.
--->
+For the Gaoshan sample, a slightly enriched δ^13^C may suggest minor contributions of pigs [@Evershed2002], in correspondence with the predominance of domestic pigs at the site [@He2020].
+However, determining changes in culinary practices over time and regional differences would require a larger sample size.
+While a geochemical approach analyzing organic residues may provide a limited perspective on local diets, it contributes to improving our current understanding of direct food practices through organic residues in ceramic vessels.
+As a preliminary study, our data reveals food contents cooked in pottery from both animal and plant sources, offering valuable insights into culinary practices during the time when agriculture spread into the Yuan Basin in southwest China.
 
 # Conclusion
-Our case study on organic residues in pottery fragments from the sites at Guijiabao and Gaoshan in southwestern China reveals the presence of lipids that can be analyzed to determine food sources.
-The foods cooked in the ceramics included ruminant fats and terrestrial plants during the Neolithic period, when millet and rice spread into southwestern China.
-The source of terrestrial plants could have been C~3~ plants, implying the cooking of rice or other wild weeds in the pots.
-However, the evidence of direct cooking of either C~3~ or C~4~ plants shows that the amount of C~3~/C~4~ plants cooked is not in proportion to the large quantity of plant remains found at the Guijiabao site, especially C~4~ plants such as millet.
-Instead, ruminant fats from local animals, such as wild deer, could have been a common food source during the Neolithic period.
-This suggests that hunting might have been an important part of local subsistence strategy, supplemented by rice/millet farming when they were introduced.
-Our results from the Guijiabao site could represent a subsistence strategy for adapting to the natural environment of the Hengduan Mountains.
-We recognize the limitations of our interpretation of the local diet due to the small sample size.
-Nevertheless, our study could offer an approach using organic residue analysis to determine food choices and explore the extent of the adoption of farming when rice/millet agriculture first spread into the Yanyuan area during the Neolithic period.
-A more systematic sampling of potsherds across different phases and pottery types may give a clear picture of diet and subsistence in southwestern China during the Neolithic period of transition.
+This paper presents a preliminary result investigating pottery use at Guijiabao in the Yanyuan area, middle Hengduan mountain range, through organic residue analyses to explore culinary practices.
+Despite the small sample size, our results reveal a relatively diverse range of food categories cooked in pottery during the Neolithic period, including ruminant adipose fats, plants such as broomcorn millet, and non-ruminant sources.
+In the historical period, mixtures of ruminant adipose fats and non-ruminant fats were identified.
+Non-ruminant fats could have originated from C3 plants, pigs, or freshwater resources.
+
+While a mixed crop of millet and rice is generally accepted as the main subsistence in southwest China, our results indicate that ruminant adipose fats were more commonly cooked in pottery.
+The prevalence of ruminant fats suggests the significance of meat consumption, a aspect often overlooked after the introduction of crops.
+The discovery of hunting tools at Guijiabao further indicates that hunting might have played a crucial role in the local subsistence strategy, complementing rice/millet farming.
+TThis provides valuable insights into subsistence strategies in a highland setting in the Hengduan Mountains ranges.
+Although our preliminary data has limitations for a comprehensive interpretation, the results demonstrate the feasibility of lipid identifications with compound-specific stable isotope analysis to determine food content processed in ceramic vessels at Guijiabao.
+A more systematic sampling of potsherds across different phases and pottery types will contribute to a clearer understanding of diet and subsistence in southwestern China in the future.
 
@@
 We also want to express our gratitude to the anonymous reviewers for their comments that helped us improve this paper.
-This research is sponsored by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).
+The research was supported by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).
 

### 2024-01-23 analysis/code/site-regional-map-Ch.R
edit map

@@ -5,7 +5,7 @@ library(rnaturalearth)
 library(rnaturalearthdata)
 library(here)
 library(ggspatial)
-# devtools::install_github('3wen/legendMap')
+#devtools::install_github('3wen/legendMap')
 library(legendMap)
 library(tmaptools)
 library(shadowtext)
@@ -28,10 +28,10 @@ site_location <-
 
 China_SE_Asia <-
   ggplot(data = world) +
-  geom_sf(fill= "antiquewhite") +
+  geom_sf(fill = "antiquewhite") + #antiquewhite
   geom_rect(xmin = 99.5, xmax = 105.5, ymin = 25.5, ymax = 31.5,
             fill = NA, colour = "red", size = 0.5) +
-  geom_shadowtext(data= world_points,
+  geom_shadowtext(data= world_points, #geom_shadowtext
                   aes(x = X, y = Y,
                       label = brk_name),
                   color='black',
@@ -46,15 +46,50 @@ China_SE_Asia <-
   theme(axis.title.x = element_blank(),
         axis.title.y = element_blank())
 
-# Topographic map
+# leaflet
+library(leaflet)
+
+# Background 2: World Imagery
+satellite <- leaflet() %>%
+  addTiles() %>%
+  setView( lng = 101, lat = 33, zoom = 5) %>%
+  addProviderTiles("Esri.WorldImagery")
+
+library(png)
+satellite <- readPNG(here::here("analysis","figures", "satellite_sw_china.png"),
+                     native = FALSE, info = FALSE)
+
+region_satellite_map <-
+  China_map %>%
+  ggmap() +
+  geom_rect(xmin = 99, xmax = 105.5, ymin = 25, ymax = 31.5,
+            fill = NA, colour = "red", size = 0.5) +
+  coord_sf(xlim = c(88, 116.5), ylim = c(13.5, 40), expand = FALSE) +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank())
+
+# Topographic map, get map from the Stadia Maps
 library(ggmap)
+register_stadiamaps("707a2ccb-7cfa-4aca-b2e0-5c517c4b9d66") # API
+# from https://stackoverflow.com/questions/77432892/problem-visualizing-maps-with-staten-maps
 # we don't want to download every time, so let's save the map locally
-# from https://stackoverflow.com/a/52710855/1036500
-#GJB_map <- ggmap(get_stamenmap(rbind(as.numeric(c(99, 25,
[…]

### 2024-01-23 analysis/code/isotope-analysis.R
edit the isotopic figure

@@ -117,6 +117,8 @@ isotope_C16_C18 <-
            label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
   annotate("text", x = -21.5, y = -24.5, angle = 45, vjust = 1.5,
            label = bquote(Delta*{}^13*"C = -3.1 \u2030")) +
+  annotate("text", x = c(-26, -27), y = c(-23.5, -32),
+           label = c("Non-Ruminant fat", "Ruminant fat")) +
   annotation_raster(tem3, ymin = -38, ymax= -17.9 ,
                     xmin = -36, xmax = -18.3)
 

### 2024-01-24 analysis/paper/paper.Rmd
minor edits

@@
 ---
-title: "Preliminary insights into the use of pottery and culinary pratices at Guijiabao site in southwestern China"
+title: "Preliminary insights into the use of pottery and culinary pratices at Guijiabao site in southwest China"
 author:
@@
 
-     The emergence of mixed agriculture of millet and rice in Neolithic southwestern China has been a topic of extensive discussion.
+     Mixed crops of millet and rice are thought to have been introduced to southwest China in the Neolithic period.
 Despite the recovery of both millet and rice grains, the varying proportions observed across the area underscore the complexity of the process of agricutural dispersal.
-Further investigation is needed to understand how new crops were integrated into southwestern China, characterized by diverse environmental conditions and landscapes.
-This study contributes to this understanding by exploring subsistence strategies in southwestern Sichuan through a preliminary analysis of lipid residues in pottery from the Guijiabao site using organic geochemical analyses.
+Further investigation is needed to understand how new crops were integrated into southwest China, characterized by diverse environmental conditions and landscapes.
+This study contributes to this understanding by exploring subsistence strategies in southwest Sichuan through a preliminary analysis of lipid residues in pottery from the Guijiabao site using organic geochemical analyses.
 Situated in the middle Hengduan mountain ranges, the Guijiabao site presents well-preserved stratigraphic records spanning from the Neolithic to the historical period, providing valuable insights into local subsistence practices in a highland environment.
@@
 This suggests that animals played a substantial role in the local diet, alongside the consumption of millet.
-These culinary practices also indicate that hunting may have remained a significant subsistence activity in the high-altitude region of southwestern China, even after the introduction of crops.
+These culinary practices also indicate that hunting may have remained a significant subsistence activity in the high-altitude region of southwest China, even after the introduction of crops.
 
 keywords: |
-  Southwestern China; Neolithic period; Lipid residue analysis; Compound-specific carbon isotope analysis; Pottery; Culinary practices
+  Southwest China; Neolithic period; Lipid residue analysis; Compound-specific carbon isotope analysis; Pottery; Culinary practices
 highlights: |
-  These are the highlights. 
+
+  Lipid residue analysis suggested the pottery was used as cooking pots at Guijiabao and Gaoshan.
+
+  The millet biomarker for broomcorn millet was identified in a Guijiabao potsherd.
+
+  Ruminant adipose fats were identified in the majority of potsherds.
+
+  Animals could be an important food source in a highland setting alongside crops.
 ---
@@
 The movement of Neolithic populations has long been discussed in archaeology to understand the processes of agricultural diffusion, regional variations, and their consequences [@Bellwood2023; @Davison2006].
-In southwest China, the emergence of farming practices was thought to be associated with the migrating population from central or northern China in the Neolithic period [@Deng2019].
-Crop remains found in prehistoric southwest China, including foxtail millet, broomcorn millet, and rice, can be traced back to as early as 5000 BP in southwest Sichuan and around 4600 BP in western Yunnan [@Jiang2016; @Deng2019; @YunnanMuseum1981; @Zhang2010; @Dal2022].
+In southwest China, the emergence of farming practices is thought to be associated with migrating populations from central or north China in the Neolithic period [@Deng2019].
+Crop remains found in prehistoric southwest China, including foxtail millet (*Setaria italica*), broomcorn millet (*Panicum miliaceum*), and rice (*Oryza sativa*), can be traced back to as early as 5000 BP in southwest Sichuan and around 4600 BP in west Yunnan [@Jiang2016; @Deng2019; @YunnanMuseum1981; @Zhang2010; @Dal2022].
 Despite different views on the routes of agricultural dispersal [@Guedes2013; @Guedes2018; @Liu2022; @Huan2022], it is generally believed that mixed farming was the main agricultural practice in southwest China [@Deng2019].
-However, more studies are still needed to understand how migrant communities interacted with new environments, especially in an area with a high diversity of landscapes, such as high mountains, plateaus, basins, and terraces.
+However, more studies are still needed to understand how farming communities interacted with diverse environments, especially in an area characterized by high landscape diversity, including high mountains, plateaus, basins, and terraces.
 
 <!--Guijaibao as a research area-->
-This study focuses mainly on archaeological sites located in the Hengduan Mountain ranges, a region in southwest China connecting the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south.
-As a critical passage for population movement, the Hengduan Mountain ranges and valleys facilitate not only material exchange [@Tong1986; @Hein2014; @Hein2016; @Ma2021], but also the spread of agriculture even further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017].
-The Guijiabao site in Yanyuan Basin, the middle Hengduan Mountain ranges, are a key site with the earliest Neolithic human occupation in this region dating to 5000 B.P., continuing to the historical period [@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022].
-The complete stratigraphy at Guijiabao provides a unique opportunity to investigate questions relative to subsistence since the Neolithic period in a basin surrounded by high mountains.
-For example, what is the role of agriculture in local environments?
+This study primarily focuses on the Guijiabao site located in the Hengduan Mountain ranges, a region in southwest China that connects the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south.
+As a crucial passage for population movement, the Hengduan Mountain ranges and valleys not only facilitate material exchange [@Tong1986; @Hein2014; @Hein2016; @Ma2021], but also contribute to the spread of agriculture further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017].
+Guijiabao is a key site with the earliest Neolithic human occupation in this region, dating to 5000 BP.
+and continuing to the historical period about 500 BP.
+[@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022].
+The complete stratigraphy at Guijiabao provides a unique opportunity to explore questions related to subsistence since the Neolithic period in a mountainous area.
+For instance, what role did agriculture play in local environments?
 What other resources were obtained in addition to crops?
@@
 <!--justifying organic residue analysis-->
-Although the remains of animal bones and plants offer evidence of subsistence, their preservation heavily depends on environmental conditions.
-Geographical regions with seasonal heavy precipitation and acidic sediments, such as Southwest Sichuan and the middle Hunduan Mountain [@Li1989], tend to lead to poor preservation of archaeological remains.
-In such cases, organic residue analysis is a powerful tool for studying ancient food practices by directly examining residues in pottery.
-Organic residues, specifically lipids, can be well preserved in the porous fabric of pottery for thousands of years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017; @Dudd1999; @Miller2020].
-The growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies [@Kwak2020], and culinary practices across a region over time [@Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
+The increasing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies [@Kwak2020], and culinary practices across a region over time [@Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
+Although remains of animal bones and plants offer evidence of subsistence, their preservation heavily depends on environmental conditions.
+Geographical regions with seasonal heavy precipitation and acidic sediments, such as Southwest Sichuan and the middle Hungduan Mountain [@Li1989], often result in poor preservation of archaeological remains.
+In such cases, organic residue analysis is a powerful tool for investigating ancient food practices by directly examining residues in pottery.
+Organic residues, particularly lipids, exhibit remarkable preservation in the porous fabric of pottery for thousands of years.
+This preservation is attributed to the protection provided by clay matrices, preventing lipids from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017; @Dudd1999; @Miller2020].
 
 <!--regional application-->
-By identifying lipid compositions and conducting compound-specific carbon isotope analysis, archaeologists are able to distinguish between plant oils, ruminant fats, dairy products, and aquatic/marine resources [@Copley2005; @Craig2007; @Evershed1994].
+By identifying lipid compositions and conducting compound-specific carbon isotope analysis, archaeologists are able to distinguish lipid origins between plant oils, ruminant fats, dairy products, and aquatic/marine resources [@Copley2005; @Craig2007; @Evershed1994].
 Recent studies have successfully identified various plant types processed in archaeological vessels, including broomcorn millet [@Bossard2013; @Heron2016; @Ganzarolli2018], starchy plants [@Shoda2018], and cereals [@Hammann2018; @Colonese2017].
-The increasing use of ceramics in China highlights the robustness of organic residue analysis for comprehending regional cooking practices [@Shoda2018; @Lyu2024; @Han2022], discerning distinct subsistence strategies between low- and high-altitude zones [@Zhang2022], and exploring the impact of nomadic pastoralism on northwest China [@Sun2023].
-In this study, we aim to investigate pottery use and culinary practices in the middle Hengduan mountains, focusing on ceramics from the Guijiabao and Gaoshan sites to understand local subsistence.
+Several applications to ancient pottery in China also highlight the robustness of this approach in understanding regional cooking practices [@Shoda2018; @Lyu2024; @Han2022], discerning distinct subsistence strategies between low- and high-altitude zones in Tibetan Plateau [@Zhang2022], and exploring the impact of nomadic pastoralism on northwest China [@Sun2023].
+In this study, we aim to investigate pottery use and culinary practices in the middle Hengduan mountains, with a specific focus on ceramics from the Guijiabao site to gain insights into local subsistence.
 
@@
 
-```{r SW-china-map, fig.cap = "The archaeological sites Guijiabao and Gaoshan and other locations in southwestern China that are mentioned in this study. Map data is from naturalearthdata.com"}
+```{r SW-china-map, fig.cap = "The archaeological sites, Guijiabao and Gaoshan, and other locations in southwest China that are mentioned in this study.
+Map data is from Leaflet and naturalearthdata.com"}
 library(here)
-knitr::include_graphics(here("analysis", "figures", "Sichuan-sites-map.png"))
+knitr::include_graphics(here("analysis", "figures", "Sichuan-sites-map3.png"))
 ```
@@
 <!--Yanyuan area and Guijiabao site-->
-The Guijiabao site is located in the southwest of Yanyuan Basin, a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)), with an elevation of 2400 meters.
-The first survey at Guijiabao in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean and ground houses, post holes, ash pits, ditches, and burials, suggesting a settlement site [@Hao2022; @Guijiabao2017a].
-The site has been occupied since the Neolithic period (5000-3700 BP), continued to the Bronze Age (3200-2700 BP) and the historical period (1200-500 BP) corresponding to the late North Song to early Ming dynasties, based on radiocarbon dates [@Hao2022; @Huan2022].
-The Neolithic period can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the artifact assemblages and ceramic styles.
+Guijiabao is located in the southwest of Yanyuan County in Sichuan Province, southwest China (Figure \@ref(fig:SW-china-map)), with an elevation of 2399 m [@Guijiabao2017a].
+Yanyuan is a pull-apart basin surrounded by mountains in the central east part of the Hengduan Mountain ranges.
+The initial survey at Guijiabao in 2015 and subsequent excavations from 2016 to 2019 uncovered abundant archaeological materials and features, such semi-subterranean and ground houses, post holes, ash pits, ditches, and burials, suggesting a settlement site [@Hao2022; @Guijiabao2017a].
+According to radiocarbon dates, the site had been occupied from the Neolithic period (5000-3700 BP), continued to the Bronze Age (3200-2700 BP) and the historical period from the late North Song to early Ming dynasties (1200-500 BP) [@Hao2022; @Huan2022].
+The Neolithic period can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, distinguished by artifact assemblages and ceramic styles.
 
 <!--the major finding at GJB-->
-Ceramics are the major finding at Guijiabao.
-The common vessel forms include jars, urns, bowls, and basins.
+The common ceramic vessel forms at Guijiabao include pots, jars, urns, bowls, and basins, with some exhibiting charred surfaces, indicating their potential use for cooking purposes.
+This research also attempts to confirm that these pottery vessels were used as cooking or storage wares.
 The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with cord-marked or basket-impressed patterns [@Hao2022; @Guijiabao2017a].
-Ceramics with charred surfaces suggest that they might be used for cooking.
-This research also attempts to confirm that these pottery vessels were used as cooking or storage wares.
 Some ceramic decorations present northern styles, while others with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of close relationships or interaction between Guijiabao and neighboring areas, the northern and southern parts of the Hengduan Mountains [@Guijiabao2017a].
-The flotation analysis of soil samples from the 2015 excavation demonstrates the predominance of cultivated broomcorn millet (*Panicum miliaceum*) and foxtail millet (*Setaria italica*) at Guijiabao during the Neolithic period [@Guijiabao2016].
-Rice grains (*Oryza sativa*) were also identified in later excavations from the Neolithic period based on macro botanical evidence and phytolith analysis [@Hao2022; @Huan2022], though in a smaller number than millet.
+The flotation analysis of soil samples from the 2015 excavation demonstrates the predominance of cultivated broomcorn millet and foxtail millet at Guijiabao during the Neolithic period [@Guijiabao2016].
+Rice grains were also identified in later excavations from the Neolithic period based on macro botanical evidence and phytolith analysis [@Hao2022; @Huan2022], though in a smaller number than millet.
 
 <!--info about Gaoshan-->
-To compare the culinary practices in different ecological zones, we examined a sample from the Gaoshan site in southwestern Chengdu Plain, an alluvial plain in Sichuan close to the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
+To compare the culinary practices in different ecological zones, we examined a sample from the Gaoshan site in southwest Chengdu Plain, an alluvial plain in Sichuan close to the Tibetan Plateau (Figure \@ref(fig:SW-china-map)).
 Excavations at the Gaoshan site reveal abundant materials, including ceramics, stone artifacts, animal and plant remains, and archaeological features such as house foundations, walls, burials, ash pits, ditches, and a well [@Gaoshan2017; @He2020; @JiangGaoshan2017; @ZhouGaoshan2015].
@@
 
-While mixed agriculture of millet and rice was practiced in southwestern China, variations in their proportions across the region have been observed.
+While mixed agriculture of millet and rice was practiced in southwest China, variations in their proportions across the region have been observed.
 Macro-botanical and phytolith analyses of the Guijiabao site indicate that millet was the primary crop, aligning with findings at the Yingpanshan site in Sichuan [@Zhao2011].
@@
 As a preliminary study to assess the feasibility of employing organic geochemical approaches, we collected five pottery fragments from cultural layers in the residential area of Guijiabao (sample code starting with SYG) (Table \@ref(tab:sherds-table)).
-For comparison, one fragment was sampled from Goashan (sample code starting with CDG).
-Those pottery fragments represent typical types with well-defined contexts, with their radiocarbon dates ranging from the Neolithic period (ca.
-4500 BP) to the historical period (ca.
-1000 BP).
+For comparison, one fragment was sampled from Gaoshan (sample code starting with CDG).
+Those pottery fragments represent typical types with well-defined contexts, with their radiocarbon dates ranging from the early phase of Neolithic period (ca.
+5000-4500 BP) to the historical period (ca.
+1200-500 BP).
 The pottery samples are mostly base fragments and fine sand-tempered, with a few displaying impressed decorations.
@@
 
-quant_FA_number <-
-  quant_FA %>%
-  pivot_longer(!number&!sample&!`g extracted`&!`%reco`,
-               names_to = "FA", 
-               values_to = "weight") %>% 
-  filter(str_starts(FA, "ng") & !sample =="") %>% 
-  mutate(FA = str_remove(FA, "ng "),
-         weight = round(weight/`%reco`/`g extracted`/1000, 2)) %>% 
-  select(-number, -`%reco`, -`g extracted`) %>% 
-  pivot_wider(names_from = sample, 
-              values_from = weight) %>% 
-  filter(!FA %in%c("C34:0", "C21:0"))
-
-FA_sum_min <- sum(quant_FA_number$"SYG-TN13-E22-2#1", na.rm = T)
-FA_sum_max <- sum(quant_FA_number$"SYG-TN13-E22-2#2")
-```
-
-The total lipid concentrations of potsherds at Guijiabao and Gaoshan range from `r FA_sum_min` to `r FA_sum_max` μg/g (Table \@ref(tab:FA-composition-table)).
-Lipid yields from all potsherds are above interpretable concentrations (5 μg/g) [@Evershed2008experimental], suggesting those ceramics should be associated with food processing.
-The main compounds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid) and C~18:0~ (stearic acid), commonly derived from animal fats and plant oils (Figure \@ref(fig:chromatograms-plots-four)).
-Unsaturated fatty acid, C~18:1~, was detected in all samples and dominant in three, but we are unable to distinguish the C~18:1~ types based on current methods.
-The most interesting finding is that milliacin, the compound specific to broomcorn millet [@Heron2016; @Bossard2013; @Ganzarolli2018], was identified in one Neolithic sample (SYG-TN13-E22-2#2) from Guijiabao (Figure \@ref(fig:chromatograms-specific-std)).
-This sample has a wider range of fatty acids from C~12:0~ to C~30:0~ (`r FA_sum_max` μg/g), where long-chain fatty acids could originate from plant sources [@Post1996; @Copley2005].
-Overall, the dominance of C~16:0~ and C~18:0~ in most samples suggests that the lipids extracted from our samples were greatly degraded.
-
-```{r FA-composition-table, cache= TRUE}
-quant_FA_tidy <-
+quant_FA_n <-
   quant_FA %>%
@@
   select(-number, -`%reco`, -`g extracted`) %>% 
-  mutate(weight = round(weight, 2),
-         weight = ifelse(is.na(weight), "-", weight),
+  mutate(weight = round(weight, 2))
+
+FA_sum <- quant_FA_n %>% filter(FA == " total") 
+FA_sum_min <- FA_sum %>% slice(which.min(weight)) %>% pull(weight)
+FA_sum_max <- FA_sum %>% slice(which.max(weight)) %>% pull(weight)
+FA_mean <- round(mean(FA_sum$weight), 2)
+```
+
+The total lipid concentrations of potsherds at Guijiabao and Gaoshan range from `r FA_sum_min` to `r FA_sum_max` μg/g, with an average of `r FA_mean` μg/g (Table \@ref(tab:FA-composition-table)).
+Lipid yields from all potsherds are above interpretable concentrations (5 μg/g) [@Evershed2008experimental], suggesting those ceramics should be associated with food processing.
+The main compounds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid) and C~18:0~ (stearic acid), commonly derived from animal fats and plant oils (Figure \@ref(fig:chromatograms-plots-four)).
+Unsaturated fatty acid, C~18:1~, was detected in all samples and dominant in three, but we are unable to distinguish the C~18:1~ types based on current methods.
+The most interesting finding is that milliacin, the compound specific to broomcorn millet [@Heron2016; @Bossard2013; @Ganzarolli2018], was identified in one Neolithic sample (SYG-TN13-E22-2#2) from Guijiabao (Figure \@ref(fig:chromatograms-specific-std)).
+This sample has a wider range of fatty acids from C~12:0~ to C~30:0~ (`r FA_sum_max` μg/g), where long-chain fatty acids could originate from plant sources [@Post1996; @Copley2005].
+Overall, the dominance of C~16:0~ and C~18:0~ in most samples suggests that the lipids extracted from our samples were greatly degraded.
+
+```{r FA-composition-table, cache= TRUE}
+quant_FA_table <-
+  quant_FA_n %>% 
+  mutate(weight = ifelse(is.na(weight), "-", weight),
          weight = ifelse(weight == 0, "-", weight)) %>% 
@@
 
-knitr::kable(quant_FA_tidy,
+knitr::kable(quant_FA_table,
              caption = "Quantitative data of fatty acid (FA) composition (μg/g) for each pottery sample based on GC-FID analysis")
@@
            label = bquote(Delta*{}^13*"C = -3.1 \u2030")) +
+  annotate("text", x = c(-26, -27), y = c(-23.5, -32),
+           label = c("Non-Ruminant fat", "Ruminant fat")) +
   annotation_raster(tem3, ymin = -38, ymax= -17.9, xmin = -36, xmax = -18.3) +
@@
            label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
-  annotate("text", x = c(-32, -21), y = c(1.3, 1.3),
+  annotate("text", x = c(-31, -21), y = c(1.3, 1.3),
            label = c("C[3]~diet", "C[4]~diet"), parse= TRUE) +
-  annotate("segment", x = -29, y = 1.3, xend = -24, yend = 1.3,
+  annotate("segment", x = -29, y = 1.3, xend = -23, yend = 1.3,
            arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
@@
 To identify the sources of degraded lipids, we conducted compound-specific carbon isotope analysis of C~16:0~ and C~18:0~ using GC-C-IRMS.
-This technique has proven useful to distinguish fat origin between ruminant, non-ruminant, ruminant dairy products, C~3~ plants, C~4~ plants, and marine resources based on their stable carbon isotope values (hereafter δ13C ) [@Dunne2012; @Evershed2008].
-For example, the δ13C values from marine sources or C~4~ plants are more enriched than those from ruminant adipose fat and C~3~ plants due to different metabolic or photosynthetic pathways [@Copley2003; @Meier2002; @Cramp2014].
+This technique has proven useful to distinguish fat origin between ruminant, non-ruminant, ruminant dairy products, C~3~ plants, C~4~ plants, and marine resources based on their stable carbon isotope values (hereafter δ^13^C) [@Dunne2012; @Evershed2008].
+For example, the δ^13^C values from marine sources or C~4~ plants are more enriched than those from ruminant adipose fat and C~3~ plants due to different metabolic or photosynthetic pathways [@Copley2003; @Meier2002; @Cramp2014].
 The values of δ^13^C~16:0~ and δ^13^C~18:0~ were plotted based on global modern references for the assessment of their origin.
@@
 
-The isotope data shows that δ^13^C~16:0~ values range from `r rg_C16_isotope[1]` ‰ to `r rg_C16_isotope[2]` ‰, δ^13^C~18:0~ values range from `r rg_C18_isotope[1]` ‰ to `r rg_C18_isotope[2]`, and Δ^13^C values are between `r delta_range[2]` ‰ and `r delta_range[1]` ‰.
-A half of the samples (n= 3) at Guijiabao and Gaushan fall within the global reference range for ruminant adipose fat, where one sample from Guijiabao (SYG-TN13-E22-2#2) has a more enriched δ^13^C~16:0~ (`r E22_22_isotope[1]`) than others.
+The isotope data shows that δ^13^C~16:0~ values range from `r rg_C16_isotope[1]` to `r rg_C16_isotope[2]` ‰, δ^13^C~18:0~ values range from `r rg_C18_isotope[1]` to `r rg_C18_isotope[2]` ‰, and Δ^13^C values are between `r delta_range[2]` and `r delta_range[1]` ‰.
+A half of the samples (n= 3) at Guijiabao and Gaoshan fall within the global reference range for ruminant adipose fat, where one sample from Guijiabao (SYG-TN13-E22-2#2) has a more enriched δ^13^C~16:0~ (`r E22_22_isotope[1]` ‰) than others.
 The enriched δ^13^C suggests that the diet of ruminants contains both C~4~ and C~3~ plants (Figure \@ref(fig:isotopic-plots): B), while the depleted δ^13^C is related to a more C~3~ plant diet.
@@
 One sample (SYG-TN13-E23-3) with Δ^13^C of `r E23_isotope[3]` falls within the range for non-ruminant fats (Figure \@ref(fig:isotopic-plots): B).
-Its relatively depleted δ^13^C~ values (δ^13^C~16:0~ of `r E23_isotope[1]` ‰ and δ^13^C~18:0~ of `r `E23_isotope[2]` ‰) reflect input of C~3~ plants, terrestrial animals, or freshwater resources [@Shoda2018; @Lucquin2016].
-The two samples (SYG-TN13-E22-1 and SYG-3) dating to the historical period clustered within a narrow range of δ^13^C~16:0~ (`r SYG3_isotope[1]` ‰ to `r E22_1_isotope[1]`) and δ^13^C~18:0~ (`r SYG3_isotope[2]` ‰ to `r E22_1_isotope[2]`) near the reference line, showing a mixing source of ruminant and non-ruminant fat [@Copley2003].
-A mixture of lipids from different sources is possible because δ13C values could be influenced by multiple distinct products, such as ruminant fats and plants, in the same cooking event or separate cooking events over time [@Evershed2002].
+Its relatively depleted δ^13^C~ values (δ^13^C~16:0~ of `r E23_isotope[1]` ‰ and δ^13^C~18:0~ of `r E23_isotope[2]` ‰) reflect input of C~3~ plants, terrestrial animals, or freshwater resources [@Shoda2018; @Lucquin2016].
+The two samples (SYG-TN13-E22-1 and SYG-3) dating to the historical period clustered within a narrow range of δ^13^C~16:0~ (`r SYG3_isotope[1]` to `r E22_1_isotope[1]` ‰) and δ^13^C~18:0~ (`r SYG3_isotope[2]` to `r E22_1_isotope[2]` ‰) near the reference line, showing a mixing source of ruminant and non-ruminant fat [@Copley2003].
+A mixture of lipids from different sources is possible because δ^13^C values could be influenced by multiple distinct products, such as ruminant fats and plants, in the same cooking event or separate cooking events over time [@Evershed2002].
 
 ```{r isotopic-plots, echo=FALSE, fig.cap= "A: The distribution of stable carbon isotopes for C~16:0~ and C~18:0~.
-The ellipses represent modern animal references fed on a C~3~diet (Copley et al.
+The ellipses represent modern animal references fed on a C~3~ diet (Copley et al.
 2003).
@@
 <!--discussion about possible ruminant fat origin-->
-Neolithic potsherds from Guijiabao and Gaushan (ca.
-4500 BP) reveal food contents, including ruminant adipose fat (n= 3) and non-ruminant fat (n= 1).
-The ruminant adipose (Δ^13^C from `r E22_21_isotope[3]` to `r E22_22_isotope [3]` ) could have originated from goat/sheep, cattle, and deer based on a modern database [@Evershed2002], alongside archaeological contexts in northwestern China [@Sun2023].
+Neolithic potsherds from Guijiabao and Gaoshan (ca.
+5000-4500 BP) reveal food contents, including ruminant adipose fat (n= 3) and non-ruminant fat (n= 1).
+The ruminant adipose (Δ^13^C from `r E22_21_isotope[3]` to `r E22_22_isotope [3]` ‰) could have originated from goat/sheep, cattle, and deer based on a modern database [@Evershed2002], alongside archaeological contexts in northwestern China [@Sun2023].
 At Guijiabao, the ruminants' diet includes C~3~ plants (SYG-TN13-E22-2#1), and a mixture of C~3~ and C~4~ plants (SYG-TN13-E22-2#2).
@@
 In addition, the non-ruminant sample has a higher ratio of C~18:1~ and C~16:0~ to C~18:0~, associated with a greater likelihood of plant oils [@Mayyas2018; @Copley2005].
-However, we acknowledge that C18:1 might also originate from animal fats, or there is a potential for modern contamination, as monounsaturated fatty acids typically have limited preservation in archaeological contexts [@Whelton2021].
+However, we acknowledge that C~18:1~ might also originate from animal fats, or there is a potential for modern contamination, as monounsaturated fatty acids typically have limited preservation in archaeological contexts [@Whelton2021].
 A further detection of biomarkers for starchy plants with a specific derivatization method for lipid compounds will be necessary to assess the presence of rice oil [@Shoda2018].
@@
 The isotopic values of the non-ruminant sample align with those of reference fats from modern wild boars in western Japan [@Lucquin2016] and in the southeast Tibetan Plateau [@Zhang2022].
-Although horses could be another non-ruminant fat source, it is less likely to have been part of the local diet in southwestern China during the Neolithic period.
+Although horses could be another non-ruminant fat source, it is less likely to have been part of the local diet in southwest China during the Neolithic period.
 This is because wild horse bones are primarily concentrated at the Pleistocene and Neolithic sites in northern China, with only sporadic occurrences in the south [@Yuan2006].
-Moreover, substantial evidence of domesticated horses in southwestern China dates back to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], a period approximately 3000 years later than the Neolithic period.
+Moreover, substantial evidence of domesticated horses in southwest China dates back to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], a period approximately 3000 years later than the Neolithic period.
 Ongoing research on faunal remains at Guijiabao will provide more details for further discussion.
@@
 The two historical potsherds (ca.
-1000 BP) from Guijiabao demonstrates mixtures of ruminant adipose and non-ruminant fats, implying a potential increase in the exploitation of C3 plants, pigs, or freshwater resources compared to the Neolithic period.
+1200-500 BP) from Guijiabao demonstrates mixtures of ruminant adipose and non-ruminant fats, implying a potential increase in the exploitation of C~3~ plants, pigs, or freshwater resources compared to the Neolithic period.
 In addition to rice, the source of C~3~ plants could also have included wheat, since it was introduced into the region during the historical period [@JiangGaoshan2017].
@@
 In the historical period, mixtures of ruminant adipose fats and non-ruminant fats were identified.
-Non-ruminant fats could have originated from C3 plants, pigs, or freshwater resources.
+Non-ruminant fats could have originated from C~3~ plants, pigs, or freshwater resources.
 
@@
 The prevalence of ruminant fats suggests the significance of meat consumption, a aspect often overlooked after the introduction of crops.
-The discovery of hunting tools at Guijiabao further indicates that hunting might have played a crucial role in the local subsistence strategy, complementing rice/millet farming.
-TThis provides valuable insights into subsistence strategies in a highland setting in the Hengduan Mountains ranges.
+The discovery of hunting tools at Guijiabao further indicates that hunting might have played a crucial role in the local subsistence strategy, complementing millet/rice farming.
+This provides valuable insights into subsistence strategies in a highland setting in the Hengduan Mountains ranges.
 Although our preliminary data has limitations for a comprehensive interpretation, the results demonstrate the feasibility of lipid identifications with compound-specific stable isotope analysis to determine food content processed in ceramic vessels at Guijiabao.
-A more systematic sampling of potsherds across different phases and pottery types will contribute to a clearer understanding of diet and subsistence in southwestern China in the future.
+A more systematic sampling of potsherds across different phases and pottery types will contribute to a clearer understanding of diet and subsistence in southwest China in the future.
 

### 2024-01-24 analysis/code/site-regional-map-Ch.R
edit maps

@@ -46,28 +46,6 @@ China_SE_Asia <-
   theme(axis.title.x = element_blank(),
         axis.title.y = element_blank())
 
-# leaflet
-library(leaflet)
-
-# Background 2: World Imagery
-satellite <- leaflet() %>%
-  addTiles() %>%
-  setView( lng = 101, lat = 33, zoom = 5) %>%
-  addProviderTiles("Esri.WorldImagery")
-
-library(png)
-satellite <- readPNG(here::here("analysis","figures", "satellite_sw_china.png"),
-                     native = FALSE, info = FALSE)
-
-region_satellite_map <-
-  China_map %>%
-  ggmap() +
-  geom_rect(xmin = 99, xmax = 105.5, ymin = 25, ymax = 31.5,
-            fill = NA, colour = "red", size = 0.5) +
-  coord_sf(xlim = c(88, 116.5), ylim = c(13.5, 40), expand = FALSE) +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank())
-
 # Topographic map, get map from the Stadia Maps
 library(ggmap)
 register_stadiamaps("707a2ccb-7cfa-4aca-b2e0-5c517c4b9d66") # API
@@ -164,3 +142,30 @@ ggsave(here::here("analysis","figures", "Sichuan-sites-map2.png"),
        dpi = 300,
        units = "in")
 
+# leaflet
+library(leaflet)
+
+# Background 2: World Imagery
+satellite <- leaflet() %>%
+  addTiles() %>%
+  setView(lng = 99.5, lat = 30.5, zoom = 5) %>%
+  addProviderTiles("Esri.WorldImagery")
+
+library(magick)
+satellite <- image_read(here::here("analysis","figures", "satellite_sw_china.png"))
+satellite_crop <- image_crop(satellite, geometry = "1000x748+84-20") #1084,768
+
+satellite_map <-
+  ggdraw() +
+  draw_image(satellite_crop)
+
+plt1 <-
+  plot_grid(satellite_map,
+            China_map_with_site,
+            ncol = 2)
+
+ggsave(here::here("analysis","figures", "Sichuan-sites-map3.png"),
+       width = 6,
+       height = 3,
+       dpi = 300,
+       units = "in")

### 2024-02-15 analysis/code/isotope-analysis.R
edit plots

@@ -128,7 +128,7 @@ big_delta_C16 <-
   geom_point(size = 2, alpha = 0.9, aes(shape = name, color = period)) +
   theme_minimal() +
   labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
-       y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
+       y = bquote(Delta*{}^13*"C")) +
   scale_x_continuous(limits = c(-33, -20),
                      breaks = seq(-33, -20, 2)) +
   ylim(-2.5, 1.5) +

### 2024-02-15 analysis/code/site-regional-map-Ch.R

@@ -91,7 +91,7 @@ China_map_with_site <-
            fontface = "italic", color = "grey1", size = 2.5) +
   annotate(geom = "text", x = 102.7, y = 27.5, label = "Yanyuan\nBasin",
            fontface = "italic", color = "grey1", size = 2.5) +
-  annotate(geom = "text", x = 101.5, y = 29.3, label = "Hengduan\nMountain",
+  annotate(geom = "text", x = 101.5, y = 29.3, label = "Hengduan\nMountains",
            fontface = "italic", color = "grey1", size = 2.5) +
   annotate("segment", x = 102.2, xend = 101.8, y = 27.5, yend = 27.5,
            color = "grey5", arrow = arrow(length = unit(.15,"cm"))) +
@@ -159,6 +159,7 @@ satellite_map <-
   ggdraw() +
   draw_image(satellite_crop)
 
+library(cowplot)
 plt1 <-
   plot_grid(satellite_map,
             China_map_with_site,

### 2024-02-15 analysis/paper/paper.Rmd
edits based on KC comments

@@
 This study contributes to this understanding by exploring subsistence strategies in southwest Sichuan through a preliminary analysis of lipid residues in pottery from the Guijiabao site using organic geochemical analyses.
-Situated in the middle Hengduan mountain ranges, the Guijiabao site presents well-preserved stratigraphic records spanning from the Neolithic to the historical period, providing valuable insights into local subsistence practices in a highland environment.
+The site, situated in the middle Hengduan mountain ranges, presents well-preserved stratigraphic records spanning from the Neolithic to the historical period, providing valuable insights into local subsistence practices in a highland environment.
 Results reveal the detection of ruminant adipose fats in the majority of potsherds, with a few showing plant inputs such as broomcorn millet.
@@
 
-  The millet biomarker for broomcorn millet was identified in a Guijiabao potsherd.
+  The biomarker for broomcorn millet was identified in a Guijiabao potsherd.
 
@@
 <!-- intro to the topic and research aim-->
-The movement of Neolithic populations has long been discussed in archaeology to understand the processes of agricultural diffusion, regional variations, and their consequences [@Bellwood2023; @Davison2006].
-In southwest China, the emergence of farming practices is thought to be associated with migrating populations from central or north China in the Neolithic period [@Deng2019].
+The movement of Neolithic populations has long been discussed in archaeology to understand the processes of agricultural diffusion [@Bellwood2005].
+In southwest China, the emergence of farming practices is thought to be associated with migrating populations from central or north China in the Neolithic period [@Zhao2011; @Deng2019].
 Crop remains found in prehistoric southwest China, including foxtail millet (*Setaria italica*), broomcorn millet (*Panicum miliaceum*), and rice (*Oryza sativa*), can be traced back to as early as 5000 BP in southwest Sichuan and around 4600 BP in west Yunnan [@Jiang2016; @Deng2019; @YunnanMuseum1981; @Zhang2010; @Dal2022].
@@
 This study primarily focuses on the Guijiabao site located in the Hengduan Mountain ranges, a region in southwest China that connects the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south.
-As a crucial passage for population movement, the Hengduan Mountain ranges and valleys not only facilitate material exchange [@Tong1986; @Hein2014; @Hein2016; @Ma2021], but also contribute to the spread of agriculture further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017].
-Guijiabao is a key site with the earliest Neolithic human occupation in this region, dating to 5000 BP.
-and continuing to the historical period about 500 BP.
+As a crucial passage for population movement, the Hengduan Mountain ranges and valleys not only facilitate material exchange [@Tong1986; @Hein2016; -@Hein2014; @Ma2021], but also contribute to the spread of agriculture further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017].
+Guijiabao is a key site with the so far earliest Neolithic human occupation in this region, dating to 5000 BP and continuing to the historical period about 500 BP.
 [@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022].
-The complete stratigraphy at Guijiabao provides a unique opportunity to explore questions related to subsistence since the Neolithic period in a mountainous area.
+The occupational history at Guijiabao provides a unique opportunity to explore questions related to subsistence since the Neolithic period in a mountainous area.
 For instance, what role did agriculture play in local environments?
 What other resources were obtained in addition to crops?
-In this study, we present our preliminary analyses of archaeological potsherds from the Guijiabao site and the Gaoshan site in southwest China using organic residue analysis of lipids.
+In this study, we present our preliminary analyses of archaeological potsherds from the Guijiabao site using organic residue analysis of lipids.
+For comparison, we also applied the same method to a sample from the Gaoshan site on the Chengdu Plain.
 
 <!--justifying organic residue analysis-->
-The increasing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies [@Kwak2020], and culinary practices across a region over time [@Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
+The increasing application of organic residue analysis demonstrates its efficacy in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies [@Kwak2020], and culinary practices across a region over time [@Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].
 Although remains of animal bones and plants offer evidence of subsistence, their preservation heavily depends on environmental conditions.
-Geographical regions with seasonal heavy precipitation and acidic sediments, such as Southwest Sichuan and the middle Hungduan Mountain [@Li1989], often result in poor preservation of archaeological remains.
-In such cases, organic residue analysis is a powerful tool for investigating ancient food practices by directly examining residues in pottery.
+Geographical regions with seasonal heavy precipitation and acidic sediments like Sichuan [@Li1989], often result in poor preservation of archaeological remains.
+In such cases, organic residue analysis is a potent instrument for investigating ancient food practices by directly examining residues in pottery; however, it has never been applied to materials from this region.
 Organic residues, particularly lipids, exhibit remarkable preservation in the porous fabric of pottery for thousands of years.
@@
 According to radiocarbon dates, the site had been occupied from the Neolithic period (5000-3700 BP), continued to the Bronze Age (3200-2700 BP) and the historical period from the late North Song to early Ming dynasties (1200-500 BP) [@Hao2022; @Huan2022].
-The Neolithic period can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, distinguished by artifact assemblages and ceramic styles.
+The Neolithic period can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, differing in artifact assemblages and ceramic styles.
 
@@
 This research also attempts to confirm that these pottery vessels were used as cooking or storage wares.
-The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with cord-marked or basket-impressed patterns [@Hao2022; @Guijiabao2017a].
-Some ceramic decorations present northern styles, while others with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of close relationships or interaction between Guijiabao and neighboring areas, the northern and southern parts of the Hengduan Mountains [@Guijiabao2017a].
-The flotation analysis of soil samples from the 2015 excavation demonstrates the predominance of cultivated broomcorn millet and foxtail millet at Guijiabao during the Neolithic period [@Guijiabao2016].
+The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish-brown coarse ware with cord-marked or basket-impressed patterns [@Hao2022; @Guijiabao2017a].
+Some ceramic decorations present northern styles, while others with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of close relationships between Guijiabao and both of its northern and southern neighboring areas [@Guijiabao2017a].
+The flotation analysis of soil samples from the 2015 excavation at Guijiabao demonstrates the predominance of cultivated broomcorn millet and foxtail millet during the Neolithic [@Guijiabao2016].
 Rice grains were also identified in later excavations from the Neolithic period based on macro botanical evidence and phytolith analysis [@Hao2022; @Huan2022], though in a smaller number than millet.
@@
 Macro-botanical and phytolith analyses of the Guijiabao site indicate that millet was the primary crop, aligning with findings at the Yingpanshan site in Sichuan [@Zhao2011].
-Conversely, botanical evidence from numerous sites in the Chengdu plain suggests a prevalence of rice, as seen at the Gaoshan and Baodun sites [@JiangGaoshan2017], and the Baiyangcun site in Yunnan [@Jiang2016; @Henglanshan2014].
+Conversely, botanical evidence from numerous sites in the Chengdu plain suggests a prevalence of rice, as seen at the Gaoshan and Baodun sites [@Guedes2013; @JiangGaoshan2017], and the Baiyangcun site in Yunnan [@Jiang2016; @Henglanshan2014].
 Moreover, while most research focuses on agricultural practices, there is a lack of clarity regarding other resources utilized by early farmers in new environments.
@@
 
-As a preliminary study to assess the feasibility of employing organic geochemical approaches, we collected five pottery fragments from cultural layers in the residential area of Guijiabao (sample code starting with SYG) (Table \@ref(tab:sherds-table)).
-For comparison, one fragment was sampled from Gaoshan (sample code starting with CDG).
+As a preliminary study to assess the feasibility of employing organic geochemical approaches, we collected five pottery fragments from cultural layers in the residential area of Guijiabao (Table \@ref(tab:sherds-table)).
+For comparison, one fragment was sampled from Gaoshan.
 Those pottery fragments represent typical types with well-defined contexts, with their radiocarbon dates ranging from the early phase of Neolithic period (ca.
@@
 1200-500 BP).
-The pottery samples are mostly base fragments and fine sand-tempered, with a few displaying impressed decorations.
-By examining potsherds from the early and late periods at Guijiabao, we aim to explore pottery use across time periods and the potential of utilizing an organic geochemical approach.
+The pottery fragments are mostly bases and fine sand-tempered, with a few displaying impressed decorations.
+By examining potsherds from the early and late periods at Guijiabao, we aim to investigate the applicability of this approach to determine the pottery use across time periods.
 
@@
 
-The data generated by these analyses and R code used for visualizations are openly available at https://github.com/LiYingWang/GJBpottery to enable research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].
+The data generated by these analyses and R code used for visualizations are openly available at https://github.com/LiYingWang/GJBpottery for research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].
 
@@
 This sample has a wider range of fatty acids from C~12:0~ to C~30:0~ (`r FA_sum_max` μg/g), where long-chain fatty acids could originate from plant sources [@Post1996; @Copley2005].
-Overall, the dominance of C~16:0~ and C~18:0~ in most samples suggests that the lipids extracted from our samples were greatly degraded.
+Overall, the dominance of C~16:0~ and C~18:0~ in most samples suggests that the lipids extracted from our samples were significantly degraded.
 
@@
 
-```{r chromatograms-plots-four, echo=FALSE, fig.cap= "Partial total ion cuurent (TIC) chromatograms of pottery samples from Guijiabao and Gaoshan.
+```{r chromatograms-plots-four, echo=FALSE, fig.cap= "Partial total ion current (TIC) chromatograms of pottery samples from Guijiabao and Gaoshan.
 'Cn:x' refers to fatty acid with n carbon atoms and x double bonds.
@@
 
-The isotope data shows that δ^13^C~16:0~ values range from `r rg_C16_isotope[1]` to `r rg_C16_isotope[2]` ‰, δ^13^C~18:0~ values range from `r rg_C18_isotope[1]` to `r rg_C18_isotope[2]` ‰, and Δ^13^C values are between `r delta_range[2]` and `r delta_range[1]` ‰.
-A half of the samples (n= 3) at Guijiabao and Gaoshan fall within the global reference range for ruminant adipose fat, where one sample from Guijiabao (SYG-TN13-E22-2#2) has a more enriched δ^13^C~16:0~ (`r E22_22_isotope[1]` ‰) than others.
+The isotope data shows that δ^13^C~16:0~ values range from `r rg_C16_isotope[1]` to `r rg_C16_isotope[2]` ‰, δ^13^C~18:0~ values vary from `r rg_C18_isotope[1]` to `r rg_C18_isotope[2]` ‰, and Δ^13^C values are between `r delta_range[2]` and `r delta_range[1]` ‰.
+Three potsherds from Guijiabao and Gaoshan fall within the global reference range for ruminant adipose fat, where one sample from Guijiabao (SYG-TN13-E22-2#2) has a more enriched δ^13^C~16:0~ (`r E22_22_isotope[1]` ‰) than others.
 The enriched δ^13^C suggests that the diet of ruminants contains both C~4~ and C~3~ plants (Figure \@ref(fig:isotopic-plots): B), while the depleted δ^13^C is related to a more C~3~ plant diet.
@@
 
-One sample (SYG-TN13-E23-3) with Δ^13^C of `r E23_isotope[3]` falls within the range for non-ruminant fats (Figure \@ref(fig:isotopic-plots): B).
+One specimen (SYG-TN13-E23-3) with Δ^13^C of `r E23_isotope[3]` falls within the range for non-ruminant fats (Figure \@ref(fig:isotopic-plots): B).
 Its relatively depleted δ^13^C~ values (δ^13^C~16:0~ of `r E23_isotope[1]` ‰ and δ^13^C~18:0~ of `r E23_isotope[2]` ‰) reflect input of C~3~ plants, terrestrial animals, or freshwater resources [@Shoda2018; @Lucquin2016].
-The two samples (SYG-TN13-E22-1 and SYG-3) dating to the historical period clustered within a narrow range of δ^13^C~16:0~ (`r SYG3_isotope[1]` to `r E22_1_isotope[1]` ‰) and δ^13^C~18:0~ (`r SYG3_isotope[2]` to `r E22_1_isotope[2]` ‰) near the reference line, showing a mixing source of ruminant and non-ruminant fat [@Copley2003].
+The two specimens (SYG-TN13-E22-1 and SYG-3) dating to the historical period clustered within a narrow range of δ^13^C~16:0~ (`r SYG3_isotope[1]` to `r E22_1_isotope[1]` ‰) and δ^13^C~18:0~ (`r SYG3_isotope[2]` to `r E22_1_isotope[2]` ‰) near the reference line, showing a mixing source of ruminant and non-ruminant fat [@Copley2003].
 A mixture of lipids from different sources is possible because δ^13^C values could be influenced by multiple distinct products, such as ruminant fats and plants, in the same cooking event or separate cooking events over time [@Evershed2002].
@@
 <!--A brief summary of the lipid yield and function of pottery-->
-Lipids extracted from potsherds at Guijiabao and Gaoshan all exceed interpretative amounts, with their profiles revealing the processing of both animal and plant products This confirms the role of the pottery as vessels for cooking or food storage at both sites.
-The generally low lipid concentrations (<20 μg/g) in five out of the six samples may be attributed to the bottom parts of vessels, which might absorb fewer residues compared to other parts, such as the rim, during cooking events [@Evershed2008experimental; @Kwak2020; @Charters1993].
+Lipids extracted from potsherds at Guijiabao and Gaoshan all exceed interpretative amounts, with their profiles revealing the processing of both animal and plant products.
+This confirms the role of the pottery as cooking vessels or food storage at both sites.
+The generally low lipid concentrations (<20 μg/g) in five out of the six samples may be attributed to the bottom parts of vessels, which might absorb fewer residues compared to other parts, such as the rims, during cooking events [@Evershed2008experimental; @Kwak2020; @Charters1993].
 Additional factors contributing to the low lipid yields could be post-depositional processes [@Correa2014], or diverse methods of food preparation, including grilling or air-drying [@Zhang2022].
-We recognize challenges for interpreting our results due to the small sample size and the absence of a direct local reference database from faunal bones or modern animals.
-Furthermore, the detailed contexts of faunal remains at Guijiabao are still in progress, thus they are not fully accessible.
+We recognize challenges for interpreting our results due to the small sample size and the lack of a direct local reference from faunal bones at Guijiabao.
 These limitations constrain a thorough comparison of various types of evidence at the site.
@@
 5000-4500 BP) reveal food contents, including ruminant adipose fat (n= 3) and non-ruminant fat (n= 1).
-The ruminant adipose (Δ^13^C from `r E22_21_isotope[3]` to `r E22_22_isotope [3]` ‰) could have originated from goat/sheep, cattle, and deer based on a modern database [@Evershed2002], alongside archaeological contexts in northwestern China [@Sun2023].
+The ruminant adipose (Δ^13^C from `r E22_21_isotope[3]` to `r E22_22_isotope [3]` ‰) could have originated from deer, goat/sheep, or cattle based on the faunal remains from the site [@Zhou2019] and archaeological contexts in northwestern China [@Sun2023], alongside a modern database [@Evershed2002].
 At Guijiabao, the ruminants' diet includes C~3~ plants (SYG-TN13-E22-2#1), and a mixture of C~3~ and C~4~ plants (SYG-TN13-E22-2#2).
@@
 <!--Plant as possible non-ruminant fat origin-->
-The non-ruminant fat (SYG-TN13-E23-3) could originate from C~3~ plants and freshwater resources, based on similar isotopic ranges reported for Neolithic pottery in eastern China [@Shoda2018; @Lyu2024], and a global database [@Steele2010].
-Although fish bones and net sinkers were found at Guijiabao [@Guijiabao2017b], lipid molecular evidence for fish, such as ω-(o-alkylphenyl)alkanoic acids and isoprenoid fatty acids [@Cramp2014], was absence in our samples.
-This suggests that fish might not have been commonly consumed, or alternative cooking methods for fish were employed.
+The non-ruminant fat (SYG-TN13-E23-3) could originate from C~3~ plants or freshwater resources, based on similar isotopic ranges reported for Neolithic pottery in eastern China [@Shoda2018; @Lyu2024], and a global database [@Steele2010].
+However, although fish bones and net sinkers were found at Guijiabao [@Guijiabao2017b], lipid molecular evidence for fish, such as ω-(o-alkylphenyl)alkanoic acids and isoprenoid fatty acids [@Cramp2014], was absence in our samples.
+This excludes the possibility of using fish fat to explain our sample.
 Possible source C~3~ plants could be rice or wild weeds based on botanical evidence [@Guijiabao2016; @Hao2022; @Huan2022].
 Due to a small portion of seeds of wild weeds in the floral assemblage [@Guijiabao2016], coupled with rice phytolichs identified from the residential area [@Huan2022], rice appears to be the more likely major source of C~3~ plants.
-In addition, the non-ruminant sample has a higher ratio of C~18:1~ and C~16:0~ to C~18:0~, associated with a greater likelihood of plant oils [@Mayyas2018; @Copley2005].
-However, we acknowledge that C~18:1~ might also originate from animal fats, or there is a potential for modern contamination, as monounsaturated fatty acids typically have limited preservation in archaeological contexts [@Whelton2021].
+In addition, the non-ruminant specimen has a higher ratio of C~18:1~ and C~16:0~ to C~18:0~, indicating a possibility of plant oils [@Mayyas2018; @Copley2005].
+However, we acknowledge that C~18:1~ might also originate from animal fats, or there is a potential for modern contamination, as monounsaturated fatty acids typically have weak preservation in archaeological contexts [@Whelton2021].
 A further detection of biomarkers for starchy plants with a specific derivatization method for lipid compounds will be necessary to assess the presence of rice oil [@Shoda2018].
@@
 The isotopic values of the non-ruminant sample align with those of reference fats from modern wild boars in western Japan [@Lucquin2016] and in the southeast Tibetan Plateau [@Zhang2022].
-Although horses could be another non-ruminant fat source, it is less likely to have been part of the local diet in southwest China during the Neolithic period.
-This is because wild horse bones are primarily concentrated at the Pleistocene and Neolithic sites in northern China, with only sporadic occurrences in the south [@Yuan2006].
-Moreover, substantial evidence of domesticated horses in southwest China dates back to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], a period approximately 3000 years later than the Neolithic period.
+Although horses could be another non-ruminant fat source, it is less likely to have been part of the local diet in southwest China during the Neolithic period, given the only sporadic occurrences in the south [@Yuan2006].
+Domesticated horses probably did not appear in southwest China until the Dian culture period (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], a period approximately 3000 years later than the Neolithic Guijiabao.
 Ongoing research on faunal remains at Guijiabao will provide more details for further discussion.
@@
 Most studies have primarily focused on mixed farming of millet and rice as the main economy in southwest China after the introduction of agricultural in the Neolithic period [@Deng2019].
-At Guijaibao, domesticated crops, especially broomcorn millet, constitute the majority of plants based on macro- and micro- botanical analyses [@Guijiabao2016; @Huan2022], and nitrogen and carbon isotope analysis of human bones suggests a more C~4~ diet [@Lin2022].
+At Guijaibao, domesticated crops, especially broomcorn millet, constitute the majority of plants based on macro- and micro- botanical analyses [@Guijiabao2016; @Huan2022].
+Nitrogen and carbon isotope analysis of human bones also suggests a more C~4~ diet [@Lin2022].
 Additionally, stone knives found at the site might be potentially be used for farming practices [@Zhou2019; @Hao2022; @Lin2022].
-In our study, a Neolithic potsherd containing the lipid compound of broomcorn millet [@Heron2016], provides direct evidence of millet processing at Guijiabao.
-Although a depleted δ^13^C~16:0~ in the non-ruminant sample suggest the possibility of C~3~plants related to rice, further analysis is needed for confirmation.
-It is important to note that lipid signatures of plants tend to be weak due to microbial degradation or being masked by animal fats [@Colonese2017; @Evershed2008experimental], possibly explaining the low lipid yield of plants in one sample.
+In our study, a Neolithic potsherd containing the lipid compound of broomcorn millet provides direct evidence of millet processing at Guijiabao.
+Our non-ruminant specimen (SYG-TN13-E23-3), noted above, also indicates the presence of rice, though further analysis is needed for confirmation.
+It is important to note that lipid signatures of plants tend to be weak due to microbial degradation or being masked by animal fats [@Colonese2017; @Evershed2008experimental], possibly explaining why only one specimen in our sample had a non-ruminant signal.
 Nevertheless, the frequent processing of ruminant meats in pottery, with unearthed hunting tools, suggests that animal hunting could be an essential subsistence strategy alongside crop consumption.
-This may also reflect a subsistence strategy adapted to highlands with cold environments where meat consumption is common [@Zhang2022].
+This may also be a strategy adapted to highlands with cold environments where meat consumption is common [@Zhang2022].
 
@@
 1200-500 BP) from Guijiabao demonstrates mixtures of ruminant adipose and non-ruminant fats, implying a potential increase in the exploitation of C~3~ plants, pigs, or freshwater resources compared to the Neolithic period.
-In addition to rice, the source of C~3~ plants could also have included wheat, since it was introduced into the region during the historical period [@JiangGaoshan2017].
+In addition to rice, the source of C~3~ plants could also have included wheat, which was introduced into the region during the late Bronze Age (ca.
+2,800–2000 cal.
+BP) [@Huan2022].
 For the Gaoshan sample, a slightly enriched δ^13^C may suggest minor contributions of pigs [@Evershed2002], in correspondence with the predominance of domestic pigs at the site [@He2020].
-However, determining changes in culinary practices over time and regional differences would require a larger sample size.
-While a geochemical approach analyzing organic residues may provide a limited perspective on local diets, it contributes to improving our current understanding of direct food practices through organic residues in ceramic vessels.
-As a preliminary study, our data reveals food contents cooked in pottery from both animal and plant sources, offering valuable insights into culinary practices during the time when agriculture spread into the Yuan Basin in southwest China.
+While determining changes in culinary practices over time and regional differences would require a larger sample size, it does contribute to improving our current understanding of direct food practices using organic residues in ceramic vessels.
+As a preliminary study, our data reveals food contents cooked in pottery from both animal and plant sources, offering valuable insights into culinary practices during the time when agriculture spread into the southwest China.
 
 # Conclusion
-This paper presents a preliminary result investigating pottery use at Guijiabao in the Yanyuan area, middle Hengduan mountain range, through organic residue analyses to explore culinary practices.
+This paper presents a preliminary result investigating pottery use and culinary practices at Guijiabao in the Yanyuan area, middle Hengduan mountain range, through organic residue analyses.
 Despite the small sample size, our results reveal a relatively diverse range of food categories cooked in pottery during the Neolithic period, including ruminant adipose fats, plants such as broomcorn millet, and non-ruminant sources.
 In the historical period, mixtures of ruminant adipose fats and non-ruminant fats were identified.
-Non-ruminant fats could have originated from C~3~ plants, pigs, or freshwater resources.
+Non-ruminant fats could have originated from C3 plants, pigs, or freshwater resources.
 
 While a mixed crop of millet and rice is generally accepted as the main subsistence in southwest China, our results indicate that ruminant adipose fats were more commonly cooked in pottery.
-The prevalence of ruminant fats suggests the significance of meat consumption, a aspect often overlooked after the introduction of crops.
+The prevalence of ruminant fats suggests the significance of meat consumption, an aspect often overlooked after the introduction of crops.
 The discovery of hunting tools at Guijiabao further indicates that hunting might have played a crucial role in the local subsistence strategy, complementing millet/rice farming.
-This provides valuable insights into subsistence strategies in a highland setting in the Hengduan Mountains ranges.
-Although our preliminary data has limitations for a comprehensive interpretation, the results demonstrate the feasibility of lipid identifications with compound-specific stable isotope analysis to determine food content processed in ceramic vessels at Guijiabao.
+This provides insights into subsistence practices in the highland setting of the Hengduan Mountains.
+Although our preliminary data has limitations for a comprehensive interpretation, the results demonstrate the feasibility of lipid identifications with compound-specific stable isotope analysis to estimate food content processed in ceramic vessels at Guijiabao.
 A more systematic sampling of potsherds across different phases and pottery types will contribute to a clearer understanding of diet and subsistence in southwest China in the future.
@@
 We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines.
-We also want to express our gratitude to the anonymous reviewers for their comments that helped us improve this paper.
+We also want to express our gratitude to the anonymous reviewers for their comments that help us improve this paper.
 The research was supported by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).

### 2024-02-16 analysis/paper/paper.Rmd
minor edits

@@
 The isotopic values of the non-ruminant sample align with those of reference fats from modern wild boars in western Japan [@Lucquin2016] and in the southeast Tibetan Plateau [@Zhang2022].
-Although horses could be another non-ruminant fat source, it is less likely to have been part of the local diet in southwest China during the Neolithic period, given the only sporadic occurrences in the south [@Yuan2006].
+Although horses could be another non-ruminant fat source, it is less likely to have been part of the local diet in southwest China during the Neolithic period, given the only sporadic occurrences in the south [@Zhang2022].
 Domesticated horses probably did not appear in southwest China until the Dian culture period (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], a period approximately 3000 years later than the Neolithic Guijiabao.

### 2024-02-16 analysis/paper/paper.Rmd
submitted to ARA

@@
 ---
-title: "Preliminary insights into the use of pottery and culinary pratices at Guijiabao site in southwest China"
+title: "Preliminary insights into the use of pottery and culinary practices at Guijiabao site in southwest China"
 author:
@@
      Mixed crops of millet and rice are thought to have been introduced to southwest China in the Neolithic period.
-Despite the recovery of both millet and rice grains, the varying proportions observed across the area underscore the complexity of the process of agricutural dispersal.
+Despite the recovery of both millet and rice grains, the varying proportions observed across the area underscore the complexity of the process of agricultural dispersal.
 Further investigation is needed to understand how new crops were integrated into southwest China, characterized by diverse environmental conditions and landscapes.

### 2024-04-10 analysis/paper/paper.Rmd
add citations

@@
 4700-3700 BP) in the Chengdu Plain [@Gaoshan2017].
-The primary subsistence strategy at the Gaoshan site could have been rice cultivation with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020].
+The primary subsistence strategy at the Gaoshan site could have been rice farming with pig husbandry, supplemented by millet cultivation [@JiangGaoshan2017; @Lee2020; @He2020], which gradually decreased in importance than rice [@Wang2023].
 It is suggested that the spread of rice farming into Sichuan occurred at least 4700 years ago through population migration from the Middle and Lower Yangzi valleys [@Zhang2010].
@@
 While mixed agriculture of millet and rice was practiced in southwest China, variations in their proportions across the region have been observed.
-Macro-botanical and phytolith analyses of the Guijiabao site indicate that millet was the primary crop, aligning with findings at the Yingpanshan site in Sichuan [@Zhao2011].
+Macro-botanical and phytolith analyses of the Guijiabao site indicate that millet was the primary crop, aligning with findings at the Yingpanshan site and Guiyuanqiao site in Sichuan [@Zhao2011; @Guedes2015].
 Conversely, botanical evidence from numerous sites in the Chengdu plain suggests a prevalence of rice, as seen at the Gaoshan and Baodun sites [@Guedes2013; @JiangGaoshan2017], and the Baiyangcun site in Yunnan [@Jiang2016; @Henglanshan2014].
@@
 Nevertheless, the frequent processing of ruminant meats in pottery, with unearthed hunting tools, suggests that animal hunting could be an essential subsistence strategy alongside crop consumption.
-This may also be a strategy adapted to highlands with cold environments where meat consumption is common [@Zhang2022].
+This may also be a strategy adapted to highlands with cold environments where meat consumption is common [@Zhang2022], or to reduce the impact of potential crop failure [@Wang2023tibetan].
 

### 2024-04-13 analysis/paper/paper.Rmd
revising based on reviewer comments

@@
 BP) [@Huan2022].
-For the Gaoshan sample, a slightly enriched δ^13^C may suggest minor contributions of pigs [@Evershed2002], in correspondence with the predominance of domestic pigs at the site [@He2020].
+It is surprising that ruminant adipose fat, rather than non-ruminant fat, was detected in the Gaoshan potsherd, given the predominance of domestic pigs in the faunal assemblage at the site [@He2020].
+One possibility might be our Gaoshan potsherd is an outlier, potentially not representative of the broader dietary patterns.
+Another plausible explanation could be that the isotopic values of ruminant adipose fat result from a mixture of non-ruminant adipose fats and ruminant milk fats.
+This possibility may be supported by the slightly enriched δ^13^C in the Gaoshan sample, suggesting minor contributions from non-ruminant fats, such as those from pigs [@Evershed2002].
+
 While determining changes in culinary practices over time and regional differences would require a larger sample size, it does contribute to improving our current understanding of direct food practices using organic residues in ceramic vessels.

### 2024-04-16 analysis/paper/paper.Rmd
edit the discussion section

@@
 The non-ruminant fat (SYG-TN13-E23-3) could originate from C~3~ plants or freshwater resources, based on similar isotopic ranges reported for Neolithic pottery in eastern China [@Shoda2018; @Lyu2024], and a global database [@Steele2010].
-However, although fish bones and net sinkers were found at Guijiabao [@Guijiabao2017b], lipid molecular evidence for fish, such as ω-(o-alkylphenyl)alkanoic acids and isoprenoid fatty acids [@Cramp2014], was absence in our samples.
-This excludes the possibility of using fish fat to explain our sample.
-Possible source C~3~ plants could be rice or wild weeds based on botanical evidence [@Guijiabao2016; @Hao2022; @Huan2022].
-Due to a small portion of seeds of wild weeds in the floral assemblage [@Guijiabao2016], coupled with rice phytolichs identified from the residential area [@Huan2022], rice appears to be the more likely major source of C~3~ plants.
-In addition, the non-ruminant specimen has a higher ratio of C~18:1~ and C~16:0~ to C~18:0~, indicating a possibility of plant oils [@Mayyas2018; @Copley2005].
-However, we acknowledge that C~18:1~ might also originate from animal fats, or there is a potential for modern contamination, as monounsaturated fatty acids typically have weak preservation in archaeological contexts [@Whelton2021].
+Possible source of C~3~ plants could include rice and wild weeds, supported by the presence of rice phytolichs and a small portion of carbonized seeds of wild weeds, such as *Pennisetum*, identified from the residential area at Guijiabao [@Guijiabao2016; @Hao2022; @Huan2022].
 A further detection of biomarkers for starchy plants with a specific derivatization method for lipid compounds will be necessary to assess the presence of rice oil [@Shoda2018].
+Although freshwater resources might be another source of non-ruminant fats according to the finding of fish bones and net sinkers at Guijiabao [@Guijiabao2017b], lipid molecular evidence for fish, such as ω-(o-alkylphenyl)alkanoic acids and isoprenoid fatty acids [@Cramp2014], was absent in our samples.
+The absence of these molecules could result from either their actual absence, or poor preservation of those compounds in archaeological contexts in general.
+Thus, we are unable to exclude the possibility of freshwater resources.
 
@@
 This possibility may be supported by the slightly enriched δ^13^C in the Gaoshan sample, suggesting minor contributions from non-ruminant fats, such as those from pigs [@Evershed2002].
-
 While determining changes in culinary practices over time and regional differences would require a larger sample size, it does contribute to improving our current understanding of direct food practices using organic residues in ceramic vessels.

### 2024-04-28 analysis/code/isotope-analysis.R
fixed a point not showing

@@ -131,7 +131,7 @@ big_delta_C16 <-
        y = bquote(Delta*{}^13*"C")) +
   scale_x_continuous(limits = c(-33, -20),
                      breaks = seq(-33, -20, 2)) +
-  ylim(-2.5, 1.5) +
+  ylim(-2.7, 1.5) +
   geom_hline(yintercept = -0.7, linetype = "dashed") +
   annotate("text", x = -31.5, y = -0.9,
            label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +

### 2024-04-28 analysis/paper/paper.Rmd
minor edits based on reviewers' comments

@@
 Further investigation is needed to understand how new crops were integrated into southwest China, characterized by diverse environmental conditions and landscapes.
-This study contributes to this understanding by exploring subsistence strategies in southwest Sichuan through a preliminary analysis of lipid residues in pottery from the Guijiabao site using organic geochemical analyses.
+This study explores pottery use in southwest Sichuan through a preliminary analysis of lipid residues in pottery from the Guijiabao site using organic geochemical analyses.
 The site, situated in the middle Hengduan mountain ranges, presents well-preserved stratigraphic records spanning from the Neolithic to the historical period, providing valuable insights into local subsistence practices in a highland environment.
-Results reveal the detection of ruminant adipose fats in the majority of potsherds, with a few showing plant inputs such as broomcorn millet.
-This suggests that animals played a substantial role in the local diet, alongside the consumption of millet.
-These culinary practices also indicate that hunting may have remained a significant subsistence activity in the high-altitude region of southwest China, even after the introduction of crops.
+The results reveal that broomcorn millet was processed in a pot, with the majority of potsherds showing mixed sources including ruminant adipose fats.
+This suggests that animals might have played an important role in the local diet, alongside the consumption of millet.
+These culinary practices also indicate a subsistence strategy relying on hunting to a certain degree in the high-altitude region of southwest China, even after the introduction of crops.
 
@@
 <!--Yanyuan area and Guijiabao site-->
-Guijiabao is located in the southwest of Yanyuan County in Sichuan Province, southwest China (Figure \@ref(fig:SW-china-map)), with an elevation of 2399 m [@Guijiabao2017a].
+Guijiabao is located in the southwest of Yanyuan County in Sichuan Province, southwest China (Figure \@ref(fig:SW-china-map)), with an elevation of around 2400 m [@Guijiabao2017a].
 Yanyuan is a pull-apart basin surrounded by mountains in the central east part of the Hengduan Mountain ranges.
@@
 The flotation analysis of soil samples from the 2015 excavation at Guijiabao demonstrates the predominance of cultivated broomcorn millet and foxtail millet during the Neolithic [@Guijiabao2016].
-Rice grains were also identified in later excavations from the Neolithic period based on macro botanical evidence and phytolith analysis [@Hao2022; @Huan2022], though in a smaller number than millet.
+Rice grains were also identified in later excavations from the Neolithic period based on macrobotanical evidence and phytolith analysis [@Hao2022; @Huan2022], though in a smaller number than millet.
+In addition to plants, faunal remains such as deer, birds, pigs, horses, and fish were also found at the site [@Zhou2019], but a detailed analysis of their temporal and spatial distribution is still ongoing.
 
@@
 
-The surface of each pottery sample was removed to prevent external contamination from the environment and handling.
-Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C) [@Mcclure2018].
+The surface (~2mm) of each potsherd was removed to prevent external contamination from the environment and handling [@Craig2004].
+Approximately 3.5 grams of powder from the interior potsherd surface were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C) [@Mcclure2018].
 To obtain a higher yield, free fatty acids were recovered from bound lipids through saponification, which was conducted by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating at 70 °C overnight [@Kaluzna2016; @Craig2004].
@@
 The direct input from C~4~ plants could be another reason for the mixed signal, as we have identified millet in the same sample.
-The domestication status of these ruminants detected in the samples remains unclear.
-However, considering the finding of hunting tools at Guijiabao, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a], with the abundance of wild ruminants in the natural settings of the Hengduan mountain ranges, such as deer, mutton, and yaks [@Zhang2022; @Zhang2016; @Chen2020; @He2020], it is plausible that subsistence at Guijiabao involving hunting to a certain degree.
+The husbandry of ruminant animals detected in the samples remains unclear.
+It is also likely that those potsherds showing ruminant signal represent mixtures of non-ruminant adipose fats, such as those from aquatic resources, and ruminant milk fats.
+Considering the finding of hunting tools at Guijiabao, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a], with the abundance of wild ruminants in the natural settings of the Hengduan mountain ranges, such as deer, mutton, and yaks [@Zhang2022; @Zhang2016; @Chen2020; @He2020], it is plausible that subsistence at Guijiabao involving hunting to a certain degree.
 
@@
 
-While a mixed crop of millet and rice is generally accepted as the main subsistence in southwest China, our results indicate that ruminant adipose fats were more commonly cooked in pottery.
-The prevalence of ruminant fats suggests the significance of meat consumption, an aspect often overlooked after the introduction of crops.
-The discovery of hunting tools at Guijiabao further indicates that hunting might have played a crucial role in the local subsistence strategy, complementing millet/rice farming.
-This provides insights into subsistence practices in the highland setting of the Hengduan Mountains.
+While a mixed crop of millet and rice is generally regarded as the main subsistence in southwest China, our findings show that other food sources, such as ruminant adipose fats, were also cooked in pottery.
+The detection of ruminant fats in Guijiabao samples might suggest the importance of meat consumption, an aspect often overlooked after the introduction of crops.
+Our results also correspond to an abundance of hunting tools at the site, implying that hunting and fishing might have played a crucial part in the local subsistence strategy, supplementing millet/rice farming.
+This sheds light on subsistence practices in the highlands of the Hengduan Mountains.
 Although our preliminary data has limitations for a comprehensive interpretation, the results demonstrate the feasibility of lipid identifications with compound-specific stable isotope analysis to estimate food content processed in ceramic vessels at Guijiabao.

### 2024-04-28 analysis/paper/paper.Rmd
minor edits

@@
 
-<!--Guijaibao as a research area-->
+<!--Guijiabao as a research area-->
 This study primarily focuses on the Guijiabao site located in the Hengduan Mountain ranges, a region in southwest China that connects the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south.
@@
 At Guijiabao, the ruminants' diet includes C~3~ plants (SYG-TN13-E22-2#1), and a mixture of C~3~ and C~4~ plants (SYG-TN13-E22-2#2).
-A C~3~ diet is supported by the predominance of C~3~ weeds at Guijiabao, such as *Pennisetum* and *Chenopodiaceae* [@Guijiabao2016].
+A C~3~ diet is supported by the predominance of C~3~ weeds, such as *Pennisetum* and *Chenopodiaceae* [@Guijiabao2016].
 Some species in *Chenopodiaceae* can undergo C~4~ photosynthesis in arid environments [@Rudov2020; @Chasan2022], which may explain the mixture of plant types in SYG-TN13-E22-2#2.
@@
 It is also likely that those potsherds showing ruminant signal represent mixtures of non-ruminant adipose fats, such as those from aquatic resources, and ruminant milk fats.
-Considering the finding of hunting tools at Guijiabao, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a], with the abundance of wild ruminants in the natural settings of the Hengduan mountain ranges, such as deer, mutton, and yaks [@Zhang2022; @Zhang2016; @Chen2020; @He2020], it is plausible that subsistence at Guijiabao involving hunting to a certain degree.
+Considering the finding of hunting tools at Guijiabao, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a], with the abundance of wild ruminants in the natural settings of the Hengduan mountain ranges, such as deer, mutton, and yaks [@Zhang2022; @Zhang2016; @Chen2020; @He2020], it is plausible that subsistence at the site involving hunting to a certain degree.
 
@@
 Domesticated horses probably did not appear in southwest China until the Dian culture period (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], a period approximately 3000 years later than the Neolithic Guijiabao.
-Ongoing research on faunal remains at Guijiabao will provide more details for further discussion.
+Ongoing research on faunal remains from the site will provide more details for further discussion.
 
@@
 Most studies have primarily focused on mixed farming of millet and rice as the main economy in southwest China after the introduction of agricultural in the Neolithic period [@Deng2019].
-At Guijaibao, domesticated crops, especially broomcorn millet, constitute the majority of plants based on macro- and micro- botanical analyses [@Guijiabao2016; @Huan2022].
+At Guijiabao, domesticated crops, especially broomcorn millet, constitute the majority of plants based on macro- and micro- botanical analyses [@Guijiabao2016; @Huan2022].
 Nitrogen and carbon isotope analysis of human bones also suggests a more C~4~ diet [@Lin2022].
@@
 In our study, a Neolithic potsherd containing the lipid compound of broomcorn millet provides direct evidence of millet processing at Guijiabao.
-Our non-ruminant specimen (SYG-TN13-E23-3), noted above, also indicates the presence of rice, though further analysis is needed for confirmation.
+Our non-ruminant specimen (SYG-TN13-E23-3), noted above, might indicate the processing of rice, though further analysis is needed for confirmation.
 It is important to note that lipid signatures of plants tend to be weak due to microbial degradation or being masked by animal fats [@Colonese2017; @Evershed2008experimental], possibly explaining why only one specimen in our sample had a non-ruminant signal.
-Nevertheless, the frequent processing of ruminant meats in pottery, with unearthed hunting tools, suggests that animal hunting could be an essential subsistence strategy alongside crop consumption.
+Nevertheless, the processing of ruminant meats in pottery, with unearthed hunting tools, suggests that animal hunting could be an essential subsistence strategy alongside crop consumption.
 This may also be a strategy adapted to highlands with cold environments where meat consumption is common [@Zhang2022], or to reduce the impact of potential crop failure [@Wang2023tibetan].
@@
 This possibility may be supported by the slightly enriched δ^13^C in the Gaoshan sample, suggesting minor contributions from non-ruminant fats, such as those from pigs [@Evershed2002].
-While determining changes in culinary practices over time and regional differences would require a larger sample size, it does contribute to improving our current understanding of direct food practices using organic residues in ceramic vessels.
-As a preliminary study, our data reveals food contents cooked in pottery from both animal and plant sources, offering valuable insights into culinary practices during the time when agriculture spread into the southwest China.
+Although determining changes in culinary practices over time and regional differences would require a larger sample size, our results contribute to improving our understanding of direct food practices during the spread of agriculture into southwest China through organic residues in ceramic vessels.
 
@@
 We thank the Chengdu Municipal Institute of Cultural Relics and Archaeology for their permission to access and analyze the pottery samples.
-We thank Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for sample preparation and for conducting lipid extraction and analysis.
-We thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines.
-We also want to express our gratitude to the anonymous reviewers for their comments that help us improve this paper.
-The research was supported by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).
+Special thanks to Julian Sachs in the Department of Oceanography at the University of Washington for his support in letting us use their laboratory for lipid extraction and analysis.
+We also thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines.
+Thanks to Yin-Ru Chiang in the Biodiversity Research Center, and the Institute of Plant and Microbial Biology at Academia Sinica, Taiwan for their support during the experiment.
+We appreciate the valuable comments provided by the anonymous reviewers, which helped improve this paper.
+This work was supported by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).
 

### 2024-04-29 analysis/paper/paper.Rmd
minor edit

@@
 # Conclusion
-This paper presents a preliminary result investigating pottery use and culinary practices at Guijiabao in the Yanyuan area, middle Hengduan mountain range, through organic residue analyses.
+This paper presents preliminary results investigating pottery use and culinary practices at Guijiabao in the Yanyuan area, middle Hengduan mountain range, through organic residue analyses.
 Despite the small sample size, our results reveal a relatively diverse range of food categories cooked in pottery during the Neolithic period, including ruminant adipose fats, plants such as broomcorn millet, and non-ruminant sources.

### 2024-05-01 analysis/paper/paper.Rmd
revised and submitted

@@
 The husbandry of ruminant animals detected in the samples remains unclear.
-It is also likely that those potsherds showing ruminant signal represent mixtures of non-ruminant adipose fats, such as those from aquatic resources, and ruminant milk fats.
+It is also likely that those potsherds showing ruminant signal represent mixtures of non-ruminant adipose fats, such as those from aquatic resources, and ruminant milk fats [@Evershed2008].
 Considering the finding of hunting tools at Guijiabao, such as stone balls and arrowheads [@Hao2022; @Guijiabao2020a], with the abundance of wild ruminants in the natural settings of the Hengduan mountain ranges, such as deer, mutton, and yaks [@Zhang2022; @Zhang2016; @Chen2020; @He2020], it is plausible that subsistence at the site involving hunting to a certain degree.
@@
 We appreciate the valuable comments provided by the anonymous reviewers, which helped improve this paper.
-This work was supported by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2).
+This work was supported by the National Science and Technology Council (MOST 109-2410-H-001-009-MY2) and the Institute of History and Philology, Academia Sinica.
 

### 2024-05-24 analysis/paper/paper.Rmd
udpated code for figures

@@
 # plot the first group
-chroma_four <-
+chroma_five <-
   meth_all %>%
@@
                      expand = c(0, 0)) +
-  scale_x_continuous(limits = c(18, 47),
+  scale_x_continuous(limits = c(17, 48),
                      breaks = seq(15, 50, 1),
@@
 
-ggsave(plot = chroma_four,
-       here::here("analysis","figures", "chromatograms_four.png"),
+ggsave(plot = chroma_five,
+       here::here("analysis","figures", "chromatograms_five.png"),
        bg = "white",
@@
   dplyr::mutate(delta = `13C C18:0`-`13C C16:0`) %>%
-  mutate(name = case_when(C13 %in% c("Pot 1") ~ "CDG-062",
+  mutate(label = case_when(C13 %in% c("Pot 1") ~ "CDG-062",
                           C13 %in% c("Pot 2") ~ "SYG-3",
@@
                           C13 %in% c("Pot 6") ~ "SYG-TN13-E23-3")) %>%
-  mutate(name = factor(name, levels = c("SYG-TN13-E22-1",
+  mutate(label = factor(label, levels = c("SYG-TN13-E22-1",
                                         "SYG-TN13-E22-2#1",
@@
                                         "CDG-062"))) %>%
-  mutate(period = ifelse(name %in% c("SYG-TN13-E22-1", "SYG-3" ), "Historical", "Neolithic"))
+  mutate(period = ifelse(label %in% c("SYG-TN13-E22-1", "SYG-3" ), "Historical", "Neolithic"))
 
@@
   ggplot(aes(`13C C16:0`,`13C C18:0`)) +
-  geom_point(size = 2, alpha = 0.9, aes(shape = name, color = period)) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = label, color = period)) +
   #ggrepel::geom_text_repel(aes(label = C13)) +
@@
   ggplot(aes(`13C C16:0`, delta)) +
-  geom_point(size = 2, alpha = 0.9, aes(shape = name, color = period)) +
-  labs(x = expression(paste(delta*{}^13* C[16:0], "\u2030")),
-       y = expression(paste(delta*{}^13* C[18:0], "\u2030"))) +
+  geom_point(size = 2, alpha = 0.9, aes(shape = label, color = period)) +
+  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
+       y = bquote(Delta*{}^13*"C")) +
   scale_x_continuous(limits = c(-33, -20), breaks = seq(-33, -20, 2)) +

### 2024-05-24 analysis/code/GCFID-quant.R

@@ -106,7 +106,7 @@ meth_all %>%
   scale_y_continuous(limits = c(0, 1160000),
                      breaks = seq(0, 1100000, 200000),
                      expand = c(0, 0)) +
-  scale_x_continuous(limits = c(10, 47),
+  scale_x_continuous(limits = c(17, 48),
                      breaks = seq(10, 50, 1),
                      labels = insert_minor(seq(10, 50, 5), 4),
                      expand = c(0, 0)) +
@@ -115,7 +115,7 @@ meth_all %>%
   theme(axis.line.y = element_blank(), strip.background = element_blank(), # remove title box
         axis.ticks.y=element_blank(), axis.text.y = element_blank())
 
-ggsave(here::here("analysis","figures", "chromatograms_four.png"),
+ggsave(here::here("analysis","figures", "chromatograms_five.png"),
        bg = "white", width = 7, height = 8, dpi = 300, units = "in")
 
 # plot the one with high yield (GC program 1)

### 2024-05-25 analysis/paper/paper.Rmd
after proof

@@
 Recent studies have successfully identified various plant types processed in archaeological vessels, including broomcorn millet [@Bossard2013; @Heron2016; @Ganzarolli2018], starchy plants [@Shoda2018], and cereals [@Hammann2018; @Colonese2017].
-Several applications to ancient pottery in China also highlight the robustness of this approach in understanding regional cooking practices [@Shoda2018; @Lyu2024; @Han2022], discerning distinct subsistence strategies between low- and high-altitude zones in Tibetan Plateau [@Zhang2022], and exploring the impact of nomadic pastoralism on northwest China [@Sun2023].
+Several applications to ancient pottery in China also highlight the robustness of this approach in understanding regional cooking practices [@Shoda2018; @Lyu2024; @Han2022], discerning distinct subsistence strategies between low- and high-altitude zones in the Tibetan Plateau [@Zhang2022], and exploring the impact of nomadic pastoralism in northwest China [@Sun2023].
 In this study, we aim to investigate pottery use and culinary practices in the middle Hengduan mountains, with a specific focus on ceramics from the Guijiabao site to gain insights into local subsistence.
@@
 
-The surface (~2mm) of each potsherd was removed to prevent external contamination from the environment and handling [@Craig2004].
+The surface (~1mm) of each potsherd was removed to prevent external contamination from the environment and handling [@Craig2004].
 Approximately 3.5 grams of powder from the interior potsherd surface were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C) [@Mcclure2018].
@@
 Separation was carried out using the same model of column for GC-MS.
-Helium was used as a gas carrier with a constant flow rate of 1.24 ml/minute.
+Helium was used as a gas carrier with a constant flow rate of 1.24 ml/min.
 The oven temperature was set at 90 °C for 1.5 minutes, then raised at a rate of 5 °C/min to 320 °C and held for 18 minutes.
@@
 5000-4500 BP) reveal food contents, including ruminant adipose fat (n= 3) and non-ruminant fat (n= 1).
-The ruminant adipose (Δ^13^C from `r E22_21_isotope[3]` to `r E22_22_isotope [3]` ‰) could have originated from deer, goat/sheep, or cattle based on the faunal remains from the site [@Zhou2019] and archaeological contexts in northwestern China [@Sun2023], alongside a modern database [@Evershed2002].
+The ruminant adipose (Δ^13^C from `r E22_21_isotope[3]` to `r E22_22_isotope [3]` ‰) could have originated from deer, goat/sheep, or cattle based on the faunal remains from the site [@Zhou2019] and archaeological contexts in northwest China [@Sun2023], alongside a modern database [@Evershed2002].
 At Guijiabao, the ruminants' diet includes C~3~ plants (SYG-TN13-E22-2#1), and a mixture of C~3~ and C~4~ plants (SYG-TN13-E22-2#2).
@@
 We also thank Matthew Wolhowe for his suggestions and assistance with the GC-based machines.
-Thanks to Yin-Ru Chiang in the Biodiversity Research Center, and the Institute of Plant and Microbial Biology at Academia Sinica, Taiwan for their support during the experiment.
+Thanks to Yin-Ru Chiang and his team in the Biodiversity Research Center, and the Institute of Plant and Microbial Biology at Academia Sinica, Taiwan for their suggestions during the experiment.
 We appreciate the valuable comments provided by the anonymous reviewers, which helped improve this paper.

### 2024-05-25 analysis/code/site-regional-map-Ch.R
edit code for maps

@@ -3,7 +3,6 @@ theme_set(theme_bw(base_size = 6))
 library(sf)
 library(rnaturalearth)
 library(rnaturalearthdata)
-library(here)
 library(ggspatial)
 #devtools::install_github('3wen/legendMap')
 library(legendMap)
@@ -11,78 +10,34 @@ library(tmaptools)
 library(shadowtext)
 library(tidyverse)
 
-world <- ne_countries(scale = "medium", returnclass = "sf")
-# class(world)
-world_points <- sf::st_point_on_surface(world)
-world_points <- cbind(world, st_coordinates(st_point_on_surface(world$geometry))) %>%
-  filter(brk_name %in% c("Thailand", "China", "Myanmar", "Laos",
-                         "Vietnam", "Cambodia", "Bangladesh", "Bhutan")) %>%
-  mutate(Y = ifelse(brk_name %in% c("Bhutan", "Bangladesh"), Y+2, Y)) %>%
-  mutate(Y = ifelse(brk_name %in% c("Thailand", "Myanmar"), Y+3.5, Y))
-
-# add site location
+# points for site location
 site_location <-
   data.frame(location = c("Guijuabao", "Gaoshan"), # Chengdu(104.0587, 30.5899)
              lon = c(101.3617, 103.3446),
              lat = c(27.2657, 30.2709))
 
-China_SE_Asia <-
-  ggplot(data = world) +
-  geom_sf(fill = "antiquewhite") + #antiquewhite
-  geom_rect(xmin = 99.5, xmax = 105.5, ymin = 25.5, ymax = 31.5,
-            fill = NA, colour = "red", size = 0.5) +
-  geom_shadowtext(data= world_points, #geom_shadowtext
-                  aes(x = X, y = Y,
-                      label = brk_name),
-                  color='black',
-                  bg.colour='white',
-                  size = 2,
-                  position = position_nudge(y = - 1.7, x = 0.5)) +
-  #annotate(geom = "text", x = 102.3, y = 9.4, label = "Gulf of\nTailand",
-           #fontface = "italic", color = "grey22", size = 2) +
-  coord_sf(xlim = c(85, 120), ylim = c(18, 45), expand = FALSE) + #add datum = NA to remove
-  scale_x_continuous(breaks = seq(80, 125, by = 10)) +
-  scale_y_continuous(breaks = seq(5, 45, by = 10)) +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank())
-
[…]

### 2024-05-25 analysis/paper/paper.Rmd
fixed the issue of knitting

@@
 library(here)
-knitr::include_graphics(here("analysis", "figures", "Sichuan-sites-map3.png"))
+knitr::include_graphics(here("analysis", "figures", "Sichuan-sites-map.png"))
 ```
@@
 Lipid yields from all potsherds are above interpretable concentrations (5 μg/g) [@Evershed2008experimental], suggesting those ceramics should be associated with food processing.
-The main compounds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid) and C~18:0~ (stearic acid), commonly derived from animal fats and plant oils (Figure \@ref(fig:chromatograms-plots-four)).
+The main compounds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid) and C~18:0~ (stearic acid), commonly derived from animal fats and plant oils (Figure \@ref(fig:chromatograms-plots-five)).
 Unsaturated fatty acid, C~18:1~, was detected in all samples and dominant in three, but we are unable to distinguish the C~18:1~ types based on current methods.
@@
 
-```{r chromatogram-four, cache= TRUE}
+```{r chromatogram-five, cache= TRUE}
 # to insert ticks in plot
@@
 
-```{r chromatograms-plots-four, echo=FALSE, fig.cap= "Partial total ion current (TIC) chromatograms of pottery samples from Guijiabao and Gaoshan.
+```{r chromatograms-plots-five, echo=FALSE, fig.cap= "Partial total ion current (TIC) chromatograms of pottery samples from Guijiabao and Gaoshan.
 'Cn:x' refers to fatty acid with n carbon atoms and x double bonds.
@@
 
-knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
+knitr::include_graphics(here::here("analysis/figures/chromatograms_five.png"))
 ```
@@
 
-# filter the Meth corrected data
+# filter the MeOH corrected data
 delta_meth_cor <-

### 2024-05-26 analysis/paper/paper.Rmd
updated the repository for data

@@
 
-The data generated by these analyses and R code used for visualizations are openly available at https://github.com/LiYingWang/GJBpottery for research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].
+The data generated by these analyses and R code used for visualizations are openly available at https://osf.io/46bpf/ for research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].
 
